### 2019-09-18 OnlineAppendix.tex
knitr and tex files

\section{Observational Study}
\subsection{Covariate Balance Plot}
\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{balancePlot.pdf}
\caption{Standardized differences on baseline covariates between
  students in the treatment group with $H=0$ and $H=1$, before and
  after propensity score matching. The dotted vertical lines are at
  $\pm$0.25 and $\pm$0.05, corresponding to What Works Clearinghouse
  standards.}
\label{fig:balance}
\end{figure}

Figure \ref{fig:balance} shows covariate balance before and after the
match.
The red squares in the figure represent covariate standardized differences
between the two groups before adjustment; apparently, $H=1$ and $H=0$
students varied on a number of characteristics.
$H=1$ students tended to have lower pretest scores, were less likely
to be fluent in English, gifted, white, or Asian, and in ninth grade,
but more likely to need special education.

\subsection{Continuous Measure of Hint Use}

As an alternative to treating $H$ as binary, we may model hint requests continuously, via
the IRT parameter $\eta$.
Doing so effectively assumes a stronger version of matched
ignorability: within matched sets potential outcomes $Y(\eta)$ are
independent of $\eta$.
Adding covariates to the regression weakens this assumption, by using
the regression model to adjust for observed confounding between $\eta$
and $Y(\eta)$ left over after the match.

We hypothesized that the effect of variance in $\eta$ may not be
monotonic: that there is an optimal proclivity to request a hint that
is greater than zero.
Students asking for too many  hints may be wheel spinning
\citep[c.f.][]{wheelSpinning} or not putting in sufficient effort to
learn the material, while students who request too few hints are not
taking advantage of a useful learning resource.
To estimate the effect of $\eta$ on posttest
scores---i.e. $Y(\eta=r_1)-Y(\eta=r_0)$ for $r_1$ and $r_2$ two
hypothetical values of $\eta$---we regressed $Y$ on matched set
indicators, the full set of covariates, and penalized splines for
pretest and $\eta$, using the \texttt{gam} function of the
\texttt{mgcv} package in \texttt{R} \citep{mgcv}.

Figure \ref{fig:gam} plots estimates of $\eta$ against adjusted
posttest scores, with the estimated effects of all the regressors
other than $\eta$ subtracted out.
The estimated effect of $\eta$ is plotted as a line, with the dotted
lines giving an approximate 95\% confidence region.
Our hypothesis of a non-monotonic $\eta$ effect was not borne out:
the fitted line is monotonically negative throughout.

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{continuousMatched.pdf}
\caption{Results from a model estimating the effect of continuous hint
  parameter $\eta$ on posttest scores. Each point on the graph plots a
  the posterior mode of $\eta_i$ for subject $i$ in the treatment
  group against $Y_{adj}$, $i$'s posttest score minus the model's
  predictions based on all regressors other than $\eta$. The slope of the
  regression line is the estimated effect of local changes in $\eta$
  on $Y$.}
\label{fig:gam}
\end{figure}

### 2019-09-18 paper.Rnw
knitr and tex files

\documentclass{article}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

%\usepackage{endfloat}

\doublespacing

\newcommand{\yti}{Y_{Ti}}
\newcommand{\yci}{Y_{Ci}}
\newcommand{\uti}{U_{Ti}}
\newcommand{\uci}{U_{Ci}}
\newcommand{\hcont}{\bar{h}}
\newcommand{\etat}{\hcont(1)}
\newcommand{\hcontti}{\hcont(1)_{i}}

\newcommand{\mti}{\bar{m}_{Ti}}
\newcommand{\byt}{\bm{Y_T}}
\newcommand{\byc}{\bm{Y_C}}
\newcommand{\bmt}{\bm{\bar{m}_T}}
\newcommand{\bmi}{\bm{m}_i}
\newcommand{\bsi}{\bm{s}_i}

\newcommand{\EE}{\mathbb{E}}

% see
% https://tex.stackexchange.com/questions/42726/align-but-show-one-equation-number-at-the-end
\newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}

\newenvironment{ass}[2][Assumption:]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}\hskip \labelsep {\bfseries #2}.]}{\end{trivlist}}

\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}

\newcommand{\indicator}[1]{\mathbf{1}_{\left[ {#1} \right] }}

\title{Student Log-Data from a Randomized Evaluation of Educational
  Technology: A Causal Case Study}

\author[1]{Adam C Sales}
\author[2]{John F Pane}
\affil[1]{University of Texas, Austin, TX}
\affil[2]{RAND Corporation, Pittsburgh, PA}

%\affiliation{University of Texas College of Education and RAND Corporation}

<<init,include=FALSE>>=
library(knitr)
opts_chunk$set(
echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
)

@

<<runStuff,include=FALSE,eval=FALSE>>=
source('R/dataAnalysis.r')
big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
save(list=ls()[!big],file='output/analysis.RData')
@

<<loadStuff,include=FALSE>>=
load('output/analysis.RData')
@

\begin{document}
\maketitle

\section{Introduction}
In the 2007--2008 and 2008--2009 school years, the RAND corporation
conducted one of the first effectiveness trials awarded through
competitive grant programs sponsored by the US Department of
Education's Institute of Education Sciences.  The randomized
controlled trial (RCT) was
designed to estimate the effect of school-wide adoption of the
Cognitive Tutor Algebra I (CTA1) curriculum, whose centerpiece is a
computerized ``tutor'' that uses cognitive science principles to teach
Algebra I \citep{anderson1985intelligent}.
The study \citep{pane2014effectiveness} found no effects in its first year, but in
the second year of implementation students in schools randomized to
the CTA1 condition outperformed the control group on the posttest by a fifth of a
standard deviation.

Over the course of the CTA1 study, the software administrators
gathered computer log data for students in the treatment group.
They assembled a dataset that records which problems each student
worked, along with timestamps, the numbers of hints requested, and the
number of errors committed for each problem.
Log data from experiments like this present an unprecedented
opportunity for education researchers:
to use the fine-grained and rich data to
help understand how, why, and when educational technology works.

Along the same lines, log data presents a challenge, since it differs in
structure and size from the type of data commonly encountered in
studies of causal mechanisms.
While statistical models of log data are nearly as old as the data itself
\citep[e.g.][]{corbett1994knowledge}, data from an RCT differs in
two principal ways from more traditional domains.
RCTs feature randomization of treatment assignment, and a post-test
(or some other outcome) administered to both users and non-users.
These factors allow for causal modeling.

As educational technology booms, so do RCTs in the mold of the RAND
CTA1 study. A recent systematic review \citep{escueta2017education} cites 29
published reports of RCTs of ``computer assisted learning'' programs,
all but one of which was published since 2001. We are aware of
numerous other such RCTs planned or ongoing. The current case-study is
about the use of hints in CTA1, but the huge variation in products
under study, each with their own unique designs, opens a wide set of opportunities to understand how features of those
products drive their effects. As one example, RAND is currently
completing an efficacy study of a different algebra tutoring system,
ALEKS. When students begin working on a particular topic in ALEKS, the
system first presents instructional material, followed by problem
solving activities, and then, if students are unsuccessful at solving
the problems, more instructional material. Some students diligently
read the initial instructional material while others skip it and dive
directly into problem solving. Is skipping the material a productive
strategy for students in terms of learning algebra, or should ALEKS be
redesigned to prevent such skipping? Researchers should not stop at
estimating overall average treatment effects, but make use of the
wealth of log data that these RCTs produce automatically, as a
bi-product. Our hope is that this paper can serve as a starting point
for researchers interested in mining this rich resource.

This paper will illustrate three approaches to this type of research
using log data from the CTA1 study---specifically, data on students' use
of hints within the software.
The first approach discards the control group, and analyzes
hint-requesting in the treatment group as an observational study
\citep[c.f.][]{rosenbaum2002observational}. The
second approach applies the framework of causal mediation analysis
\citep{vanderweele2015explanation,hong2015causality,imai2011unpacking} to estimate the role of
hint-requesting in the overall CTA1 treatment effect. However, since
students in the control group were unable to request hints, not all
mediational estimands are identified, and estimation methods must be
modified.
Finally, we will conduct a principal stratification analysis
\citep{frangakis} to learn if students requesting different amounts of
hints experienced different treatment effects.

Our intention is to develop, demonstrate, and contrast these three
approaches to modeling log data.
We hope that this paper can serve as a starting point for
applied educational researchers interested in mining this rich
resource, and that this paper helps spur methodological interest in
this problem.

After a brief overview of the three causal approaches we will discuss,
the next section will provide background for the case-study,
describing the CTA1 curriculum, the RAND study, and our focus on
hints.
The following three sections will present the observational study,
causal mediation, and principal stratification approaches to the hint
data.
Section \ref{sec:synthesis} will compare the three methods directly,
and discuss how they may be applied to data from other educational
technology RCTs.
Section \ref{sec:conclusion} will conclude.

\subsection{Brief Overview of Three Causal Approaches}
\begin{figure}
\centering
\begin{tikzpicture}[scale=1,transform shape]
	%\draw[help lines] (0,0) grid (15,3);
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (1,0) {$H$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(0:3){$Y$};
	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
	%\draw[->,line width=2pt](x)--(z);
	%\draw[->,line width=2pt](x)--(y);
	\draw[->,line width=2pt](z)--(y);
	\node at (2.5,-1) {(a)};

	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (5,0) {$Z$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(8,0){$Y$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (m) at (6.5,2){$H$};
	\draw[->,line width=2pt](z)--(m);
	\draw[->,line width=2pt](m)--(y);
	\draw[->,line width=2pt](z)--(y);
	\node at (5.4,1.2) {(1)};
	\node at (7.6,1.2) {(2)};
	\node at (6.5,.3) {(3)};
	\node at (6.5,-1) {(b)};

        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (10,0) {$Z$};
	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(12,0){$Y$};
        \node (mid) at (11,0) {};
        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (h)
        at (11,2) {$H(1)$};
	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
	%\draw[->,line width=2pt](x)--(z);
	%\draw[->,line width=2pt](x)--(y);
	\draw[->,line width=2pt](z)--(y);
        \draw[->,line width=2pt](h)--(mid);
	\node at (11,-1) {(c)};

\end{tikzpicture}
\caption{A schematic representation of three causal models for
  log data: (a) an observational study estimating the effects of hints
  $H$ on posttest scores $Y$, (b) a mediation analysis estimating
  natural direct effects of treatment assignment $Z$ on $Y$ and
  indirect effects mediated by $H$, and (c) principal stratification.}
\label{mediationFig}
\end{figure}

Figure \ref{mediationFig} is a schematic representation of the three
approaches to causal modeling of log data that we will discuss in this
paper.
First, an observational study to estimate the effect of hint usage $H$
on posttest scores $Y$, is represented by figure (a).
This approach, which we will discuss in Section
\ref{sec:observational}, discards the control group from the RCT and
uses techniques from observational studies to determine if and how requesting
a greater number of hints affects post-test scores.
For the sake of illustration, we will use techniques similar to those
discussed in \citet{rosenbaum2002observational},
 \citet{hansen2009b}, and \citet{ho:etal:2007}, combining a propensity
 score matching design with methods for analyzing experiments and
 parametric modeling.
The conceptual and methodological issues we will discuss extend to
other covariate control methods as well.

Next, as represented in Figure (b), we will conduct a mediation analysis, estimating natural
indirect effects of CTA1 treatment assignment $Z$, as mediated by
$H$. That is, we will estimate the component of $Z$'s effect on $Y$ that comes about
due to $Z$'s effect on $H$---path (1)---and $H$'s effect on $Y$, path
(2).
Notice that path (2) is identical to figure (a)---the effect of $H$ on
$Y$.
We will use this equivalence to estimate and interpret indirect
effects.
As part of the mediation analysis, we will also estimate path (3), the
natural direct effect of $Z$ on $Y$---that is, the component of $Z$'s
effect on $Y$ that is not due to hints $H$.

Finally, we will estimate a principal stratification model, as represented in Figure
(c).
In the principal stratification analysis, we do not model $H$ as an
intervention that could, in theory, be randomized.
Instead, we posit a variable ``$H_i(1)$,''  subject $i$'s penchant for
requesting hints \emph{if} $i$ is assigned to the treatment
condition.
Unlike $H$, $H(1)$ is well defined (if
unobserved) for all participants in the RCT.
Principal stratification treats $H(1)$ as an effect moderator or
interaction variable: $H(1)$ does not affect $Y$, but the effect of
$Z$ on $Y$ may vary as a function of $H(1)$.
This analysis will largely follow the method of \citet{aoas}, which in
turn built heavily on \citet{schwartz2011bayesian} and
\citet{jin2008principal}.

All three of these approaches are designed to model intermediate
variables---those variables, like hint usage,
that were possibly affected by $Z$ but were measured prior to $Y$.
Though they were developed for more traditional data scenarios, the
novel scenario of computer log data from an educational technology RCT
presents new and interesting challenges and opportunities.

\section{Case Study: The Role of Hints in the Cognitive Tutor Effect}

\subsection{The Cognitive Tutor}
CTA1 is one of a series of complete mathematics curricula developed by
Carnegie Learning, Inc., which include both textbook materials and an
automated computer-based Cognitive Tutor
\citep{anderson1995cognitive,pane2014effectiveness}.

The computerized tutor was originally built to test the ACT and ACT-R
theories of skill acquisition \citep{anderson2013architecture}.
These are elaborate theories describing the necessary components of
cognitive skills in, say, mathematics or computer programming, and the
process of acquiring those skills.
The Algebra I tutor guides students through a sequence of Algebra I
problems nested in sections within units, and organized by the specific sets of skills they require.
Students move through the curriculum as the tutor's internal learning
model determines that they have mastered the requisite skills.

As students work through specific problems, they receive two types
of instruction from the tutor.
When they err---that is, depart from one of the pre-specified correct
solution paths for that problem---they receive an error message
alerting them to their mistake and helping them return to the correct
solution path.
Students may also request help, in the form of hints, that are also
tailored to the specific context and solution path they are on.

\subsection{Data}\label{sec:data}

Since our goal was to better understand the CTA1
treatment effect, we focused our analysis on data from high school students in
the second year of the CTA1 trial, for whom the treatment effect was
most evident.

We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
Table \ref{tab:covariateBalance} describes the covariates we used, including
missingness information, control and treatment means, and standardized
differences \citep[c.f.][]{kalton1968standardization}.
We singly-imputed missing values with the Random Forest routine implemented
by the \texttt{missForest} package in \texttt{R}
\citep{missForest,rcite},  which estimated the ``out of box'' imputation
errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.

\begin{table}[ht]
\centering
\include{output/covariateTable}
\caption{Missingness information, control (``BaU'' or ``Business as
  Usual'') and treatment (``CTA1'') means, and balance for the
  covariates included in this study, from the high school year two
  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
  categorical variables (Race/Ethnicity, Sex, and Special Education)
  and standardized root mean squared error for Pretest, which is
  continuous. %The overall p-value accounts for clustering at the
  %school level and blocked random sampling.
  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
\label{tab:covariateBalance}
\end{table}

Over the course of the effectiveness study, Carnegie Learning gathered
log data for each worked problem.
The resulting dataset lists the problem name, section, and unit of
each problem, the numbers of hints requested and errors committed, and time-stamps.
Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
Treatment schools with mastery data missing
for 90\% or more students were omitted from the analysis, along with
their entire randomization block.
Of the remaining 2,390 students, 87\% (2,091) had log
data.
The students with missing log data were dropped from the observational
study and mediation analysis, but included in the principal
stratification study.
Problem log data for sections that were not part of the
standard CTA1 Algebra I curriculum and problems worked by fewer than 100
students were omitted from the
dataset.

All told, the main analysis included $n=$5308 students, 2390 of whom were assigned to the CTA1 condition and 2918 of whom were assigned to control.
The students were nested within 116 teachers, in 43 schools across five states.

\subsection{The Role of Hints\footnote{This subsection draws heavily
  on comments from an anonymous reviewer.}}
Students using the Cognitive Tutor Algebra software work algebra
problems on a computer, instead of on paper, as is typical.
Each problem in the software is associated with a set of skills that
the student must master, and a path (or set of paths) to solving the
problem.
As soon as a student errs in working through the problem, the
software provides feedback tailored to the specific mistake, and
designed to guide the student back to a correct solution path.
Similarly, when a student gets stuck, and is unable to determine the
next step in solving the problem, he or she can ask for a hint.
Like the error feedback, these hints are tailored to the specific
algebra skills corresponding to the next step on the solution path.
Thus, hints and error feedback are crucial aspects of the tutor's
pedagogy.

There are good reasons to believe that the availability of
hints---help on demand---may increase student learning, but there is
also reason to be skeptical \cite[See][for an overview of the theory]{aleven2016help}.
First of all, as, e.g., \citet{anderson1995cognitive} points out, ``if
students never receive help of any sort, they are in danger of becoming
permanently stuck on some problem'' (p. 190).
That is, hints give students who are stuck a way to continue working
after getting stuck.
Hints may also boost learning by ``helping students identify relevant features in
problems'' \citep[][p. 6]{aleven2016help},  that is, point students
towards aspects of the problem that are ``most important for the goals
of learning'' \citep[][p. 782]{koedinger2012knowledge}.
The final available ``bottom out'' hint typically gives the student
the answer.
Students who bottom out then see a worked example, which can be an
effective tool for learning \citep[e.g.][]{sweller1985use}.
Finally, hints can help improve students' meta-cognitive states, such
as their self-knowledge of skills.

On the other hand, all of these theoretical mechanisms rely on
students to play their part.
In order for hints to be instructive, students need to reflect on
their content, for instance with ``self-explanation''
\citep[c.f.][]{renkl1998learning}.
In the same vein, hints need to be well-designed in order to be
effective \citep{mckendree1990effective}.
This suggests that hints may be helpful for some students but not for
others.
They may even be harmful: as \citet{koedinger2007exploring} put it
(p. 241),
``many lines of research and theory suggest the importance of \dots
withholding information from students so that they can exercise, test,
or reason toward new knowledge on their own.''
\citet{paas1994variability} interpret this dilemma in terms of balancing
``cognitive load,'' that is, optimally allocating a student's
attention to the most relevant aspects of a problem.

The empirical literature on the value of hints in intelligent tutors
is mixed \citep[see][for a summary]{goldin2012learner}.
\citet{anderson1995cognitive} compared the performance of students
randomly assigned to different versions of the Cognitive Tutor with
different feedback structures, and found that students who were able
to request hints finished the lesson faster than those who were not,
but could not detect an effect on learning.
\citet{singh2011feedback} used a similar design, comparing two
versions of the ASSISTments intelligent tutor, one in which hints and
immediate error messages were available, and one which did not provide
feedback. Students in the feedback condition showed larger gains on a
posttest, though it is unclear whether these gains were due to hints
or error feedback.
On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
availability, but instead used statistical models to control for
observed confounding.
Both papers used students' subsequent performance within the tutor to
estimate hint effects.
\citet{beck2008does} found conflicting results using different
analytical methods to control for confounding.
\citet{goldin2012learner} investigated two sources of heterogeneity in
the effect of a hint request: by student and by the ``level'' of the
hint (the first, second, or third ``bottom out''  hint available for a given problem).
They found that, indeed, the effect of asking for a hint varies
between students, but on average it is negative for the first two
hints requested on a problem, but positive for the bottom out hint.

\subsection{Measuring Hint Usage in the CTA1 Data}
Unlike previous studies of hint usage, the CTA1 study was not designed
to estimate the effect of hint usage; its purpose was to estimate the
effect of CTA1 as a package.
Like \citet{beck2008does} and \citet{goldin2012learner}, but unlike
\citet{anderson1995cognitive} and \citet{singh2011feedback}, hint
usage was not randomized.
On the other hand, the CTA1 study includes a posttest, like the
randomized studies but unlike the observational studies cited here.
In this regard, the CTA1 study is similar to other large-scale
evaluations of educational technology which do not randomize usage but
do include a post-test.

The CTA1 study was of a much longer duration than
any study of hint effects that we are aware of: students had access to the tutor for an entire
school year before the posttest.
This factor allows for much more data for each student than shorter
studies, as well as more variability in the types of problems and
contexts in which students may request hints.
This variability, in turn, allows researchers to fit flexible models
and ask nuanced questions about hint usage, for instance, exploring differing
patterns of usage among different students.
This sort of analysis is beyond the scope of the current study, which
focuses on comparing and contrasting causal modeling approaches for
these data.
What is relevant for our purposes is that we may narrow down the type
of hint data we model to align with our research focus and to avoid
some modeling pitfalls.

Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set one if the student requested any
hints on that problem, and zero otherwise.
(More mundanely, we do not have access to data
describing the number of hints available for every problem, which
would hinder interpretation of any other measure of hint usage.)

Lastly, any observational study of hint usage must contend with the
fact that students are unlikely to request hints on problems that they
already know how to solve (other than by accident).
Hence, any measure of hint usage is necessarily also a measure of
student ability---students who know more algebra will almost certainly
request fewer hints.
This leads to two related problems: first, student ability
confounds any statistical relationship between hint request frequency
and posttest scores.
Secondly,
a student's decision to request a hint implies both that she
found the problem at least somewhat difficult, and that, in this case,
she responded to that difficulty by requesting a hint.
Our interest is in the second implication, not the first.
Conversely, the significance of a student's decision to forgo a hint
depends on how challenging he or she finds the problem.
Hint request frequency combines students' hint requests (or non-requests) from both
problems they find challenging and trivial, so is a poor measure of
students' use of the hint feature.

Ideally, we would solve this problem by only including data on
problem-student pairs in which the problem was challenging to the
student.
Of course, no direct measure of problem challenge is available.
Instead, we considered a problem to be challenging if the student
either requested a hint or made an error on that problem.
Then, let
\begin{equation*}
\bar{h}_i=\frac{\text{\# problems on which }i\text{ requested a
    hint}}{\text{\# problems on which }i\text{ either requested a hint
    or made an error}}
\end{equation*}
measure hint frequency.

This measure is inversely related to the ratio of the percentage (of all
problems) a student makes an error without having requested a hint,
to the percentage of problems a student requests a hint
\footnote{Technically, if $h$ is the event that a student requests
a hint on a problem and $e$ is the event that the student makes
an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(h\text{ and not }e)/Pr(h)\}$}.
Making an error on a problem without having requested a hint reflects
a hesitance to request a hint when one might have been warranted.
$\bar{h}$ ignores the class of challenging problems on which students
are challenged and figure the answer out (or guess correctly) without
requesting a hint.
This class of problems is inherently interesting to our question;
however, it is unidentified.

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)} of students as
high-hint users, so we chose $c=0.6$, approximately the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

\section{An Observational Study Within an Experiment}\label{sec:observational}

How does requesting hints within CTA1 affect learning?
More precisely, would students have achieved higher posttest scores
had they requested hints more (or less) often?

Let $Y_i$ represent subject $i$'s posttest score, and let $Z_i$ represent
$i$'s treatment status (i.e. the CTA1 or control group).
Following \citet{neyman} and \citet{rubin}, let $Y_i(Z=1)$ be
the posttest score student $i$ would exhibit were $i$ (perhaps
counterfactually) assigned to the treatment condition, and
$Y_i(Z=0)$ the score were $i$ assigned to control.
Then the CTA1 treatment effect for student $i$ is $Y_i(Z=1)-Y_i(Z=0)$.
A similar structure applies to hint request for students in the
treatment group.
For these students, let the potential outcomes be $Y_i(Z=1,H=1)$ and
$Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
For a study measuring hint requests as a continuous variable, we may
define $Y_i(\hcont)$, the posttest score $i$ would exhibit were $i$ to
request hints with proclivity $\hcont$.
This structure implicitly assumes ``non-interference,'' that one
student's treatment assignment $Z$ or hint proclivity $H$ does not
affect another student's posttest scores.

Individual treatment effects are (typically) unidentified, since only
one of the relevant potential outcomes is ever observed for each
subject.
For instance, $Y(H=1)$ is unobserved for members of the
treatment group with $H=0$.
If $Pr(H_i=1)$ were known for each $i$ (as would be the case if $H$
were randomized) then the average treatment effect (ATE) of $H$ on $Y$,
$\EE[Y(H=1)]-\EE[Y(H=0)]$, could be estimated without bias or further
assumptions.
Of course, the distribution of $H_i$ is unknown, and is presumably a
function of $i$'s individual characteristics.
However, if these relevant characteristics were all measured and
available, the ATE of $H$ on $Y$ would be identified.

Let $\bm{x}_i$ represent a vector of baseline covariates for subject $i$.
In our study of hints, $\bm{x}$ includes indicators for state, school
id, class id, grade (9th or higher), race (White/Asian,
Black/Multi-Racial, or Hispanic/American Indian/Alaskan Native),
special education status, economic disadvantage, and English language
learner status, as well as pre-test scores.
Then we assume \citep[c.f.][]{rosenbaum1983central}:
\begin{ass}{Strong Ignorability}
\begin{equation*}
 \{Y(H=1),Y(H=0)\}\independent H |\bm{x}
\end{equation*}
\end{ass}
that conditional on $\bm{x}$, potential outcomes $Y(H=1)$ and $Y(H=0)$
are independent of realized $H$.
Under strong ignorability, one may compare subjects with identical
covariates $\bm{x}$ to estimate causal effects.

Unfortunately, this sort of exact matching is impossible in our finite
sample.
Instead, we estimated ``propensity scores'' \citep{rosenbaum1983central}:
\begin{equation*}
\pi_i=Pr(H_i=1|\bm{x})
\end{equation*}
the probability of a treatment-group subject requesting frequent hints, conditional on
his or her covariate vector $\bm{x}$.
\citet{rosenbaum1983central} shows that conditioning on $\pi$ is
equivalent to conditioning on $\bm{x}$.
We estimated $\pi$ using a Bayesian multilevel logistic regression,
with weakly informative priors, using the \texttt{rstanarm} package in
\texttt{R} \citep{rstanarm}.
The propensity score model was:
\begin{equation*}
\pi_i=logit^{-1}(\alpha+\bm{\tilde{x}}\bm{\beta}+\epsilon^{state}+\epsilon^{school}+\epsilon^{class})
\end{equation*}
where $\bm{\tilde{x}}$ is the $\bm{x}$ vector excluding indicators for
state, school, and class, which were each included using random
intercepts modeled as normally distributed.

We constructed an optimal full matching design \citep{hansen2004} to
condition on estimated propensity scores $\hat{\pi}$ using the
\texttt{optmatch} package in \texttt{R} \citep{optmatch}.
We matched $H=0$ subjects to $H=1$ subjects in such a way as
to minimize the overall distance between subjects in matched sets in
the logit-transformed propensity scores, $logit(\hat{\pi})$.
We constrained the match so that students could only be matched within
schools.
This was for two reasons: first, schools determine a number of factors
that in turn may impact posttest scores, including baseline
student characteristics, pedagogical styles, and CTA1 usage patterns
\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
analysis in the following section.
Each matched set was allowed to contain any positive number of $H=1$
and any positive number of $H=0$ subjects, resulting in matches of
variable size and composition.
For instance, one matched set included 30 $H=1$ subjects matched to a
single $H=0$ subject, and another matched 21 $H=0$ subjects to a
single $H=1$ subject.

Table \ref{fig:balance} in the online appendix shows that
matching largely eliminated mean differences in pre-treatment
covariates between $H=1$ and $H=0$ students.
The only notable exception is in pretest scores: $H=1$ students tended
to have slightly higher pretest scores than their matched comparisons
with $H=0$.
The p-value from an omnibus covariate balance test
\citep{covBal} is 0.26.

Let $m_i$ be a categorical variable denoting the matched set that $i$
belongs to.
Then we may re-state the ignorability condition with reference to this
match \citep[c.f.][]{rebar}:
\begin{ass}{Matched Ignorability}
\begin{equation*}
 \{Y(H=1),Y(H=0)\}\independent H |m
\end{equation*}
\end{ass}
% Matched ignorability formalizes a ``latent experiment'' design: by
% assuming matched ignorability, we imagine a randomized trial in which
% members of the CTA1 group were divided into strata defined by $m$, and
% randomized to $H=1$ or $H=0$ within those strata.
Under matched ignorability, the mean of the outcomes between $H=1$ and
$H=0$ students within each match is unbiased for the average treatment
effect for students in that match.
The (weighted) mean of those estimated treatment effects estimates the
overall treatment effect in the sample.

% The classic approach to estimating effects from blocked randomization
% designs is a fixed-effects regression \citep{fisher1926arrangement},
% regressing outcomes on an indicator for treatment assignment and a set
% of indicators for matched sets.
% The results from this analysis are displayed on the row labeled
% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
% errors and 95\% confidence intervals.
% They suggest that a greater-than-median proclivity to request hints
% \emph{hurts} student learning by a large amount, roughly equal to the
% entire treatment effect of CTA1 estimated in
% \citet{pane2014effectiveness}.
Table \ref{tab:matchResults} gives three treatment effect estimates
under matched ignorability.
The rows labeled ATE and TOT weight the matched sets to
estimate the average effect of $H$ over the entire treatment
group and over those members of the treatment group with $H=1$,
respectively.
The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
racial categories, grade, ESL status, economic disadvantage, and sex,
along with linear and quadratic terms for pre-test (the standard error
and confidence interval use the ``HC3'' sandwich estimator,
\citealt{sandwichPackage}).
This regression adjusts for imbalances in these covariates within
matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
the others.

\begin{table}
\centering
\begin{tabular}{crllcll}
&&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
\hline
\multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
&ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
&TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
&ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
\hline
\multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
&[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
\hline
\end{tabular}
\caption{Estimates with and without regression adjustment and
  sensitivity analysis of the effect of hint usage.}
\label{tab:matchResults}
\end{table}

Matched ignorability is a very strong assumption, so these results
should not be taken at face value---especially if weaker students ask for hints at
higher rates, and also score lower on the posttest.
While adjusting for pre-test scores via matching and/or regression
may account for some of that bias, pre-test may not capture all of the
relevant pre-treatment information.
To account for the possibility of unmeasured confounding, we conducted
a sensitivity analysis following the method of \citet{hhh}.
This method imagines a hypothetical missing covariate $U$ and estimates
how the the omission of $U$ alters the estimate
and its standard error.
The result is a ``sensitivity interval'' \citep[c.f.][]{rosenbaum2002observational} that,
with 95\% confidence, contains the true effect accounting for both
sampling uncertainty and uncertainty due to possible confounding.
$U$ is characterized by two sensitivity parameters:
$\rho^2$ is the squared partial correlation between $U$ and $Y$, after
accounting for the observed covariates in the model, and $T_Z$ is the
t-statistic from an ordinary least squares regression of $H$ on $U$, along with the observed covariates.
These sensitivity parameters may be benchmarked by estimating their
counterparts when observed covariates are left out of the model.
The row of Table \ref{tab:matchResults} labeled ``[Pretest]''
imagines a $U$ that predicts $Y$ and $H$ to a similar extent as
pretest.
The resulting sensitivity interval is quite wide, including
unbelievably large negative effects as well as substantial positive
effects.
A slightly less conservative sensitivity analysis, labeled
``[Pretest/2]'' allows $U$ to
predict $H$ and $\rho$ half as well as pretest scores.
This sensitivity interval is wide, but entirely negative.

\subsection{The Observational Study Approach: Interpretation and Discussion}
The observational study approach to log data seeks to estimate the
effect of a usage parameter, estimated from  computer log data, on
outcomes.
This approach does not exploit randomization---in fact, it discards
the control group entirely.
However, it depends heavily on another feature of RCTs: namely, an
outcome measured for all participants, in this case a posttest.
Causal identification derives not from randomization, but from an
untestable ignorability assumption, coupled with a sensitivity analysis.
In our example, we operationalized strong ignorability with a
propensity-score matching design, but other observational study
designs may apply as well.

We found that hint requests have a negative effect: students who tend
to request hints more often than the median score lower on the
posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after control for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.

Despite being hobbled by the lack of randomization, and the resultant
threat of confounding, the observational study approach provides weak
evidence that hints are harmful to posttest scores---a potentially
important finding.
This result, if confirmed in other scenarios and other analyses, may
inform student behavior in the tutor and future tutor designs.
However, since it ignores the control group from the CTA1 RCT, it does
not shed light on the CTA1 causal mechanisms---it does not explain the
CTA1 effect.
However, its design and results figure prominently in mediation
analysis, which does.

\section{Mediation Analysis}\label{sec:mediation}
The goal of causal mediation analysis is to decompose a treatment
effect into an ``indirect'' or ``mediated''
effect and a ``direct'' effect.
The indirect effect captures the component of the effect that is due
to the mediator: the treatment affected the outcome by affecting the
mediator, which itself, in turn, affected the outcome.
In contrast, the direct effect is the component of the effect that
operates via other mechanisms.

More precisely, express a subject's potential outcomes as a function
of two variables, treatment assignment $Z$ and hint requests $H$:
$Y(Z,H)$.
Now, if $Z$ and $H$ are both binary, as in our example, subjects
each have four potential outcomes:
\begin{equation*}
Y(0,0);\;Y(1,0);\;Y(0,1);\;Y(1,1)
\end{equation*}
representing the outcomes they would exhibit were they assigned to
control and requested few hints, assigned to
treatment and requested few hints, assigned to control and requested
many hints, or assigned to treatment and requested many hints.
In the CTA1 trial, however, the potential outcome $Y(0,1)$ is
meaningless: students in the control group had no access to CTA1, and
therefore could not request hints.
This will be an important factor going forward.

Since $H$ is itself affected by treatment assignment, it
too has potential values:
\begin{equation*}
H(0);\;H(1)
\end{equation*}
representing the hints that would be requested under treatment and
control.
In the CTA1 study, $H(0)=0$ for all subjects.

Combining potential values for $Y$ and for $H$ yields the fundamental
building blocks of causal mediation analysis
\citep[e.g.][]{vanderweele2015explanation,sales2017mediation}.
Let $Y(z,H(z'))$
represent the outcome that a subject would express given treatment
assignment $z$, but the hint behavior that he or she would have
expressed under assignment $z'$.
When $z=z'$, the original two potential outcomes for $Y$ emerge:
$Y(1,H(1))=Y(Z=1)$ is the outcome exhibited under the treatment, when $H$
takes its potential treatment value; similarly, $Y(0,H(0))=Y(Z=0)$. On
the other hand, $Y(1,H(0))$ and $Y(0,H(1))$ are strictly
counterfactual. $Y(1,H(0))$ gives the outcome value
that would result for a subject assigned to the treatment condition
but who nevertheless requested hints as he or she would have under
control. $Y(0,H(1))$ gives the potential outcome for a subject
assigned to control but who nevertheless requested hints as he or she
would have under treatment.
This last quantity is problematic in our case, since $H(1)$ can take the value 1,
and as we have seen $Y(0,1)$ is undefined.

This framework facilitates precise definitions of direct and indirect
effects; in fact, there are two versions of each \cite[e.g.][]{imai2011unpacking}:
\begin{align*}
\delta(1)&=Y(1,H(1))-Y(1,H(0))\\
\delta(0)&=Y(0,H(1))-Y(0,H(0))\\
\end{align*}
are the indirect effects, contrasting potential outcomes when $H$
varies as it would with varying treatment assignment, but holding the
assignment itself constant at either 1 or 0.
\begin{align*}
\xi(1)&=Y(1,H(1))-Y(0,H(1))\\
\xi(0)&=Y(1,H(0))-Y(0,H(0))\\
\end{align*}
represent the direct effects: holding the value of $H$ constant,
either at its potential value under treatment or control, while
varying treatment assignment.
The total treatment effect, $Y(1,H(1))-Y(0,H(0))$, can then be
decomposed in two ways: as $\delta(1)+\xi(0)$ or as
$\delta(0)+\xi(1)$.
These decompositions can potentially reveal the role hints play in the
CTA1 treatment effect: $\delta(1)$ gives the extent to which
assignment to CTA1 affects a student's post-test score by (possibly)
causing him or her to request hints.
$\xi(0)$ gives the effect of assignment to CTA1 if hints were,
perhaps counterfactually, held at their control level, $H=0$.

\subsection{Identification and Estimation of Direct and Indirect
  Effects}

Since the potential outcome $Y(0,1)$ is undefined in the CTA1 hints
study, any expression (potentially) including $Y(0,1)$ is also
undefined; this includes $\delta(0)$ and $\xi(1)$.
Indeed, since hint behavior is unobserved in the control group
altogether, conventional approaches to mediation analysis do not
apply.
Further, were we to operationalize hint behavior as a continuous
variable, as may seem natural, mediation analysis in this case would
also be nearly impossible.
This is because students assigned to the control group request exactly
zero hints, whereas merely nine members of the treatment group
requested no hints over the course of the study, and these students
barely used the tutor at all.
That being the case, the data provide little to no information about
the distribution of potential outcomes when the software is used
but no hints are requested.

Our solution was to dichotomize hint requests with a median split, as
described above, essentially re-defining the causal question.
Half of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the (treatment group)
median.
To make sense of the respective roles of $H$ and $\hcont$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\hcont$.
In the control group all students are assigned $H=0$ and $\hcont=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\hcont$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\hcont=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
individual level, as $\delta(1)_i$ and $\xi(0)_i$.
However, as in the usual causal inference case, individual effects are
not identified without strong assumptions, and the target of
estimation is instead taken to be super-population average effects,
$\EE[\delta(1)]$ and $\EE[\xi(0)]$. Our approach to estimating these
is to estimate the means of each of the component potential outcomes,
$\EE Y(1,H(1))$, $\EE Y(1,H(0))$, and $\EE Y(0,H(0))$.
The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
Y(0,H(0))=Y(Z=0)$ are non-parametrically identified due to the
randomization of treatment assignment.

On the other hand, randomization alone does not identify the strictly counterfactual
potential outcome $\EE Y(1,H(0))$.
In a typical mediation analysis of an RCT, identification of the
distribution of $H(0)$ in the population would derive from
randomization, along with an observation of $H$ in the control group.
In contrast, in the CTA1 study, $H(0)$ is not observed but is
determined by the experimental design: since the CTA1 software was
unavailable to the control group, $H(0)=0$ for all experimental
participants.
Therefore, $Y(1,H(0))\equiv Y(1,0)$.
For treated subjects $i$ for whom $H_i=0$, then, $Y_i(1,H(0))=Y_i$,
the observed outcome is equal to the counterfactual outcome.
For these subjects, there is no indirect effect.

Treated subjects with $H=1$ present more of a challenge.
To estimate indirect effects for these subjects, we may rely on the
same causal setup as in the observational approach:
matched ignorability and the matching design of section
\ref{sec:observational}.
If $Y_i(1,H(0))$ is independent of $Z_i$ within matched sets, then we
may estimate counterfactual potential outcomes for $H=1$ subjects
using the observed outcomes of their matches with $H=0$.
Specifically, let
\begin{equation*}
\hat{Y}_i(1,H(0))=\begin{cases}
 Y_i & H_i=0\\[3ex]
\frac{\displaystyle\sum_{j:M_j=M_i}
  Y_j(1-H_j)}{\displaystyle\sum_{j:M_j=M_i} 1-H_j} & H_i=1
\end{cases}
\end{equation*}
For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
average outcome of their matched pairs with $H=0$.
Then, the average indirect effect in the treatment group can be
estimated as:
\begin{equation*}
 \hat{\Delta}\equiv\widehat{\EE[\delta(0)|Z=1]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
\end{equation*}
(Typically, the overall average indirect effect, $\EE[\delta(0)]$, is
estimated instead of the average indirect effect of the treatment
group, $\EE[\delta(0)|Z=1]$.
We estimated the latter since data from the control group was not used
in our estimation strategy, and estimating the overall average would
simply complicate our standard error estimation.)

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the the difference between the average outcome for students in
the treatment group with $H=1$, and the average outcome under the
counterfactual scenario in which they all had $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
\end{equation*}
$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
\end{equation*}
This can be re-written (omitting conditioning on $Z$):
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
\end{align*}
Since $Pr(H=1)=1/2$ ($H$ was defined with a median split),
$\hat{\Delta}=TOT/2$.
Estimated thusly, the indirect effect is negative: $-0.11$, with
a standard error of 0.03.
This analysis suggests that the availability of excessive hints
actually \emph{lowers} the treatment effect.
CTA1 works not because of hints, but despite them.

What would the effect be without hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct effect of $0.16\pm
0.24$.
Including covariates in the regression produces a direct effect of
$0.19\pm 0.21$.

\subsection{Mediation Analysis: Interpretation and Discussion}
Here we have used mediation analysis as a formal framework to
interpret the results of the observational study in Section
\ref{sec:observational} in terms of the overall CTA1 treatment effect.
Doing so requires some conceptual and statistical acrobatics, such as
relying on a dichotomized measurement of hint requests and strictly
counterfactual potential outcomes, but can nevertheless yield
interesting and interpretable results.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.

Like the observational study, these results are subject to
confounding; indeed, the sensitivity analysis in the Section
\ref{sec:observational} suggests that these results, too, may be
explained due to a missing confounder.

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well; modeling hint requests continuously with the $\hcont$ parameter
from (\ref{eq:rasch}) gives potential values $\hcont(1)$ and $\hcont(0)$,
though only $\hcont(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\hcont(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\hcont(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\hcont$ and posttest scores, the
focus of PS is the relationship between $\hcont$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\hcont(1)=r]$,
despite the fact that $\hcont(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\hcont(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\hcont$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\hcont$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\hcont$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

Our approach to PS estimation is model-based and Bayesian, following
\citet{page2012principal}, \citet{feller2016compared},
\citet{schwartz2011bayesian} and others.
The PS model consists of three sub-models, all depending on a
 vector of parameters $\bm{\theta}$ that contains regression
 coefficients, variance components, and treatment effects.
The sub-models are:
$p(h;\eta(1),\bm{\theta})$, giving the distribution of actual hint
requests as a function of $\eta(1)$, $p(\eta(1)|\bm{x},\bm{\theta})$
giving the distribution of $\eta(1)$ conditional on covariates, and
$p(Y|\eta(1),Z,\bm{x},\bm{\theta})$ giving the conditional
distribution of outcomes.
With these three models in hand, and a prior distribution
$p(\bm{\theta})$, posterior inference for $\bm{\theta}$ proceeds based
on the following structure:
\begin{align*}
p(\bm{\theta}|\bm{Y},\bm{Z},&\bm{X},\{\bm{h}_i\}_{i:Z_i=1})\propto\\
p(\bm{\theta})&\displaystyle\prod_{i: Z_i=1}\int p(Y|\bm{x}_i,Z=1,\bm{\theta},\eta(1))p(\bm{h}_{i}|\eta(1)(1),\bm{\theta})p(\eta(1)|\bm{x}_i,\bm{\theta})d\eta(1)\\
\times&\displaystyle\prod_{i:Z_i=0} \int p(\yci|\bm{x}_i,\bm{\theta},\eta(1))p(\eta(1)|\bm{x}_i,\bm{\theta})d\eta(1).\numberthis\label{eq:posteriorEta}
\end{align*}
where $\bm{h}_i$ is the vector of hint request data for subject $i$,
ranging over all challenging problems.
In other words, we estimate parameters by integrating over unknown
(latent) $\eta(1)$ values, using the distribution of $\eta$ estimated
in the treatment group.
This is, in essence, an infinite mixture distribution, with outcome
distribution $p(Y|\cdot)$ and mixing proportions $p(\eta(1)|\cdot)$.
Unlike in typical PS setups, $\eta(1)$ is unobserved for both $Z=1$
and $Z=0$ treatment groups, but is estimated using both $\bm{h}$ and
$\bm{x}$ in the treatment group, but only $\bm{x}$ in the control
group.

\subsection{Identification and Estimation}

First, students' proclivity to request hints is measured by the $\eta$
parameter in (\ref{eq:rasch}).
In the next level, $\eta(1)$ is modeled as a function of baseline
covariates $\bm{x}$:
\begin{equation}\label{eq:rasch2}
\eta(1)|\left(\bm{x}_i,\bm{\theta}\right) \sim
\mathcal{N}\left(\bm{x_i}\bm{\beta}^M+\epsilon^{Mt}_{t[i]}+\epsilon^{Ms}_{s[i]},
\sigma^M\right)
\end{equation}
where $\bm{\beta}^U$ is a vector of coefficients.
Since students were nested within teachers, who were nested within
schools, we included normally-distributed school ($\epsilon_{Us}$) and
teacher ($\epsilon_{Ut}$) random intercepts.
The covariates in the model, $\bm{x}_i$, were detailed in Table
\ref{tab:covariateBalance}; preliminary model checking suggested
including a quadratic term for pretest, which was added as a column of
$\bm{x}_i$.

We modeled students' post-test scores $Y$ as
conditionally normal:
\begin{multline}\label{eq:outcomeSubmodel}
 Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim \\ \mathcal{N}\left(
\beta^Y_{0b[i]}+\bm{x}_i^T\bm{\beta}^Y+a\eta(1)+Z_i\tau(\eta(1))+\epsilon^{Yt}_{t[i]}+\epsilon^{Ys}_{s[i]},\sigma^Y_{Z[i]}\right)
\end{multline}
where $\beta^Y_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\beta}^Y$ are the
covariate coefficients, and $\epsilon^{Yt}$, and
$\epsilon^{Ys}$ are normally-distributed teacher and school random
intercepts.
The residual variance $\sigma^Y$ varies with treatment assignment $Z$;
this captures measurement error in $Y$, treatment effect heterogeneity
that is not linearly related to
$\eta(1)$, and other between-student variation in $Y$ that is not predicted by
the mean model.

Finally, we modeled treatment effects
$\tau(\eta(1))$ as linear:
\begin{equation}\label{eq:tauModel}
\EE[Y_{T}-Y_{C}|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)
\end{equation}
While more complex models for $\tau(\eta(1))$ are theoretically
possible (for instance, \citet{jin2008principal} uses a quadratic
model), in our experience non-linear effect models do not perform as
well in model checks as the the linear model (\ref{eq:tauModel}).

Covariates $\bm{X}$ were standardized prior to fitting.
Prior distributions for the block fixed effects $\beta^Y_b$ and covariate coefficients
$\bm{\beta^Y}$ and $\bm{\beta^M}$ were normal with mean zero and
standard deviation 2;
priors for treatment effects and the coefficient on $\eta(1)$ were standard
normal.
The rest of the parameters received Stan's default uniform priors.
In all cases, we expected true parameter values to be much smaller in
magnitude than the prior standard deviation.

Figure \ref{fig:psResults} shows the estimated linear function
$\hat{\tau}{\eta(0)}$.
The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
$\tau(1)$ varies; the black line shows the estimate (i.e. posterior
mean) and the red lines are random draws from the posterior
distribution, showing estimation uncertainty.
Eighty-eight percent of posterior draws of the slope of the
$\hat{\tau}(\cdot)$ function were positive, implying a posterior
probability of 0.88 that students who asked for hints
with greater regularity benefited more from the CTA1 curriculum.
The left hand panel plots one posterior draw of the $\bm{\eta(1)}$
alongside observed outcomes $Y$.
The figure also includes estimated regression lines for the two
treatment groups.
Although, as observed in Sections \ref{sec:observational} and
\ref{sec:mediation}, $\eta(1)$ is anticorrelated with $Y$, the slope
is steeper in the control group than in the treatment group.
The distance between the two regression lines is the treatment effect,
which grows with $\eta(1)$.

\begin{figure}
\centering
\includegraphics[width=0.95\textwidth]{output/psModel.pdf}
\caption{Results from the PS model. On the left, the estimated
  CTA1 treatment effect is plotted against hint request proclivity
  $\eta(1)$. The black line is the posterior mean, and the red lines
  are random posterior draws. The right panel plots observed posttest
  scores against one random draw of $\eta(1)$, along with regression
  lines for the treatment and control groups.}
\label{fig:psResults}
\end{figure}

\subsection{Principal Stratification: Interpretation and Discussion}
Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Unlike those two methods, PS understands $\eta(1)$ as a
student-specific baseline characteristic---albeit one that only
manifests in the treatment group.
Students who request more hints need more help, and thus score lower
on the posttest.
On the other hand, the PS analysis suggests that it's these students
who also receive more help---perhaps due to CTA1 hints, or other
features of the curriculum---and hence benefit more from CTA1.

Can these results be reconciled with the results from the previous two
sections?
Does requesting hints help or hurt?
Technically, PS has nothing to say about the effect of hints---$\eta(1)$
is modeled as a student characteristic, not an intervention.
(See \citealt{jin2008principal} for a nice discussion of this point.)
Still, doesn't the correlation between treatment effects and hint
requests suggest that hints may be beneficial?
Does that contradict the results in Sections \ref{sec:observational}
and \ref{sec:mediation} that hints requests lower students' posttest
scores?
In fact, the results of the three methods can be reconciled.
For instance, it may be that requesting hints indeed lowers test
scores---a negative indirect effect---but that those students who are
likely to request more hints also tend to experience greater direct
effects.
That is, it may be that the observational and mediational results are
correct---that requesting hints lowers test scores---but that the PS
results are also correct---students who request more hints tend to
experience higher treatment effects, due to other mechanisms.

On the other hand, the observational and mediational results may be
due not to causation but to confounding.
In contrast, PS does not depend on an ignorability assumption beyond
what is guaranteed by randomization.
Perhaps by recognizing $\eta(1)$ as hopelessly confounded and focusing
attention elsewhere, PS is able to arrive at the more reliable
conclusion that---whatever the mechanism may be---students who ask
more hints benefit more from the software.

Though PS does not depend on ignorability, it does depend on a highly
intricate and parametric modeling structure.
If any part of the model is sufficiently misspecified, the results may
be wrong.
Perhaps worse, \citet{feller2016principal} argued that even well-specified
models can give severely biased estimates.
Extensive model interrogation and checking is necessary for PS
analysis, but the complexity of the PS model makes this process
particularly difficult.
\citet{aoas} gives some examples of potentially fruitful model
checking procedures.

In contrast, the matching estimator in Section \ref{sec:observational}
is highly transparent.
Whatever intricacies went into formulating and fitting a propensity
score model or matching procedure are inconsequential if they produce
a convincing match.
The validity of the matching estimate rests entirely on the quality of
the match, which can be examined by checking covariate balance and
other relatively simple metrics.
The uncertainty in matching estimators due to possible confounding
might be offset by the simplicity in modeling assumptions.

\section{Which Strategy is Best?}

\begin{table}
\begin{tabular}{p{1in}p{2in}p{2in}}
& Observational Study/Mediation Analysis &Principal Stratification\\
\hline
Research question & Does $H$ help $Z$ affect $Y$? & Does $H$ help predict
                                           treatment effect of $Z$ on $Y$?\\
\hline
Role of $H$ & Causal & Predictive\\
\hline
Untestable assumption & $H$ is unconfounded & Distribution of $Y$
                                              given $H_T$ in control
                                              group\\
\hline
Other difficulties & Discrete $H$ necessary for mediation (in our
                     framework) & Model fitting can be opaque and
                                  erratic\\
\end{tabular}
\end{table}

\section{Discussion: The Essential Role of Measurement}\label{sec:conclusion}
This case study has focused on three approaches to causal modeling of
hint requests during an RCT of the Cognitive Tutor program.
We hope we have demonstrated some of the potential and some of the
challenges that new computerized data bring to these old problems.

In the hints case study, the observational study and mediation
analysis differ from principal stratification in their approach to
potential confounding of the statistical relationship between hint
requests and posttest scores.
Matching-based estimators attempt to adjust for confounding,
and sensitivity analysis accounts for confounding.
In contrast, principal stratification includes it in the model,
estimating the relationship between (potential) hint usage and
treatment effects, as opposed to the relationship between hints and
outcomes.

A second challenge is extrapolation in mediation analysis.
In RCTs comparing computerized educational technology with traditional
pedagogy, members of the control group do not have computer log data
or access to the mechanisms that the log data measures.
This forms a serious barrier to mediation analysis.
However, by dichotomizing the mediator and matching, we were able to
re-define the causal question in such a way as to make it (possibly) answerable.

Lurking behind all three cases is the equally important issue of
measurement.
This issue arose in our case study in our decision to use an IRT model
to measure hint usage, instead of a simple proportion.
Doing so is an important step in the right direction, of thinking
carefully about measurement issues and using as much available data as
possible.
The Rasch model also opens the door to a host of exciting
possibilities in causal models of log data.
Specifically, other model-based dimension reduction or cluster analysis techniques
could be brought to bear, defining new intermediate variables.

% \section*{Acknowledgments}
% This material is based upon work supported by the National Science Foundation under Grant Number 1420374. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the National Science Foundation.

\bibliographystyle{plainnat}
\bibliography{ct}

\end{document}

### 2019-09-18 paper.tex
knitr and tex files

--- 2019-09-18 paper.Rnw
@@
 \documentclass{article}
-\usepackage[stable]{footmisc}
 \usepackage{fullpage}
@@
 \newcommand{\uci}{U_{Ci}}
-\newcommand{\hcont}{\bar{h}}
-\newcommand{\etat}{\hcont(1)}
-\newcommand{\hcontti}{\hcont(1)_{i}}
+\newcommand{\etat}{\eta(1)}
+\newcommand{\etati}{\eta(1)_{i}}
 
@@
 %\affiliation{University of Texas College of Education and RAND Corporation}
-
-<<init,include=FALSE>>=
-library(knitr)
-opts_chunk$set(
-echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
-)
-
-@
-
-<<runStuff,include=FALSE,eval=FALSE>>=
-source('R/dataAnalysis.r')
-big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
-save(list=ls()[!big],file='output/analysis.RData')
-@
-
-<<loadStuff,include=FALSE>>=
-load('output/analysis.RData')
-@
 
@@
 
+The CTA1 effectiveness trial is a prominant example of a large and
+growing number of randomized evaluations of educational technology
+products (see \citealt{escueta2017education} for a recent systematic review).
+In many, if not all, experimenters were able to
+gather log data from students and/or teachers in the treatment group,
+capturing the varying ways they used the technology.
+These datasets could go a long way in helping researchers and
+developers better understand how or for whom the software works.
+
 This paper will illustrate three approaches to this type of research
@@
 this problem.
+Ulitimately, we believe that education researchers should not stop at
+estimating overall average treatment effects, but make use of the
+wealth of log data that these RCTs produce.
 
@@
 $Y$.
-We will use this equivalence to estimate and interpret indirect
+We will use this equivalency to estimate and interpret indirect
 effects.
@@
 Students may also request help, in the form of hints, that are also
-tailored to the specific context and solution path they are on.
+tailored to the specific context and solution path of the problem.
+
+\subsection{The Role of Hints}\footnote{This subsection draws heavily
+  on comments from an anonymous reviewer.}
+Students using the Cognitive Tutor Algebra software work algebra
+problems on a computer, instead of on paper, as is typical.
+But the Cognitive Tutor is not simply an online repository of
+problems.
+Each problem in the software is associated with a set of skills that
+the student must master, and a path (or set of paths) to solving the
+problem.
+As soon as a student errs in working through the problem, the
+software provides feedback tailored to the specific mistake, and
+designed to guide the student back to a correct solution path.
+Similarly, when a student gets stuck, and is unable to determine the
+next step in solving the problem, he or she can ask for a hint.
+Like the error feedback, these hints are tailored to the specific
+algebra skills corresponding to the next step on the solution path.
+Thus, hints and error feedback are crucial aspects of the tutor's
+pedagogy.
+
+On the other hand, the ready availability of hints may prevent some
+students from trying to figure out algebra problems themselves,
+instead asking for a hint as soon as things get hard.
+If hints prevent student effort, they may be detrimental---after all,
+students working through traditional worksheets or textbook problems
+do not have recourse to hints and may put more effort into their
+schoolwork.
+\citet{koedinger2007exploring} refer to this as the ``assistance
+dilemma.''
+
+The RAND CTA1 effectiveness trial provides a valuable opportunity to
+study the role of hints---on average, do they tend to help students
+learn or to prevent them from learning?
+What role does the availability of hints play in CTA1's effectiveness
+when compared to traditional instruction?
+Does requesting hints make CTA1 more effective or less?
 
@@
 \centering
-\include{output/covariateTable}
-\caption{Missingness information, control (``BaU'' or ``Business as
-  Usual'') and treatment (``CTA1'') means, and balance for the
-  covariates included in this study, from the high school year two
-  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
+\include{covariateTable}
+\caption{Missingness information and balance for the covariates included in this study, from
+  the CTA1 Effectiveness experiment, high
+  school, year two. Imputation error is percent falsely classified for
   categorical variables (Race/Ethnicity, Sex, and Special Education)
@@
 
-\subsection{The Role of Hints\footnote{This subsection draws heavily
-  on comments from an anonymous reviewer.}}
-Students using the Cognitive Tutor Algebra software work algebra
-problems on a computer, instead of on paper, as is typical.
-Each problem in the software is associated with a set of skills that
-the student must master, and a path (or set of paths) to solving the
-problem.
-As soon as a student errs in working through the problem, the
-software provides feedback tailored to the specific mistake, and
-designed to guide the student back to a correct solution path.
-Similarly, when a student gets stuck, and is unable to determine the
-next step in solving the problem, he or she can ask for a hint.
-Like the error feedback, these hints are tailored to the specific
-algebra skills corresponding to the next step on the solution path.
-Thus, hints and error feedback are crucial aspects of the tutor's
-pedagogy.
-
-There are good reasons to believe that the availability of
-hints---help on demand---may increase student learning, but there is
-also reason to be skeptical \cite[See][for an overview of the theory]{aleven2016help}.
-First of all, as, e.g., \citet{anderson1995cognitive} points out, ``if
-students never receive help of any sort, they are in danger of becoming
-permanently stuck on some problem'' (p. 190).
-That is, hints give students who are stuck a way to continue working
-after getting stuck.
-Hints may also boost learning by ``helping students identify relevant features in
-problems'' \citep[][p. 6]{aleven2016help},  that is, point students
-towards aspects of the problem that are ``most important for the goals
-of learning'' \citep[][p. 782]{koedinger2012knowledge}.
-The final available ``bottom out'' hint typically gives the student
-the answer.
-Students who bottom out then see a worked example, which can be an
-effective tool for learning \citep[e.g.][]{sweller1985use}.
-Finally, hints can help improve students' meta-cognitive states, such
-as their self-knowledge of skills.
-
-On the other hand, all of these theoretical mechanisms rely on
-students to play their part.
-In order for hints to be instructive, students need to reflect on
-their content, for instance with ``self-explanation''
-\citep[c.f.][]{renkl1998learning}.
-In the same vein, hints need to be well-designed in order to be
-effective \citep{mckendree1990effective}.
-This suggests that hints may be helpful for some students but not for
-others.
-They may even be harmful: as \citet{koedinger2007exploring} put it
-(p. 241),
-``many lines of research and theory suggest the importance of \dots
-withholding information from students so that they can exercise, test,
-or reason toward new knowledge on their own.''
-\citet{paas1994variability} interpret this dilemma in terms of balancing
-``cognitive load,'' that is, optimally allocating a student's
-attention to the most relevant aspects of a problem.
-
-The empirical literature on the value of hints in intelligent tutors
-is mixed \citep[see][for a summary]{goldin2012learner}.
-\citet{anderson1995cognitive} compared the performance of students
-randomly assigned to different versions of the Cognitive Tutor with
-different feedback structures, and found that students who were able
-to request hints finished the lesson faster than those who were not,
-but could not detect an effect on learning.
-\citet{singh2011feedback} used a similar design, comparing two
-versions of the ASSISTments intelligent tutor, one in which hints and
-immediate error messages were available, and one which did not provide
-feedback. Students in the feedback condition showed larger gains on a
-posttest, though it is unclear whether these gains were due to hints
-or error feedback.
-On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
-availability, but instead used statistical models to control for
-observed confounding.
-Both papers used students' subsequent performance within the tutor to
-estimate hint effects.
-\citet{beck2008does} found conflicting results using different
-analytical methods to control for confounding.
-\citet{goldin2012learner} investigated two sources of heterogeneity in
-the effect of a hint request: by student and by the ``level'' of the
-hint (the first, second, or third ``bottom out''  hint available for a given problem).
-They found that, indeed, the effect of asking for a hint varies
-between students, but on average it is negative for the first two
-hints requested on a problem, but positive for the bottom out hint.
-
-\subsection{Measuring Hint Usage in the CTA1 Data}
-Unlike previous studies of hint usage, the CTA1 study was not designed
-to estimate the effect of hint usage; its purpose was to estimate the
-effect of CTA1 as a package.
-Like \citet{beck2008does} and \citet{goldin2012learner}, but unlike
-\citet{anderson1995cognitive} and \citet{singh2011feedback}, hint
-usage was not randomized.
-On the other hand, the CTA1 study includes a posttest, like the
-randomized studies but unlike the observational studies cited here.
-In this regard, the CTA1 study is similar to other large-scale
-evaluations of educational technology which do not randomize usage but
-do include a post-test.
-
-The CTA1 study was of a much longer duration than
-any study of hint effects that we are aware of: students had access to the tutor for an entire
-school year before the posttest.
-This factor allows for much more data for each student than shorter
-studies, as well as more variability in the types of problems and
-contexts in which students may request hints.
-This variability, in turn, allows researchers to fit flexible models
-and ask nuanced questions about hint usage, for instance, exploring differing
-patterns of usage among different students.
-This sort of analysis is beyond the scope of the current study, which
-focuses on comparing and contrasting causal modeling approaches for
-these data.
-What is relevant for our purposes is that we may narrow down the type
-of hint data we model to align with our research focus and to avoid
-some modeling pitfalls.
-
-Since our posttest measured Algebra I skills, we only
-considered worked problems from the Algebra I curriculum.
-Next, although \citet{goldin2012learner} demonstrated the importance
-of hint level, our focus is on the overall average effect of
-requesting any hints; therefore, the variable measuring a student's
-hint request on a problem was set one if the student requested any
-hints on that problem, and zero otherwise.
-(More mundanely, we do not have access to data
-describing the number of hints available for every problem, which
-would hinder interpretation of any other measure of hint usage.)
-
-Lastly, any observational study of hint usage must contend with the
-fact that students are unlikely to request hints on problems that they
-already know how to solve (other than by accident).
-Hence, any measure of hint usage is necessarily also a measure of
-student ability---students who know more algebra will almost certainly
-request fewer hints.
-This leads to two related problems: first, student ability
-confounds any statistical relationship between hint request frequency
-and posttest scores.
-Secondly,
-a student's decision to request a hint implies both that she
-found the problem at least somewhat difficult, and that, in this case,
-she responded to that difficulty by requesting a hint.
-Our interest is in the second implication, not the first.
-Conversely, the significance of a student's decision to forgo a hint
-depends on how challenging he or she finds the problem.
-Hint request frequency combines students' hint requests (or non-requests) from both
-problems they find challenging and trivial, so is a poor measure of
-students' use of the hint feature.
-
-Ideally, we would solve this problem by only including data on
-problem-student pairs in which the problem was challenging to the
-student.
-Of course, no direct measure of problem challenge is available.
-Instead, we considered a problem to be challenging if the student
-either requested a hint or made an error on that problem.
-Then, let
-\begin{equation*}
-\bar{h}_i=\frac{\text{\# problems on which }i\text{ requested a
-    hint}}{\text{\# problems on which }i\text{ either requested a hint
-    or made an error}}
-\end{equation*}
-measure hint frequency.
-
-This measure is inversely related to the ratio of the percentage (of all
-problems) a student makes an error without having requested a hint,
-to the percentage of problems a student requests a hint
-\footnote{Technically, if $h$ is the event that a student requests
-a hint on a problem and $e$ is the event that the student makes
-an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(h\text{ and not }e)/Pr(h)\}$}.
-Making an error on a problem without having requested a hint reflects
-a hesitance to request a hint when one might have been warranted.
-$\bar{h}$ ignores the class of challenging problems on which students
-are challenged and figure the answer out (or guess correctly) without
-requesting a hint.
-This class of problems is inherently interesting to our question;
-however, it is unidentified.
-
-The observational study and (especially) the mediation analysis we
-describe below require a dichotomous measurement of hint usage: is a
-student a high or low hint user?
-To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
-that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
-and 0 otherwise.
-To choose $c$, we fit a two-component modified Rasch mixture model
-\citep{rasch} to hint request data, described in an online appendix.
-The model classified approximately \Sexpr{round(mixProp*100)} of students as
-high-hint users, so we chose $c=0.6$, approximately the
-\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
-This definition agreed with the model's classification in
-approximately \Sexpr{round(agree*100)}\% of cases.
+\section{Measuring Hint Requests}\label{sec:modelingHints}
+In our CTA1 log data, 2091 students requested a total of 801,518 hints on
+926,314 problems.
+Some students never requested a hint, and one student requested over
+three thousand hints; some students requested at least one hint on every problem
+they worked.
+
+Clearly, several factors played a role in this variation.
+First, academically weaker students---those whom we would expect to
+score lower on the posttest regardless of CTA1 use---probably struggle
+with CTA1 problems more often, and probably request more hints.
+This presents a serious confounding threat, which we will attempt to
+address in a number of ways throughout the paper.
+Our first line of defense is a slight restructuring of the causal
+questions, reflected in the measurement stage.
+By ``hint use'' we refer to a student's use of hints on problems that
+he or she finds challenging.
+Specifically, we consider only worked problems on which
+the student either requested a hint or committed an error.
+This effectively filters out the problems that each student finds
+trivial; the fact that stronger students struggle less often is no
+longer a driving factor.
+(In an equivalent formulation, the question addresses each student's
+likelihood of committing an error without having asked for a
+hint---i.e. deciding to attempt a difficult problem without help.)
+This strategy removes some of the confounding between
+student ability and hint requests, though (as we shall see) a
+substantial amount remains.
+
+Factors beyond student ability also play a role in the variation
+in the numbers of hints students requested.
+For one, students who worked a larger number of CTA1 problems also
+requested more hints.
+Secondly, not all problems are equivalent.
+Problems vary in their structure, and hence in the number of hints
+available---some problems allow an indefinite number of hint requests.
+Problems from some sections never elicited hints, whereas problems
+from other sections nearly always did.
+A students' hint requests are a function of their learning
+style within the tutoring software, which parts of the software they
+worked, and how much they worked.
+
+A focus on hint requests, as opposed to these other
+factors, requires some care.
+Modeling hint requests per problem as binary---did a student request
+at least one hint?---puts problems with many or few
+available hints on equal footing.
+Problem difficulty is another matter.
+What is difficult for one student may not be difficult for others.
+That students do not request hints on problems they find trivial is
+uninteresting; what is more interesting is how they react when a
+problem presents a challenge.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.95\textwidth]{rasch.jpg}
+\caption{The proportion of problems in which each student requested a
+  hint, or the IRT student parameter $\eta$ from model
+  (\ref{eq:rasch}), as a function of each student's average section difficulty. For each
+  section, ``section difficulty'' is defined as the proportion of
+  \emph{that section's} worked problems that elicited a hint request,
+  and a student's average section difficulty is the average section
+  difficulty over that student's worked problems.}
+\label{fig:mbar}
+\end{figure}
+
+To capture the effect of hint requesting behavior, as distinct from
+total amount of usage, a first step may be to calculate the proportion
+of each student's challenging problems that resulted in a hint
+request.
+However, this presents additional issues, one of which can be seen in Figure
+\ref{fig:mbar}.
+First, even after dichotomizing hints and restricting our focus to
+instances in which students appear to be challenged, worked problems
+may differ from each other in relevant ways.
+For instance, multi-step problems which provide more occasions for hint
+requests may elicit at least one hint more frequently.
+For each section of the tutor, let ``section difficulty'' be the proportion of that
+section's challenging problems that elicited a hint request; let a
+student's ``average section difficulty'' be the average section
+difficulty over all challenging problems that student worked.
+On the left of Figure \ref{fig:mbar}, the proportion of problems on
+which each student requested at least one hint is plotted against that
+student's average section difficulty.
+It is apparent, if unsurprising, that students who worked on sections of the
+curriculum that typically elicited more hints tended to request hints
+more often.
+The Spearman correlation between average section difficulty and hint
+request proportion is 0.35.
+How do we equate the hint records of students who worked problems in different
+sections?
+
+% Further, hint request proportions are related to the number of
+% challenging problems each student worked, as can be seen on the right
+% panel of Figure \ref{fig:mbar}.
+% This is due to measurement error.
+% The variance of a sample proportion is inversely proportional to the
+% size of the sample; therefore, it is unsurprising that the extremes of
+% hint request proportions are composed primarily of students who worked
+% very few problems.
+% For instance, nearly all students who requested a hint on more than
+% 75\% of the problems that challenged them worked fewer than 250
+% challenging problems.
+
+Item response theory \citep[e.g.][]{irtHandbook} provides an attractive alternative.
+Under a modified Rasch model \citep[e.g.][]{rasch}, for instance,
+\begin{equation}\label{eq:rasch}
+Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
+\end{equation}
+where $h_{ip}$ is the event that student $i$ requests a hint on
+problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
+section-level parameter for $s[p]$, the section that problem $p$ was
+drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
+function.
+Typically, the student parameter $\eta_i$ measures student ability;
+here, instead, it measures a student's proclivity to request a hint.
+The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
+dataset many problems were worked by very few students, so problem
+level parameters would be hard to estimate.
+Instead, we included the section-level parameter $\delta_{s[p]}$,
+which measures the extent to which students tend to request hints on
+problems in section $s$.
+This essentially assumes that all problems in each section are equally
+likely to elicit a hint.
+Replacing student hint request proportions with the posterior mode for
+$\eta$ from model (\ref{eq:rasch}), fit using the \texttt{lme4}
+package in \texttt{R} \citep{lme4}, resolves the section-difficulty conundrum: the right
+side of Figure \ref{fig:mbar}, in which $\eta$ is plotted instead of
+hint request proportion, shows no substantial relationship between
+section difficulty and hint requests.
+
+In some analyses, we dichotomize $\eta$ with a median split.
+Let $H_i=1$ if $\eta_i>median(\bm{\eta})$ and 0 otherwise.
+While in dichotomizing $\eta$ we lose valuable information, doing so
+simplifies the observational matching design of the following section,
+and makes the mediation analysis possible.
 
@@
 For a study measuring hint requests as a continuous variable, we may
-define $Y_i(\hcont)$, the posttest score $i$ would exhibit were $i$ to
-request hints with proclivity $\hcont$.
+define $Y_i(\eta)$, the posttest score $i$ would exhibit were $i$ to
+request hints with proclivity $\eta$.
 This structure implicitly assumes ``non-interference,'' that one
@@
 assumptions.
-Of course, the distribution of $H_i$ is unknown, and is presumably a
+Of course, the distribution of $H_i$ is unknown, and is presumeably a
 function of $i$'s individual characteristics.
@@
 median.
-To make sense of the respective roles of $H$ and $\hcont$, the
+To make sense of the respective roles of $H$ and $\eta$, the
 continuous measure on which it is based, we may imagine a process in
 which students are assigned $H$, and under that constraint choose a
-value for $\hcont$.
-In the control group all students are assigned $H=0$ and $\hcont=0$, but
+value for $\eta$.
+In the control group all students are assigned $H=0$ and $\eta=0$, but
 in the treatment group half of students have $H=1$ and half have
-$H=0$, and $\hcont$ varies within those groups.
+$H=0$, and $\eta$ varies within those groups.
 That structure implies a ``treatment mediator interaction''
@@
 Of course, in our example treatment mediator interactions are not
-identified, since $H=0$ and $\hcont=0$ in the control group, and very
+identified, since $H=0$ and $\eta=0$ in the control group, and very
 few treatment students asked for 0 hints.
@@
 As in mediation analysis, hint behavior has potential outcomes as
-well; modeling hint requests continuously with the $\hcont$ parameter
-from (\ref{eq:rasch}) gives potential values $\hcont(1)$ and $\hcont(0)$,
-though only $\hcont(1)$ is relevant in the CTA1 study.
+well; modeling hint requests continuously with the $\eta$ parameter
+from (\ref{eq:rasch}) gives potential values $\eta(1)$ and $\eta(0)$,
+though only $\eta(1)$ is relevant in the CTA1 study.
 
@@
 \begin{equation*}
-\tau(r)=\EE[Y(1)-Y(0)|\hcont(1)=r]
+\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
 \end{equation*}
 This is the treatment effect for subjects who \emph{would} request
-hints at level $\hcont(1)=r$, if assigned to the treatment condition.
+hints at level $\eta(1)=r$, if assigned to the treatment condition.
 Whereas the focus of the observational study and mediation
-analysis was the relationship between $\hcont$ and posttest scores, the
-focus of PS is the relationship between $\hcont$ and treatment effects.
-Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\hcont(1)=r]$,
-despite the fact that $\hcont(1)$ is never observed at the same time as
+analysis was the relationship between $\eta$ and posttest scores, the
+focus of PS is the relationship between $\eta$ and treatment effects.
+Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\eta(1)=r]$,
+despite the fact that $\eta(1)$ is never observed at the same time as
 $Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
 strictly counterfactual quantities, as in mediation analysis.
-$\hcont(1)$ is defined for subjects in both randomized groups, but only
+$\eta(1)$ is defined for subjects in both randomized groups, but only
 observed in the treatment group.
 
-In previous sections $\hcont$ was dichotomized into $H$; such
-dichotomization is not necessary here.
-However, $\hcont$ has some disadvantages as a continuous measure of
-hint usage.
-For one, since it is essentially a sample mean over a subset of each student's
-worked problems, much of its variance is driven by the total number of
-problems students worked---in particular, its extreme values belong to
-those students who barely used the tutor at all.
-Further, it does not account for varying difficulty of the tutor's
-problems.
-As an alternative, we fit a
-modified Rasch model \citep[e.g.][]{rasch},
-\begin{equation}\label{eq:rasch}
-Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
-\end{equation}
-where $h_{ip}$ is the event that student $i$ requests a hint on
-problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
-section-level parameter for $s[p]$, the section that problem $p$ was
-drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
-function.
-Typically, the student parameter $\eta_i$ measures student ability;
-here, instead, it measures a student's proclivity to request a hint.
-The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
-dataset many problems were worked by very few students, so problem
-level parameters would be hard to estimate.
-Instead, we included the section-level parameter $\delta_{s[p]}$,
-which measures the extent to which students tend to request hints on
-problems in section $s$.
-This essentially assumes that all problems in each section are equally
-likely to elicit a hint.
-
-We incorporated the model for hint requests, (\ref{eq:rasch}), into
-the larger principal stratification model, replacing $\eta$ for
-$\hcont$.
-Then, the principal stratification estimand is
-\begin{equation*}
-\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
-\end{equation*}
-A full treatment of this variant of principal stratification, including a
+In typical PS studies, the intermediate variable is modeled as fixed
+and known.
+We chose, instead, to measure hint behavior as a latent variable from
+Rasch model (\ref{eq:rasch}), due to the reasons described in Section
+\ref{sec:modelingHints} and because of the measurement error
+due to the varying number of problems each subject worked.
+A full treatment of this distinction and its implications, including a
 discussion of identification and estimation, may be found in
@@
 \centering
-\includegraphics[width=0.95\textwidth]{output/psModel.pdf}
+\includegraphics[width=0.95\textwidth]{psModel.pdf}
 \caption{Results from the PS model. On the left, the estimated

### 2019-09-18 R/covariateTable.r
R code

@@ -1,0 +1,64 @@
+## covariate means/balance
+miss <- NULL
+for(i in c('race','sex','spec','xirt')) miss <- rbind(miss,
+ c(sum(is.na(covs[[i]])),mean(is.na(covs[[i]])),error[i,'error']))
+miss <- as.data.frame(miss)
+miss$`Error Type` <- c('PFC','PFC','PFC','SRMSE')
+rownames(miss) <- c('Race/Ethnicity','Sex','Special Education','Pretest')
+names(miss)[1:3] <- c('# Missing','% Missing','Imputation Error')
+miss[,2] <- as.integer(round(miss[,2]*100))
+miss[,1] <- as.integer(miss[,1])
+miss['Pretest','Imputation Error'] <- sqrt(miss['Pretest','Imputation Error'])/sd(covs$xirt,na.rm=TRUE)
+
+
+covs2 <- covs[rownames(dat),] ### same as dat but with NAs
+
+balTest <- balanceTest(treatment~race+sex+spec+xirt+cluster(schoolid2)+strata(pair),data=dat,report=c('adj.means','std.diffs','chisquare.test'),include.NA.flags=T)
+
+## use "unstrat" for group means cuz no one knows/cares about "adjusted means"
+trtMeans <- round(balTest$results[,'Treatment','Unstrat']*100)
+ctlMeans <- round(balTest$results[,'Control','Unstrat']*100)
+stdDiff <- sprintf('%.2f',balTest$results[,'std.diff','pair'])
+names(stdDiff) <- dimnames(balTest$results)[[1]]
+
+impErr <- sprintf('%.2f',miss[,3])
+names(impErr) <- rownames(miss)
+
+results <- list()
+results$balP <- sprintf('%.2f',balTest$overall['pair','p.value'])
+
+
+cat('
+\\begin{tabular}{cccrccc}
+  \\hline
+  \\hline
+ &\\% Miss.& Imp. Err.&Levels& BaU & CTA1 & Std. Diff.\\\\
+  \\hline
+  \\hline
+\\multirow{3}{*}{Ethnicity}&\\multirow{3}{*}{',miss['Race/Ethnicity',2],'\\%}
+  &\\multirow{3}{*}{',impErr['Race/Ethnicity'],'}& White/Asian & ',
[…]

### 2019-09-18 R/dataAnalysis.r

@@ -1,0 +1,303 @@
+library(gridExtra)
+library(optmatch)
+library(RItools)
+library(sandwich)
+library(clubSandwich)
+library(lmtest)
+library(mgcv)
+library(tidyverse)
+library(rstan)
+library(lme4)
+traceplot <- rstan::traceplot
+select <- dplyr::select
+
+omar <- par()$mar
+
+
+
+source('R/prelimStan.r')
+source('~/Box Sync/rcode/matchingFunctions.r')
+source('~/Box Sync/rcode/hhh.r')
+source('~/Box Sync/rcode/printXbal.r')
+
+source('R/covariateTable.r')
+
+source('R/evaluateSmooths.r')
+
+overallHintInfo <-
+  xOrig%>%
+  filter(field_id%in%dat$field_id)%>%
+  mutate(hint=nhints1>0,err=nerrs1>0)%>%
+  xtabs(~hint+err,data=.)
+
+
+
+sdat2 <- with(
+  sdat,
+  list(
+    N=nsecWorked,
+    nstud=sum(Z),
+    nsec=nsec,
+    studentM=as.numeric(as.factor(studentM)),
+    section=section,
+    hint=grad
+  )
+)
+
+###########################################################################################
+### defining "high hint users"
+###########################################################################################
+
+### raw means
+hintAvg <-
+  with(sdat2,tibble(student=studentM,section=section,hint=hint,field_id=x$field_id))%>%
+  group_by(section)%>%
+  mutate(nprobSec=n(),nstudSec=n_distinct(student),propHintSec=mean(hint))%>%
+  group_by(student)%>%
+  mutate(nprobStud=n(),nsecStud=n_distinct(section),propHintStud=mean(hint),secDiffStud=mean(propHintSec))
+
+hintAvgStud <- hintAvg%>%group_by(student)%>%slice(1L)
+
+hintAvgStud%>%
+  gather("xaxis","x",nprobStud,secDiffStud)%>%
+  mutate(nsec=ifelse(xaxis=='nprobStud', median(nsecStud), nsecStud))%>%
+  ggplot(aes(x,propHintStud,size=nsec))+
+  geom_point()+
+  facet_wrap(~xaxis,scales="free_x")
+
+
+### rasch mixture model?
+##NOT RUN
+## source('R/raschMixture.r')
[…]

### 2019-09-18 R/evaluateSmooths.r

@@ -1,0 +1,1 @@
+# From: https://stackoverflow.com/questions/19735149/is-it-possible-to-plot-the-smooth-components-of-a-gam-fit-with-ggplot2

EvaluateSmooths = function(model, select=NULL, x=NULL, n=100) {
  if (is.null(select)) {
    select = 1:length(model$smooth)
  }
  do.call(rbind, lapply(select, function(i) {
    smooth = model$smooth[[i]]
    data = model$model
    
    if (is.null(x)) {
      min = min(data[smooth$term])
      max = max(data[smooth$term])
      x = seq(min, max, length=n)
    }
    if (smooth$by == "NA") {
      by.level = "NA"
    } else {
      by.level = smooth$by.level
    }
    range = data.frame(x=x, by=by.level)
    names(range) = c(smooth$term, smooth$by)
    
    mat = PredictMat(smooth, range)
    par = smooth$first.para:smooth$last.para
    
    y = mat %*% model$coefficients[par]
    
    se = sqrt(rowSums(
      (mat %*% model$Vp[par, par, drop = FALSE]) * mat
    ))
    
    return(data.frame(
      label=smooth$label
      , x.var=smooth$term
      , x.val=x
      , by.var=smooth$by
      , by.val=by.level
      , value = y
      , se = se
    ))
  }))
}

# Modified from above
# EvaluateData = function(model, select = NULL) {
#   
#   do.call(rbind, lapply(select, function(i) {
#     
#     smooth = model$smooth[[i]]
#     
#     data = model$model
#     
#     data.frame(
#       x.var = smooth$term,
#       x.val = data[, 1],
#       value = data[, smooth$term]
#     )
#     
#   } ) )
#   
# }

### 2019-09-18 R/matching.r

@@ -1,0 +1,64 @@
+
+#### record of some of the process of coming up with a match
+
+psmod0 <- lmer(qlogis(propHint)~(1|state)+(1|schoolid2)+(1|classid2)+sex+grade+race+ns(xirt,5)+spec+esl+gradeMIS+raceMIS+sexMIS+frlMIS,data=subset(dat1,propHint<1&propHint>0))
+
+psmod3.2 <- ## stan_
+  glmer(M~(1|state)+(1|schoolid2)+(1|classid2)+sex+grade+race+ns(xirt,5)+spec+esl+gradeMIS+raceMIS+sexMIS+frlMIS,family=binomial,data=dat1)
+
+psmod1 <-   glmer(M~(1|state)+(1|schoolid2)+(1|classid2)+grade+race+xirt+spec+esl,family=binomial,data=dat1)
+#save(psmod3.2,file='stanPSmod2.RData')
+#load('stanPSmod.RData')
+
+
+dist1 <- match_on(M~predict(psmod3.2,type='link'),data=dat1)
+m1 <- fullmatch(dist1,data=dat1)
+
+dist2 <- match_on(M~predict(psmod3.2,type='link')+xirt,data=dat1)
+m2 <- fullmatch(dist2,data=dat1)
+balMod <- xBalance(M~grade+race+xirt+spec+esl,strata=list(unmatched=NULL,m2=~m2),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
+
+dist3 <- match_on(M~predict(psmod3.2,type='link'),within=exactMatch(M~schoolid2,data=dat1),data=dat1)
+m3 <- fullmatch(dist3,data=dat1)
+
+balMod <- xBalance(M~grade+race+xirt+spec+esl+schoolid2+state,strata=list(unmatched=NULL,m2=~m2,m3=~m3),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
+
+plot(balMod,colors=subwayPalette)
+abline(v=c(-.05,-.25,.05,.25),lty='dotted')
+
+
+dist4.3 <- match_on(M~predict(psmod3.2,type='link'),within=exactMatch(M~schoolid2,data=dat1),data=dat1)
+m4.3 <- fullmatch(dist4.3,data=dat1)
[…]

### 2019-09-18 R/medExplore.r

@@ -1,0 +1,74 @@
+#### mediation?
+
+### estimate E[Y(1,M(0))]
+dat10 <- subset(dat1,!M)
+tab <- xtabs(~match+M,data=dat1)
+K <- 1+tab[,2]/tab[,1]
+dat10$ww <- K[dat10$match]
+out5 <- lm(Y~1,data=dat10,weights=ww)
+
+
+### SE depends on question. for
+vvv <- vcovCR(out5,dat10$schoolid2,'CR2')
+se1 <- sqrt(vvv[1,1])
+
+
+out5 <- lmer(Y~(1|schoolid2)+(1|classid2),data=dat10)
+muhat <- weighted.mean(dat10$Y,dat10$ww)
+cls <- dat10%>%group_by(classid2)%>%summarize(ww=sum(ww)^2)
+scl <- dat10%>%group_by(schoolid2)%>%summarize(ww=sum(ww)^2)
+vvv <- as.data.frame(VarCorr(out5))
+varMu <- (vvv$vcov[1]*sum(cls$ww)+vvv$vcov[2]*sum(scl$ww)+vvv$vcov[3]*sum(dat10$ww^2))/nrow(dat1)^2
+se2 <- sqrt(varMu)
+
+
+### direct effect
+ddd1 <- dat10
+ddd1$Y <- with(ddd1,Y*ww/sum(ww)*nrow(ddd1))
+ddd1$xirt <- with(ddd1,xirt*ww/sum(ww)*nrow(ddd1))
+datDir1 <- rbind(subset(dat,treatment==0),ddd1[,names(dat)])
+dir1 <- lm(Y~treatment+pair,data=datDir1)
+vvv <- vcovCR(dir1,datDir1$schoolid2,'CR2')
+se1 <- sqrt(vvv[1,1])
+
+dir2 <- lmer(Y~treatment+pair+(1|classid2)+(1|schoolid2),data=datDir1)
+dir3 <- lmer(Y~treatment+poly(xirt,2)+race+grade+esl+frl+sex+pair+(1|classid2)+(1|schoolid2),data=datDir1)
+
+### mediated effect
+### E[Y(1,m(1))-Y(1,m(0))]
+ddd2 <- dat1
+adj <- lm(Y~M+schoolid2+poly(xirt,2)+race+grade+esl+frl+sex,data=dat1)
+ddd2$Ytilde <- ddd2$Y-model.matrix(adj)[,-c(1:2)]%*%coef(adj)[-c(1:2)]
[…]

### 2019-09-18 R/prelimStan.r

@@ -1,0 +1,146 @@
+### only look at year 2
+### and only at students who weren't in  year 1
+library(splines)
+library(rstan)
+library(tidyverse)
+memory.limit(50000)
+
+rstan_options(auto_write = TRUE)
+options(mc.cores = parallel::detectCores())
+
+discard <- TRUE
+afterError <- FALSE
+
+#save(list=ls(),file=paste0('prev',Sys.time(),'.RData'))
+#rm(list=ls())
+
+load('~/Box Sync/CT/data/RANDstudyData/HSdata.RData')
+load('~/Box Sync/CT/data/problemLevelUsageData/probLevelData.RData')
+
+for(nn in c('curriculum','unit','section','Prob1')) x[[nn]] <- tolower(x[[nn]])
+
+datOrig <- dat
+xOrig <- x
+
+
+## ### estimated eta using only treatment group
+## library(jagstools)
+## load('~/Google Drive/CTmodels/fullUsage.RData')
+## usageMod <- mod
+## U <- jagsresults(usageMod,'studEff')[,1]
+
+#dataPrep <- function(dat,x,discard=TRUE,afterError=FALSE){
+ dat <- dat[!dat$field_id%in%dat$field_id[dat$year==1],]
+ #print(table(dat$year))
+
+ dat <- droplevels(dat)
+
+ x <- x[x$field_id%in%dat$field_id[dat$treatment==1],]
+
+ ### just look at algebra I sections--- likely different advance patterns in other curricula
+### make sure to keep algebra i units that are also part of other curricula
+ algUnit <- unique(x$unit[x$curriculum=='algebra i'])
+ x <- subset(x,unit%in%algUnit)
+
+### take out bridge to algebra, algebra ii, geometry
+ x <- x[-grep('ii|geo|bridge',x$curriculum),]
+
+ x$hint <- x$nhints1>0 #| x$nerrs1>0
+
+ x <- subset(x,nhints1>0 |x$nerrs1>0)
+
+ x <- droplevels(x)
+
+
+ ### discard some pairs
+ ## discard treatment schools with no usage data
+ ## (do a robustness check with everything left in afterwards)
+ percUse <- function(scl)
[…]

### 2019-09-18 R/psDiagnostic.r

@@ -1,0 +1,97 @@
+library(bayesplot)
+### make Yrep
+set.seed(613)
+samp <- sample(1:nrow(draws$studEff),size=1000)
+draws1k <- lapply(draws, function(x) if(NCOL(x)==1) x[samp] else x[samp,])
+Ymean <- with(c(sdat,draws1k),
+              teacherEffY[,teacher]+schoolEffY[,school]+pairEffect[,pair]+sweep(studEff,1,a1,'*')+
+              sweep(sweep(studEff,1,b1,'*'),2,Z,'*')+betaY%*%t(X))
+
+Yrep <- sapply(1:1000,function(i) rnorm(sdat$nstud,Ymean[i,],draws1k$sigY[i,sdat$Z+1]))
+
+## overall Y
+ppc_dens_overlay(sdat$Y,t(Yrep))+ggtitle('Pooled Treatment and Control')
+ggsave('figure/ppcYoverall.jpg')
+
+ppc_dens_overlay(sdat$Y[sdat$Z==1],t(Yrep)[,sdat$Z==1])+ggtitle('Treatment Group')
+ggsave('figure/ppcYtrt.jpg')
+
+ppc_dens_overlay(sdat$Y[sdat$Z==0],t(Yrep)[,sdat$Z==0])+ggtitle('Control Group')
+ggsave('figure/ppcYctl.jpg')
+
+### residual plots
+fitd <- rowMeans(Yrep)
+qplot(fitd,sdat$Y-fitd,xlab='Fitted Values',ylab='Residuals',main='Overall')
+ggsave('figure/residPlotOverall.jpg')
+
+qplot(fitd[sdat$Z==1],sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],xlab='Fitted Values',ylab='Residuals',main='Treatment Group')
+ggsave('figure/residPlotTrt.jpg')
+
+qplot(fitd[sdat$Z==0],sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],xlab='Fitted Values',ylab='Residuals',main='Control Group')
+ggsave('figure/residPlotCtl.jpg')
+
+
+pdf('figure/qqPlots.pdf', height=3,width=6)
+par(mfrow=c(1,2))
+qqnorm(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],main='Treatment Group')
[…]

### 2019-09-18 R/raschMixture.r

@@ -1,0 +1,67 @@
+
+raschMix <- stan('R/raschMix.stan',data=sdat2,chains=5,iter=2000)
+save(raschMix,file='output/mixtureModel2.RData')
+ss <- summary(raschMix)
+studEffSumm <- ss$summary[grep('studEff',rownames(ss$summary)),]
+mean(studEffSumm[,'Rhat']<1.01)
+draws <- rstan::extract(raschMix,permute=TRUE)
+
+rhat <- function(mat){
+  n <- nrow(mat)
+  W <- apply(mat,2,var)
+  W <- mean(W)
+  B <- var(colMeans(mat))
+  sqrt(((1-1/n)*W+B)/W)
+}
+
+
+post <- function(theta,p1,mu0,mu1,sigma){
+  p0 <- 1-p1
+  post0 <- dnorm(theta,mu0,sigma[1])*p0
+  post1 <- dnorm(theta,mu1,sigma[2])*p1
+  post1/(post0+post1)
+}
+
+classProbs <- with(draws,
+  vapply(1:nrow(studEff),
+    function(i)
+      post(
+        theta=studEff[i,],
+        p1=p1[i],
+        mu0=mu0[i],
+        mu1=mu1[i],
+        sigma=sigma[i,]
+      ),
+    numeric(ncol(studEff))
+  )
+)
+
+classProbsOverall <- rowMeans(classProbs)
+
+
+hintAvgStud$classProb <- classProbsOverall[1:nrow(hintAvgStud)]
+hintAvgStud$studEff <- colMeans(draws$studEff[,1:nrow(hintAvgStud)])
+
+## mm <- glmer(hint~(1|field_id)+(1|section),family=binomial,data=x)
+## save(mm,file='hardHintRaschNew.RData')
+
+## re <- ranef(mm)
+## bbb <- x%>%group_by(field_id)%>%summarize(nprob=n(),hint=mean(nhints1>0),err=mean(nerrs1>0), hintHard=mean(nhints1[nerrs1>0]>0),nhard=sum(nerrs1>0|nhints1>0))
+
+## bbb$theta <- re[[1]][as.character(bbb$field_id),1]
+## bbb$theta <- bbb$theta+fixef(mm)[1]
+## med <- median(bbb$theta,na.rm=TRUE)
+## bbb$M <- bbb$theta>med
+## dat <- merge(dat,bbb,all.x=TRUE)
+
+save(hintAvgStud,file='output/raschMixOutput.RData')
[…]

### 2019-09-18 R/fakeModels.r
diagnostic checks

@@ -1,0 +1,121 @@
+source('R/prelimStan.r')
+
+################################################
+## Fake Models
+##############################################
+
+### use data, E[studEff] from main model
+load('output/hintPS.RData')
+dat$U <- colMeans(extract(psmod1,'studEff')[[1]])
+### U has a couple outliers:
+dat$U[dat$U< -4] <- -4
+
+effs <- extract(psmod1,c('b0','b1'))
+
+rm(psmod1); gc()
+
+##### make fake data:
+
+### first delete control schools
+datF <- subset(dat,treatment==1)
+
+### now double the dataset
+datF$schoolid2 <- as.character(datF$schoolid2)
+datF$teachid2 <- as.character(datF$teachid2)
+dat2 <- datF
+dat2$schoolid2 <- paste0(dat2$schoolid2,'Fake')
+dat2$teachid2 <- paste0(dat2$teachid2,'Fake')
+dat2$field_id <- dat2$field_id*100+99
+dat2$treatment <- 0
+
+## ### take bootstrap samples within each school
+## for(scl in unique(datF$schoolid2)){
+##     ind <- which(datF$schoolid2==scl)
+##     datF[ind,] <- datF[sample(ind,length(ind),replace=TRUE),]
+##     dat2[ind,] <- dat2[sample(ind,length(ind),replace=TRUE),]
+## }
+
+datF <- rbind(datF,dat2)
+
+datF$schoolid2 <- factor(datF$schoolid2)
+datF$teachid2 <- factor(datF$teachid2)
+
+### now delete usage data for "control" group
+xF <- x[x$field_id%in%datF$field_id[datF$treatment==1],]
+
+#########################################################
+### Compile Stan data as before
+#####################################################
+sdatF <- makeStanDat(datF,xF)
+
+
+noEff <- stan('R/prinStratStan.stan',data=sdatF,iter=3000,chains=6)
+
+##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: NO EFFECT\n',rep('-',40),'\n\n\n')
+##print(noEff,c('a0','a1','b0','b1'),c(0.05,0.95))
+save(list=ls(),file='output/noEffect.RData'); rm(noEff); gc()
+
+
+
+
+######## constant effect
[…]

### 2019-09-18 R/psDiagnostic.r

@@ -3,35 +3,29 @@ library(bayesplot)
 set.seed(613)
 samp <- sample(1:nrow(draws$studEff),size=1000)
 draws1k <- lapply(draws, function(x) if(NCOL(x)==1) x[samp] else x[samp,])
-Ymean <- with(c(sdat,draws1k),
-              teacherEffY[,teacher]+schoolEffY[,school]+pairEffect[,pair]+sweep(studEff,1,a1,'*')+
-              sweep(sweep(studEff,1,b1,'*'),2,Z,'*')+betaY%*%t(X))
 
-Yrep <- sapply(1:1000,function(i) rnorm(sdat$nstud,Ymean[i,],draws1k$sigY[i,sdat$Z+1]))
-
-## overall Y
-ppc_dens_overlay(sdat$Y,t(Yrep))+ggtitle('Pooled Treatment and Control')
-ggsave('figure/ppcYoverall.jpg')
+ppc_dens_overlay(sdat$Y,draws1k$Yrep)+ggtitle('Pooled Treatment and Control')
+ggsave('output/ppcYoverall.jpg')
 
 ppc_dens_overlay(sdat$Y[sdat$Z==1],t(Yrep)[,sdat$Z==1])+ggtitle('Treatment Group')
-ggsave('figure/ppcYtrt.jpg')
+ggsave('output/ppcYtrt.jpg')
 
 ppc_dens_overlay(sdat$Y[sdat$Z==0],t(Yrep)[,sdat$Z==0])+ggtitle('Control Group')
-ggsave('figure/ppcYctl.jpg')
+ggsave('output/ppcYctl.jpg')
 
 ### residual plots
 fitd <- rowMeans(Yrep)
 qplot(fitd,sdat$Y-fitd,xlab='Fitted Values',ylab='Residuals',main='Overall')
-ggsave('figure/residPlotOverall.jpg')
+ggsave('output/residPlotOverall.jpg')
 
 qplot(fitd[sdat$Z==1],sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],xlab='Fitted Values',ylab='Residuals',main='Treatment Group')
-ggsave('figure/residPlotTrt.jpg')
+ggsave('output/residPlotTrt.jpg')
 
 qplot(fitd[sdat$Z==0],sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],xlab='Fitted Values',ylab='Residuals',main='Control Group')
[…]

### 2019-09-18 R/fakeModels.r
fake models

@@ -66,7 +66,7 @@ sdatFConst <- within(sdatF,{
  }
 )
 
-constEff <- stan('R/psmod.stan',data=sdatFConst,iter=3000,chains=6)
+constEff <- stan('R/prinStratStan.stan',data=sdatFConst,iter=3000,chains=6)
 
 ##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: CONSTANT EFFECT ATE=0.18\n',rep('-',40),'\n\n\n')
 ##print(constEff,c('a0','a1','b0','b1'),c(0.05,0.95))
@@ -84,7 +84,7 @@ datLin$Y[datLin$treatment==1] <- datLin$Y[datLin$treatment==1]+datLin$te[datLin$
 
 sdatFlin <- makeStanDat(datLin,advanceF)
 
-linEff <- stan('R/psmod.stan',data=sdatFlin,iter=3000,chains=6)
+linEff <- stan('R/prinStratStan.stan',data=sdatFlin,iter=3000,chains=6)
 ##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: LINEAR EFFECT b1=',round(mean(effs$b1),2),'\n',rep('-',40),'\n\n\n')
 ##print(linEff,c('a0','a1','b0','b1'),c(0.05,0.95))
 
@@ -106,7 +106,7 @@ sdatFquad <- makeStanDat(datQuad,advanceF)
 
 
 
-quadEff <- stan('R/psmod.stan',data=sdatFquad,iter=3000,chains=6)
+quadEff <- stan('R/prinStratStan.stan',data=sdatFquad,iter=3000,chains=6)
 
 ##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: QUADRATIC EFFECT',rep('-',40),'\n\n\n')
 ##print(constEff,c('a0','a1','b0','b1'),c(0.05,0.95))

### 2019-09-19 R/dataAnalysis.r
fixed bugs, helper functions file

@@ -14,11 +14,14 @@ select <- dplyr::select
 omar <- par()$mar
 
 
-
-source('R/prelimStan.r')
+source('R/helperFunctions.r')
 source('~/Box Sync/rcode/matchingFunctions.r')
 source('~/Box Sync/rcode/hhh.r')
 source('~/Box Sync/rcode/printXbal.r')
+
+source('R/prelimStan.r')
+
+sdat <- makeStanDat(dat,x)
 
 source('R/covariateTable.r')
 
@@ -105,7 +108,7 @@ dev.off()
 
 dat1$match <- m3
 
-save(dat1,file='output/dat1.RData')
+save(dat1,file='data/dat1.RData')
 
 ##########################################################################################
 ### estimate treatment effects (obs. study)
@@ -203,39 +206,6 @@ draws <- rstan::extract(psmod1)
 ######## main effect plot
 #########################
 
-
-pdMod <- function(mod,row=1,column=1,func){
-    draws <- rstan::extract(mod)
-    samp <- seq(1,length(draws$b1),length=1000)
-    Usamp <- draws$studEff[samp,]
-    iqr <- apply(Usamp,1,IQR)
-    studEff95 <- quantile(Usamp,c(0.025,0.975))
-    Usamp[Usamp<studEff95[1] | Usamp>studEff95[2]] <- NA
-    trtEff <- sweep(sweep(Usamp,1,draws$b1[samp],'*'),1,draws$b0[samp],'+')
-
-
-
-    if(missing(func)){
-        func <- function(x) mean(draws$b0)+mean(draws$b1)*x
-        knownTruth <- FALSE
-    } else knownTruth <- TRUE
-    truth <- curve(func,from=studEff95[1],to=studEff95[2],n=length(samp)/3)
-    avg <- curve(mean(draws$b0)+x*mean(draws$b1),
-                 from=studEff95[1],to=studEff95[2],n=length(samp)/3)
-    postDraw <- curve(mean(draws$b0)+x*mean(draws$b1),
-                      from=studEff95[1],to=studEff95[2],n=length(samp)-length(truth$x)-length(avg$x))
-    x <- c(postDraw$x,truth$x,avg$x)
-    y <- c(postDraw$y,truth$y,avg$y)
[…]

### 2019-09-19 R/helperFunctions.r

@@ -1,0 +1,89 @@
+
+makeStanDat <- function(dat,x,xInteract=FALSE,missingUsage=TRUE){
+  stanDat <- list()
+
+  if(!missingUsage)
+    dat <- filter(dat,field_id%in%x$field_id|dat$treatment==0)
+
+  x <- droplevels(x)
+ dat <- droplevels(dat)
+
+ x$section <- as.factor(x$section)
+
+ stanDat$nsecWorked <- nrow(x)
+ stanDat$nstud <- nrow(dat)
+ stanDat$nteacher <- nlevels(dat$teachid2)
+ stanDat$nschool <- nlevels(dat$schoolid2)
+ stanDat$nsec <- nlevels(x$section)
+ stanDat$npair <- nlevels(dat$pair)
+
+ stanDat$teacher <- as.numeric(as.factor(dat$teachid2))
+ stanDat$pair <- as.numeric(dat$pair)
+ stanDat$school <- as.numeric(dat$school)
+ stanDat$studentM <- seq(stanDat$nstud)[match(x$field_id,dat$field_id)]
+ stanDat$section <- as.numeric(x$section)
+
+ stanDat$grad <- as.numeric(x$hint)
+
+ X <- model.matrix(~poly(xirt,2)+race+sex+spec,data=dat)[,-1]
+ if(xInteract) X <- model.matrix(~poly(xirt,2)*(race+sex+spec)+(race+sex+spec)^2+state,data=dat)[,-1]
+ stanDat$X <- scale(X)
+ stanDat$ncov <- ncol(X)
+
+
+ stanDat$Z <- as.numeric(dat$treatment)
+
+ stanDat$Y <- dat$Y
+
+ stanDat
+}
+
+
+pdMod <- function(mod,row=1,column=1,func){
+    draws <- rstan::extract(mod)
+    samp <- seq(1,length(draws$b1),length=1000)
+    Usamp <- draws$studEff[samp,]
+    iqr <- apply(Usamp,1,IQR)
+    studEff95 <- quantile(Usamp,c(0.025,0.975))
+    Usamp[Usamp<studEff95[1] | Usamp>studEff95[2]] <- NA
+    trtEff <- sweep(sweep(Usamp,1,draws$b1[samp],'*'),1,draws$b0[samp],'+')
+
+
+
+    if(missing(func)){
+        func <- function(x) mean(draws$b0)+mean(draws$b1)*x
+        knownTruth <- FALSE
+    } else knownTruth <- TRUE
[…]

### 2019-09-19 R/prelimStan.r

@@ -93,53 +93,13 @@ xOrig <- x
     mutate(prevErr=lag(nerrs1))%>%
     filter(!is.na(prevErr),prevErr>0)
 
- save(dat,x,file='hintsData.RData')
+ save(dat,x,file='data/hintsData.RData')
 
 
 
 
 
-makeStanDat <- function(dat,x,xInteract=FALSE,missingUsage=TRUE){
-  stanDat <- list()
 
-  if(!missingUsage)
-    dat <- filter(dat,field_id%in%x$field_id)
-
-  x <- droplevels(x)
- dat <- droplevels(dat)
-
- x$section <- as.factor(x$section)
-
- stanDat$nsecWorked <- nrow(x)
- stanDat$nstud <- nrow(dat)
- stanDat$nteacher <- nlevels(dat$teachid2)
- stanDat$nschool <- nlevels(dat$schoolid2)
- stanDat$nsec <- nlevels(x$section)
- stanDat$npair <- nlevels(dat$pair)
-
- stanDat$teacher <- as.numeric(as.factor(dat$teachid2))
- stanDat$pair <- as.numeric(dat$pair)
- stanDat$school <- as.numeric(dat$school)
- stanDat$studentM <- seq(stanDat$nstud)[match(x$field_id,dat$field_id)]
- stanDat$section <- as.numeric(x$section)
-
- stanDat$grad <- as.numeric(x$hint)
-
- X <- model.matrix(~poly(xirt,2)+race+sex+spec,data=dat)[,-1]
- if(xInteract) X <- model.matrix(~poly(xirt,2)*(race+sex+spec)+(race+sex+spec)^2+state,data=dat)[,-1]
- stanDat$X <- scale(X)
- stanDat$ncov <- ncol(X)
-
-
- stanDat$Z <- as.numeric(dat$treatment)
-
- stanDat$Y <- dat$Y
-
- stanDat
-}
-
-
-sdat <- makeStanDat(dat,x)
 
 
 

### 2019-09-19 R/psDiagnostic.r

@@ -1,4 +1,6 @@
 library(bayesplot)
+
+
 ### make Yrep
 set.seed(613)
 samp <- sample(1:nrow(draws$studEff),size=1000)
@@ -56,6 +58,7 @@ ggsave('output/mbarPPC.jpg')
 
 ### drop treatment cases w/o usage data (same dataset as obs study and mediation analysis)
 sdatDrop <- makeStanDat(dat,x,missingUsage=FALSE)
+
 psmodDrop <- stan('R/prinStratStan.stan',data=sdatDrop)
 save(psmodDrop,file='output/psmodDrop.RData')
 

### 2019-09-19 R/fakeConstEff.r
broke up fake models into separate files

@@ -1,0 +1,21 @@
+if(!exists('datF')|!exists('xF'))
+  source('R/fakePrep.r')
+
+sdatF <- makeStanDat(datF,xF)
+
+functions <- currentCode()
+######## constant effect
+########
+sdatFConst <- within(sdatF,{
+ te <- rnorm(sum(Z),0.18,0.1)
+ Y[Z==1] <- Y[Z==1]+ te
+ }
+)
+
+constEff <- stan('R/prinStratStan.stan',data=sdatFConst,iter=3000,chains=6)
+
+modSum(constEff)
+
+##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: CONSTANT EFFECT ATE=0.18\n',rep('-',40),'\n\n\n')
+##print(constEff,c('a0','a1','b0','b1'),c(0.05,0.95))
+save(constEff,sdatF,functions,file='output/constEff.RData')

### 2019-09-19 R/fakeLinEff.r

@@ -1,0 +1,24 @@
+if(!exists('datF')|!exists('xF'))
+  source('R/fakePrep.r')
+
+functions <- currentCode()
+
+##################################
+######### linear effect ##########
+##################################
+datLin <- datF
+datLin$te <- mean(effs$b0)+mean(effs$b1)*datF$U
+datLin$Y[datLin$treatment==1] <- datLin$Y[datLin$treatment==1]+datLin$te[datLin$treatment==1]
+
+sdatFlin <- makeStanDat(datLin,advanceF)
+
+linEff <- stan('R/prinStratStan.stan',data=sdatFlin,iter=3000,chains=6)
+##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: LINEAR EFFECT b1=',round(mean(effs$b1),2),'\n',rep('-',40),'\n\n\n')
+##print(linEff,c('a0','a1','b0','b1'),c(0.05,0.95))
+
+modSum(linEff)
+
+save(linEff,sdatFlin,functions,file='output/linEff.RData');rm(linEff);gc();
+
+
+

### 2019-09-19 R/fakeModels.r

@@ -1,119 +1,15 @@
-source('R/prelimStan.r')
+### run all four fake models
 
-################################################
-## Fake Models
-##############################################
-
-### use data, E[studEff] from main model
-load('output/hintPS.RData')
-dat$U <- colMeans(extract(psmod1,'studEff')[[1]])
-### U has a couple outliers:
-dat$U[dat$U< -4] <- -4
-
-effs <- extract(psmod1,c('b0','b1'))
-
-rm(psmod1); gc()
-
-##### make fake data:
-
-### first delete control schools
-datF <- subset(dat,treatment==1)
-
-### now double the dataset
-datF$schoolid2 <- as.character(datF$schoolid2)
-datF$teachid2 <- as.character(datF$teachid2)
-dat2 <- datF
-dat2$schoolid2 <- paste0(dat2$schoolid2,'Fake')
-dat2$teachid2 <- paste0(dat2$teachid2,'Fake')
-dat2$field_id <- dat2$field_id*100+99
-dat2$treatment <- 0
-
-## ### take bootstrap samples within each school
-## for(scl in unique(datF$schoolid2)){
-##     ind <- which(datF$schoolid2==scl)
-##     datF[ind,] <- datF[sample(ind,length(ind),replace=TRUE),]
-##     dat2[ind,] <- dat2[sample(ind,length(ind),replace=TRUE),]
-## }
-
-datF <- rbind(datF,dat2)
-
-datF$schoolid2 <- factor(datF$schoolid2)
-datF$teachid2 <- factor(datF$teachid2)
-
-### now delete usage data for "control" group
-xF <- x[x$field_id%in%datF$field_id[datF$treatment==1],]
-
-#########################################################
-### Compile Stan data as before
-#####################################################
-sdatF <- makeStanDat(datF,xF)
-
-
-noEff <- stan('R/prinStratStan.stan',data=sdatF,iter=3000,chains=6)
-
-##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: NO EFFECT\n',rep('-',40),'\n\n\n')
-##print(noEff,c('a0','a1','b0','b1'),c(0.05,0.95))
-save(list=ls(),file='output/noEffect.RData'); rm(noEff); gc()
-
-
-
-
[…]

### 2019-09-19 R/fakeNoEff.r

@@ -1,0 +1,22 @@
+if(!exists('datF')|!exists('xF'))
+  source('R/fakePrep.r')
+
+functions <- currentCode()
+
+#########################################################
+### Compile Stan data as before
+#####################################################
+sdatF <- makeStanDat(datF,xF)
+
+
+noEff <- stan('R/prinStratStan.stan',data=sdatF,iter=3000,chains=6)
+
+modSum(noEff)
+
+##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: NO EFFECT\n',rep('-',40),'\n\n\n')
+##print(noEff,c('a0','a1','b0','b1'),c(0.05,0.95))
+save(noEff,sdatF,functions,file='output/noEffect.RData'); rm(noEff); gc()
+
+
+
+

### 2019-09-19 R/fakePrep.r

@@ -1,0 +1,54 @@
+library(rstan)
+library(tidyverse)
+
+options(mc.cores = 6)
+
+#source('R/prelimStan.r')
+source('R/helperFunctions.r')
+
+load('data/hintsData.RData')
+################################################
+## Fake Models
+##############################################
+
+### use data, E[studEff] from main model
+load('output/hintPS.RData')
+dat$U <- colMeans(rstan::extract(psmod1,'studEff')[[1]])
+### U has a couple outliers:
+## newRng <- quantile(dat$U,c(0.25,0.75))+c(-1,1)*2*IQR(dat$U)
+## dat$U[dat$U>newRng[2]] <- newRng[2]
+## dat$U[dat$U<newRng[1]] <- newRng[1]
+
+effs <- rstan::extract(psmod1,c('b0','b1'))
+
+rm(psmod1); gc()
+
+##### make fake data:
+
+### first delete control schools
+datF <- subset(dat,treatment==1)
+
+### now double the dataset
+datF$schoolid2 <- as.character(datF$schoolid2)
+datF$teachid2 <- as.character(datF$teachid2)
+dat2 <- datF
+dat2$schoolid2 <- paste0(dat2$schoolid2,'Fake')
+dat2$teachid2 <- paste0(dat2$teachid2,'Fake')
+dat2$field_id <- dat2$field_id*100+99
+dat2$treatment <- 0
+
+## ### take bootstrap samples within each school
+## for(scl in unique(datF$schoolid2)){
+##     ind <- which(datF$schoolid2==scl)
+##     datF[ind,] <- datF[sample(ind,length(ind),replace=TRUE),]
+##     dat2[ind,] <- dat2[sample(ind,length(ind),replace=TRUE),]
+## }
+
+datF <- rbind(datF,dat2)
+
+datF$schoolid2 <- factor(datF$schoolid2)
+datF$teachid2 <- factor(datF$teachid2)
+
+### now delete usage data for "control" group
+xF <- x[x$field_id%in%datF$field_id[datF$treatment==1],]
+

### 2019-09-19 R/fakeQuadEff.r

@@ -1,0 +1,33 @@
+if(!exists('datF')|!exists('xF'))
+  source('R/fakePrep.r')
+
+functions <- currentCode()
+
+########### quadratic effects
+datQuad <- within(datF,{
+    te <- -(U-mean(U))^2
+    te <- te/sd(te)*0.1
+    te <- te-mean(te)+0.13
+    Y[treatment==1] <- Y[treatment==1]+te[treatment==1]
+})
+
+sdatFquad <- makeStanDat(datQuad,advanceF)
+
+
+
+quadEff <- stan('R/prinStratStan.stan',data=sdatFquad,iter=3000,chains=6)
+
+##cat('\n\n\n\n',rep('-',40),'\n','TRUTH: QUADRATIC EFFECT',rep('-',40),'\n\n\n')
+##print(constEff,c('a0','a1','b0','b1'),c(0.05,0.95))
+
+modSum(quadEff)
+
+save(quadEff,sdatFquad,functions,file='output/quadEff.RData'); rm(quadEff); gc();
+
+
+
+
+
+
+
+

### 2019-09-19 R/helperFunctions.r

@@ -87,3 +87,13 @@ modSum <- function(mod){
     }
   }
 }
+
+
+currentCode <- function(){
+  functions <- list()
+  for(ll in ls(name=.GlobalEnv))
+    if(is.function(get(ll,envir=.GlobalEnv)))
+      functions[[ll]] <- get(ll,envir=.GlobalEnv)
+  functions$timestamp <- Sys.time()
+  functions
+}

### 2019-09-19 R/diagnosticPlots.r
tinkered w diagnostics

@@ -1,0 +1,63 @@
+require(tidyverse)
+source('R/helperFunctions.r')
+
+library(bayesplot)
+
+if(!exists('draws')){
+  load('output/hintPS.RData')
+  draws <- rstan::extract(psmod1)
+}
+if(!exists('sdat')){
+  load('data/hintsData.RData')
+  sdat <- makeStanDat(dat,x)
+}
+
+### make Yrep
+set.seed(613)
+samp <- sample(1:nrow(draws$studEff),size=1000)
+draws1k <- lapply(draws, function(x) if(NCOL(x)==1) x[samp] else x[samp,])
+
+ppc_dens_overlay(sdat$Y,draws1k$Yrep)+ggtitle('Pooled Treatment and Control')
+ggsave('output/ppcYoverall.jpg')
+
+ppc_dens_overlay(sdat$Y[sdat$Z==1],t(Yrep)[,sdat$Z==1])+ggtitle('Treatment Group')
+ggsave('output/ppcYtrt.jpg')
+
+ppc_dens_overlay(sdat$Y[sdat$Z==0],t(Yrep)[,sdat$Z==0])+ggtitle('Control Group')
+ggsave('output/ppcYctl.jpg')
+
+### residual plots
+fitd <- rowMeans(Yrep)
+qplot(fitd,sdat$Y-fitd,xlab='Fitted Values',ylab='Residuals',main='Overall')
+ggsave('output/residPlotOverall.jpg')
+
+qplot(fitd[sdat$Z==1],sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],xlab='Fitted Values',ylab='Residuals',main='Treatment Group')
+ggsave('output/residPlotTrt.jpg')
+
+qplot(fitd[sdat$Z==0],sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],xlab='Fitted Values',ylab='Residuals',main='Control Group')
+ggsave('output/residPlotCtl.jpg')
+
+
+pdf('output/qqPlots.pdf', height=3,width=6)
+par(mfrow=c(1,2))
+qqnorm(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],main='Treatment Group')
+qqline(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1])
+qqnorm(sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],main='Control Group')
[…]

### 2019-09-19 R/fakePrep.r

@@ -52,3 +52,4 @@ datF$teachid2 <- factor(datF$teachid2)
 ### now delete usage data for "control" group
 xF <- x[x$field_id%in%datF$field_id[datF$treatment==1],]
 
+

### 2019-09-19 R/psDiagnostic.r

@@ -1,66 +1,7 @@
-library(bayesplot)
 
+source('R/diagnosticPlots.r')
 
-### make Yrep
-set.seed(613)
-samp <- sample(1:nrow(draws$studEff),size=1000)
-draws1k <- lapply(draws, function(x) if(NCOL(x)==1) x[samp] else x[samp,])
-
-ppc_dens_overlay(sdat$Y,draws1k$Yrep)+ggtitle('Pooled Treatment and Control')
-ggsave('output/ppcYoverall.jpg')
-
-ppc_dens_overlay(sdat$Y[sdat$Z==1],t(Yrep)[,sdat$Z==1])+ggtitle('Treatment Group')
-ggsave('output/ppcYtrt.jpg')
-
-ppc_dens_overlay(sdat$Y[sdat$Z==0],t(Yrep)[,sdat$Z==0])+ggtitle('Control Group')
-ggsave('output/ppcYctl.jpg')
-
-### residual plots
-fitd <- rowMeans(Yrep)
-qplot(fitd,sdat$Y-fitd,xlab='Fitted Values',ylab='Residuals',main='Overall')
-ggsave('output/residPlotOverall.jpg')
-
-qplot(fitd[sdat$Z==1],sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],xlab='Fitted Values',ylab='Residuals',main='Treatment Group')
-ggsave('output/residPlotTrt.jpg')
-
-qplot(fitd[sdat$Z==0],sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],xlab='Fitted Values',ylab='Residuals',main='Control Group')
-ggsave('output/residPlotCtl.jpg')
-
-
-pdf('output/qqPlots.pdf', height=3,width=6)
-par(mfrow=c(1,2))
-qqnorm(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],main='Treatment Group')
-qqline(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1])
-qqnorm(sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],main='Control Group')
-qqline(sdat$Y[sdat$Z==0]-fitd[sdat$Z==0])
-dev.off()
-
-
-### usage model
-set.seed(613)
-samp <- sample(1:1000,9)
-lp <- with(c(sdat,draws1k),studEff[,studentM]+secEff[,section])
[…]

### 2019-09-19 R/robustnessChecks.r

@@ -1,0 +1,26 @@
+require(rstan)
+require(tidyverse)
+
+source('R/helperFunctions.r')
+
+if(is.null(getOption('mc.cores'))) options(mc.cores = 4)
+if(getOption('mc.cores')<4)  options(mc.cores = 4)
+
+functions <- currentCode()
+
+############################
+## robustness checks
+############################
+
+load('data/hintsData.RData')
+
+### drop treatment cases w/o usage data (same dataset as obs study and mediation analysis)
+sdatDrop <- makeStanDat(dat,x,missingUsage=FALSE)
+
+psmodDrop <- stan('R/prinStratStan.stan',data=sdatDrop)
+
+modSum(psmodDrop)
+
+psmodDropSumm <- summary(psmodDrop)$summary
+
+save(psmodDrop,sdatDrop,functions,file='output/psmodDrop.RData'); rm(psmodDrop); gc()

### 2019-10-11 paper.Rnw
main model drops missing log data

@@
 their entire randomization block.
-Of the remaining 2,390 students, 87\% (2,091) had log
+Of the remaining \Sexpr{Nt} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
 data.
-The students with missing log data were dropped from the observational
-study and mediation analysis, but included in the principal
-stratification study.
+For the sake of simplicity, the treatment students without log data
+were dropped from the study.\footnote{Including these students in a
+  principal stratification model is straightforward \citep{aoas}; see
+  the online supplement for results from a model including them, which
+  were broadly similar to the results presented here. Including
+  subjects with missing log data in a mediation or observational study
+  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
 Problem log data for sections that were not part of the
@@
 
-All told, the main analysis included $n=$5308 students, 2390 of whom were assigned to the CTA1 condition and 2918 of whom were assigned to control.
-The students were nested within 116 teachers, in 43 schools across five states.
+All told, the main analysis included $n=$\Sexpr{prettyNum(sdat$nstud,',')} students, \Sexpr{prettyNum(sum(sdat$Z),',')} of whom were assigned to the CTA1 condition and \Sexpr{prettyNum(with(sdat,nstud-sum(Z)),',')} of whom were assigned to control.
+The students were nested within \Sexpr{sdat[['nteacher']]} teachers, in \Sexpr{sdat[['nschool']]} schools across five states.
 

### 2019-10-11 R/dataAnalysis.r

@@ -21,7 +21,13 @@ source('~/Box Sync/rcode/printXbal.r')
 
 source('R/prelimStan.r')
 
-sdat <- makeStanDat(dat,x)
+## usage data availability
+
+Nt <- sum(dat$treatment)
+
+obsUse=sum(dat$field_id[dat$treatment==1]%in%x$field_id)
+
+sdat <- makeStanDat(dat,x,missingUsage=FALSE)
 
 source('R/covariateTable.r')
 
@@ -191,11 +197,15 @@ dir2.1ci <- coefci(dir2.1,'treatment',vcov.=dir2.1vcv)
 
 
 ########## principal stratification
+if(is.null(getOption('mc.cores'))) options(mc.cores = 6)
+if(getOption('mc.cores')<6)  options(mc.cores = 6)
+
+functions <- currentCode()
 
 ### NOT RUN
-## psmod1 <- stan('R/prinStratStan.stan',data=sdat)
-## save(psmod1,file='output/hintPS.RData')
-load('output/hintPS.RData')
+psmod1 <- stan('R/prinStratStan.stan',data=sdat,chains=6,iter=6000)
+save(psmod1,sdat,functions,file='fitModels/hintPS.RData')
+load('fitModels/hintPS.RData')
 
 
 

### 2019-10-11 R/robustnessChecks.r

@@ -15,12 +15,12 @@ functions <- currentCode()
 load('data/hintsData.RData')
 
 ### drop treatment cases w/o usage data (same dataset as obs study and mediation analysis)
-sdatDrop <- makeStanDat(dat,x,missingUsage=FALSE)
+sdatFull <- makeStanDat(dat,x,missingUsage=TRUE)
 
-psmodDrop <- stan('R/prinStratStan.stan',data=sdatDrop,chains=6,iter=3000)
+psmodFull <- stan('R/prinStratStan.stan',data=sdatFull,chains=6,iter=6000)
 
-modSum(psmodDrop)
+modSum(psmodFull)
 
-psmodDropSumm <- summary(psmodDrop)$summary
+psmodFullSumm <- summary(psmodFull)$summary
 
-save(psmodDrop,sdatDrop,functions,file='fitModels/psmodDrop.RData'); rm(psmodDrop); gc()
+save(psmodFull,sdatFull,functions,file='fitModels/psmodFull.RData'); rm(psmodFull); gc()

### 2019-10-28 paper.Rnw
as of 10-25

@@
 two principal ways from more traditional domains.
-RCTs feature randomization of treatment assignment, and a post-test
+RCTs feature randomization of treatment assignment, and a posttest
 (or some other outcome) administered to both users and non-users.
@@
 uses techniques from observational studies to determine if and how requesting
-a greater number of hints affects post-test scores.
+a greater number of hints affects posttest scores.
 For the sake of illustration, we will use techniques similar to those
@@
 Students using the Cognitive Tutor Algebra software work algebra
-problems on a computer, instead of on paper, as is typical.
+problems on a computer.
 Each problem in the software is associated with a set of skills that
@@
 permanently stuck on some problem'' (p. 190).
-That is, hints give students who are stuck a way to continue working
-after getting stuck.
+That is, hints give students who are stuck a way to continue working.
 Hints may also boost learning by ``helping students identify relevant features in
@@
 of learning'' \citep[][p. 782]{koedinger2012knowledge}.
-The final available ``bottom out'' hint typically gives the student
-the answer.
-Students who bottom out then see a worked example, which can be an
+Hints can help improve students' self-awareness and
+problem-solving strategy, such as by clarifying which skills they have
+and have not mastered.
+
+Most problems are associated with several hints, arranged in a
+sequence so that a student who remains stuck after one hint may
+request another.
+The last hint in the sequence will show a complete solution to the
+problem---essentially showing students a worked example, which can be an
 effective tool for learning \citep[e.g.][]{sweller1985use}.
-Finally, hints can help improve students' meta-cognitive states, such
-as their self-knowledge of skills.
 
@@
 In order for hints to be instructive, students need to reflect on
-their content, for instance with ``self-explanation''
-\citep[c.f.][]{renkl1998learning}.
-In the same vein, hints need to be well-designed in order to be
-effective \citep{mckendree1990effective}.
+their content in productive ways.
 This suggests that hints may be helpful for some students but not for
 others.
-They may even be harmful: as \citet{koedinger2007exploring} put it
+In the same vein, hints need to be well-designed in order to be
+effective \citep{mckendree1990effective}, so some hints may be more
+helpful than others.
+
+Hints may even be harmful: as \citet{koedinger2007exploring} put it
 (p. 241),
@@
 the effect of a hint request: by student and by the ``level'' of the
-hint (the first, second, or third ``bottom out''  hint available for a given problem).
+hint (the first, second, or final hint available for a given problem).
 They found that, indeed, the effect of asking for a hint varies
 between students, but on average it is negative for the first two
-hints requested on a problem, but positive for the bottom out hint.
+hints requested on a problem, but positive for the final hint.
 
@@
 to estimate the effect of hint usage; its purpose was to estimate the
-effect of CTA1 as a package.
-Like \citet{beck2008does} and \citet{goldin2012learner}, but unlike
-\citet{anderson1995cognitive} and \citet{singh2011feedback}, hint
-usage was not randomized.
-On the other hand, the CTA1 study includes a posttest, like the
-randomized studies but unlike the observational studies cited here.
+effect of CTA1 as a package, including the associated textbook and
+recommended classroom practices.
+Access to the full CTA1 curriculum was randomized, rather than any
+specific aspect of usage, and effectiveness was measured with a
+standardized posttest covering a broad range of Algebra I skills.
 In this regard, the CTA1 study is similar to other large-scale
 evaluations of educational technology which do not randomize usage but
-do include a post-test.
+do include a posttest.
 
 The CTA1 study was of a much longer duration than
-any study of hint effects that we are aware of: students had access to the tutor for an entire
+any of the studies of hint effects we are aware of: students had access to the tutor for an entire
 school year before the posttest.
@@
 hints on that problem, and zero otherwise.
-(More mundanely, we do not have access to data
-describing the number of hints available for every problem, which
-would hinder interpretation of any other measure of hint usage.)
+% (More mundanely, we do not have access to data
+% describing the number of hints available for every problem, which
+% would hinder interpretation of any other measure of hint usage.)
 
@@
 fact that students are unlikely to request hints on problems that they
-already know how to solve (other than by accident).
+already know how to solve.
 Hence, any measure of hint usage is necessarily also a measure of
@@
 Conversely, the significance of a student's decision to forgo a hint
-depends on how challenging he or she finds the problem.
+depends in part on how challenging he or she finds the problem.
 Hint request frequency combines students' hint requests (or non-requests) from both
 problems they find challenging and trivial, so is a poor measure of
-students' use of the hint feature.
+students' disposition to use the hint feature.
 
@@
 problems) a student makes an error without having requested a hint,
-to the percentage of problems a student requests a hint
+to the percentage of problems a student requests a hint.%
 \footnote{Technically, if $h$ is the event that a student requests
 a hint on a problem and $e$ is the event that the student makes
-an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(h\text{ and not }e)/Pr(h)\}$}.
+an error, then $Pr(h|h\text{ or }e)=1/\{1+Pr(e\text{ and not }h)/Pr(h)\}$.}
 Making an error on a problem without having requested a hint reflects
@@
 \citep{rasch} to hint request data, described in an online appendix.
-The model classified approximately \Sexpr{round(mixProp*100)} of students as
+The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
 high-hint users, so we chose $c=0.6$, approximately the
@@
 Following \citet{neyman} and \citet{rubin}, let $Y_i(Z=1)$ be
-the posttest score student $i$ would exhibit were $i$ (perhaps
+the posttest score student $i$ would achieve were $i$ (perhaps
 counterfactually) assigned to the treatment condition, and
@@
 In our study of hints, $\bm{x}$ includes indicators for state, school
-id, class id, grade (9th or higher), race (White/Asian,
-Black/Multi-Racial, or Hispanic/American Indian/Alaskan Native),
-special education status, economic disadvantage, and English language
-learner status, as well as pre-test scores.
+id, class id, and the variables described in Table \ref{tab:covariateBalance}.
 Then we assume \citep[c.f.][]{rosenbaum1983central}:
@@
 We estimated $\pi$ using a Bayesian multilevel logistic regression,
-with weakly informative priors, using the \texttt{rstanarm} package in
-\texttt{R} \citep{rstanarm}.
+using the \texttt{lme4} package in
+\texttt{R} \citep{lme4}.
 The propensity score model was:
@@
 state, school, and class, which were each included using random
-intercepts modeled as normally distributed.
+intercepts modeled as normally distributed, including missingness
+indicators for grade, race, sex, and economic disadvantage, and
+a natural spline with five degrees of freedom for pretest.
 
@@
 
-Table \ref{fig:balance} in the online appendix shows that
+Figure \ref{fig:balance} in the online appendix shows that
 matching largely eliminated mean differences in pre-treatment
@@
 The p-value from an omnibus covariate balance test
-\citep{covBal} is 0.26.
+\citep{covBal} is \Sexpr{prettyNum(round(balMod[['overall']]['matched','p.value'],2))}.
 
@@
 racial categories, grade, ESL status, economic disadvantage, and sex,
-along with linear and quadratic terms for pre-test (the standard error
+along with a natural spline with five degrees of freedom for pretest (the standard error
 and confidence interval use the ``HC3'' sandwich estimator,
@@
 CTA1 treatment effect: $\delta(1)$ gives the extent to which
-assignment to CTA1 affects a student's post-test score by (possibly)
+assignment to CTA1 affects a student's posttest score by (possibly)
 causing him or her to request hints.
@@
 
-We modeled students' post-test scores $Y$ as
+We modeled students' posttest scores $Y$ as
 conditionally normal:

### 2019-10-28 R/covariateTable.r

@@ -1,19 +1,25 @@
 ## covariate means/balance
-miss <- NULL
-for(i in c('race','sex','spec','xirt')) miss <- rbind(miss,
- c(sum(is.na(covs[[i]])),mean(is.na(covs[[i]])),error[i,'error']))
-miss <- as.data.frame(miss)
-miss$`Error Type` <- c('PFC','PFC','PFC','SRMSE')
-rownames(miss) <- c('Race/Ethnicity','Sex','Special Education','Pretest')
-names(miss)[1:3] <- c('# Missing','% Missing','Imputation Error')
-miss[,2] <- as.integer(round(miss[,2]*100))
-miss[,1] <- as.integer(miss[,1])
-miss['Pretest','Imputation Error'] <- sqrt(miss['Pretest','Imputation Error'])/sd(covs$xirt,na.rm=TRUE)
-
 
 covs2 <- covs[rownames(dat),] ### same as dat but with NAs
 
-balTest <- balanceTest(treatment~race+sex+spec+xirt+cluster(schoolid2)+strata(pair),data=dat,report=c('adj.means','std.diffs','chisquare.test'),include.NA.flags=T)
+miss <- NULL
+varbs <- c('grade','esl','frl','race','sex','spec','xirt')
+for(i in varbs)
+  miss <- rbind(miss,
+    c(sum(is.na(covs2[[i]])),mean(is.na(covs2[[i]])),error[i,'error']))
+miss <- as.data.frame(miss)
+#miss$`Error Type` <- c('PFC','PFC','PFC','SRMSE')
+#rownames(miss) <- c('Race/Ethnicity','Sex','Special Education','Pretest')
+rownames(miss) <- varbs
+names(miss)[1:3] <- c('# Missing','% Missing','Imputation Error')
+miss[,2] <- as.integer(round(miss[,2]*100))
+miss[,1] <- as.integer(miss[,1])
+miss['xirt','Imputation Error'] <- sqrt(miss['xirt','Imputation Error'])/sd(covs2$xirt,na.rm=TRUE)
+
+
+
+
[…]

### 2019-10-28 paper.Rnw
obs. study & mediation sections

@@
 )
-
+pn <- function(...) prettyNum(...,big.mark=',')
+library(xtable)
 @
@@
 source('R/dataAnalysis.r')
-big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
-save(list=ls()[!big],file='output/analysis.RData')
 @
 
-<<loadStuff,include=FALSE>>=
+<<loadStuff,cache=FALSE>>=
 load('output/analysis.RData')
@@
 
-All told, the main analysis included $n=$\Sexpr{prettyNum(sdat$nstud,',')} students, \Sexpr{prettyNum(sum(sdat$Z),',')} of whom were assigned to the CTA1 condition and \Sexpr{prettyNum(with(sdat,nstud-sum(Z)),',')} of whom were assigned to control.
-The students were nested within \Sexpr{sdat[['nteacher']]} teachers, in \Sexpr{sdat[['nschool']]} schools across five states.
+All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
+The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
 
@@
 The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
-high-hint users, so we chose $c=0.6$, approximately the
-\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
+high-hint users, so we chose $c$ as the
+\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
+approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
 This definition agreed with the model's classification in
@@
 variable size and composition.
-For instance, one matched set included 30 $H=1$ subjects matched to a
-single $H=0$ subject, and another matched 21 $H=0$ subjects to a
+For instance, one matched set included \Sexpr{max(sapply(levels(m3),function(ll) sum(optmatch:::getZfromMatch(m3)[m3==ll])))} $H=1$ subjects matched to a
+single $H=0$ subject, and another matched \Sexpr{max(sapply(levels(m3),function(ll) sum(!optmatch:::getZfromMatch(m3)[m3==ll])))} $H=0$ subjects to a
 single $H=1$ subject.
 
-Figure \ref{fig:balance} in the online appendix shows that
+Figure \ref{fig:balance} and Table \ref{tab:balance} in the online appendix show that
 matching largely eliminated mean differences in pre-treatment
@@
 along with a natural spline with five degrees of freedom for pretest (the standard error
-and confidence interval use the ``HC3'' sandwich estimator,
+and confidence interval using the ``HC3'' sandwich estimator,
 \citealt{sandwichPackage}).
@@
 
-\begin{table}
-\centering
-\begin{tabular}{crllcll}
-&&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
-\hline
-\multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
-&ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
-&TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
-&ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
-\hline
-\multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
-&[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
-\hline
-\end{tabular}
-\caption{Estimates with and without regression adjustment and
-  sensitivity analysis of the effect of hint usage.}
-\label{tab:matchResults}
-\end{table}
+<<obsStudyEst,results='asis',echo=FALSE>>=
+
+matchTEtab <-
+  cbind(
+    data.frame(Effect=c('ANOVA','ATE','TOT','ANCOVA')),
+    rbind(anovat,atet,tott,ancovat)[,c('Estimate','Std. Error')],
+    CI=vapply(list(anovaci,ateci,totci,ancovaci),printci,'a'),
+    `$T_Z$`=0,
+    `$\\rho^2$`=0
+  )
+
+for(tt in c(ttt['xirt'],ttt['xirt']/2))
+  matchTEtab <- add_case(matchTEtab,
+    Effect=ifelse(tt==ttt['xirt'],"[Pretest]","[Pretest/2]"),
+    Estimate=ancovat[1],
+    CI=printci(interval(tt,rrr['xirt'],b=ancovat[1], se=ancovat[2],df=ancova$df)),
+    `$\\rho^2$`=rrr['xirt'],
+    `$T_Z$`=tt)
+
+matchTEtab <- cbind(
+  ` `=c('\\multirow{4}{*}{Estimate}',rep('',3),'\\multirow{2}{*}{Sensitivity}',''),
+  matchTEtab
+)
+
+matchTEtab <- add_case(matchTEtab,` `='\\hline',.before=5)#,rep(NA,ncol(matchTEtab)-1)),matchTEtab[5:6,])
+
+print(xtable(matchTEtab,
+  align='rllrrrrr',
+  caption='Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.',
+  label='tab:matchResults'),
+  include.rownames=FALSE,sanitize.text.function =function(x) x)
+@
+
+% \begin{table}
+% \centering
+% \begin{tabular}{crllcll}
+% &&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
+% \hline
+% \multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
+% &ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
+% &TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
+% &ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
+% \hline
+% \multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
+% &[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
+% \hline
+% \end{tabular}
+% \caption{Estimates with and without regression adjustment and
+%   sensitivity analysis of the effect of hint usage.}
+% \label{tab:matchResults}
+% \end{table}
 
@@
 
-We found that hint requests have a negative effect: students who tend
-to request hints more often than the median score lower on the
-posttest, after controlling for observed covariates.
+We found that hint requests have a negative effect: students who
+requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
+percentile score lower on the posttest, after controlling for observed covariates.
 When hint usage is expressed numerically, we find a monotonically
@@
 anticorrelated with hint requests.
-This could even be the case after control for pretest scores and other
+This could even be the case after controlling for pretest scores and other
 covariates, if these failed to capture all of the relevant dimensions
@@
 This is because students assigned to the control group request exactly
-zero hints, whereas merely nine members of the treatment group
+zero hints, whereas merely
+\Sexpr{english(sum(hintAvgStud[['propHintStud']]==0))}
+members of the treatment group
 requested no hints over the course of the study, and these students
@@
 
-Our solution was to dichotomize hint requests with a median split, as
+Our solution was to dichotomize hint requests, as
 described above, essentially re-defining the causal question.
-Half of the treatment group and all of the control group has
-$H=0$---requesting hints at a rate less than the (treatment group)
-median.
+\Sexpr{round((mixProp)*100)}\% of the treatment group and all of the control group has
+$H=0$---requesting hints at a rate less than the
+\Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
 To make sense of the respective roles of $H$ and $\hcont$, the
@@
 \end{align*}
-Since $Pr(H=1)=1/2$ ($H$ was defined with a median split),
-$\hat{\Delta}=TOT/2$.
-Estimated thusly, the indirect effect is negative: $-0.11$, with
-a standard error of 0.03.
+Since $Pr(H=1)=$\Sexpr{round(1-mixProp,2)},
+$\hat{\Delta}=$\Sexpr{round(1-mixProp,2)}$TOT$.
+Estimated thusly, the indirect effect is negative: \Sexpr{round(tott[1]*(1-mixProp),2)}, with
+a standard error of \Sexpr{round(tott[2]*(1-mixProp),2)}.
 This analysis suggests that the availability of excessive hints
@@
 Following the latter route, and using the \texttt{clubSandwich}
-package in \texttt{R} \citep{clubsandwich}, we estimated a direct effect of $0.16\pm
-0.24$.
+package in \texttt{R} \citep{clubsandwich}, we estimated a direct
+effect of
+\Sexpr{citopm(dir1.1ci)}.
 Including covariates in the regression produces a direct effect of
-$0.19\pm 0.21$.
+\Sexpr{citopm(dir2.1ci)}.
 

### 2019-10-28 paper.tex
as of 10-25
obs. study & mediation sections

--- 2019-09-18 paper.Rnw
@@
-\documentclass{article}
+\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[stable]{footmisc}
@@
 
-<<init,include=FALSE>>=
-library(knitr)
-opts_chunk$set(
-echo=FALSE,results='asis',warning = FALSE,message = FALSE,cache=TRUE,error = FALSE
-)
-
-@
-
-<<runStuff,include=FALSE,eval=FALSE>>=
-source('R/dataAnalysis.r')
-big <- sapply(ls(),function(x) object.size(get(x))>2*1e7)
-save(list=ls()[!big],file='output/analysis.RData')
-@
-
-<<loadStuff,include=FALSE>>=
-load('output/analysis.RData')
-@
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 their entire randomization block.
-Of the remaining 2,390 students, 87\% (2,091) had log
+Of the remaining 2390 students, 87\% (2,091) had log
 data.
-The students with missing log data were dropped from the observational
-study and mediation analysis, but included in the principal
-stratification study.
+For the sake of simplicity, the treatment students without log data
+were dropped from the study.\footnote{Including these students in a
+  principal stratification model is straightforward \citep{aoas}; see
+  the online supplement for results from a model including them, which
+  were broadly similar to the results presented here. Including
+  subjects with missing log data in a mediation or observational study
+  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
 Problem log data for sections that were not part of the
@@
 
-All told, the main analysis included $n=$5308 students, 2390 of whom were assigned to the CTA1 condition and 2918 of whom were assigned to control.
+All told, the main analysis included $n=$5,009 students, 2,091 of whom were assigned to the CTA1 condition and 2,918 of whom were assigned to control.
 The students were nested within 116 teachers, in 43 schools across five states.
@@
 \citep{rasch} to hint request data, described in an online appendix.
-The model classified approximately \Sexpr{round(mixProp*100)} of students as
-high-hint users, so we chose $c=0.6$, approximately the
-\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$.
+The model classified approximately 30\% of students as
+high-hint users, so we chose $c$ as the
+70th percentile of $\bar{h}$,
+approximately 0.6.
 This definition agreed with the model's classification in
-approximately \Sexpr{round(agree*100)}\% of cases.
+approximately 93\% of cases.
 
@@
 variable size and composition.
-For instance, one matched set included 30 $H=1$ subjects matched to a
-single $H=0$ subject, and another matched 21 $H=0$ subjects to a
+For instance, one matched set included 44 $H=1$ subjects matched to a
+single $H=0$ subject, and another matched 117 $H=0$ subjects to a
 single $H=1$ subject.
 
-Table \ref{fig:balance} in the online appendix shows that
+Figure \ref{fig:balance} and Table \ref{tab:balance} in the online appendix show that
 matching largely eliminated mean differences in pre-treatment
@@
 The p-value from an omnibus covariate balance test
-\citep{covBal} is 0.26.
+\citep{covBal} is 0.39.
 
@@
 racial categories, grade, ESL status, economic disadvantage, and sex,
-along with linear and quadratic terms for pre-test (the standard error
-and confidence interval use the ``HC3'' sandwich estimator,
+along with a natural spline with five degrees of freedom for pretest (the standard error
+and confidence interval using the ``HC3'' sandwich estimator,
 \citealt{sandwichPackage}).
@@
 
-\begin{table}
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Mon Oct 28 14:48:44 2019
+\begin{table}[ht]
 \centering
-\begin{tabular}{crllcll}
-&&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
+\begin{tabular}{llrrlrr}
   \hline
-\multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
-&ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
-&TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
-&ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
+  & Effect & Estimate & Std. Error & CI & $T_Z$ & $\rho^2$ \\ 
   \hline
-\multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
-&[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
+\multirow{4}{*}{Estimate} & ANOVA & -0.18 & 0.05 & [-0.27,-0.08] & 0.00 & 0.00 \\ 
+   & ATE & -0.21 & 0.05 & [-0.32,-0.11] & 0.00 & 0.00 \\ 
+   & TOT & -0.18 & 0.06 & [-0.30,-0.07] & 0.00 & 0.00 \\ 
+   & ANCOVA & -0.19 & 0.05 & [-0.29,-0.10] & 0.00 & 0.00 \\ 
+  \hline &  &  &  &  &  &  \\ 
+  \multirow{2}{*}{Sensitivity} & [Pretest] & -0.19 &  & [-0.47,0.09] & -9.86 & 0.18 \\ 
+   & [Pretest/2] & -0.19 &  & [-0.37,-0.01] & -4.93 & 0.18 \\ 
    \hline
 \end{tabular}
-\caption{Estimates with and without regression adjustment and
-  sensitivity analysis of the effect of hint usage.}
+\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
 \label{tab:matchResults}
 \end{table}
+
+% \begin{table}
+% \centering
+% \begin{tabular}{crllcll}
+% &&Est.&SE&95\% CI&$\rho^2$&$T_Z$\\
+% \hline
+% \multirow{4}{*}{Estimate}%&ANOVA& -0.22 &0.04 &[-0.30,-0.13]&0&0\\
+% &ATE&-0.24&0.05&[-0.33,-0.14]&0&0\\
+% &TOT&-0.21&0.06&[-0.33,-0.10]&0&0\\
+% &ANCOVA&-0.22&0.04&[-0.31,-0.14]&0&0\\
+% \hline
+% \multirow{2}{*}{Sensitivity}&[Pretest]&-0.22&&[-0.55,0.10]&0.17&-13.14\\
+% &[Pretest/2]&-0.22&&[-0.40,-0.05]&0.08&-6.57\\
+% \hline
+% \end{tabular}
+% \caption{Estimates with and without regression adjustment and
+%   sensitivity analysis of the effect of hint usage.}
+% \label{tab:matchResults}
+% \end{table}
 

### 2019-10-28 R/dataAnalysis.r

@@ -10,16 +10,27 @@ library(rstan)
 library(lme4)
 traceplot <- rstan::traceplot
 select <- dplyr::select
+options(stringsAsFactors=FALSE)
 
 omar <- par()$mar
 
+### this is a tool to skip the most computationally-expensive steps but run everything else:
+runFull <- FALSE
+cat('\n\n----------------------------------------------------\n')
+cat('Warning: runFull=',runFull,'\n')
+cat('----------------------------------------------------\n\n')
 
 source('R/helperFunctions.r')
 source('~/Box Sync/rcode/matchingFunctions.r')
 source('~/Box Sync/rcode/hhh.r')
 source('~/Box Sync/rcode/printXbal.r')
 
-source('R/prelimStan.r')
+if(runFull) source('R/prelimStan.r') else load('data/hintsData.RData')
+
+## for printing nice CIs
+printci <- function(ci) paste0('[',paste(sprintf('%.2f',ci),collapse=','),']')
+
+citopm <- function(ci) paste(sprintf('%.2f',c(median(ci),(ci[2]-ci[1])/2)),collapse='$\\pm$')
 
 ## usage data availability
 
@@ -28,6 +39,11 @@ Nt <- sum(dat$treatment)
 obsUse=sum(dat$field_id[dat$treatment==1]%in%x$field_id)
 
 sdat <- makeStanDat(dat,x,missingUsage=FALSE)
+
+Ntot <- sdat$nstud
+NtFinal <- sum(sdat$Z)
+nteacher <- sdat$nteacher
+nschool <- sdat$nschool
 
 source('R/covariateTable.r')
 
@@ -53,9 +69,11 @@ sdat2 <- with(
   )
 )
 
+cat('
 ###########################################################################################
 ### defining "high hint users"
 ###########################################################################################
+',as.character(Sys.time()),'\n')
 
 ### raw means
 hintAvg <-
@@ -77,8 +95,8 @@ hintAvgStud%>%
 
 ### rasch mixture model?
 ##NOT RUN
-## source('R/raschMixture.r')
-load('output/raschMixOutput.RData')
+if(runFull) source('R/raschMixture.r') else load('output/raschMixOutput.RData')
+
[…]

### 2019-10-28 R/helperFunctions.r

@@ -99,7 +99,7 @@ currentCode <- function(){
 }
 
 pdMod <- function(mod,row=1,column=1,func){
-    draws <- extract(mod)
+    draws <- rstan::extract(mod)
     samp <- seq(1,length(draws$b1),length=1000)
     Usamp <- draws$studEff[samp,]
     iqr <- apply(Usamp,1,IQR)

### 2019-10-29 paper.Rnw
appendix plus

@@
 \usepackage{authblk}
-
+\usepackage[section]{placeins}
 %\usepackage{endfloat}
@@
 
+\renewenvironment{knitrout}{\begin{singlespace}}{\end{singlespace}}
+
 \title{Student Log-Data from a Randomized Evaluation of Educational
@@
 library(xtable)
+library(english)
+
 @
@@
 
-<<loadStuff,cache=FALSE>>=
+<<loadStuff,cache=TRUE>>=
 load('output/analysis.RData')
@@
 approaches to modeling log data.
-We hope that this paper can serve as a starting point for
-applied educational researchers interested in mining this rich
-resource, and that this paper helps spur methodological interest in
-this problem.
+A rich, and often polemic, literature surrounds principal
+stratification and mediation analysis
+\citep[e.g.][]{rubin2004direct,vanderweele2011principal,pearl2011principal,mealli2012refreshing,vanderweele2012comments}.
+\citet{vanderweele2008simple}, \citet{jo2008causal} and others have
+derived formal equivalencies between parameters under certain
+conditions.
+Our contribution will build on this literature in two ways: first, by
+providing a detailed demonstration of matching, mediation analysis,
+and principal stratification on the same dataset, and second by
+focusing on the applicability of these three methods in the specific
+context of studying implementation data from an EdTech randomized
+experiment.
 
@@
 To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
-that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
+that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
 and 0 otherwise.
@@
 $Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
-For a study measuring hint requests as a continuous variable, we may
-define $Y_i(\hcont)$, the posttest score $i$ would exhibit were $i$ to
-request hints with proclivity $\hcont$.
+% For a study measuring hint requests as a continuous variable, we may
+% define $Y_i(\bar{h})$, the posttest score $i$ would exhibit were $i$ to
+% request hints with proclivity $\bar{h}$.
 This structure implicitly assumes ``non-interference,'' that one
@@
 
-\subsection{The Observational Study Approach: Interpretation and Discussion}
-The observational study approach to log data seeks to estimate the
-effect of a usage parameter, estimated from  computer log data, on
-outcomes.
-This approach does not exploit randomization---in fact, it discards
-the control group entirely.
-However, it depends heavily on another feature of RCTs: namely, an
-outcome measured for all participants, in this case a posttest.
-Causal identification derives not from randomization, but from an
-untestable ignorability assumption, coupled with a sensitivity analysis.
-In our example, we operationalized strong ignorability with a
-propensity-score matching design, but other observational study
-designs may apply as well.
-
-We found that hint requests have a negative effect: students who
-requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
-percentile score lower on the posttest, after controlling for observed covariates.
-When hint usage is expressed numerically, we find a monotonically
-decreasing relationship between hint requests and posttest scores---as
-far as we can tell, the optimal strategy is to never request a hint.
-However, this observed effect may be due to confounding.
-If, say, weaker students ask for hints more often and score lower
-on the posttest regardless of hint usage, posttest scores will be
-anticorrelated with hint requests.
-This could even be the case after controlling for pretest scores and other
-covariates, if these failed to capture all of the relevant dimensions
-of student ability.
-We found that an unobserved confounder that predicts both hint usage
-and posttest scores just as strongly as the pretest could explain the
-entire observed relationship.
-The data, coupled with the existence of such a missing covariate, are
-consistent with both positive and negative effects of hint requests.
-On the other hand, covariates that are as important as pretests are
-quite rare in practice---this confounding scenario might not be
-plausible after all.
-A less extreme missing confounder, predicting $Y$ and $H$ with half
-the strength of the pretest, cannot explain the observed relationship,
-and is only consistent with negative effects of hint requests.
-
-Despite being hobbled by the lack of randomization, and the resultant
-threat of confounding, the observational study approach provides weak
-evidence that hints are harmful to posttest scores---a potentially
-important finding.
-This result, if confirmed in other scenarios and other analyses, may
-inform student behavior in the tutor and future tutor designs.
-However, since it ignores the control group from the CTA1 RCT, it does
-not shed light on the CTA1 causal mechanisms---it does not explain the
-CTA1 effect.
-However, its design and results figure prominently in mediation
-analysis, which does.
-
 \section{Mediation Analysis}\label{sec:mediation}
@@
 \Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
-To make sense of the respective roles of $H$ and $\hcont$, the
+To make sense of the respective roles of $H$ and $\bar{h}$, the
 continuous measure on which it is based, we may imagine a process in
 which students are assigned $H$, and under that constraint choose a
-value for $\hcont$.
-In the control group all students are assigned $H=0$ and $\hcont=0$, but
+value for $\bar{h}$.
+In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
 in the treatment group half of students have $H=1$ and half have
-$H=0$, and $\hcont$ varies within those groups.
+$H=0$, and $\bar{h}$ varies within those groups.
 That structure implies a ``treatment mediator interaction''
@@
 Of course, in our example treatment mediator interactions are not
-identified, since $H=0$ and $\hcont=0$ in the control group, and very
+identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
 few treatment students asked for 0 hints.
@@
 
-\subsection{Mediation Analysis: Interpretation and Discussion}
-Here we have used mediation analysis as a formal framework to
-interpret the results of the observational study in Section
-\ref{sec:observational} in terms of the overall CTA1 treatment effect.
-Doing so requires some conceptual and statistical acrobatics, such as
-relying on a dichotomized measurement of hint requests and strictly
-counterfactual potential outcomes, but can nevertheless yield
-interesting and interpretable results.
-Specifically, our analysis showed that CTA1 affected students'
-posttest scores in (at least) two opposite ways: by allowing hints, it
-lowered their posttest scores, but its other mechanisms increased
-their scores by an even greater amount.
-
-Like the observational study, these results are subject to
-confounding; indeed, the sensitivity analysis in the Section
-\ref{sec:observational} suggests that these results, too, may be
-explained due to a missing confounder.
-
 \section{Principal Stratification}\label{sec:principalStratification}
@@
 As in mediation analysis, hint behavior has potential outcomes as
-well; modeling hint requests continuously with the $\hcont$ parameter
-from (\ref{eq:rasch}) gives potential values $\hcont(1)$ and $\hcont(0)$,
-though only $\hcont(1)$ is relevant in the CTA1 study.
+well.
+Specifically, the rate at which as student requests hints, $\bar{h}$,
+has potential values $\bar{h}(1)$ and $\bar{h}(0)$,
+though only $\bar{h}(1)$ is relevant in the CTA1 study.
 
@@
 \begin{equation*}
-\tau(r)=\EE[Y(1)-Y(0)|\hcont(1)=r]
+\tau(r)=\EE[Y(1)-Y(0)|\bar{h}(1)=r]
 \end{equation*}
 This is the treatment effect for subjects who \emph{would} request
-hints at level $\hcont(1)=r$, if assigned to the treatment condition.
+hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
 Whereas the focus of the observational study and mediation
-analysis was the relationship between $\hcont$ and posttest scores, the
-focus of PS is the relationship between $\hcont$ and treatment effects.
-Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\hcont(1)=r]$,
-despite the fact that $\hcont(1)$ is never observed at the same time as
+analysis was the relationship between $\bar{h}$ and posttest scores, the
+focus of PS is the relationship between $\bar{h}$ and treatment effects.
+Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\bar{h}(1)=r]$,
+despite the fact that $\bar{h}(1)$ is never observed at the same time as
 $Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
 strictly counterfactual quantities, as in mediation analysis.
-$\hcont(1)$ is defined for subjects in both randomized groups, but only
+$\bar{h}(1)$ is defined for subjects in both randomized groups, but only
 observed in the treatment group.
 
-In previous sections $\hcont$ was dichotomized into $H$; such
+In previous sections $\bar{h}$ was dichotomized into $H$; such
 dichotomization is not necessary here.
-However, $\hcont$ has some disadvantages as a continuous measure of
+However, $\bar{h}$ has some disadvantages as a continuous measure of
 hint usage.
@@
 the larger principal stratification model, replacing $\eta$ for
-$\hcont$.
+$\bar{h}$.
 Then, the principal stratification estimand is
@@
 conditionally normal:
-\begin{multline}\label{eq:outcomeSubmodel}
- Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim \\ \mathcal{N}\left(
+\begin{equation}\label{eq:outcomeSubmodel}
+ Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
 \beta^Y_{0b[i]}+\bm{x}_i^T\bm{\beta}^Y+a\eta(1)+Z_i\tau(\eta(1))+\epsilon^{Yt}_{t[i]}+\epsilon^{Ys}_{s[i]},\sigma^Y_{Z[i]}\right)
-\end{multline}
+\end{equation}
 where $\beta^Y_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\beta}^Y$ are the
@@
 
-\subsection{Principal Stratification: Interpretation and Discussion}
+\section{Comparing Strategies}\label{sec:synthesis}
+The goals, assumptions, and processes of matching, mediation analysis,
+and principal stratification, when applied to implementation data from
+an EdTech RCT, each differ from each other, but also overlap in
+interesting ways.
+In the case discussed here, the observational study and mediation
+approaches appear to give the opposite conclusion from principal
+stratification.
+This section will compare and contrast the three methods in terms of
+ their necessary identification assumptions, their model fitting
+ mechanics, and interpretation of their results.
+
+The fundamental difference between the observational study and
+mediation approach, on the one hand, and principal stratification on
+the other, is in the role of $H$, the intermediate variable.
+An observational study or mediation analysis treats $H$ as a causal
+agent, an intervention or exposure that affects $Y$.
+In mediation analysis, it is also an (intermediate) outcome, affected
+by $Z$.
+The approach of principal stratification is quite nearly opposite:
+$H$ is not an exposure, that ``happens to'' subjects, but rather its
+potential value $H_T$ is a characteristic of the subjects.
+Observed $H$ reveals what was there all along.
+
+\subsection{Causal Identification}
+
+Observational studies, and mediation analyses that build on them,
+require ignorability assumptions such as Strong Ignorability or
+Matched Ignorability.
+These may be particularly problematic in our study, since students who
+are more likely to request hints are also more likely to struggle with
+the material, and hence score lower on the posttest.
+We addressed this concern in three ways: first, we focused our attention
+on the set of problems in the tutor in which students either requested
+a hint or made an error, second, with propensity score matching, and
+third with sensitivity analysis.
+None of these approaches is bulletproof---they each require further
+assumptions about how students request hints.
+Eliminating problems in which students showed no signs of struggling
+is likely to reduce the association between hint requests and
+underlying academic ability, but not eliminate it.
+For instance, stronger students may be better able to figure out a
+difficult problem without access to hints or error feedback than
+weaker students.
+Matching depends both on the adequacy of the observed covariates to
+capture baseline differences between $H=0$ and $H=1$ students, and the
+sensitivity analyses we performed assumed that any un-measured
+covariate would predict $H$ no better than pretest, or no better than
+a covariate half as important as pretest.
+These assumptions are impossible to verify from the data at hand;
+instead, they require a keen understanding of student practice within
+the tutor.
+
+Mediation analysis builds on the matching design, and also requires an
+ignorability assumption; it must
+additionally overcome the hurdle of the almost complete lack of
+overlap between hint request behavior in the control group (i.e. zero)
+and in the treatment group.
+Our approach of dichotomizing hint request behavior as $H$ provides a
+way forward, at the cost of ignoring information that could be gleaned
+from more continuous measures of hint requests.
+
+In contrast, no ignorabilty assumptions are necessary for principal
+stratification.
+That said, an assumption similar to ignorability may play an important
+role for interpretation.
+Further, principal stratification estimators require a model for $Y(0)$
+conditional on $\eta(1)$, which is unobserved.
+On the other hand, model misspecification is, in principle,
+detectable---an assumption of a particular family of models is not
+``untestable.''
+Still, the data will, in general, be unable to distinguish between certain
+alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
+turn lead to qualitatively different conclusions.
+In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
+distributed similarly to $Y(1)$ conditional on $\eta(1)$.
+
+An additional contrast of note between matching-based studies and
+principal stratification relates to transparency.
+However complex the process of devising a match may be, the resulting
+design is simple and clear---subjects with $H=1$ are compared against
+matches with $H=0$.
+This transparency may help researchers assess the plausibility of the
+match, and diagnose potential problems with resulting estimates.
+Though PS does not depend on ignorability, it does depend on a highly
+intricate and parametric modeling structure.
+If any part of the model is sufficiently misspecified, the results may
+be wrong.
+Perhaps worse, \citet{feller2016principal} argued that even well-specified
+models can give severely biased estimates.
+Extensive model interrogation and checking is necessary for PS
+analysis, but the complexity of the PS model makes this process
+particularly difficult.
+\citet{aoas} gives some examples of potentially fruitful model
+checking procedures, which are also carried out in the online supplement.
+
+\subsection{Results and Interpretation}
+
+Our observational study found that hint requests have a negative effect: students who
+requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
+percentile score lower on the posttest, after controlling for observed covariates---as
+far as we can tell, the optimal strategy is to never request a hint.
+However, this observed effect may be due to confounding.
+Mediation analysis essentially interprets this result in terms of the
+overall CTA1 treatment effect.
+Specifically, our analysis showed that CTA1 affected students'
+posttest scores in (at least) two opposite ways: by allowing hints, it
+lowered their posttest scores, but its other mechanisms increased
+their scores by an even greater amount.
+
 Like in the observational study and in mediation analysis, the
@@
 
-On the other hand, the observational and mediational results may be
-due not to causation but to confounding.
-In contrast, PS does not depend on an ignorability assumption beyond
-what is guaranteed by randomization.
-Perhaps by recognizing $\eta(1)$ as hopelessly confounded and focusing
-attention elsewhere, PS is able to arrive at the more reliable
-conclusion that---whatever the mechanism may be---students who ask
-more hints benefit more from the software.
-
-Though PS does not depend on ignorability, it does depend on a highly
-intricate and parametric modeling structure.
-If any part of the model is sufficiently misspecified, the results may
-be wrong.
-Perhaps worse, \citet{feller2016principal} argued that even well-specified
-models can give severely biased estimates.
-Extensive model interrogation and checking is necessary for PS
-analysis, but the complexity of the PS model makes this process
-particularly difficult.
-\citet{aoas} gives some examples of potentially fruitful model
-checking procedures.
-
-In contrast, the matching estimator in Section \ref{sec:observational}
-is highly transparent.
-Whatever intricacies went into formulating and fitting a propensity
-score model or matching procedure are inconsequential if they produce
-a convincing match.
-The validity of the matching estimate rests entirely on the quality of
-the match, which can be examined by checking covariate balance and
-other relatively simple metrics.
-The uncertainty in matching estimators due to possible confounding
-might be offset by the simplicity in modeling assumptions.
-
-\section{Which Strategy is Best?}
-
-\begin{table}
-\begin{tabular}{p{1in}p{2in}p{2in}}
-& Observational Study/Mediation Analysis &Principal Stratification\\
-\hline
-Research question & Does $H$ help $Z$ affect $Y$? & Does $H$ help predict
-                                           treatment effect of $Z$ on $Y$?\\
-\hline
-Role of $H$ & Causal & Predictive\\
-\hline
-Untestable assumption & $H$ is unconfounded & Distribution of $Y$
-                                              given $H_T$ in control
-                                              group\\
-\hline
-Other difficulties & Discrete $H$ necessary for mediation (in our
-                     framework) & Model fitting can be opaque and
-                                  erratic\\
-\end{tabular}
-\end{table}
+\subsection{Choosing Between Methods}
+
+The previous discussion suggests three criteria for choosing between
+the three approaches we've discussed here.
+First is the question the researcher seeks to answer.
+In their most straightforward interpretations, principal stratification
+is a method for examining treatment effect heterogeneity, whereas
+mediation analysis is a method for assessing causal mechanisms.
+That said, PS may also shed light on potential causal mechanisms---if,
+indeed, treatment effects are higher for students who request more
+hints, then hints may play a role in the tutor's effectiveness.
+
+A second criterion is the researcher's willingness to make
+ignorability assumptions.
+A researcher's confidence in her understanding of a data generating
+process and the quality of observed covariates can translate into
+confidence about the results of an observational study or mediation
+analysis.
+Conversely, researchers unwilling to consider untestable ignorability
+assumptions will find PS an attractive alternative.
+
+The final criterion is a researcher's comfort with complex statistical
+models.
+Researchers who are able and willing to experiment with a range of
+models for a dataset, and perhaps devise tests of model fit tailored
+to a specific problem, may be confident in the fit of a PS model.
+Researchers who prefer transparency and non-parametric or
+semi-parametric analysis may prefer matching studies.
+
+Additional philosophical concerns come in to play as well.
+For instance, unlike observational studies and PS, mediation analysis
+depends on strictly counterfactual quantities, such as $\EE Y(0,H(1))$,
+that can never occur.
+Analogous quantities in PS, such as $\EE [Y(0)|H(1)]$ are
+unobservable, but nevertheless refer to averages of potential outcomes
+among actual experimental subjects.
+These concerns are important but beyond the scope of the current
+paper; for more discussion, see the citations in the introduction.
+
+% \begin{table}
+% \begin{tabular}{p{1in}p{1.5}p{2in}p{2in}}
+% & Observational Study&Mediation Analysis &Principal Stratification\\
+% \hline
+% Research question & How does $H$ affect $Y$& What is the role of $Z$'s
+%                                              effect on $H$ in its
+%                                              effect on $Y$? & Does
+%                                                               $Z$'s
+%                                                               effect
+%                                                               on $H$
+%                                                               predict its
+%                                                               effect
+%                                                               on $Y$?\\
+
+% \hline
+% Role of $H$ & Causal & Causal & Predictive\\
+% \hline
+% Data used & treatment group & indirect effect: treatment group; direct
+%                               effect: full dataset & full dataset\\
+% \hline
+% Untestable assumption &$H$ is unconfounded& $H$ is unconfounded & Distribution of $Y$
+%                                               given $H_T$ in control
+%                                               group\\
+% \hline
+% Other difficulties & Discrete $H$ necessary for mediation (in our
+%                      framework) & Model fitting can be opaque and
+%                                   erratic\\
+% \end{tabular}
+% \caption{Some contrasts between
+% \end{table}
 
@@
 A second challenge is extrapolation in mediation analysis.
-In RCTs comparing computerized educational technology with traditional
+in RCTs comparing computerized educational technology with traditional
 pedagogy, members of the control group do not have computer log data
@@
 
+\newpage
+\appendix
+
+\section{Rasch Mixture Model of Hint Usage}
+
+To model hint requests with a mixture model, we modeled each hint
+request as in \eqref{eq:rasch}.
+Then, instead of modeling $\eta_i$ as normal, we assigned it a
+two-component normal mixture model:
+\begin{equation*}
+\eta\sim
+p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
+\end{equation*}
+where we constrained $\mu_0<\mu_1$.
+
+\subsection{Stan Code}
+<<raschMixture,results='markup'>>=
+  raschMix <- readLines('R/raschMix.stan')
+  raschMix <- gsub('studEff','eta',raschMix)
+  raschMix <- gsub('secEff','delta',raschMix)
+cat(raschMix,sep="\n")
+@
+%\clearpage
+\section{Covariate Balance Before and After Matching}
+<<covariateBalanceTable,results='asis'>>=
+library(RItools)
+print(xtable(balMod,caption='Covariate balance (standardized differences and Z-scores) before and after the propensity score match.',label='tab:balance'),add.to.row=list(pos=list(-1),command='&\\multicolumn{3}{c}{Unmatched}&\\multicolumn{3}{c}{Matched}\\\\\n'))
+@
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/balancePlot.pdf}
+\caption{Covariate balance (standardized differences) before and after
+  the propensity score match.}
+\label{fig:balance}
+\end{figure}
+
+%\clearpage
+\section{Stan Code for PS Model}
+
+%\clearpage
+\subsection{Stan Code}
+<<psCode,results='markup',cache=FALSE>>=
+  pscode <- readLines('R/prinStratStan.stan')
+  pscode <- gsub('studEff','eta',pscode)
+  pscode <- gsub('secEff','delta',pscode)
+pscode <- gsub('grad','hint',pscode)
+cat(pscode,sep="\n")
+@
+
+%\clearpage
+\section{Diagnostic Plots for Principal Stratification}
+
+<<makeDiagnosticPlots,eval=FALSE>>=
+source('R/diagnosticPlots.r')
+@
+
+\subsection{Posterior Predictive Density Plots}
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/ppcYoverall.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/ppcYtrt.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/ppcYctl.jpg}
+\end{figure}
+
+%\clearpage
+\subsection{Residual Plots}
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/residPlotOverall.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/residPlotTrt.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/residPlotCtl.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/binnedplot.pdf}
+\caption{Binned residual plot for model of hint requests.}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/qqPlots.pdf}
+\caption{Q-Q plots for outcomes and fitted values}
+\end{figure}
+
 \end{document}

### 2019-10-29 paper.tex
appendix plus

@@
 \newcommand{\indicator}[1]{\mathbf{1}_{\left[ {#1} \right] }}
+
+\renewenvironment{knitrout}{\begin{singlespace}}{\end{singlespace}}
 
@@
 % latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Mon Oct 28 14:48:44 2019
+% Mon Oct 28 20:04:28 2019
 \begin{table}[ht]
 \centering
-\begin{tabular}{llrrlrr}
+\begin{tabular}{llrrrrr}
   \hline
@@
 
-\subsection{The Observational Study Approach: Interpretation and Discussion}
-The observational study approach to log data seeks to estimate the
-effect of a usage parameter, estimated from  computer log data, on
-outcomes.
-This approach does not exploit randomization---in fact, it discards
-the control group entirely.
-However, it depends heavily on another feature of RCTs: namely, an
-outcome measured for all participants, in this case a posttest.
-Causal identification derives not from randomization, but from an
-untestable ignorability assumption, coupled with a sensitivity analysis.
-In our example, we operationalized strong ignorability with a
-propensity-score matching design, but other observational study
-designs may apply as well.
-
-We found that hint requests have a negative effect: students who tend
-to request hints more often than the median score lower on the
-posttest, after controlling for observed covariates.
-When hint usage is expressed numerically, we find a monotonically
-decreasing relationship between hint requests and posttest scores---as
-far as we can tell, the optimal strategy is to never request a hint.
-However, this observed effect may be due to confounding.
-If, say, weaker students ask for hints more often and score lower
-on the posttest regardless of hint usage, posttest scores will be
-anticorrelated with hint requests.
-This could even be the case after controlling for pretest scores and other
-covariates, if these failed to capture all of the relevant dimensions
-of student ability.
-We found that an unobserved confounder that predicts both hint usage
-and posttest scores just as strongly as the pretest could explain the
-entire observed relationship.
-The data, coupled with the existence of such a missing covariate, are
-consistent with both positive and negative effects of hint requests.
-On the other hand, covariates that are as important as pretests are
-quite rare in practice---this confounding scenario might not be
-plausible after all.
-A less extreme missing confounder, predicting $Y$ and $H$ with half
-the strength of the pretest, cannot explain the observed relationship,
-and is only consistent with negative effects of hint requests.
-
-Despite being hobbled by the lack of randomization, and the resultant
-threat of confounding, the observational study approach provides weak
-evidence that hints are harmful to posttest scores---a potentially
-important finding.
-This result, if confirmed in other scenarios and other analyses, may
-inform student behavior in the tutor and future tutor designs.
-However, since it ignores the control group from the CTA1 RCT, it does
-not shed light on the CTA1 causal mechanisms---it does not explain the
-CTA1 effect.
-However, its design and results figure prominently in mediation
-analysis, which does.
-
 \section{Mediation Analysis}\label{sec:mediation}
@@
 This is because students assigned to the control group request exactly
-zero hints, whereas merely nine members of the treatment group
+zero hints, whereas merely
+5
+members of the treatment group
 requested no hints over the course of the study, and these students
@@
 
-Our solution was to dichotomize hint requests with a median split, as
+Our solution was to dichotomize hint requests, as
 described above, essentially re-defining the causal question.
-Half of the treatment group and all of the control group has
-$H=0$---requesting hints at a rate less than the (treatment group)
-median.
-To make sense of the respective roles of $H$ and $\hcont$, the
+30\% of the treatment group and all of the control group has
+$H=0$---requesting hints at a rate less than the
+70th percentile of the treatment group.
+To make sense of the respective roles of $H$ and $\bar{h}$, the
 continuous measure on which it is based, we may imagine a process in
 which students are assigned $H$, and under that constraint choose a
-value for $\hcont$.
-In the control group all students are assigned $H=0$ and $\hcont=0$, but
+value for $\bar{h}$.
+In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
 in the treatment group half of students have $H=1$ and half have
-$H=0$, and $\hcont$ varies within those groups.
+$H=0$, and $\bar{h}$ varies within those groups.
 That structure implies a ``treatment mediator interaction''
@@
 \end{align*}
-Since $Pr(H=1)=1/2$ ($H$ was defined with a median split),
-$\hat{\Delta}=TOT/2$.
-Estimated thusly, the indirect effect is negative: $-0.11$, with
-a standard error of 0.03.
+Since $Pr(H=1)=$0.7,
+$\hat{\Delta}=$0.7$TOT$.
+Estimated thusly, the indirect effect is negative: -0.1, with
+a standard error of 0.04.
 This analysis suggests that the availability of excessive hints
@@
 Following the latter route, and using the \texttt{clubSandwich}
-package in \texttt{R} \citep{clubsandwich}, we estimated a direct effect of $0.16\pm
-0.24$.
+package in \texttt{R} \citep{clubsandwich}, we estimated a direct
+effect of
+0.14$\pm$0.23.
 Including covariates in the regression produces a direct effect of
-$0.19\pm 0.21$.
-
-\subsection{Mediation Analysis: Interpretation and Discussion}
-Here we have used mediation analysis as a formal framework to
-interpret the results of the observational study in Section
-\ref{sec:observational} in terms of the overall CTA1 treatment effect.
-Doing so requires some conceptual and statistical acrobatics, such as
-relying on a dichotomized measurement of hint requests and strictly
-counterfactual potential outcomes, but can nevertheless yield
-interesting and interpretable results.
-Specifically, our analysis showed that CTA1 affected students'
-posttest scores in (at least) two opposite ways: by allowing hints, it
-lowered their posttest scores, but its other mechanisms increased
-their scores by an even greater amount.
-
-Like the observational study, these results are subject to
-confounding; indeed, the sensitivity analysis in the Section
-\ref{sec:observational} suggests that these results, too, may be
-explained due to a missing confounder.
+0.18$\pm$0.19.
 
@@
 
-\subsection{Principal Stratification: Interpretation and Discussion}
+\section{Comparing Strategies}\label{sec:synthesis}
+The goals, assumptions, and processes of matching, mediation analysis,
+and principal stratification, when applied to implementation data from
+an EdTech RCT, each differ from each other, but also overlap in
+interesting ways.
+In the case discussed here, the observational study and mediation
+approaches appear to give the opposite conclusion from principal
+stratification.
+This section will compare and contrast the three methods in terms of
+ their necessary identification assumptions, their model fitting
+ mechanics, and interpretation of their results.
+
+The fundamental difference between the observational study and
+mediation approach, on the one hand, and principal stratification on
+the other, is in the role of $H$, the intermediate variable.
+An observational study or mediation analysis treats $H$ as a causal
+agent, an intervention or exposure that affects $Y$.
+In mediation analysis, it is also an (intermediate) outcome, affected
+by $Z$.
+The approach of principal stratification is quite nearly opposite:
+$H$ is not an exposure, that ``happens to'' subjects, but rather its
+potential value $H_T$ is a characteristic of the subjects.
+Observed $H$ reveals what was there all along.
+
+\subsection{Causal Identification}
+
+Observational studies, and mediation analyses that build on them,
+require ignorability assumptions such as Strong Ignorability or
+Matched Ignorability.
+These may be particularly problematic in our study, since students who
+are more likely to request hints are also more likely to struggle with
+the material, and hence score lower on the posttest.
+We addressed this concern in three ways: first, we focused our attention
+on the set of problems in the tutor in which students either requested
+a hint or made an error, second, with propensity score matching, and
+third with sensitivity analysis.
+None of these approaches is bulletproof---they each require further
+assumptions about how students request hints.
+Eliminating problems in which students showed no signs of struggling
+is likely to reduce the association between hint requests and
+underlying academic ability, but not eliminate it.
+For instance, stronger students may be better able to figure out a
+difficult problem without access to hints or error feedback than
+weaker students.
+Matching depends both on the adequacy of the observed covariates to
+capture baseline differences between $H=0$ and $H=1$ students, and the
+sensitivity analyses we performed assumed that any un-measured
+covariate would predict $H$ no better than pretest, or no better than
+a covariate half as important as pretest.
+These assumptions are impossible to verify from the data at hand;
+instead, they require a keen understanding of student practice within
+the tutor.
+
+Mediation analysis builds on the matching design, and also requires an
+ignorability assumption; it must
+additionally overcome the hurdle of the almost complete lack of
+overlap between hint request behavior in the control group (i.e. zero)
+and in the treatment group.
+Our approach of dichotomizing hint request behavior as $H$ provides a
+way forward, at the cost of ignoring information that could be gleaned
+from more continuous measures of hint requests.
+
+In contrast, no ignorabilty assumptions are necessary for principal
+stratification.
+That said, an assumption similar to ignorability may play an important
+role for interpretation.
+Further, principal stratification estimators require a model for $Y(0)$
+conditional on $\eta(1)$, which is unobserved.
+On the other hand, model misspecification is, in principle,
+detectable---an assumption of a particular family of models is not
+``untestable.''
+Still, the data will, in general, be unable to distinguish between certain
+alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
+turn lead to qualitatively different conclusions.
+In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
+distributed similarly to $Y(1)$ conditional on $\eta(1)$.
+
+An additional contrast of note between matching-based studies and
+principal stratification relates to transparency.
+However complex the process of devising a match may be, the resulting
+design is simple and clear---subjects with $H=1$ are compared against
+matches with $H=0$.
+This transparency may help researchers assess the plausibility of the
+match, and diagnose potential problems with resulting estimates.
+Though PS does not depend on ignorability, it does depend on a highly
+intricate and parametric modeling structure.
+If any part of the model is sufficiently misspecified, the results may
+be wrong.
+Perhaps worse, \citet{feller2016principal} argued that even well-specified
+models can give severely biased estimates.
+Extensive model interrogation and checking is necessary for PS
+analysis, but the complexity of the PS model makes this process
+particularly difficult.
+\citet{aoas} gives some examples of potentially fruitful model
+checking procedures, which are also carried out in the online supplement.
+
+\subsection{Results and Interpretation}
+
+Our observational study found that hint requests have a negative effect: students who
+requested hints more often than the 70th
+percentile score lower on the posttest, after controlling for observed covariates---as
+far as we can tell, the optimal strategy is to never request a hint.
+However, this observed effect may be due to confounding.
+Mediation analysis essentially interprets this result in terms of the
+overall CTA1 treatment effect.
+Specifically, our analysis showed that CTA1 affected students'
+posttest scores in (at least) two opposite ways: by allowing hints, it
+lowered their posttest scores, but its other mechanisms increased
+their scores by an even greater amount.
+
 Like in the observational study and in mediation analysis, the
@@
 
+\newpage
+\appendix
+
+\section{Rasch Mixture Model of Hint Usage}
+
+To model hint requests with a mixture model, we modeled each hint
+request as in \eqref{eq:rasch}.
+Then, instead of modeling $\eta_i$ as normal, we assigned it a
+two-component normal mixture model:
+\begin{equation*}
+\eta\sim
+p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
+\end{equation*}
+where we constrained $\mu_0<\mu_1$.
+
+\subsection{Stan Code}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## data{
+## //Sample sizes
+##  int<lower=1> N; // total number of obs (long format)
+##  int<lower=1> nstud; // # students
+##  int<lower=1> nsec; // # sections (problems nested in sections)
+## 
+## // indices
+##  int<lower=1,upper=nstud> studentM[N]; //for each obs, which student?
+##  int<lower=1,upper=nsec> section[N]; // for each obs, which section?
+## 
+## // data data
+##  int<lower=0,upper=1> hint[N]; // for each obs, was a hint requested?
+## 
+## }
+## parameters{
+## 
+##  vector[nstud] eta; // ability parameter
+##  real<lower=0,upper=1> p1; // mixing proportion
+##  //ordered[2] mu;             // locations of mixture components
+##  real<upper=0> mu0; //location of mixture component 0
+##  vector<lower=0>[2] sigma;  // scales of mixture components
+## 
+##  real delta[nsec]; // section difficulty parameter
+##  //real<lower=0> sigSec; // scale of difficulty parameters
+## 
+## }
+## transformed parameters{
+##  real<lower=0> mu1; // location of mixture component 1
+##  mu1= -(1-p1)/p1*mu0;
+## }
+## model{
+##  // logs of mixing proportions to save time
+##  real log_p1=log(p1);
+##  real log_p0=log(1-p1);
+## 
+## // linear predictor for logit model
+##  real linPred[N];
+## 
+## // priors
+##  sigma ~ lognormal(0, 2);
+##  mu0 ~ normal(0, 2);
+## // sigSec ~ lognormal(0,2);
+##  p1~ beta(2,2); // I'd rather the groups have similar sizes
+## 
+## // model for section difficulty parameters
+##  delta~normal(0,3);//sigSec);
+## 
+## // Rasch model for hint requests
+##  for(i in 1:N)
+##   linPred[i]= eta[studentM[i]]-delta[section[i]];
+## 
+##  hint~bernoulli_logit(linPred);
+## 
+## 
+## // mixture model for ability parameters
+##  for(n in 1:nstud)
+##   target += log_sum_exp(log_p0+normal_lpdf(eta[n]|mu0,sigma[1]),
+## 			log_p1+normal_lpdf(eta[n]|mu1,sigma[2]));
+## 
+## }
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+%\clearpage
+\section{Covariate Balance Before and After Matching}
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Tue Oct 29 14:14:20 2019
+\begin{table}[ht]
+\centering
+\begin{tabular}{lrrrrrr}
+  &\multicolumn{3}{c}{Unmatched}&\multicolumn{3}{c}{Matched}\\
+ \hline
+ & std.diff & z &  & std.diff & z &  \\ 
+  \hline
+grade9 & -0.34 & -7.05 & *** & 0.03 & 0.44 &     \\ 
+  gradehigher & 0.34 & 7.05 & *** & -0.03 & -0.44 &     \\ 
+  raceWhiteAsian & -0.64 & -12.83 & *** & 0.03 & 0.75 &     \\ 
+  raceBlackMulti & 0.23 & 4.71 & *** & 0.01 & 0.24 &     \\ 
+  raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
+  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
+  spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 
+  specspeced & 0.34 & 7.07 & *** & -0.10 & -1.44 &     \\ 
+  specgifted & -0.14 & -3.03 & **  & 0.06 & 1.16 &     \\ 
+  esl0 & -0.37 & -7.70 & *** & 0.04 & 0.63 &     \\ 
+  esl1 & 0.37 & 7.70 & *** & -0.04 & -0.63 &     \\ 
+  stateCT & 0.58 & 11.81 & *** & -0.00 & 0.00 &     \\ 
+  stateKY & -0.36 & -7.51 & *** & -0.00 & 0.00 &     \\ 
+  stateLA & -0.08 & -1.68 & .   & -0.00 & 0.00 &     \\ 
+  stateMI & -0.36 & -7.51 & *** & 0.00 & 0.00 &     \\ 
+  stateTX & 0.43 & 8.77 & *** & 0.00 & 0.00 &     \\ 
+   \hline
+\end{tabular}
+\caption{Covariate balance (standardized differences and Z-scores) before and after the propensity score match.} 
+\label{tab:balance}
+\end{table}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/balancePlot.pdf}
+\caption{Covariate balance (standardized differences) before and after
+  the propensity score match.}
+\label{fig:balance}
+\end{figure}
+
+%\clearpage
+\section{Stan Code for PS Model}
+
+%\clearpage
+\subsection{Stan Code}
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{verbatim}
+## data{
+## //Sample sizes
+##  int<lower=1> nsecWorked;
+##  int<lower=1> ncov;
+##  int<lower=1> nstud;
+##  int<lower=1> nteacher;
+##  int<lower=1> nsec;
+##  int<lower=1> nschool;
+##  int<lower=1> npair;
+## 
+## // indices
+##  int<lower=1,upper=nteacher> teacher[nstud];
+##  int<lower=1,upper=npair> pair[nstud];
+##  int<lower=1,upper=nschool> school[nstud];
+##  int<lower=1,upper=nstud> studentM[nsecWorked];
+##  int<lower=1,upper=nsec> section[nsecWorked];
+## 
+## // data data
+##  int<lower=0,upper=1> hint[nsecWorked];
+##  matrix[nstud,ncov] X;
+##  int<lower=0,upper=1> Z[nstud];
+##  real Y[nstud];
+## 
+## }
+## parameters{
+## 
+##  vector[nstud] eta;
+## 
+##  vector[ncov] betaU;
+##  vector[ncov] betaY;
+## 
+##  real a1;
+##  real b0;
+##  real b1;
+## 
+##  real teacherEffY[nteacher];
+##  real teacherEffU[nteacher];
+##  real pairEffect[npair];
+##  real schoolEffU[nschool];
+##  real schoolEffY[nschool];
+##  real delta[nsec];
+## 
+##  real<lower=0> sigTchY;
+##  real<lower=0> sigSclY;
+##  real<lower=0> sigY[2];
+##  real<lower=0> sigTchU;
+##  real<lower=0> sigSclU;
+##  real<lower=0> sigU;
+## }
+## 
+## model{
+##  real linPred[nsecWorked];
+##  vector[nstud] muY;
+##  vector[nstud] muU;
+##  real useEff[nstud];
+##  real trtEff[nstud];
+##  real sigYI[nstud];
+## 
+## 
+## // hint model
+##  for(i in 1:nsecWorked)
+##   linPred[i]= delta[section[i]]+eta[studentM[i]];
+## 
+##  for(i in 1:nstud){
+##   useEff[i]=a1*eta[i];
+##   trtEff[i]=b0+b1*eta[i];
+##   muU[i]=teacherEffU[teacher[i]]+schoolEffU[school[i]];
+##   muY[i]=teacherEffY[teacher[i]]+schoolEffY[school[i]]+
+##          pairEffect[pair[i]]+useEff[i]+Z[i]*trtEff[i];
+##   sigYI[i]=sigY[Z[i]+1];
+##  }
+## 
+##  //priors
+##  betaY~normal(0,2);
+##  betaU~normal(0,2);
+##  pairEffect~normal(0,2);
+## 
+##  a1~normal(0,1);
+##  b0~normal(0,1);
+##  b1~normal(0,1);
+## 
+## 
+##  schoolEffY~normal(0,sigSclY);
+##  schoolEffU~normal(0,sigSclU);
+##  teacherEffU~normal(0,sigTchU);
+##  teacherEffY~normal(0,sigTchY);
+## 
+##  hint~bernoulli_logit(linPred);
+## 
+##  eta~normal(muU+X*betaU,sigU);
+##  Y~normal(muY+X*betaY,sigYI);
+## }
+## generated quantities{
+## // int<lower=0,upper=1> hintRep[nsecWorked];
+##  real Yrep[nstud];
+## 
+## // hintRep=bernoulli_logit_rng(linPred);
+##  for(i in 1:nstud)
+##   Yrep[i] = normal_rng(teacherEffY[teacher[i]]+schoolEffY[school[i]]+
+##             pairEffect[pair[i]]+a1*eta[i]+Z[i]*(b0+b1*eta[i])+
+## 	    X[i,]*betaY,sigY[Z[i]+1]);
+## }
+\end{verbatim}
+\end{kframe}
+\end{knitrout}
+
+%\clearpage
+\section{Diagnostic Plots for Principal Stratification}
+
+\subsection{Posterior Predictive Density Plots}
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/ppcYoverall.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/ppcYtrt.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/ppcYctl.jpg}
+\end{figure}
+
+%\clearpage
+\subsection{Residual Plots}
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/residPlotOverall.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/residPlotTrt.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/residPlotCtl.jpg}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/binnedplot.pdf}
+\caption{Binned residual plot for model of hint requests.}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{output/qqPlots.pdf}
+\caption{Q-Q plots for outcomes and fitted values}
+\end{figure}
+
 \end{document}

### 2019-10-29 R/dataAnalysis.r

@@ -159,12 +159,13 @@ dat1 <-
   mutate(ntot=n(),ntrt=sum(M),nctl=ntot-ntrt,eff=mean(Y[M])-mean(Y[!M]),winv=(1/nctl+1/ntrt))
 
 
+ate <- update(anova,weights=winv)
 atevcv <- vcovHC(ate)
 atet <- coeftest(ate,atevcv)[2,]
 ateci <- coefci(ate,'MTRUE',vcov.=atevcv)
 
 
-tot <- update(ancova,weights=ntrt*winv)
+tot <- update(anova,weights=ntrt*winv)
 totvcv <- vcovHC(tot)
 tott <- coeftest(tot,totvcv)[2,]
 totci <- coefci(tot,'MTRUE',vcov.=totvcv)
@@ -228,12 +229,14 @@ if(getOption('mc.cores')<6)  options(mc.cores = 6)
 
 functions <- currentCode()
 
+cat(ifelse(runFull,'running','loading'),'ps model',as.character(Sys.time()),'\n')
 ### NOT RUN
 if(runFull){
   psmod1 <- stan('R/prinStratStan.stan',data=sdat,chains=6,iter=6000)
   save(psmod1,sdat,functions,file='fitModels/hintPS.RData')
 } else load('fitModels/hintPS.RData')
 
+cat('extracting ps model',as.character(Sys.time()),'\n')
 draws <- rstan::extract(psmod1)
 
 cat('

### 2019-10-29 R/diagnosticPlots.r

@@ -4,7 +4,7 @@ source('R/helperFunctions.r')
 library(bayesplot)
 
 if(!exists('draws')){
-  load('output/hintPS.RData')
+  load('fitModels/hintPS.RData')
   draws <- rstan::extract(psmod1)
 }
 if(!exists('sdat')){
@@ -20,14 +20,14 @@ draws1k <- lapply(draws, function(x) if(NCOL(x)==1) x[samp] else x[samp,])
 ppc_dens_overlay(sdat$Y,draws1k$Yrep)+ggtitle('Pooled Treatment and Control')
 ggsave('output/ppcYoverall.jpg')
 
-ppc_dens_overlay(sdat$Y[sdat$Z==1],t(Yrep)[,sdat$Z==1])+ggtitle('Treatment Group')
+ppc_dens_overlay(sdat$Y[sdat$Z==1],draws1k$Yrep[,sdat$Z==1])+ggtitle('Treatment Group')
 ggsave('output/ppcYtrt.jpg')
 
-ppc_dens_overlay(sdat$Y[sdat$Z==0],t(Yrep)[,sdat$Z==0])+ggtitle('Control Group')
+ppc_dens_overlay(sdat$Y[sdat$Z==0],draws1k$Yrep[,sdat$Z==0])+ggtitle('Control Group')
 ggsave('output/ppcYctl.jpg')
 
 ### residual plots
-fitd <- rowMeans(Yrep)
+fitd <- colMeans(draws$Yrep)
 qplot(fitd,sdat$Y-fitd,xlab='Fitted Values',ylab='Residuals',main='Overall')
 ggsave('output/residPlotOverall.jpg')
 
@@ -50,14 +50,14 @@ dev.off()
 ### usage model
 set.seed(613)
 samp <- sample(1:1000,9)
-lp <- with(c(sdat,draws1k),studEff[,studentM]+secEff[,section])
+lp <- with(c(sdat,draws1k),studEff[samp,studentM]+secEff[samp,section])
 prob <- exp(lp)/(1+exp(lp))
-p <- ppc_error_binned(sdat$grad,prob[samp,])
+p <- ppc_error_binned(sdat$grad,prob)
 ggplot2::ggsave('output/binnedplot.pdf',p)
 
-### ppc m-bar
-mbarRep <- apply(prob,1,function(p) aggregate(rbinom(length(sdat$grad),1,p),by=list(sdat$studentM),FUN=mean)$x)
[…]

### 2019-10-29 R/prelimStan.r

@@ -93,7 +93,7 @@ xOrig <- x
     mutate(prevErr=lag(nerrs1))%>%
     filter(!is.na(prevErr),prevErr>0)
 
- save(dat,x,file='data/hintsData.RData')
+ save(dat,x,covs,file='data/hintsData.RData')
 
 
 

### 2019-11-28 paper.Rnw
John's suggestions thru Sec 2.1

@@
 \section{Introduction}
-In the 2007--2008 and 2008--2009 school years, the RAND corporation
+From 2007 through 2010, the RAND corporation
 conducted one of the first effectiveness trials awarded through
@@
 The study \citep{pane2014effectiveness} found no effects in its first year, but in
-the second year of implementation students in schools randomized to
+the second year of implementation students in high schools randomized to
 the CTA1 condition outperformed the control group on the posttest by a fifth of a
-standard deviation.
+standard deviation.\footnote{\citep{pane2014effectiveness} only
+  reports results from the first two years of implementation; the
+  third year included only middle schools, not high schools.}
 
@@
 under study, each with their own unique designs, opens a wide set of opportunities to understand how features of those
-products drive their effects. As one example, RAND is currently
-completing an efficacy study of a different algebra tutoring system,
+products drive their effects. As one example, RAND is recently
+completed an efficacy study of a different algebra tutoring system,
 ALEKS. When students begin working on a particular topic in ALEKS, the
@@
 The first approach discards the control group, and analyzes
-hint-requesting in the treatment group as an observational study
+hint-requesting in the treatment group as an observational study,
+using a matching design
 \citep[c.f.][]{rosenbaum2002observational}. The
@@
 modified.
-Finally, we will conduct a principal stratification analysis
+Finally, we conduct a principal stratification analysis
 \citep{frangakis} to learn if students requesting different amounts of
@@
 Our contribution will build on this literature in two ways: first, by
-providing a detailed demonstration of matching, mediation analysis,
+providing a detailed demonstration and comparison of matching, mediation analysis,
 and principal stratification on the same dataset, and second by
 focusing on the applicability of these three methods in the specific
-context of studying implementation data from an EdTech randomized
-experiment.
-
-After a brief overview of the three causal approaches we will discuss,
+context of studying implementation data from an randomized
+experiment studying educational technology (EdTech).
+
+After a brief overview of the three causal approaches,
 the next section will provide background for the case-study,
@@
 The computerized tutor was originally built to test the ACT and ACT-R
-theories of skill acquisition \citep{anderson2013architecture}.
-These are elaborate theories describing the necessary components of
+theories of cognition \citep{anderson2013architecture}.
+These are elaborate theories that describe, among other things, the necessary components of
 cognitive skills in, say, mathematics or computer programming, and the

### 2019-11-28 paper.Rnw
data section to the end of Sec 2

@@
 
-As students work through specific problems, they receive two types
-of instruction from the tutor.
-When they err---that is, depart from one of the pre-specified correct
-solution paths for that problem---they receive an error message
-alerting them to their mistake and helping them return to the correct
-solution path.
-Students may also request help, in the form of hints, that are also
-tailored to the specific context and solution path they are on.
-
-\subsection{Data}\label{sec:data}
-
-Since our goal was to better understand the CTA1
-treatment effect, we focused our analysis on data from high school students in
-the second year of the CTA1 trial, for whom the treatment effect was
-most evident.
-
-We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
-Table \ref{tab:covariateBalance} describes the covariates we used, including
-missingness information, control and treatment means, and standardized
-differences \citep[c.f.][]{kalton1968standardization}.
-We singly-imputed missing values with the Random Forest routine implemented
-by the \texttt{missForest} package in \texttt{R}
-\citep{missForest,rcite},  which estimated the ``out of box'' imputation
-errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.
-
-\begin{table}[ht]
-\centering
-\include{output/covariateTable}
-\caption{Missingness information, control (``BaU'' or ``Business as
-  Usual'') and treatment (``CTA1'') means, and balance for the
-  covariates included in this study, from the high school year two
-  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
-  categorical variables (Race/Ethnicity, Sex, and Special Education)
-  and standardized root mean squared error for Pretest, which is
-  continuous. %The overall p-value accounts for clustering at the
-  %school level and blocked random sampling.
-  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
-\label{tab:covariateBalance}
-\end{table}
-
-Over the course of the effectiveness study, Carnegie Learning gathered
-log data for each worked problem.
-The resulting dataset lists the problem name, section, and unit of
-each problem, the numbers of hints requested and errors committed, and time-stamps.
-Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
-Treatment schools with mastery data missing
-for 90\% or more students were omitted from the analysis, along with
-their entire randomization block.
-Of the remaining \Sexpr{Nt} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
-data.
-For the sake of simplicity, the treatment students without log data
-were dropped from the study.\footnote{Including these students in a
-  principal stratification model is straightforward \citep{aoas}; see
-  the online supplement for results from a model including them, which
-  were broadly similar to the results presented here. Including
-  subjects with missing log data in a mediation or observational study
-  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
-Problem log data for sections that were not part of the
-standard CTA1 Algebra I curriculum and problems worked by fewer than 100
-students were omitted from the
-dataset.
-
-All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
-The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
-
 \subsection{The Role of Hints\footnote{This subsection draws heavily
@@
 problem.
-As soon as a student errs in working through the problem, the
-software provides feedback tailored to the specific mistake, and
+As soon as a student errs in working through a problem---that is, departs from one of the
+solution paths---the
+software shows the student an error message.
+This feedback is tailored to the specific mistake, and
 designed to guide the student back to a correct solution path.
@@
 \subsection{Measuring Hint Usage in the CTA1 Data}
+\marginpar{make sure this makes sense before ``Data'' section}
 Unlike previous studies of hint usage, the CTA1 study was not designed
@@
 approximately \Sexpr{round(agree*100)}\% of cases.
+
+\subsection{Data}\label{sec:data}
+
+Since our goal was to better understand the CTA1
+treatment effect, we focused our analysis on data from high school students in
+the second year of the CTA1 trial, for whom the treatment effect was
+most evident.
+
+We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
+Table \ref{tab:covariateBalance} describes the covariates we used, including
+missingness information, control and treatment means, and standardized
+differences \citep[c.f.][]{kalton1968standardization}.
+We singly-imputed missing values with the Random Forest routine implemented
+by the \texttt{missForest} package in \texttt{R}
+\citep{missForest,rcite},  which estimated the ``out of box'' imputation
+errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.
+
+\begin{table}[ht]
+\centering
+\include{output/covariateTable}
+\caption{Missingness information, control (``BaU'' or ``Business as
+  Usual'') and treatment (``CTA1'') means, and balance for the
+  covariates included in this study, from the high school year two
+  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
+  categorical variables (Race/Ethnicity, Sex, and Special Education)
+  and standardized root mean squared error for Pretest, which is
+  continuous. %The overall p-value accounts for clustering at the
+  %school level and blocked random sampling.
+  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
+\label{tab:covariateBalance}
+\end{table}
+
+Over the course of the effectiveness study, Carnegie Learning gathered
+log data for each worked problem.
+The resulting dataset lists the problem name, section, and unit of
+each problem, the numbers of hints requested and errors committed, and time-stamps.
+Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
+Treatment schools with mastery data missing
+for 90\% or more students were omitted from the analysis, along with
+their entire randomization block.
+Of the remaining \Sexpr{Nt} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
+data.
+For the sake of simplicity, the treatment students without log data
+were dropped from the study.\footnote{Including these students in a
+  principal stratification model is straightforward \citep{aoas}; see
+  the online supplement for results from a model including them, which
+  were broadly similar to the results presented here. Including
+  subjects with missing log data in a mediation or observational study
+  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
+Problem log data for sections that were not part of the
+standard CTA1 Algebra I curriculum and problems worked by fewer than 100
+students were omitted from the
+dataset.
+
+All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
+The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
 

### 2019-11-28 paper.Rnw
john's sec 2 comments

@@
 
+\citep{pane2014effectiveness} estimated effects separately in middle
+schools and high schools in each the first two years of the RCT; the
+only statistically significant treatment effect was in the high school
+sample in the second year.
 Since our goal was to better understand the CTA1
@@
 
-We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
+We merged data from two sources: computerized log data gathered by
+Carnegie Learning, and covariate, treatment and outcome data gathered by RAND.
+
+The log data lists the problem name, section, and unit of
+each problem, the numbers of hints requested and errors committed, and time-stamps.
+Log data were missing for some students, either because the log files
+were not retrievable, or because of an imperfect ability to link log
+data to other student records.
+Further, log data for sections that were not part of the
+standard CTA1 Algebra I curriculum and problems worked by fewer than 100
+students were omitted from the
+dataset.
+
+To construct our analysis dataset, we first dropped treatment schools with log data missing
+for 90\% or more students.
+Prior to randomization, schools were stratified into pairs or triples
+based on baseline covariates, and randomized between treatment and
+control condition within these randomization blocks.
+When we dropped a treatment school from our analysis, we also dropped
+the control schools in its randomization block.
+Of the remaining \Sexpr{prettyNum(Nt,',')} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
+data.
+Then, for the sake of simplicity, the treatment students without log data
+were dropped from the study.\footnote{Including these students in a
+  principal stratification model is straightforward \citep{aoas}; see
+  the online supplement for results from a model including them, which
+  were broadly similar to the results presented here. Including
+  subjects with missing log data in a mediation or observational study
+  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
+
+All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
+The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
+
 Table \ref{tab:covariateBalance} describes the covariates we used, including
 missingness information, control and treatment means, and standardized
-differences \citep[c.f.][]{kalton1968standardization}.
+differences \citep[c.f.][]{kalton1968standardization} from the final
+analysis sample.
 We singly-imputed missing values with the Random Forest routine implemented
 by the \texttt{missForest} package in \texttt{R}
-\citep{missForest,rcite},  which estimated the ``out of box'' imputation
-errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.
+\citep{missForest,rcite},  which estimated ``out of box'' imputation
+error rates as part of the random forest regression, also shown in Table \ref{tab:covariateBalance}.
 
@@
 \end{table}
-
-Over the course of the effectiveness study, Carnegie Learning gathered
-log data for each worked problem.
-The resulting dataset lists the problem name, section, and unit of
-each problem, the numbers of hints requested and errors committed, and time-stamps.
-Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
-Treatment schools with mastery data missing
-for 90\% or more students were omitted from the analysis, along with
-their entire randomization block.
-Of the remaining \Sexpr{Nt} students, \Sexpr{round(obsUse/Nt*100)}\% (\Sexpr{prettyNum(obsUse,',')}) had log
-data.
-For the sake of simplicity, the treatment students without log data
-were dropped from the study.\footnote{Including these students in a
-  principal stratification model is straightforward \citep{aoas}; see
-  the online supplement for results from a model including them, which
-  were broadly similar to the results presented here. Including
-  subjects with missing log data in a mediation or observational study
-  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
-Problem log data for sections that were not part of the
-standard CTA1 Algebra I curriculum and problems worked by fewer than 100
-students were omitted from the
-dataset.
-
-All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
-The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
 

### 2019-11-28 R/covariateTable.r

@@ -1,6 +1,5 @@
 ## covariate means/balance
-
-covs2 <- covs[rownames(dat),] ### same as dat but with NAs
+covs2 <- covs[rownames(dat)[dat$field_id%in%x$field_id|dat$treatment==0],] ### same as dat but with NAs
 
 miss <- NULL
 varbs <- c('grade','esl','frl','race','sex','spec','xirt')

### 2019-11-28 R/dataAnalysis.r

@@ -25,6 +25,7 @@ source('~/Box Sync/rcode/matchingFunctions.r')
 source('~/Box Sync/rcode/hhh.r')
 source('~/Box Sync/rcode/printXbal.r')
 
+load('~/Box Sync/CT/data/RANDstudyData/HSdata.RData') ## for "error"
 if(runFull) source('R/prelimStan.r') else load('data/hintsData.RData')
 
 ## for printing nice CIs

### 2019-12-03 paper.Rnw
jp comment 19

@@
 these data.
-What is relevant for our purposes is that we may narrow down the type
+
+For the current purpose, we narrow down the type
 of hint data we model to align with our research focus and to avoid
 some modeling pitfalls.
-
 Since our posttest measured Algebra I skills, we only

### 2019-12-03 paper.Rnw
jp comment 21
moved text about rasch model from sec 5 (principal strat) to sec 2.4
(measuring hint usage)

adjusted text in sec 5

small change to dichotomization discussion in sec 4.1 (mediation)

@@
 For the current purpose, we narrow down the type
-of hint data we model to align with our research focus and to avoid
-some modeling pitfalls.
+of hint data we model to align with our research focus.
 Since our posttest measured Algebra I skills, we only
@@
 requesting any hints; therefore, the variable measuring a student's
-hint request on a problem was set one if the student requested any
+hint request on a problem was set to one if the student requested any
 hints on that problem, and zero otherwise.
@@
 
-The observational study and (especially) the mediation analysis we
-describe below require a dichotomous measurement of hint usage: is a
+\subsubsection{Dichotomizing Hint Usage}\label{sec:dichotomizingHintUsage}
+The observational study and mediation analysis below require a dichotomous measurement of hint usage: is a
 student a high or low hint user?
@@
 and 0 otherwise.
-To choose $c$, we fit a two-component modified Rasch mixture model
-\citep{rasch} to hint request data, described in an online appendix.
-The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
+
+To choose $c$, we first fit a modified Rasch mixture model
+\citep{rasch} to hint request data.
+Specifically, we modeled the probability of a hint request in each problem
+$p$, worked by student $i$, as
+\begin{equation*}\label{eq:rasch}
+Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
+\end{equation*}
+where $h_{ip}$ is the event that student $i$ requests a hint on
+problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
+section-level parameter for $s[p]$, the section that problem $p$ was
+drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
+function.
+Typically, the student parameter $\eta_i$ measures student ability;
+here, instead, it measures a student's proclivity to request a hint.
+The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
+dataset many problems were worked by very few students, so problem
+level parameters would be hard to estimate.
+Instead, we included the section-level parameter $\delta_{s[p]}$,
+which measures the extent to which students tend to request hints on
+problems in section $s$.
+This essentially assumes that all problems within each section are equally
+likely to elicit a hint.
+
+Then, in order to dichotomize hint request behavior, we modeled
+student effects $\eta$ as
+\begin{equation}\label{eq:mixture}
+\eta\sim
+p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
+\end{equation}
+and constrained $\mu_0<\mu_1$.
+This mixture model clusters students into two categories: high hint
+users have $\eta$ drawn from a normal distribution with mean $\mu_1$
+and standard deviation $\sigma_1$, and low hint users have $\eta$
+drawn from a normal distribution with mean $\mu_0$ and standard
+deviation $\sigma_0$.
+
+Our main parameter of interest here is $p_0$, the proportion of
+students who are low-hint users.
+We estimated $p_0\approx$\Sexpr{prettyNum(round(1-mixProp,2))},
+classifying \Sexpr{round(mixProp*100)}\% of students as
 high-hint users, so we chose $c$ as the
@@
 approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
-This definition agreed with the model's classification in
+This definition agreed with the model's classification (based on
+$\eta$ as opposed to $\bar{h}$ in
 approximately \Sexpr{round(agree*100)}\% of cases.
+
+The Rasch model for hint useage \eqref{eq:rasch}, but not the mixture
+model \eqref{eq:mixture} will play an important role in the principal
+stratification analysis of Section \ref{sec:principalStratification}.
 
@@
 problems.
-As an alternative, we fit a
-modified Rasch model \citep[e.g.][]{rasch},
-\begin{equation}\label{eq:rasch}
-Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
-\end{equation}
-where $h_{ip}$ is the event that student $i$ requests a hint on
-problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
-section-level parameter for $s[p]$, the section that problem $p$ was
-drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
-function.
-Typically, the student parameter $\eta_i$ measures student ability;
-here, instead, it measures a student's proclivity to request a hint.
-The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
-dataset many problems were worked by very few students, so problem
-level parameters would be hard to estimate.
-Instead, we included the section-level parameter $\delta_{s[p]}$,
-which measures the extent to which students tend to request hints on
-problems in section $s$.
-This essentially assumes that all problems in each section are equally
-likely to elicit a hint.
+In section \ref{sec:dichotomizingHintUsage}, we showed that
+dichotomized $\bar{h}$ largely agrees with a dicotomized version of a
+more sophisticated measure of hint usage.
+However, when modeling hint usage continuously, the correspondence may
+not hold \citet[see][for a more complete discussion]{aoas}.
+
+For those reasons, we modeled hint usage at the problem level, with
+equation \eqref{eq:rasch}, but with two important differences.
+First, a conceptual difference: the $\eta$ in \eqref{eq:rasch} was
+replaced by $\eta (1)$, indicating that it is measuring
+\emph{potential} hint usage---the hint usage a student would exhibit
+were he or she assigned to the treatment condition.
+Whereas $\eta$ is only defined for students assigned to the treatment condition,
+$\eta(1)$ is defined for all the students in the study.
+Second, a modeling difference: since we are not dichotomizing $\eta(1)$
+in this analysis, we did not model it as a normal mixture, as in
+\eqref{eq:mixture}.
 
 We incorporated the model for hint requests, (\ref{eq:rasch}), into
-the larger principal stratification model, replacing $\eta$ for
-$\bar{h}$.
+the larger principal stratification model, replacing $\eta(1)$ for
+$\bar{h}(1)$.
 Then, the principal stratification estimand is

### 2019-12-03 paper.Rnw
jp comment 22-23
- took out ANOVA estimate
- added summary of results, interpreted (lack of) differences
- clarified HC standard errors
- left in 0s for rho and Tz, but explained why they're there

@@
 Our main parameter of interest here is $p_0$, the proportion of
-students who are low-hint users.
+students who are low hint users.
 We estimated $p_0\approx$\Sexpr{prettyNum(round(1-mixProp,2))},
 classifying \Sexpr{round(mixProp*100)}\% of students as
-high-hint users, so we chose $c$ as the
+high hint users, so we chose $c$ as the
 \Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
@@
 How does requesting hints within CTA1 affect learning?
-More precisely, would students have achieved higher posttest scores
-had they requested hints more (or less) often?
+More precisely, would low hint users have achieved higher posttest scores
+had they requested hints more often?
+Would high hint users have achieved higher posttest scores had the
+requested hints less often?
 
@@
 racial categories, grade, ESL status, economic disadvantage, and sex,
-along with a natural spline with five degrees of freedom for pretest (the standard error
-and confidence interval using the ``HC3'' sandwich estimator,
-\citealt{sandwichPackage}).
+along with a natural spline with five degrees of freedom for pretest.
 This regression adjusts for imbalances in these covariates within
-matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
-the others.
+matched sets \citep[c.f.][]{biasAdjust}.
+All three sets of standard errors
+and confidence intervals used the heteroskedasticity-robust ``HC3'' sandwich estimator,
+\citealt{sandwichPackage}.
+
+The three estimates all roughly agree that being a high hint user
+decreases posttest scores by nearly a quarter of a standard deviation.
+That is, requesting hints hurts posttest scores---teachers should
+recommend against students' requesting hints.
+
+The similarity of the three estimates boosts our confidence, if only
+slightly. The ``ATE'' and ``TOT'' estimates refer to different groups
+of students---ATE estimates an average effect for the entire treatment
+group, whereas TOT estimates an effect for high hint users. Had these
+been very different, we might wonder why the effect of hint requests
+varies so much between high and low hint users, and if that suggests
+selection bias.
+The fact that the additional covariate adjustment of the ANCOVA model
+made little difference in the effect estimate suggests that the
+propensity score match eliminated all or most of the confounding bias due to
+measured covariates; hence, additional adjustment was unnecessary.
 
@@
   cbind(
-    data.frame(Effect=c('ANOVA','ATE','TOT','ANCOVA')),
-    rbind(anovat,atet,tott,ancovat)[,c('Estimate','Std. Error')],
-    CI=vapply(list(anovaci,ateci,totci,ancovaci),printci,'a'),
+    data.frame(Effect=c('ATE','TOT','ANCOVA')),
+    rbind(atet,tott,ancovat)[,c('Estimate','Std. Error')],
+    CI=vapply(list(ateci,totci,ancovaci),printci,'a'),
     `$T_Z$`=0,
@@
   )
+
+stopifnot(min(matchTEtab[,1])<0.2)
+stopfinot(max(matchTEtab[,1])>0.25)
 
@@
 
-Matched ignorability is a very strong assumption, so these results
-should not be taken at face value---especially if weaker students ask for hints at
+None of these estimates addressed the possiblity of confounding from
+unmeasured covariates; instead, they relied on matched ignorability.
+This would be a particular concern if pretest scores do not completely
+capture students' mathematics ability, and if weaker students ask for hints at
 higher rates, and also score lower on the posttest.
-While adjusting for pre-test scores via matching and/or regression
-may account for some of that bias, pre-test may not capture all of the
-relevant pre-treatment information.
+Then, whatever aspects of student ability are not captures by the
+pretest, and correlate with both hint requests and posttest scores,
+would confound the estimates in the upper part of Table
+\ref{tab:matchResults}.
+
 To account for the possibility of unmeasured confounding, we conducted
@@
 counterparts when observed covariates are left out of the model.
+The first three rows of the table have both $\rho^2$ and $T_Z$ equal
+to zero, to reflect that they assume no unmeasured confounders.
 The row of Table \ref{tab:matchResults} labeled ``[Pretest]''
@@
 The resulting sensitivity interval is quite wide, including
-unbelievably large negative effects as well as substantial positive
+both large negative effects as well as substantial positive
 effects.
@@
 
-Our solution was to dichotomize hint requests, as
-described above, essentially re-defining the causal question.
-\Sexpr{round((mixProp)*100)}\% of the treatment group and all of the control group has
-$H=0$---requesting hints at a rate less than the
-\Sexpr{round((1-mixProp)*100)}th percentile of the treatment group.
+Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
+the causal question.
 To make sense of the respective roles of $H$ and $\bar{h}$, the
-continuous measure on which it is based, we may imagine a process in
+continuous measure on which it is based (see Section
+\ref{sec:dichotomizingHintUsage}),
+we may imagine a process in
 which students are assigned $H$, and under that constraint choose a

### 2019-12-03 paper.Rnw
jp comments 24-25

@@
 In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
-in the treatment group half of students have $H=1$ and half have
+in the treatment group \Sexpr{round((1-mixProp)*100)}\% of students have $H=1$ and \Sexpr{round((mixProp)*100)}\% have
 $H=0$, and $\bar{h}$ varies within those groups.

### 2019-12-03 paper.Rnw
jp comments 26-28

@@
 estimation is instead taken to be super-population average effects,
-$\EE[\delta(1)]$ and $\EE[\xi(0)]$. Our approach to estimating these
+$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$. Our approach to estimating these
 is to estimate the means of each of the component potential outcomes,
@@
 The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
-Y(0,H(0))=Y(Z=0)$ are non-parametrically identified due to the
+Y(0,H(0))=\EE Y(Z=0)$ are non-parametrically identified due to the
 randomization of treatment assignment.
@@
 In a typical mediation analysis of an RCT, identification of the
-distribution of $H(0)$ in the population would derive from
-randomization, along with an observation of $H$ in the control group.
-In contrast, in the CTA1 study, $H(0)$ is not observed but is
-determined by the experimental design: since the CTA1 software was
-unavailable to the control group, $H(0)=0$ for all experimental
-participants.
-Therefore, $Y(1,H(0))\equiv Y(1,0)$.
-For treated subjects $i$ for whom $H_i=0$, then, $Y_i(1,H(0))=Y_i$,
-the observed outcome is equal to the counterfactual outcome.
-For these subjects, there is no indirect effect.
+distribution of $H(0)$ may be estimated based on the observed
+distribution of $H$ in the control group, due to randomization.
+In contrast, in the CTA1 study, $H$ is not observed in the control
+group.
+However, $H(0)$ was determined by the experimental design: since the
+CTA1 software was unavailable to the control group, $H(0)=0$ for all
+experimental participants, and $Y(1,H(0))\equiv Y(1,0)$.
+
+For treated subjects $i$ for whom $H_i=0$, then, treatment assignment
+did not affect $H$---$H_i(0)=H_i(1)=0$.
+In other words, for these subjects,
+\begin{equation*}
+Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
+\end{equation*}
+Therefore, when $Z_i=1$ and $H_i=0$,
+$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
+indirect effect.
 
@@
 \ref{sec:observational}.
-If $Y_i(1,H(0))$ is independent of $Z_i$ within matched sets, then we
-may estimate counterfactual potential outcomes for $H=1$ subjects
-using the observed outcomes of their matches with $H=0$.
+Each member of the treatment group with $H=1$ is matched to other
+treatment-group subjects with $H=0$.
+Matched ignorability states that within these matched sets, $Y(1,0)$
+is independent of $H$.
+Therefore, if subject $i$ is matched to subject $j$, $H_i=1$, and
+$H_j=0$, we may estimate $Y_i(1,0)$ based on the observed
+$Y_j=Y_j(1,0)$.
+When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
+$Y_i(1,0)$ is estimated with the average of their observed outcomes.
+
 Specifically, let
-\begin{equation*}
-\hat{Y}_i(1,H(0))=\begin{cases}
+\begin{equation}\label{eq:estY10}
+\hat{Y}_i(1,0)=\begin{cases}
  Y_i & H_i=0\\[3ex]
@@
 \end{cases}
-\end{equation*}
+\end{equation}
 For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
 average outcome of their matched pairs with $H=0$.
-Then, the average indirect effect in the treatment group can be
-estimated as:
+
+Now, for each member $i$ of the treatment group we have observed
+$Y_i(1,H(1))=Y_i$, and estimated or inferred
+$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
+Then, we can estimate $\Delta$, the average indirect effect in the
+treatment group, by averaging over the treatment group:
 \begin{equation*}
@@
 data from only the treatment group.
-TOT is the the difference between the average outcome for students in
-the treatment group with $H=1$, and the average outcome under the
-counterfactual scenario in which they all had $H=0$.
+TOT is the the average difference between the realized outcomes for students in
+with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
 That is,
 \begin{equation*}
-TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
+TOT=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]
 \end{equation*}
+
 $\hat{\Delta}$ estimates the difference between the average outcome
@@
 \begin{equation*}
-\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
+\Delta=\EE[Y(1,H(1))-Y(1,0)|Z=1]
 \end{equation*}
-This can be re-written (omitting conditioning on $Z$):
+This can be re-written:
 \begin{align*}
 \begin{split}
-\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
-&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
+\Delta&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
+&\qquad +\EE[Y(1,0)-Y(1,0)|Z=1,H=0]Pr(H=0)
 \end{split}
 \\[2ex]
-&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
+&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)
+&=Pr(H=1)TOT
 \end{align*}

### 2019-12-17 R/dataAnalysis.r
fiddling with matching estimator

@@ -8,6 +8,7 @@ library(mgcv)
 library(tidyverse)
 library(rstan)
 library(lme4)
+library(estimatr)
 traceplot <- rstan::traceplot
 select <- dplyr::select
 options(stringsAsFactors=FALSE)
@@ -31,6 +32,12 @@ if(runFull) source('R/prelimStan.r') else load('data/hintsData.RData')
 ## for printing nice CIs
 printci <- function(ci) paste0('[',paste(sprintf('%.2f',ci),collapse=','),']')
 
+### I don't know why this is necessary...
+map_dfr <- function(.x, .f, ...){
+  out <- map(.x,.f,...)
+  as.data.frame(do.call('rbind',out))
+}
+
 citopm <- function(ci) paste(sprintf('%.2f',c(median(ci),(ci[2]-ci[1])/2)),collapse='$\\pm$')
 
 ## usage data availability
@@ -48,7 +55,7 @@ nschool <- sdat$nschool
 
 source('R/covariateTable.r')
 
-source('R/evaluateSmooths.r')
+#source('R/evaluateSmooths.r')
 
 overallHintInfo <-
   xOrig%>%
@@ -93,10 +100,12 @@ hintAvgStud%>%
   geom_point()+
   facet_wrap(~xaxis,scales="free_x")
 
-
+runFull <- FALSE
 ### rasch mixture model?
 ##NOT RUN
 if(runFull) source('R/raschMixture.r') else load('output/raschMixOutput.RData')
+
+runFull <- TRUE
 
 mixProp <- round(mean(hintAvgStud$classProb),1)
 
@@ -113,9 +122,6 @@ cat('
 ### propensity score matching
 ##########################################################################################
 ',as.character(Sys.time()),'\n')
-##NOT RUN
-## this is a record of stuff I tried w/ matching
-## source('R/matching.r')
 
 if(runFull){
   psmod3.2 <- ## stan_
@@ -144,35 +150,35 @@ cat('
 ',as.character(Sys.time()),'\n')
 
 anova <- lm(Y~M+match,data=dat1)
-anovavcv <- vcovHC(anova)
+anovavcv <- vcovHC(anova,'HC2')
[…]

### 2019-12-17 paper.Rnw
description of matching estimators

@@
 
-Let $m_i$ be a categorical variable denoting the matched set that $i$
+\subsection{Estimating the Effects of $H$ on Posttest Scores}
+
+Let $M_i$ be a categorical variable denoting the matched set that $i$
 belongs to.
@@
 \begin{equation*}
- \{Y(H=1),Y(H=0)\}\independent H |m
+ \{Y(H=1),Y(H=0)\}\independent H |M
 \end{equation*}
@@
 effect for students in that match.
-The (weighted) mean of those estimated treatment effects estimates the
-overall treatment effect in the sample.
-
-% The classic approach to estimating effects from blocked randomization
-% designs is a fixed-effects regression \citep{fisher1926arrangement},
-% regressing outcomes on an indicator for treatment assignment and a set
-% of indicators for matched sets.
-% The results from this analysis are displayed on the row labeled
-% ``ANCOVA'' in Table \ref{tab:matchResults}, with ``HC3'' standard
-% errors and 95\% confidence intervals.
-% They suggest that a greater-than-median proclivity to request hints
-% \emph{hurts} student learning by a large amount, roughly equal to the
-% entire treatment effect of CTA1 estimated in
-% \citet{pane2014effectiveness}.
-Table \ref{tab:matchResults} gives three treatment effect estimates
-under matched ignorability.
+In other words, if $\tau_m=\overline{Y(H=1)}-\overline{Y(H=0)}$ is the
+average effect in match $m$, then
+$\hat{\tau}_m=\bar{Y}_{M=m,H=1}-\bar{Y}_{M=m,H=0}$ is an unbiased
+estimate of $\tau_m$, where $\bar{Y}_{M=m,H=h}$ is the average
+observed outcome for subjects in match $m$ with hint level
+$h\in\{0,1\}$.
+
+Weighted means of those estimated treatment effects, of the form
+\begin{equation}\label{eq:tauWeightedMean}
+\hat{\tau}_w=\sum_m w_m\hat{\tau}_m/\sum_m w_m
+\end{equation}
+estimate aggregate
+treatment effects in the sample.
+To estimate the average effect of $H$ for all subjects, set $w_m=
+n_{1m}+n_{0m}$, where $n_{1m}$ and $n_{0m}$ are the number of
+$H=1$ and $H=0$ subjects in match $m$, respectively.
+Alternatively, setting $w_m=n_{1m}$ estimates the average effect of
+$H$ on those subjects for whom $H=1$---the ``treatment on the
+treated,'' or TOT, effect.
+Weights $w_m=(1/n_{1m}+1/n_{0m})^{-1}$, the harmonic mean  are
+``precision weights'' \citep[c.f.][]{schochet2015statistical}; the
+estimate using these weights is equal to the coefficient on $H$ from
+an ordinary least squares (OLS) regression of $Y$ on $H$ and a set of dummy
+variables for $M$.
+Under standard OLS assumptions, precision weights minimize the
+standard error of the weighted mean estimator.
+
+% Effect estimates may be improved by combining matching with regression
+% adjustment
+% \citep[e.g.]{rubin2000combining,rosenbaum2002covariance,ho:etal:2007}.
+% Regression can improve matching estimators both by correcting for
+% violations of Matched Ignorability due to observed covariate
+% imbalances within matched sets (hence reducing the estimate's bias),
+% and by explaining a portion of the variance in $Y$ and hence reducing
+% the estimate's standard error.
+% Here we use the regression adjustment method described in
+% \citet{hansen2009b}, which takes place in three steps: (1) regressing posttest scores for students with
+% $H=0$ on indicators for racial categories, grade, ESL status, economic disadvantage, and sex,
+% and with a natural spline with five degrees of freedom for pretest,
+% yielding a vector of coefficient estimates $\bm{\hat{\beta}}$;
+% (2) computing residuals $e_i=Y_i-\bm{\hat{\beta}}^T\bm{\tilde{x}}_i$,
+% where $\tilde{x}_i$ is the vector of covariates for subject $i$,
+% for every unit $i$ in the sample; and (3) estimating effects in
+% equation \eqref{eq:tauWeightedMean} with
+% $\hat{tau}_m=\bar{e}_{M=m,H=1}-\bar{e}_{M=m,H=0}$, that is, with $e$
+% in place of $Y$.
+
+Table \ref{tab:matchResults} gives four treatment effect estimates
+using these methods.
+The first row gives the result using precision-weights without
+regression adjustment. estimating an effect of $H$
 The rows labeled ATE and TOT weight the matched sets to
@@
 The three estimates all roughly agree that being a high hint user
-decreases posttest scores by nearly a quarter of a standard deviation.
+decreases posttest scores by roughly a fifth of a standard deviation.
 That is, requesting hints hurts posttest scores---teachers should
@@
   cbind(
-    data.frame(Effect=c('ATE','TOT','ANCOVA')),
-    rbind(atet,tott,ancovat)[,c('Estimate','Std. Error')],
-    CI=vapply(list(ateci,totci,ancovaci),printci,'a'),
-    `$T_Z$`=0,
-    `$\\rho^2$`=0
+    tibble(
+      Weights=c('Precision','Precision','ATE','TOT'),
+      Regression=c('No','Yes','Yes','Yes')
+    ),
+    map_dfr(
+      effects[c('prec_eff','prec_adj','ate_adj','tot_adj')],
+      ~tibble(
+        Estimate=unname(.$coefficients),
+        "Std. Error"=unname(.$std.error),
+        CI=printci(c(.$conf.low,.$conf.high)))
+     ),
+      `$T_Z$`=NA,
+    `$\\rho^2$`=NA
   )
 
-stopifnot(min(matchTEtab[,1])<0.2)
-stopfinot(max(matchTEtab[,1])>0.25)
+stopifnot(max(matchTEtab[,'Estimate'])<0.25)
+stopifnot(min(matchTEtab[,'Estimate'])>0.15)
 
@@
 data from only the treatment group.
-TOT is the the average difference between the realized outcomes for students in
+TOT is the average difference between the realized outcomes for students in
 with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
@@
 a standard error of \Sexpr{round(tott[2]*(1-mixProp),2)}.
-This analysis suggests that the availability of excessive hints
-actually \emph{lowers} the treatment effect.
-CTA1 works not because of hints, but despite them.
+This analysis suggests that the availability of hints
+actually \emph{lowers} the treatment effect, that CTA1 works not
+because of hints, but despite them.
 
@@
 Like indirect effects, this expression includes the average potential
-outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
+outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
 potential outcomes under the control condition.

### 2019-12-17 paper.Rnw
take out regression

@@
     tibble(
-      Weights=c('Precision','Precision','ATE','TOT'),
-      Regression=c('No','Yes','Yes','Yes')
+      Weights=c('OLS','ATE','TOT'),
+#      Regression=c('No','Yes','Yes','Yes')
     ),
     map_dfr(
-      effects[c('prec_eff','prec_adj','ate_adj','tot_adj')],
+      c('prec_eff','ate_eff','tot_eff'),
       ~tibble(
-        Estimate=unname(.$coefficients),
-        "Std. Error"=unname(.$std.error),
-        CI=printci(c(.$conf.low,.$conf.high)))
-     ),
-      `$T_Z$`=NA,
-    `$\\rho^2$`=NA
+        Estimate=unname(effects[[.]]$coefficients),
+        "Std. Error"=unname(effects[[.]]$std.error),
+        CI=printci(with(effects[[.]],c(conf.low,conf.high))),
+        SI1=printci(sens[[.]][1,]),
+        SI2=printci(sens[[.]][2,])
       )
-
-stopifnot(max(matchTEtab[,'Estimate'])<0.25)
-stopifnot(min(matchTEtab[,'Estimate'])>0.15)
-
-for(tt in c(ttt['xirt'],ttt['xirt']/2))
-  matchTEtab <- add_case(matchTEtab,
-    Effect=ifelse(tt==ttt['xirt'],"[Pretest]","[Pretest/2]"),
-    Estimate=ancovat[1],
-    CI=printci(interval(tt,rrr['xirt'],b=ancovat[1], se=ancovat[2],df=ancova$df)),
-    `$\\rho^2$`=rrr['xirt'],
-    `$T_Z$`=tt)
-
-matchTEtab <- cbind(
-  ` `=c('\\multirow{4}{*}{Estimate}',rep('',3),'\\multirow{2}{*}{Sensitivity}',''),
-  matchTEtab
     )
-
-matchTEtab <- add_case(matchTEtab,` `='\\hline',.before=5)#,rep(NA,ncol(matchTEtab)-1)),matchTEtab[5:6,])
+  )
+
+##       `$T_Z$`=NA,
+##     `$\\rho^2$`=NA
+##   )
+
+## stopifnot(max(matchTEtab[,'Estimate'])<0.25)
+## stopifnot(min(matchTEtab[,'Estimate'])>0.15)
+
+## for(tt in c(ttt['xirt'],ttt['xirt']/2))
+##   matchTEtab <- add_case(matchTEtab,
+##     Effect=ifelse(tt==ttt['xirt'],"[Pretest]","[Pretest/2]"),
+##     Estimate=ancovat[1],
+##     CI=printci(interval(tt,rrr['xirt'],b=ancovat[1], se=ancovat[2],df=ancova$df)),
+##     `$\\rho^2$`=rrr['xirt'],
+##     `$T_Z$`=tt)
+
+## matchTEtab <- cbind(
+##   ` `=c('\\multirow{4}{*}{Estimate}',rep('',3),'\\multirow{2}{*}{Sensitivity}',''),
+##   matchTEtab
+## )
+
+## matchTEtab <- add_case(matchTEtab,` `='\\hline',.before=5)#,rep(NA,ncol(matchTEtab)-1)),matchTEtab[5:6,])
 
 print(xtable(matchTEtab,
-  align='rllrrrrr',
+  align='rrllccc',
   caption='Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.',

### 2019-12-17 paper.tex
take out regression

@@
 model determines that they have mastered the requisite skills.
-
-As students work through specific problems, they receive two types
-of instruction from the tutor.
-When they err---that is, depart from one of the pre-specified correct
-solution paths for that problem---they receive an error message
-alerting them to their mistake and helping them return to the correct
-solution path.
-Students may also request help, in the form of hints, that are also
-tailored to the specific context and solution path they are on.
-
-\subsection{Data}\label{sec:data}
-
-Since our goal was to better understand the CTA1
-treatment effect, we focused our analysis on data from high school students in
-the second year of the CTA1 trial, for whom the treatment effect was
-most evident.
-
-We merged data from two sources: covariate, treatment and outcome data gathered by RAND, and computerized log data gathered by Carnegie Learning.
-Table \ref{tab:covariateBalance} describes the covariates we used, including
-missingness information, control and treatment means, and standardized
-differences \citep[c.f.][]{kalton1968standardization}.
-We singly-imputed missing values with the Random Forest routine implemented
-by the \texttt{missForest} package in \texttt{R}
-\citep{missForest,rcite},  which estimated the ``out of box'' imputation
-errors also shown in Table \ref{tab:covariateBalance} as part of the random forest regression.
-
-\begin{table}[ht]
-\centering
-\include{output/covariateTable}
-\caption{Missingness information, control (``BaU'' or ``Business as
-  Usual'') and treatment (``CTA1'') means, and balance for the
-  covariates included in this study, from the high school year two
-  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
-  categorical variables (Race/Ethnicity, Sex, and Special Education)
-  and standardized root mean squared error for Pretest, which is
-  continuous. %The overall p-value accounts for clustering at the
-  %school level and blocked random sampling.
-  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
-\label{tab:covariateBalance}
-\end{table}
-
-Over the course of the effectiveness study, Carnegie Learning gathered
-log data for each worked problem.
-The resulting dataset lists the problem name, section, and unit of
-each problem, the numbers of hints requested and errors committed, and time-stamps.
-Log data were missing for some students, either because the log files were not retrievable, or because of an imperfect ability to link log data to other student records.
-Treatment schools with mastery data missing
-for 90\% or more students were omitted from the analysis, along with
-their entire randomization block.
-Of the remaining 2390 students, 87\% (2,091) had log
-data.
-For the sake of simplicity, the treatment students without log data
-were dropped from the study.\footnote{Including these students in a
-  principal stratification model is straightforward \citep{aoas}; see
-  the online supplement for results from a model including them, which
-  were broadly similar to the results presented here. Including
-  subjects with missing log data in a mediation or observational study
-  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
-Problem log data for sections that were not part of the
-standard CTA1 Algebra I curriculum and problems worked by fewer than 100
-students were omitted from the
-dataset.
-
-All told, the main analysis included $n=$5,009 students, 2,091 of whom were assigned to the CTA1 condition and 2,918 of whom were assigned to control.
-The students were nested within 116 teachers, in 43 schools across five states.
 
@@
 these data.
-What is relevant for our purposes is that we may narrow down the type
-of hint data we model to align with our research focus and to avoid
-some modeling pitfalls.
-
+
+For the current purpose, we narrow down the type
+of hint data we model to align with our research focus.
 Since our posttest measured Algebra I skills, we only
@@
 and 0 otherwise.
-To choose $c$, we fit a two-component modified Rasch mixture model
-\citep{rasch} to hint request data, described in an online appendix.
-The model classified approximately 30\% of students as
-high-hint users, so we chose $c$ as the
+
+To choose $c$, we first fit a modified Rasch mixture model
+\citep{rasch} to hint request data.
+Specifically, we modeled the probability of a hint request in each problem
+$p$, worked by student $i$, as
+\begin{equation*}\label{eq:rasch}
+Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
+\end{equation*}
+where $h_{ip}$ is the event that student $i$ requests a hint on
+problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
+section-level parameter for $s[p]$, the section that problem $p$ was
+drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
+function.
+Typically, the student parameter $\eta_i$ measures student ability;
+here, instead, it measures a student's proclivity to request a hint.
+The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
+dataset many problems were worked by very few students, so problem
+level parameters would be hard to estimate.
+Instead, we included the section-level parameter $\delta_{s[p]}$,
+which measures the extent to which students tend to request hints on
+problems in section $s$.
+This essentially assumes that all problems within each section are equally
+likely to elicit a hint.
+
+Then, in order to dichotomize hint request behavior, we modeled
+student effects $\eta$ as
+\begin{equation}\label{eq:mixture}
+\eta\sim
+p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
+\end{equation}
+and constrained $\mu_0<\mu_1$.
+This mixture model clusters students into two categories: high hint
+users have $\eta$ drawn from a normal distribution with mean $\mu_1$
+and standard deviation $\sigma_1$, and low hint users have $\eta$
+drawn from a normal distribution with mean $\mu_0$ and standard
+deviation $\sigma_0$.
+
+Our main parameter of interest here is $p_0$, the proportion of
+students who are low hint users.
+We estimated $p_0\approx$0.7,
+classifying 30\% of students as
+high hint users, so we chose $c$ as the
 70th percentile of $\bar{h}$,
 approximately 0.6.
-This definition agreed with the model's classification in
+This definition agreed with the model's classification (based on
+$\eta$ as opposed to $\bar{h}$ in
 approximately 93\% of cases.
 
+The Rasch model for hint useage \eqref{eq:rasch}, but not the mixture
+model \eqref{eq:mixture} will play an important role in the principal
+stratification analysis of Section \ref{sec:principalStratification}.
+
+\subsection{Data}\label{sec:data}
+
+\citep{pane2014effectiveness} estimated effects separately in middle
+schools and high schools in each the first two years of the RCT; the
+only statistically significant treatment effect was in the high school
+sample in the second year.
+Since our goal was to better understand the CTA1
+treatment effect, we focused our analysis on data from high school students in
+the second year of the CTA1 trial, for whom the treatment effect was
+most evident.
+
+We merged data from two sources: computerized log data gathered by
+Carnegie Learning, and covariate, treatment and outcome data gathered by RAND.
+
+The log data lists the problem name, section, and unit of
+each problem, the numbers of hints requested and errors committed, and time-stamps.
+Log data were missing for some students, either because the log files
+were not retrievable, or because of an imperfect ability to link log
+data to other student records.
+Further, log data for sections that were not part of the
+standard CTA1 Algebra I curriculum and problems worked by fewer than 100
+students were omitted from the
+dataset.
+
+To construct our analysis dataset, we first dropped treatment schools with log data missing
+for 90\% or more students.
+Prior to randomization, schools were stratified into pairs or triples
+based on baseline covariates, and randomized between treatment and
+control condition within these randomization blocks.
+When we dropped a treatment school from our analysis, we also dropped
+the control schools in its randomization block.
+Of the remaining 2,390 students, 87\% (2,091) had log
+data.
+Then, for the sake of simplicity, the treatment students without log data
+were dropped from the study.\footnote{Including these students in a
+  principal stratification model is straightforward \citep{aoas}; see
+  the online supplement for results from a model including them, which
+  were broadly similar to the results presented here. Including
+  subjects with missing log data in a mediation or observational study
+  design can be more problematic \citep[see, e.g.][]{li2017identifiability}.}
+
+All told, the main analysis included $n=$5,009 students, 2,091 of whom were assigned to the CTA1 condition and 2,918 of whom were assigned to control.
+The students were nested within 116 teachers, in 43 schools across five states.
+
+Table \ref{tab:covariateBalance} describes the covariates we used, including
+missingness information, control and treatment means, and standardized
+differences \citep[c.f.][]{kalton1968standardization} from the final
+analysis sample.
+We singly-imputed missing values with the Random Forest routine implemented
+by the \texttt{missForest} package in \texttt{R}
+\citep{missForest,rcite},  which estimated ``out of box'' imputation
+error rates as part of the random forest regression, also shown in Table \ref{tab:covariateBalance}.
+
+\begin{table}[ht]
+\centering
+\include{output/covariateTable}
+\caption{Missingness information, control (``BaU'' or ``Business as
+  Usual'') and treatment (``CTA1'') means, and balance for the
+  covariates included in this study, from the high school year two
+  stratum of CTA1 Effectiveness experiment. Imputation error is percent falsely classified for
+  categorical variables (Race/Ethnicity, Sex, and Special Education)
+  and standardized root mean squared error for Pretest, which is
+  continuous. %The overall p-value accounts for clustering at the
+  %school level and blocked random sampling.
+  Analysis done in \texttt{R} via \texttt{RItools} \citep{ritools}.}
+\label{tab:covariateBalance}
+\end{table}
+
 \section{An Observational Study Within an Experiment}\label{sec:observational}
@@
 racial categories, grade, ESL status, economic disadvantage, and sex,
-along with a natural spline with five degrees of freedom for pretest (the standard error
-and confidence interval using the ``HC3'' sandwich estimator,
-\citealt{sandwichPackage}).
+along with a natural spline with five degrees of freedom for pretest.
 This regression adjusts for imbalances in these covariates within
-matched sets \citep[c.f.][]{biasAdjust}; its estimate is similar to
-the others.
+matched sets \citep[c.f.][]{biasAdjust}.
+All three sets of standard errors
+and confidence intervals used the heteroskedasticity-robust ``HC3'' sandwich estimator,
+\citealt{sandwichPackage}.
+
+The three estimates all roughly agree that being a high hint user
+decreases posttest scores by roughly a fifth of a standard deviation.
+That is, requesting hints hurts posttest scores---teachers should
+recommend against students' requesting hints.
+
+The similarity of the three estimates boosts our confidence, if only
+slightly. The ``ATE'' and ``TOT'' estimates refer to different groups
+of students---ATE estimates an average effect for the entire treatment
+group, whereas TOT estimates an effect for high hint users. Had these
+been very different, we might wonder why the effect of hint requests
+varies so much between high and low hint users, and if that suggests
+selection bias.
+The fact that the additional covariate adjustment of the ANCOVA model
+made little difference in the effect estimate suggests that the
+propensity score match eliminated all or most of the confounding bias due to
+measured covariates; hence, additional adjustment was unnecessary.
 
 % latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Mon Oct 28 20:04:28 2019
+% Tue Dec 17 17:28:44 2019
 \begin{table}[ht]
 \centering
-\begin{tabular}{llrrrrr}
+\begin{tabular}{rllccc}
   \hline
-  & Effect & Estimate & Std. Error & CI & $T_Z$ & $\rho^2$ \\ 
+Weights & Estimate & Std. Error & CI & SI1 & SI2 \\ 
   \hline
-\multirow{4}{*}{Estimate} & ANOVA & -0.18 & 0.05 & [-0.27,-0.08] & 0.00 & 0.00 \\ 
-   & ATE & -0.21 & 0.05 & [-0.32,-0.11] & 0.00 & 0.00 \\ 
-   & TOT & -0.18 & 0.06 & [-0.30,-0.07] & 0.00 & 0.00 \\ 
-   & ANCOVA & -0.19 & 0.05 & [-0.29,-0.10] & 0.00 & 0.00 \\ 
-  \hline &  &  &  &  &  &  \\ 
-  \multirow{2}{*}{Sensitivity} & [Pretest] & -0.19 &  & [-0.47,0.09] & -9.86 & 0.18 \\ 
-   & [Pretest/2] & -0.19 &  & [-0.37,-0.01] & -4.93 & 0.18 \\ 
+OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.31,-0.04] \\ 
+  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.31,0.06] \\ 
+  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.33,0.05] \\ 
    \hline
@@
 
-Our solution was to dichotomize hint requests, as
-described above, essentially re-defining the causal question.
-30\% of the treatment group and all of the control group has
-$H=0$---requesting hints at a rate less than the
-70th percentile of the treatment group.
+Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
+the causal question.
 To make sense of the respective roles of $H$ and $\bar{h}$, the
-continuous measure on which it is based, we may imagine a process in
+continuous measure on which it is based (see Section
+\ref{sec:dichotomizingHintUsage}),
+we may imagine a process in
 which students are assigned $H$, and under that constraint choose a
@@
 In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
-in the treatment group half of students have $H=1$ and half have
+in the treatment group 70\% of students have $H=1$ and 30\% have
 $H=0$, and $\bar{h}$ varies within those groups.
@@
 data from only the treatment group.
-TOT is the the difference between the average outcome for students in
-the treatment group with $H=1$, and the average outcome under the
-counterfactual scenario in which they all had $H=0$.
+TOT is the average difference between the realized outcomes for students in
+with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
 That is,
 \begin{equation*}
-TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
+TOT=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]
 \end{equation*}
+
 $\hat{\Delta}$ estimates the difference between the average outcome
@@
 a standard error of 0.04.
-This analysis suggests that the availability of excessive hints
-actually \emph{lowers} the treatment effect.
-CTA1 works not because of hints, but despite them.
+This analysis suggests that the availability of hints
+actually \emph{lowers} the treatment effect, that CTA1 works not
+because of hints, but despite them.
 

### 2019-12-17 R/prelimStan.r
save xOrig

@@ -93,7 +93,7 @@ xOrig <- x
     mutate(prevErr=lag(nerrs1))%>%
     filter(!is.na(prevErr),prevErr>0)
 
- save(dat,x,covs,file='data/hintsData.RData')
+ save(dat,x,xOrig,covs,file='data/hintsData.RData')
 
 
 

### 2020-01-06 paper.Rnw
updated treatment effect description

@@
 treated,'' or TOT, effect.
-Weights $w_m=(1/n_{1m}+1/n_{0m})^{-1}$, the harmonic mean  are
-``precision weights'' \citep[c.f.][]{schochet2015statistical}; the
+Weights $w_m=(1/n_{1m}+1/n_{0m})^{-1}$, the harmonic mean  are ``OLS'' or
+``precision'' \citep[c.f.][]{schochet2015statistical}; the
 estimate using these weights is equal to the coefficient on $H$ from
@@
 
-Table \ref{tab:matchResults} gives four treatment effect estimates
-using these methods.
-The first row gives the result using precision-weights without
-regression adjustment. estimating an effect of $H$
-The rows labeled ATE and TOT weight the matched sets to
-estimate the average effect of $H$ over the entire treatment
-group and over those members of the treatment group with $H=1$,
-respectively.
-The row labeled ANCOVA of Table \ref{tab:matchResults} gives the
-coefficient on $H$ from a regression of Y on $H$ along with indicators for matched set,
-racial categories, grade, ESL status, economic disadvantage, and sex,
-along with a natural spline with five degrees of freedom for pretest.
-This regression adjusts for imbalances in these covariates within
-matched sets \citep[c.f.][]{biasAdjust}.
+Table \ref{tab:matchResults} gives treatment effect estimates,
+standard errors, and confidence intervals under these three weighting
+schemes.
 All three sets of standard errors
@@
 \citealt{sandwichPackage}.
-
 The three estimates all roughly agree that being a high hint user
@@
 selection bias.
-The fact that the additional covariate adjustment of the ANCOVA model
-made little difference in the effect estimate suggests that the
-propensity score match eliminated all or most of the confounding bias due to
-measured covariates; hence, additional adjustment was unnecessary.
+% The fact that the additional covariate adjustment of the ANCOVA model
+% made little difference in the effect estimate suggests that the
+% propensity score match eliminated all or most of the confounding bias due to
+% measured covariates; hence, additional adjustment was unnecessary.
 
@@
         CI=printci(with(effects[[.]],c(conf.low,conf.high))),
-        SI1=printci(sens[[.]][1,]),
-        SI2=printci(sens[[.]][2,])
+       `[Pretest]`=printci(sens[[.]][1,]),
+        `[Pretest]/2`=printci(sens[[.]][2,])
       )
@@
 
+addtorow <- list(
+  pos=list(0,0),
+  command=c("&&&&\\multicolumn{2}{c}{Sensitivity Intervals}\\\\\n",
+    paste(paste(names(matchTEtab),collapse='&'),'\\\\\n'))
+)
+
 print(xtable(matchTEtab,
@@
   label='tab:matchResults'),
-  include.rownames=FALSE,sanitize.text.function =function(x) x)
+  include.rownames=FALSE,sanitize.text.function =function(x) x,add.to.row=addtorow,include.colnames=FALSE)
 @
@@
 pretest.
-The resulting sensitivity interval is quite wide, including
+The resulting sensitivity intervals are quite wide, including
 both large negative effects as well as substantial positive
@@
 A slightly less conservative sensitivity analysis, labeled
-``[Pretest/2]'' allows $U$ to
-predict $H$ and $\rho$ half as well as pretest scores.
-This sensitivity interval is wide, but entirely negative.
+``[Pretest]/2'' allows $U$ to
+predict $H$ and $Y$ half as well as pretest scores.
+These sensitivity intervals are wide, but the interval corresponding
+to OLS weights is entirely negative.
+
+The conclusion is that the omission of a confounder half as important as pretest
+scores cannot explain the estimated OLS-weighted ATE.
+Allowing for the possible omission of such a covariate, we infer that requesting a
+large number of hints hurts students' posttest scores.
+The size of the effect may be as small as
+\Sexpr{abs(round(sens[['prec_eff']][2,2]*100))}\% of a standard deviation
+or as large as \Sexpr{abs(round(sens[['prec_eff']][1,2]*100))}\% of a standard deviation.
 

### 2020-01-06 R/dataAnalysis.r

@@ -1,3 +1,4 @@
+library(splines)
 library(gridExtra)
 library(optmatch)
 library(RItools)

### 2020-01-06 paper.Rnw
working on comment 36

@@
 \\[2ex]
-&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)
+&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
 &=Pr(H=1)TOT
 \end{align*}
-Since $Pr(H=1)=$\Sexpr{round(1-mixProp,2)},
-$\hat{\Delta}=$\Sexpr{round(1-mixProp,2)}$TOT$.
-Estimated thusly, the indirect effect is negative: \Sexpr{round(tott[1]*(1-mixProp),2)}, with
-a standard error of \Sexpr{round(tott[2]*(1-mixProp),2)}.
+Since $Pr(H=1)=$\Sexpr{round(mixProp,2)},
+$\hat{\Delta}=$\Sexpr{round(mixProp,2)}$TOT$.
+Estimated thusly, the indirect effect is negative: \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
+a standard error of \Sexpr{round(tott[2]*(mixProp),2)}.
 This analysis suggests that the availability of hints
@@
 We incorporated the model for hint requests, (\ref{eq:rasch}), into
-the larger principal stratification model, replacing $\eta(1)$ for
+the principal stratification model, replacing $\eta(1)$ for
 $\bar{h}(1)$.
@@
 \eta(1)|\left(\bm{x}_i,\bm{\theta}\right) \sim
-\mathcal{N}\left(\bm{x_i}\bm{\beta}^M+\epsilon^{Mt}_{t[i]}+\epsilon^{Ms}_{s[i]},
-\sigma^M\right)
+\mathcal{N}\left(\bm{x_i}\bm{\beta}+\epsilon^{tch}_{t[i]}+\epsilon^{scl}_{s[i]},
+\sigma\right)
 \end{equation}
-where $\bm{\beta}^U$ is a vector of coefficients.
+where $\bm{\beta}$ is a vector of coefficients.
 Since students were nested within teachers, who were nested within
-schools, we included normally-distributed school ($\epsilon_{Us}$) and
-teacher ($\epsilon_{Ut}$) random intercepts.
+schools, we included normally-distributed school ($\epsilon^{scl}$) and
+teacher ($\epsilon^{tch}$) random intercepts.
 The covariates in the model, $\bm{x}_i$, were detailed in Table
@@
  Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
-\beta^Y_{0b[i]}+\bm{x}_i^T\bm{\beta}^Y+a\eta(1)+Z_i\tau(\eta(1))+\epsilon^{Yt}_{t[i]}+\epsilon^{Ys}_{s[i]},\sigma^Y_{Z[i]}\right)
+\gamma_{0b[i]}+\bm{x}_i^T\bm{\gamma}+a\eta(1)+Z_i\tau(\eta(1))+\zeta^{tch}_{t[i]}+\zeta^{scl}_{s[i]},\omega_{Z[i]}\right)
 \end{equation}
-where $\beta^Y_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\beta}^Y$ are the
-covariate coefficients, and $\epsilon^{Yt}$, and
-$\epsilon^{Ys}$ are normally-distributed teacher and school random
+where $\bamma_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\gamma}$ are the
+covariate coefficients, and $\zeta^{tch}$, and
+$\zeta^{scl}$ are normally-distributed teacher and school random
 intercepts.
-The residual variance $\sigma^Y$ varies with treatment assignment $Z$;
+The residual variance $\omega$ varies with treatment assignment $Z$;
 this captures measurement error in $Y$, treatment effect heterogeneity
@@
 These assumptions are impossible to verify from the data at hand;
-instead, they require a keen understanding of student practice within
+instead, judging their plausibility requires a keen understanding of student practice within
 the tutor.
@@
 stratification.
-That said, an assumption similar to ignorability may play an important
-role for interpretation.
-Further, principal stratification estimators require a model for $Y(0)$
+% That said, an assumption similar to ignorability may play an important
+% role for interpretation.
+However, principal stratification estimators require modeling
+assumptions---specifically, it requires a model for $Y(0)$
 conditional on $\eta(1)$, which is unobserved.
+This modeling assumption is testable, but only to an extent.
+Model misspecification checks can compare a fitted model to observed
+data, but
+
+Certain departures from this assumption are detectable,
 On the other hand, model misspecification is, in principle,
@@
 
-Our observational study found that hint requests have a negative effect: students who
+Our observational study suggested that hint requests have a negative effect: students who
 requested hints more often than the \Sexpr{round((1-mixProp)*100)}th

### 2020-01-07 paper.Rnw
JP comment 36

@@
 p(\bm{\theta}|\bm{Y},\bm{Z},&\bm{X},\{\bm{h}_i\}_{i:Z_i=1})\propto\\
-p(\bm{\theta})&\displaystyle\prod_{i: Z_i=1}\int p(Y|\bm{x}_i,Z=1,\bm{\theta},\eta(1))p(\bm{h}_{i}|\eta(1)(1),\bm{\theta})p(\eta(1)|\bm{x}_i,\bm{\theta})d\eta(1)\\
-\times&\displaystyle\prod_{i:Z_i=0} \int p(\yci|\bm{x}_i,\bm{\theta},\eta(1))p(\eta(1)|\bm{x}_i,\bm{\theta})d\eta(1).\numberthis\label{eq:posteriorEta}
+p(\bm{\theta})&\displaystyle\prod_{i: Z_i=1}\int p(Y_i|\bm{x}_i,Z=1,\bm{\theta},\eta(1))p(\bm{h}_{i}|\eta(1),\bm{\theta})p(\eta(1)|\bm{x}_i,\bm{\theta})d\eta(1)\\
+\times&\displaystyle\prod_{i:Z_i=0} \int p(Y_i|\bm{x}_i,Z=0,\bm{\theta},\eta(1))p(\eta(1)|\bm{x}_i,\bm{\theta})d\eta(1).\numberthis\label{eq:posteriorEta}
 \end{align*}
@@
 \begin{equation}\label{eq:tauModel}
-\EE[Y_{T}-Y_{C}|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)
+\EE[Y(1)-Y(0)|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)
 \end{equation}
@@
 However, principal stratification estimators require modeling
-assumptions---specifically, it requires a model for $Y(0)$
-conditional on $\eta(1)$, which is unobserved.
-This modeling assumption is testable, but only to an extent.
-Model misspecification checks can compare a fitted model to observed
-data, but
-
-Certain departures from this assumption are detectable,
-On the other hand, model misspecification is, in principle,
-detectable---an assumption of a particular family of models is not
-``untestable.''
-Still, the data will, in general, be unable to distinguish between certain
-alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
-turn lead to qualitatively different conclusions.
-In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
-distributed similarly to $Y(1)$ conditional on $\eta(1)$.
+assumptions---most problematically, $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ of
+\eqref{eq:posteriorEta}, the model for $Y(0)$
+conditional on $\eta(1)$, $\bm{x}$, and paramters $\bm{\theta}$.
+While any model specification can, and should, be checked against
+observed data, this process can only narrow space of acceptible
+models, it cannot be used to determine the correct form of
+$p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$.
+This is because $Y(0)$ and $\eta(1)$ are never observed simultaneously.
+Therefore, it is possible for an analyst to misspecify $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$, such that principal effect estimates are
+severely biased, without being able to detect the model
+misspecification with the data.
+That is, the data will typically be unable to distinguish between
+alternative $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ models that lead to
+qualitatively different conclusions.
+In practice, analysts make the untestable assumption that $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ is
+of the same form as $p(Y|\bm{x},Z=1,\bm{\theta},\eta(1))$.
 

### 2020-01-07 paper.Rnw
ethnicity instead of pretest/2

@@
 
-Next, as represented in Figure (b), we will conduct a mediation analysis, estimating natural
+Next, as represented in Figure (b), we will conduct a mediation
+analysis, estimating natural
 indirect effects of CTA1 treatment assignment $Z$, as mediated by
@@
 
-This measure is inversely related to the ratio of the percentage (of all
+This measure is inversely related to the ratio of the percentage (of
+all
 problems) a student makes an error without having requested a hint,
@@
        `[Pretest]`=printci(sens[[.]][1,]),
-        `[Pretest]/2`=printci(sens[[.]][2,])
+        `[Ethnicity]`=printci(sens[[.]][2,])
       )
@@
   align='rrllccc',
-  caption='Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.',
+  caption='Estimates and sensitivity analysis of the effect of hint usage.',
   label='tab:matchResults'),
@@
 counterparts when observed covariates are left out of the model.
-The first three rows of the table have both $\rho^2$ and $T_Z$ equal
-to zero, to reflect that they assume no unmeasured confounders.
-The row of Table \ref{tab:matchResults} labeled ``[Pretest]''
-imagines a $U$ that predicts $Y$ and $H$ to a similar extent as
-pretest.
+
+The last two columns of table \ref{tab:matchResults} give sensitivity intervals for each of the three weighted average treatment effects.
+The column of Table \ref{tab:matchResults} labeled ``[Pretest]''
+accounts for the possible omission of a covariate $U$ that, at most, predicts $Y$ and $H$ as well as do pretest scores.
 The resulting sensitivity intervals are quite wide, including
@@
 effects.
-A slightly less conservative sensitivity analysis, labeled
-``[Pretest]/2'' allows $U$ to
-predict $H$ and $Y$ half as well as pretest scores.
-These sensitivity intervals are wide, but the interval corresponding
+In fact, pretest is the most important of our observed covariates; it is the observed covariate whose omission would cause the most bias.
+This is unsurprising, since pretests are generally considered to be the most important covariate to measure, and studies that include pretest measures often reproduce experimental estimates \citep[e.g.][]{cook2008three,cook2009bias}.
+Therefore, a sensitivity analysis considering an omitted covariate as important as pretest scores may be too pessimistic. 
+Instead,
+the final column of Table \ref{tab:matchResults} gives sensitivity intervals for the omission of a covariate that, at most, predicts $Y$ and $H$ as well as ethnicity dummy variables, the second most important of our observed covariates.
+These sensitivity intervals are also wide, but the interval corresponding
 to OLS weights is entirely negative.
 
-The conclusion is that the omission of a confounder half as important as pretest
-scores cannot explain the estimated OLS-weighted ATE.
+The conclusion is that the omission of a confounder as important as ethicity indicators
+cannot explain the estimated OLS-weighted ATE.
 Allowing for the possible omission of such a covariate, we infer that requesting a
@@
 The size of the effect may be as small as
-\Sexpr{abs(round(sens[['prec_eff']][2,2]*100))}\% of a standard deviation
-or as large as \Sexpr{abs(round(sens[['prec_eff']][1,2]*100))}\% of a standard deviation.
+\Sexpr{abs(round(sens[['prec_eff']][2,'race']*100))}\% of a standard deviation
+or as large as \Sexpr{abs(round(sens[['prec_eff']][1,'race']*100))}\% of a standard deviation.
 
@@
 sensitivity analyses we performed assumed that any un-measured
-covariate would predict $H$ no better than pretest, or no better than
-a covariate half as important as pretest.
+covariate would predict $H$ and $Y$ no better than pretest or ethnicity indicators, respectively.
 These assumptions are impossible to verify from the data at hand;
@@
 requested hints more often than the \Sexpr{round((1-mixProp)*100)}th
-percentile score lower on the posttest, after controlling for observed covariates---as
-far as we can tell, the optimal strategy is to never request a hint.
-However, this observed effect may be due to confounding.
+percentile score lower on the posttest, after controlling for observed
+covariates.
+As far as we can tell, the optimal strategy is to request few, if any, hints.
+This observed effect may be explained by due to confounding.
 Mediation analysis essentially interprets this result in terms of the

### 2020-01-07 R/dataAnalysis.r
no regression + ethnicity instead of pretest/2

@@ -150,30 +150,37 @@ cat('
 ##########################################################################################
 ',as.character(Sys.time()),'\n')
 
-anova <- lm(Y~M+match,data=dat1)
-anovavcv <- vcovHC(anova,'HC2')
-anovat <- coeftest(anova,anovavcv)[2,]
-anovaci <- coefci(anova,'MTRUE',vcov.=anovavcv)
+## anova <- lm(Y~M+match,data=dat1)
+## anovavcv <- vcovHC(anova,'HC2')
+## anovat <- coeftest(anova,anovavcv)[2,]
+## anovaci <- coefci(anova,'MTRUE',vcov.=anovavcv)
 
-summary(ancova <- lm(Y~M+splines::ns(xirt,5)+race+grade+esl+frl+sex+match,data=dat1))$coef[1:5,]
-ancovavcv <- vcovHC(ancova,'HC2')
-ancovat <- coeftest(ancova,ancovavcv)[2,]
-ancovaci <- coefci(ancova,'MTRUE',vcov.=ancovavcv)
+## summary(ancova <- lm(Y~M+splines::ns(xirt,5)+race+grade+esl+frl+sex+match,data=dat1))$coef[1:5,]
+## ancovavcv <- vcovHC(ancova,'HC2')
+## ancovat <- coeftest(ancova,ancovavcv)[2,]
+## ancovaci <- coefci(ancova,'MTRUE',vcov.=ancovavcv)
 
 
-### ben's method
-olsC <- update(ancova,.~.-M-match,subset=!M)
-yhat <- predict(olsC,dat1)
-dat1$e <- dat1$Y-yhat
+## ### ben's method
+## olsC <- update(ancova,.~.-M-match,subset=!M)
+## yhat <- predict(olsC,dat1)
+## dat1$e <- dat1$Y-yhat
 
 effWithWeights <-
   dat1%>%
   group_by(match)%>%
-  summarize(eff=mean(Y[M])-mean(Y[!M]),adj=mean(e[M])-mean(e[!M]),n=n(),ntrt=sum(M),nctl=n-ntrt,prec=1/(1/ntrt+1/nctl))
+  summarize(
+    eff=mean(Y[M])-mean(Y[!M]),
+    #adj=mean(e[M])-mean(e[!M]),
+    n=n(),
+    ntrt=sum(M),
+    nctl=n-ntrt,
[…]

### 2020-01-07 paper.Rnw
comments 37-end, new conclusion, spellcheck

@@
 
-@
-
-<<runStuff,include=FALSE,eval=FALSE>>=
-source('R/dataAnalysis.r')
 @
@@
 
-The Rasch model for hint useage \eqref{eq:rasch}, but not the mixture
+The Rasch model for hint usage \eqref{eq:rasch}, but not the mixture
 model \eqref{eq:mixture} will play an important role in the principal
@@
         CI=printci(with(effects[[.]],c(conf.low,conf.high))),
-       `[Pretest]`=printci(sens[[.]][1,]),
-        `[Ethnicity]`=printci(sens[[.]][2,])
+       `[Pretest]`=printci(sens[[.]]['pre',]),
+        `[Ethnicity]`=printci(sens[[.]]['race',])
       )
@@
 
-None of these estimates addressed the possiblity of confounding from
+None of these estimates addressed the possibility of confounding from
 unmeasured covariates; instead, they relied on matched ignorability.
@@
 
-The conclusion is that the omission of a confounder as important as ethicity indicators
+The conclusion is that the omission of a confounder as important as ethnicity indicators
 cannot explain the estimated OLS-weighted ATE.
@@
 The size of the effect may be as small as
-\Sexpr{abs(round(sens[['prec_eff']][2,'race']*100))}\% of a standard deviation
-or as large as \Sexpr{abs(round(sens[['prec_eff']][1,'race']*100))}\% of a standard deviation.
+\Sexpr{min(abs(round(sens[['prec_eff']]['race',]*100)))}\% of a standard deviation
+or as large as \Sexpr{max(abs(round(sens[['prec_eff']]['race',]*100)))}\% of a standard deviation.
 
@@
 Estimated thusly, the indirect effect is negative: \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
-a standard error of \Sexpr{round(tott[2]*(mixProp),2)}.
+a standard error of \Sexpr{round(effects$tot_eff$std.error*(mixProp),2)}.
 This analysis suggests that the availability of hints
@@
 In section \ref{sec:dichotomizingHintUsage}, we showed that
-dichotomized $\bar{h}$ largely agrees with a dicotomized version of a
+dichotomized $\bar{h}$ largely agrees with a dichotomized version of a
 more sophisticated measure of hint usage.
@@
 \eqref{eq:posteriorEta}, the model for $Y(0)$
-conditional on $\eta(1)$, $\bm{x}$, and paramters $\bm{\theta}$.
+conditional on $\eta(1)$, $\bm{x}$, and parameters $\bm{\theta}$.
 While any model specification can, and should, be checked against
-observed data, this process can only narrow space of acceptible
+observed data, this process can only narrow space of acceptable
 models, it cannot be used to determine the correct form of
@@
 As far as we can tell, the optimal strategy is to request few, if any, hints.
-This observed effect may be explained by due to confounding.
-Mediation analysis essentially interprets this result in terms of the
+While our estimate is subject to bias from an omitted confounder, the omitted variable must be a relatively strong predictor of hint usage and of posttest scores in order to explain the observed relationship.
+An omitted confounder that is as important as pretest could explain the relationship, but an omitted confounder as important as ethnicity---our second most important covariate---could not (so long as we estimate an OLS-weighted average effect).
+
+Mediation analysis essentially interprets the observational study result in terms of the
 overall CTA1 treatment effect.
@@
 between hint requesting and posttest scores.
-Unlike those two methods, PS understands $\eta(1)$ as a
-student-specific baseline characteristic---albeit one that only
-manifests in the treatment group.
 Students who request more hints need more help, and thus score lower
 on the posttest.
-On the other hand, the PS analysis suggests that it's these students
-who also receive more help---perhaps due to CTA1 hints, or other
-features of the curriculum---and hence benefit more from CTA1.
+Fortunately, the PS analysis also suggests that these students
+receive more help----that is, they benefit more from CTA1---than do their peers who request fewer hints.
+However, it is unclear whether their larger treatment effects are due to their higher rate of hint requests, or to some other correlated factor.
 
@@
 models for a dataset, and perhaps devise tests of model fit tailored
-to a specific problem, may be confident in the fit of a PS model.
+to a specific problem, may be more confident in the fit of a PS model.
 Researchers who prefer transparency and non-parametric or
@@
 
-\section{Discussion: The Essential Role of Measurement}\label{sec:conclusion}
+\section{Discussion: Causal Inference and Measurement}\label{sec:conclusion}
 This case study has focused on three approaches to causal modeling of
@@
 We hope we have demonstrated some of the potential and some of the
-challenges that new computerized data bring to these old problems.
-
-In the hints case study, the observational study and mediation
-analysis differ from principal stratification in their approach to
-potential confounding of the statistical relationship between hint
-requests and posttest scores.
-Matching-based estimators attempt to adjust for confounding,
-and sensitivity analysis accounts for confounding.
-In contrast, principal stratification includes it in the model,
-estimating the relationship between (potential) hint usage and
-treatment effects, as opposed to the relationship between hints and
-outcomes.
-
-A second challenge is extrapolation in mediation analysis.
-in RCTs comparing computerized educational technology with traditional
-pedagogy, members of the control group do not have computer log data
-or access to the mechanisms that the log data measures.
-This forms a serious barrier to mediation analysis.
-However, by dichotomizing the mediator and matching, we were able to
-re-define the causal question in such a way as to make it (possibly) answerable.
-
-Lurking behind all three cases is the equally important issue of
-measurement.
-This issue arose in our case study in our decision to use an IRT model
-to measure hint usage, instead of a simple proportion.
-Doing so is an important step in the right direction, of thinking
-carefully about measurement issues and using as much available data as
-possible.
-The Rasch model also opens the door to a host of exciting
-possibilities in causal models of log data.
-Specifically, other model-based dimension reduction or cluster analysis techniques
-could be brought to bear, defining new intermediate variables.
+challenges that new datasets logged by online
+learning systems bring to these old problems.
+
+Measurement of students' hint request rates played a central role in all three approaches we considered.
+The results were largely driven by three measurement decisions.
+First, our decision to consider only worked Cognitive Tutor problems on which students either requested a hint or made an error (or both) reduced the relationship between hint requests and student algebra I ability.
+Second, our decision to dichotomize hint request rates in the observational study and mediation analysis allowed us to use of a traditional matching estimator and to identify mediational estimands.
+Third, our decision to use a Rasch model to measure hint usage in the PS analysis accounted for differences between students in both the number and difficulty of worked CT problems.
+
+While questions of measurement has always been important in causal inference, analysis of EdTech log data brings them to the fore.
+Log datasets from technology products are large, multivariate, and complex, so careful thought is necessary in order to measure implementation constructs of interest.
+
+Even if they are motivated by methodological concerns, decisions about measurement are inherently also decisions about causal questions---estimates using different measurements answer different questions.
+Just as researchers must choose a causal approach, such as matching, mediation, or principal stratification, they must choose a measurement approach as well.
+As our case study demonstrated, these two decisions are deeply intertwined.
+
+The size, dimension, and complexity of EdTech log data suggests some exciting opportunities for innovative combinations of measurement and causal approaches.
+We demonstrated the inclusion of an IRT model in principal stratification; this suggests the possibility of including other measurement models into causal estimators.
+Multivariate measurement models that incorporate not only hint requests, but other usage measures such as time spent, actions taken, and errors could yield deep insights about how students use EdTech products and how different usage patterns correspond to different effects.
+The development of these causal models will require simultaneous consideration of causal inference and measurement.
 

### 2020-01-07 paper.tex
updated treatment effect description
comments 37-end, new conclusion, spellcheck

@@
 \documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+% maxwidth is the original width if it is less than linewidth
+% otherwise use linewidth (to make sure the graphics do not exceed the margin)
 \makeatletter
@@
 selection bias.
-The fact that the additional covariate adjustment of the ANCOVA model
-made little difference in the effect estimate suggests that the
-propensity score match eliminated all or most of the confounding bias due to
-measured covariates; hence, additional adjustment was unnecessary.
-
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Tue Dec 17 17:28:44 2019
+% The fact that the additional covariate adjustment of the ANCOVA model
+% made little difference in the effect estimate suggests that the
+% propensity score match eliminated all or most of the confounding bias due to
+% measured covariates; hence, additional adjustment was unnecessary.
+
+% latex table generated in R 3.6.2 by xtable 1.8-4 package
+% Tue Jan  7 14:23:39 2020
 \begin{table}[ht]
@@
   \hline
-Weights & Estimate & Std. Error & CI & SI1 & SI2 \\ 
+  &&&&\multicolumn{2}{c}{Sensitivity Intervals}\\
+ Weights&Estimate&Std. Error&CI&[Pretest]&[Ethnicity] \\
  \hline
-OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.31,-0.04] \\ 
-  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.31,0.06] \\ 
-  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.33,0.05] \\ 
+OLS & -0.18 & 0.04 & [-0.25,-0.10] & [-0.41,0.06] & [-0.30,-0.05] \\ 
+  ATE & -0.12 & 0.05 & [-0.23,-0.02] & [-0.45,0.20] & [-0.29,0.04] \\ 
+  TOT & -0.14 & 0.06 & [-0.25,-0.03] & [-0.47,0.19] & [-0.31,0.03] \\ 
    \hline
 \end{tabular}
-\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
+\caption{Estimates and sensitivity analysis of the effect of hint usage.} 
 \label{tab:matchResults}
@@
 counterparts when observed covariates are left out of the model.
-The first three rows of the table have both $\rho^2$ and $T_Z$ equal
-to zero, to reflect that they assume no unmeasured confounders.
-The row of Table \ref{tab:matchResults} labeled ``[Pretest]''
-imagines a $U$ that predicts $Y$ and $H$ to a similar extent as
-pretest.
-The resulting sensitivity interval is quite wide, including
+
+The last two columns of table \ref{tab:matchResults} give sensitivity intervals for each of the three weighted average treatment effects.
+The column of Table \ref{tab:matchResults} labeled ``[Pretest]''
+accounts for the possible omission of a covariate $U$ that, at most, predicts $Y$ and $H$ as well as do pretest scores.
+The resulting sensitivity intervals are quite wide, including
 both large negative effects as well as substantial positive
 effects.
-A slightly less conservative sensitivity analysis, labeled
-``[Pretest/2]'' allows $U$ to
-predict $H$ and $\rho$ half as well as pretest scores.
-This sensitivity interval is wide, but entirely negative.
+In fact, pretest is the most important of our observed covariates; it is the observed covariate whose omission would cause the most bias.
+This is unsurprising, since pretests are generally considered to be the most important covariate to measure, and studies that include pretest measures often reproduce experimental estimates \citep[e.g.][]{cook2008three,cook2009bias}.
+Therefore, a sensitivity analysis considering an omitted covariate as important as pretest scores may be too pessimistic. 
+Instead,
+the final column of Table \ref{tab:matchResults} gives sensitivity intervals for the omission of a covariate that, at most, predicts $Y$ and $H$ as well as ethnicity dummy variables, the second most important of our observed covariates.
+These sensitivity intervals are also wide, but the interval corresponding
+to OLS weights is entirely negative.
+
+The conclusion is that the omission of a confounder as important as ethnicity indicators
+cannot explain the estimated OLS-weighted ATE.
+Allowing for the possible omission of such a covariate, we infer that requesting a
+large number of hints hurts students' posttest scores.
+The size of the effect may be as small as
+5\% of a standard deviation
+or as large as 30\% of a standard deviation.
 
@@
 \\[2ex]
-&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)
+&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
 &=Pr(H=1)TOT
 \end{align*}
-Since $Pr(H=1)=$0.7,
-$\hat{\Delta}=$0.7$TOT$.
-Estimated thusly, the indirect effect is negative: -0.1, with
-a standard error of 0.04.
+Since $Pr(H=1)=$0.3,
+$\hat{\Delta}=$0.3$TOT$.
+Estimated thusly, the indirect effect is negative: -0.04, with
+a standard error of 0.02.
 This analysis suggests that the availability of hints
@@
 sensitivity analyses we performed assumed that any un-measured
-covariate would predict $H$ no better than pretest, or no better than
-a covariate half as important as pretest.
+covariate would predict $H$ and $Y$ no better than pretest or ethnicity indicators, respectively.
 These assumptions are impossible to verify from the data at hand;
-instead, they require a keen understanding of student practice within
+instead, judging their plausibility requires a keen understanding of student practice within
 the tutor.
@@
 stratification.
-That said, an assumption similar to ignorability may play an important
-role for interpretation.
-Further, principal stratification estimators require a model for $Y(0)$
-conditional on $\eta(1)$, which is unobserved.
-On the other hand, model misspecification is, in principle,
-detectable---an assumption of a particular family of models is not
-``untestable.''
-Still, the data will, in general, be unable to distinguish between certain
-alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
-turn lead to qualitatively different conclusions.
-In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
-distributed similarly to $Y(1)$ conditional on $\eta(1)$.
+% That said, an assumption similar to ignorability may play an important
+% role for interpretation.
+However, principal stratification estimators require modeling
+assumptions---most problematically, $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ of
+\eqref{eq:posteriorEta}, the model for $Y(0)$
+conditional on $\eta(1)$, $\bm{x}$, and parameters $\bm{\theta}$.
+While any model specification can, and should, be checked against
+observed data, this process can only narrow space of acceptable
+models, it cannot be used to determine the correct form of
+$p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$.
+This is because $Y(0)$ and $\eta(1)$ are never observed simultaneously.
+Therefore, it is possible for an analyst to misspecify $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$, such that principal effect estimates are
+severely biased, without being able to detect the model
+misspecification with the data.
+That is, the data will typically be unable to distinguish between
+alternative $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ models that lead to
+qualitatively different conclusions.
+In practice, analysts make the untestable assumption that $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ is
+of the same form as $p(Y|\bm{x},Z=1,\bm{\theta},\eta(1))$.
 
@@
 
-Our observational study found that hint requests have a negative effect: students who
+Our observational study suggested that hint requests have a negative effect: students who
 requested hints more often than the 70th
-percentile score lower on the posttest, after controlling for observed covariates---as
-far as we can tell, the optimal strategy is to never request a hint.
-However, this observed effect may be due to confounding.
-Mediation analysis essentially interprets this result in terms of the
+percentile score lower on the posttest, after controlling for observed
+covariates.
+As far as we can tell, the optimal strategy is to request few, if any, hints.
+While our estimate is subject to bias from an omitted confounder, the omitted variable must be a relatively strong predictor of hint usage and of posttest scores in order to explain the observed relationship.
+An omitted confounder that is as important as pretest could explain the relationship, but an omitted confounder as important as ethnicity---our second most important covariate---could not (so long as we estimate an OLS-weighted average effect).
+
+Mediation analysis essentially interprets the observational study result in terms of the
 overall CTA1 treatment effect.
@@
 \section{Covariate Balance Before and After Matching}
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Tue Oct 29 14:14:20 2019
+% latex table generated in R 3.6.2 by xtable 1.8-4 package
+% Tue Jan  7 14:23:39 2020
 \begin{table}[ht]

### 2020-01-07 R/dataAnalysis.r

@@ -180,7 +180,7 @@ effWithWeights <-
 
 effects <- list()
 for(est in c('ate','tot','prec'))
-  #for(ca in c('eff','adj'))
+  for(ca in c('eff'))#,'adj'))
     effects[[paste0(est,'_',ca)]] <-
       lm_robust(
         as.formula(paste(ca,'~1')),

### 2020-01-29 paper.Rnw
delete erroneous footnote

@@
 the CTA1 condition outperformed the control group on the posttest by a fifth of a
-standard deviation.\footnote{\citep{pane2014effectiveness} only
-  reports results from the first two years of implementation; the
-  third year included only middle schools, not high schools.}
+standard deviation.
 
@@
 \subsection{Measuring Hint Usage in the CTA1 Data}
-\marginpar{make sure this makes sense before ``Data'' section}
+%\marginpar{make sure this makes sense before ``Data'' section}
 Unlike previous studies of hint usage, the CTA1 study was not designed
@@
 \end{equation}
-where $\bamma_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\gamma}$ are the
+where $\gamma_{0b[i]}$ is a fixed effect for $i$'s randomization block, $\bm{\gamma}$ are the
 covariate coefficients, and $\zeta^{tch}$, and

### 2020-01-29 paper.tex
delete erroneous footnote

@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Tue Jan  7 14:23:39 2020
+% Wed Jan 29 14:36:34 2020
 \begin{table}[ht]

### 2020-02-04 paper.Rnw
style edits Sec 1

@@
 gathered computer log data for students in the treatment group.
-They assembled a dataset that records which problems each student
+They assembled a dataset that recorded which problems each student
 worked, along with timestamps, the numbers of hints requested, and the
 number of errors committed for each problem.
-Log data from experiments like this present an unprecedented
-opportunity for education researchers:
-to use the fine-grained and rich data to
-help understand how, why, and when educational technology works.
+This log data is fine-grained and rich;
+education researchers could use it, and datasets like it,
+to gain insight into how, why, and when educational technology works.
 
@@
 all but one of which was published since 2001. We are aware of
-numerous other such RCTs planned or ongoing. The current case-study is
+numerous other such RCTs planned or ongoing. 
+These RCTs produce log data automatically, as a bi-product.
+The current case-study is
 about the use of hints in CTA1, but the huge variation in products
-under study, each with their own unique designs, opens a wide set of opportunities to understand how features of those
+under study, each with its own unique design, opens a wide set of opportunities to understand how features of those
 products drive their effects. As one example, RAND is recently
@@
 strategy for students in terms of learning algebra, or should ALEKS be
-redesigned to prevent such skipping? Researchers should not stop at
-estimating overall average treatment effects, but make use of the
-wealth of log data that these RCTs produce automatically, as a
-bi-product. Our hope is that this paper can serve as a starting point
-for researchers interested in mining this rich resource.
+redesigned to prevent such skipping? %Researchers should not stop at
+%estimating overall average treatment effects, but 
+Our hope is that this paper can serve as a starting point
+for researchers interested in mining the rich resource of log data.
 
@@
 First, an observational study to estimate the effect of hint usage $H$
-on posttest scores $Y$, is represented by figure (a).
+on posttest scores $Y$, is represented by panel (a).
 This approach, which we will discuss in Section
@@
 
-Next, as represented in Figure (b), we will conduct a mediation
+Next, as represented in panel (b), we will conduct a mediation
 analysis, estimating natural
@@
 (2).
-Notice that path (2) is identical to figure (a)---the effect of $H$ on
+Notice that path (2) is identical to panel (a)---the effect of $H$ on
 $Y$.
@@
 
-Finally, we will estimate a principal stratification model, as represented in Figure
+Finally, we will estimate a principal stratification model, as represented in panel
 (c).

### 2020-02-04 paper.Rnw
Johns email comments 2-4

@@
 \citep{pane2014effectiveness} estimated effects separately in middle
-schools and high schools in each the first two years of the RCT; the
+schools and high schools in each year of implementation; the
 only statistically significant treatment effect was in the high school
@@
 with greater regularity benefited more from the CTA1 curriculum.
-The left hand panel plots one posterior draw of the $\bm{\eta(1)}$
+The right hand panel plots one posterior draw of $\eta(1)$
 alongside observed outcomes $Y$.

### 2020-02-04 paper.Rnw
edits to sec 2

@@
 Each problem in the software is associated with a set of skills that
-the student must master, and a path (or set of paths) to solving the
-problem.
+the student must master in order to solve the problem.
+The software also specifies a ``solution path'' for each problem---a sequence
+of actions (such as ``multiply both sides of the equation by 3'')
+students must take in order to solve the problem.
+If there is more than one way to solve a problem, the software specifies
+multiple correct solution paths students may follow.
 As soon as a student errs in working through a problem---that is, departs from one of the
-solution paths---the
-software shows the student an error message.
+solution paths---the software shows the student an error message.
 This feedback is tailored to the specific mistake, and
@@
 or error feedback.
-On the other hand, \citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
+\citet{beck2008does} and \citet{goldin2012learner}, did not randomize hint
 availability, but instead used statistical models to control for
@@
 hint (the first, second, or final hint available for a given problem).
-They found that, indeed, the effect of asking for a hint varies
+They found that the effect of asking for a hint varies
 between students, but on average it is negative for the first two
@@
 Specifically, we modeled the probability of a hint request in each problem
-$p$, worked by student $i$, as
-\begin{equation*}\label{eq:rasch}
+$p$, worked by student $i$, in which $i$ requested a hint and/or
+committed an error, as 
+\begin{equation}\label{eq:rasch}
 Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
-\end{equation*}
+\end{equation} 
 where $h_{ip}$ is the event that student $i$ requests a hint on 
@@
 This essentially assumes that all problems within each section are equally
-likely to elicit a hint.
+likely to elicit a hint request.
 

### 2020-02-05 paper.Rnw
edits to sec 3

@@
 Let $\bm{x}_i$ represent a vector of baseline covariates for subject $i$.
-In our study of hints, $\bm{x}$ includes indicators for state, school
-id, class id, and the variables described in Table \ref{tab:covariateBalance}.
+In our study of hints, $\bm{x}$ includes indicators for state, school,
+and classroom, and the variables described in Table \ref{tab:covariateBalance}.
 Then we assume \citep[c.f.][]{rosenbaum1983central}:
@@
 \end{equation*}
-where $\bm{\tilde{x}}$ is the $\bm{x}$ vector excluding indicators for
-state, school, and class, which were each included using random
-intercepts modeled as normally distributed, including missingness
+where $\bm{\tilde{x}}$ is a vector of covariates including the
+variables in Table \ref{tab:covariateBalance}, missingness
 indicators for grade, race, sex, and economic disadvantage, and
 a natural spline with five degrees of freedom for pretest.
+State, school, and class, were each included using normally
+distributed random intercepts.
 
@@
 % randomized to $H=1$ or $H=0$ within those strata.
-Under matched ignorability, the mean of the outcomes between $H=1$ and
+Under matched ignorability, the difference of mean outcomes between $H=1$ and
 $H=0$ students within each match is unbiased for the average treatment
@@
 $H=1$ and $H=0$ subjects in match $m$, respectively.
-Alternatively, setting $w_m=n_{1m}$ estimates the average effect of
-$H$ on those subjects for whom $H=1$---the ``treatment on the
-treated,'' or TOT, effect.
-Weights $w_m=(1/n_{1m}+1/n_{0m})^{-1}$, the harmonic mean  are ``OLS'' or
+Alternatively, setting $w_m=n_{1m}$ estimates the the ``treatment on the
+treated,'' or TOT, effect---the average effect of
+$H$ on those subjects for whom $H=1$.
+Weights $w_m=(1/n_{1m}+1/n_{0m})^{-1}$ are ``OLS'' or
 ``precision'' \citep[c.f.][]{schochet2015statistical}; the
@@
 All three sets of standard errors
-and confidence intervals used the heteroskedasticity-robust ``HC3'' sandwich estimator,
-\citealt{sandwichPackage}.
+and confidence intervals used the heteroskedasticity-robust ``HC3'' sandwich estimator
+\citep{sandwichPackage}.
 The three estimates all roughly agree that being a high hint user
@@
 The similarity of the three estimates boosts our confidence, if only
-slightly. The ``ATE'' and ``TOT'' estimates refer to different groups
+slightly. The ATE and TOT estimates refer to different groups
 of students---ATE estimates an average effect for the entire treatment
@@
 selection bias.
+The OLS estimate downweights matched sets with many more $H=1$ than
+$H=0$ students or vice-versa, relative to matched sets with several of
+both.
+It is larger in magnitude and much less noisy than the other two
+estimates.
 % The fact that the additional covariate adjustment of the ANCOVA model
@@
 
-The last two columns of table \ref{tab:matchResults} give sensitivity intervals for each of the three weighted average treatment effects.
+The last two columns of Table \ref{tab:matchResults} give sensitivity intervals for each of the three weighted average treatment effects.
 The column of Table \ref{tab:matchResults} labeled ``[Pretest]''

### 2020-02-05 paper.Rnw
edits to sec 4

@@
 treatment and requested few hints, assigned to control and requested
-many hints, or assigned to treatment and requested many hints.
+many hints, or assigned to treatment and requested many hints, respectively.
 In the CTA1 trial, however, the potential outcome $Y(0,1)$ is
@@
 For treated subjects $i$ for whom $H_i=0$, then, treatment assignment
-did not affect $H$---$H_i(0)=H_i(1)=0$.
-In other words, for these subjects,
-\begin{equation*}
-Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
-\end{equation*}
+did not affect $H$, so $H_i(0)=H_i(1)=0$.
+%In other words, for these subjects,
+%\begin{equation*}
+%Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
+%\end{equation*}
 Therefore, when $Z_i=1$ and $H_i=0$,
@@
 same causal setup as in the observational approach:
-matched ignorability and the matching design of section
+matched ignorability and the matching design of Section
 \ref{sec:observational}.
@@
 data from only the treatment group.
-TOT is the average difference between the realized outcomes for students in
+TOT is the average difference between the realized outcomes for students
 with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
@@
 \end{align*}
-Since $Pr(H=1)=$\Sexpr{round(mixProp,2)},
+Since $Pr(H=1)=$\Sexpr{round(mixProp,2)} by definition,
 $\hat{\Delta}=$\Sexpr{round(mixProp,2)}$TOT$.
@@
 
-What would the effect be without hints?
+What would the effect be were all students to request few hints?
 To answer that question, we estimate ``direct'' effects,
@@
 problems.
-In section \ref{sec:dichotomizingHintUsage}, we showed that
+In Section \ref{sec:dichotomizingHintUsage}, we showed that
 dichotomized $\bar{h}$ largely agrees with a dichotomized version of a

### 2020-02-09 paper.Rnw
edits to sec 5

@@
 \citet{vanderweele2008simple}, \citet{jo2008causal} and others have
-derived formal equivalencies between parameters under certain
+derived formal equivalences between parameters under certain
 conditions.
@@
 few treatment students asked for 0 hints.
-In any event, dichotomizing hint requests likely undersells the role
+In any event, dichotomizing hint requests likely under-sells the role
 of hints, but allows a tractable mediation analysis.
@@
 hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
-Whereas the focus of the observational study and mediation
+Whereas the goal of the observational study and mediation
 analysis was the relationship between $\bar{h}$ and posttest scores, the
-focus of PS is the relationship between $\bar{h}$ and treatment effects.
-Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\bar{h}(1)=r]$,
-despite the fact that $\bar{h}(1)$ is never observed at the same time as
-$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
-strictly counterfactual quantities, as in mediation analysis.
-$\bar{h}(1)$ is defined for subjects in both randomized groups, but only
-observed in the treatment group.
+goal of PS is the relationship between $\bar{h}(1)$ and treatment effects.
+Notably, although $Y(0)$ and $\bar{h}(1)$ and $Y(0)$ are never
+simultaneously observed, the $\tau(r)$ estimand does not depend on any
+strictly-counterfactual outcomes.
+Unlike, say, $Y(0,H(1))$, $Y(0)$ and $Y(1)$ are each truly
+potential---they would each occur under treatment assignments $Z=0$
+and $Z=1$.
+That said, estimating $\tau(r)$ requires estimation of
+$\EE[Y(0)|\bar{h}(1)]$, the average of $Y(0)$ over a subset of the
+sample that is unobserved and must be inferred.
+For this reason, $\tau(r)$ is only \emph{partially} identified
+\citep[e.g.][]{mealli2016identification}; even with an infinite
+sample, principal effect estimates will still contain uncertainty.
 
@@
 However, when modeling hint usage continuously, the correspondence may
-not hold \citet[see][for a more complete discussion]{aoas}.
+not hold \citep[see][for a more complete discussion]{aoas}.
 
@@
 $\eta(1)$ is defined for all the students in the study.
-Second, a modeling difference: since we are not dichotomizing $\eta(1)$
-in this analysis, we did not model it as a normal mixture, as in
+Second, a modeling difference: since we did not dichotomize $\eta(1)$
+in this analysis, we modeled it as normal , not as a normal mixture, as in
 \eqref{eq:mixture}.
-
-We incorporated the model for hint requests, (\ref{eq:rasch}), into
-the principal stratification model, replacing $\eta(1)$ for
-$\bar{h}(1)$.
-Then, the principal stratification estimand is
+With this change, the principal stratification estimand became
 \begin{equation*}
@@
 \end{equation*}
+the treatment effect for students whose potential hint usage, if assigned to the treatment condition $\eta(1)=r$.
 A full treatment of this variant of principal stratification, including a
@@
 \citet{aoas}.
+
+\subsection{Identification and Estimation}
+Estimating $\tau(r)$ required estimating $\EE[Y(0)|\eta(1)=r]$,
+despite the fact that hint requests never observed at the same time as
+$Y(0)$ (in fact $\eta(1)$ isn't directly observed in either the
+treatment or control group).
+However, hint requests are observed, and the conditional distribution
+$\eta(1)|\bm{x}$ is identified, among those students with $Z=1$;
+because of randomization of $Z$, these inferences extend to the
+control group as well.
 
@@
 
-\subsection{Identification and Estimation}
-
 First, students' proclivity to request hints is measured by the $\eta$
@@
 Figure \ref{fig:psResults} shows the estimated linear function
-$\hat{\tau}{\eta(0)}$.
+$\hat{\tau}(r)$.
 The left panel shows the CTA1 treatment effect, $Y(1)-Y(0)$ as
-$\tau(1)$ varies; the black line shows the estimate (i.e. posterior
+$r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
 mean) and the red lines are random draws from the posterior
@@
 \end{figure}
+
+These results suggest that students who, if assigned to the treatment
+condition, would be more likely to request hints on challenging
+problems, tend to experience higher treatment effects.
+That is, it suggests that frequent hint requesters benefit more from
+assignment to the CTA1 condition than infrequent hint requesters.
 
@@
 
-In contrast, no ignorabilty assumptions are necessary for principal
+In contrast, no ignorability assumptions are necessary for principal
 stratification.

### 2020-02-09 paper.tex
edits to sec 5

@@
 all but one of which was published since 2001. We are aware of
-numerous other such RCTs planned or ongoing. The current case-study is
+numerous other such RCTs planned or ongoing.
+These RCTs produce log data automatically, as a bi-product.
+The current case-study is
 about the use of hints in CTA1, but the huge variation in products
-under study, each with their own unique designs, opens a wide set of opportunities to understand how features of those
+under study, each with its own unique design, opens a wide set of opportunities to understand how features of those
 products drive their effects. As one example, RAND is recently
@@
 strategy for students in terms of learning algebra, or should ALEKS be
-redesigned to prevent such skipping? Researchers should not stop at
-estimating overall average treatment effects, but make use of the
-wealth of log data that these RCTs produce automatically, as a
-bi-product. Our hope is that this paper can serve as a starting point
-for researchers interested in mining this rich resource.
+redesigned to prevent such skipping? %Researchers should not stop at
+%estimating overall average treatment effects, but
+Our hope is that this paper can serve as a starting point
+for researchers interested in mining the rich resource of log data.
 
@@
 Specifically, we modeled the probability of a hint request in each problem
-$p$, worked by student $i$, as
-\begin{equation*}\label{eq:rasch}
+$p$, worked by student $i$, in which $i$ requested a hint and/or
+committed an error, as
+\begin{equation}\label{eq:rasch}
 Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
-\end{equation*}
+\end{equation}
 where $h_{ip}$ is the event that student $i$ requests a hint on
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Wed Jan 29 14:36:34 2020
+% Sun Feb  9 15:15:44 2020
 \begin{table}[ht]
@@
 \end{align*}
-Since $Pr(H=1)=$0.3,
+Since $Pr(H=1)=$0.3 by definition,
 $\hat{\Delta}=$0.3$TOT$.
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Wed Jan 29 14:36:34 2020
+% Sun Feb  9 15:15:45 2020
 \begin{table}[ht]

### 2020-02-10 paper.Rnw
small edit to sec 6

@@
 approaches appear to give the opposite conclusion from principal
-stratification.
+stratification---matching and mediation suggest that exessive hint
+requests hurt students' posttest scores, while principal
+stratification suggests that students who request more hints
+experiance higher treatment effects.
 This section will compare and contrast the three methods in terms of

### 2020-02-10 paper.Rnw
fiiine I'll add a results table to sec 4

@@
 To answer that question, we estimate ``direct'' effects,
-$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
+$\Xi=\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
 Like indirect effects, this expression includes the average potential
@@
 One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
-subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
-treatment group.
+subjects in the control group and $\hat{Y}(1,0)$ (of Eq. \ref{eq:estY10}) for subjects in the
+treatment group:
+\begin{equation*}
+\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)\hat{Y}(1,0)
+\end{equation*}
 Then, regress it on a dummy variable for treatment group and dummy
-variables for randomization block.
+variables for randomization block:
+\begin{equation}\label{eq:directEffReg}
+\tilde{Y}_i=\alpha_{b[i]}+\hat{\Xi}Z_i+\epsilon_i
+\end{equation}
+where $\alpha_{b[i]}$ is a fixed intercept for subject $i$'s randomization block, $\hat{\Xi}$ is an estimate of $\Xi$ and $\epsilon_i$ is a regression error.
 The clustering of subjects within schools can be accounted for in a
@@
 \Sexpr{citopm(dir1.1ci)}.
-Including covariates in the regression produces a direct effect of
+Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
 \Sexpr{citopm(dir2.1ci)}.
+
+Table \ref{tab:mediation} summarizes our mediation results.
+
+\begin{table}
+\begin{tabular}{llll}
+Effect&Method&Estimate&CI\\
+<<mediationTable,echo=FALSE>>=
+cat(
+  '$1-p_0=\\EE[H(1)-H(0)|Z=1]$ & By Definition&',round(mixProp,2),'&n/a\\\\
+$TOT=\\EE[Y(Z=1,H=1)-Y(Z=1,H=0)|Z=1]$&Matching&',round(effects[['tot_eff']][['coefficients']],2),'&',printci(with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
+$\\Delta=\\EE[Y(1,H(1))-Y(1,H(0))|Z=1]$&$(1-p_0)TOT$&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
+$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \\eqref{eq:directEffReg}&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
+$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$& \\eqref{eq:directEffReg} + covariates&',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\n')
+@
+\end{tabular}
+\caption{Estimated effects in mediation analysis. }
+\label{tab:mediation}
+\end{table}
 

### 2020-02-10 paper.tex
fiiine I'll add a results table to sec 4

@@
 \documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
-% maxwidth is the original width if it is less than linewidth
-% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
 \makeatletter
@@
 
-% latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Sun Feb  9 15:15:44 2020
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Mon Feb 10 10:49:52 2020
 \begin{table}[ht]
@@
 0.14$\pm$0.23.
-Including covariates in the regression produces a direct effect of
+Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
 0.18$\pm$0.19.
+
+\begin{table}
+\begin{tabular}{llll}
+Effect&Method&Estimate&CI\\
+$1-p_0=\EE[H(1)-H(0)|Z=1]$ & By Definition& 0.3 &n/a\\
+$TOT=\EE[Y(Z=1,H=1)-Y(Z=1,H=0)|Z=1]$&Matching& -0.14 & [-0.25,-0.03] \\
+$\Delta=\EE[Y(1,H(1))-Y(1,H(0))|Z=1]$&$(1-p_0)TOT$& -0.04 & [-0.07,-0.01] \\
+$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \eqref{eq:directEffReg}& 0.14 & [-0.08,0.37] \\
+$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$& \eqref{eq:directEffReg} + covariates& 0.18 & [-0.01,0.38]
+
+\end{tabular}
+\caption{Estimated effects in mediation analysis. }
+\label{tab:mediation}
+\end{table}
 
@@
 \section{Covariate Balance Before and After Matching}
-% latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Sun Feb  9 15:15:45 2020
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Mon Feb 10 10:56:04 2020
 \begin{table}[ht]

### 2020-02-10 paper.Rnw
added lines to mediation table, modified caption of tab 2

@@
   align='rrllccc',
-  caption='Estimates and sensitivity analysis of the effect of hint usage.',
+  caption='Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage, under different weighting schemes.',
   label='tab:matchResults'),
@@
 \begin{tabular}{llll}
+\hline
 Effect&Method&Estimate&CI\\
+\hline
 <<mediationTable,echo=FALSE>>=
@@
 $\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \\eqref{eq:directEffReg}&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
-$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$& \\eqref{eq:directEffReg} + covariates&',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\n')
+$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$& \\eqref{eq:directEffReg} + covariates&',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\\\\\n')
 @
+\hline
 \end{tabular}

### 2020-02-10 paper.tex
added lines to mediation table, modified caption of tab 2

@@
 
+Table \ref{tab:mediation} summarizes our mediation results.
+
 \begin{table}
 \begin{tabular}{llll}
+\hline
 Effect&Method&Estimate&CI\\
+\hline
 $1-p_0=\EE[H(1)-H(0)|Z=1]$ & By Definition& 0.3 &n/a\\
@@
 $\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \eqref{eq:directEffReg}& 0.14 & [-0.08,0.37] \\
-$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$& \eqref{eq:directEffReg} + covariates& 0.18 & [-0.01,0.38]
-
+$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$& \eqref{eq:directEffReg} + covariates& 0.18 & [-0.01,0.38] \\
+
+\hline
 \end{tabular}

### 2020-02-11 paper.Rnw
final 2-11-20?

@@
 Section \ref{sec:conclusion} will conclude.
+
+The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{github.com/[redacted]}.
 
@@
 
-All told, the main analysis included $n=$\Sexpr{prettyNum(Ntot,',')} students, \Sexpr{pn(NtFinal)} of whom were assigned to the CTA1 condition and \Sexpr{pn(Ntot-NtFinal)} of whom were assigned to control.
-The students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
+All told, the analyses presented here were all based on \Sexpr{pn(NtFinal)} students assigned to the CTA1 condition; the estimation of direct effects in Section \ref{sec:mediation} and the principal stratification also relied on the \Sexpr{pn(Ntot-NtFinal)} students assigned to control.
+Together, the \Sexpr{pn(Ntot)} students were nested within \Sexpr{pn(nteacher)} teachers, in \Sexpr{pn(nschool)} schools across five states.
+%The observational study (Section \ref{sec:observational} and part of the mediation analysis \ref{sec:mediation} used only the treatment (CTA1) group.
 
@@
 single $H=1$ subject.
+All students in the CTA1 group of the dataset were included in the match.
 
@@
   align='rrllccc',
-  caption='Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage, under different weighting schemes.',
+  caption='Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage on posttest scores, under different weighting schemes.',
   label='tab:matchResults'),

### 2020-02-11 paper.tex
final 2-11-20?

@@
 
-All told, the main analysis included $n=$5,009 students, 2,091 of whom were assigned to the CTA1 condition and 2,918 of whom were assigned to control.
-The students were nested within 116 teachers, in 43 schools across five states.
+All told, the analyses presented here were all based on 2,091 students assigned to the CTA1 condition; the estimation of direct effects in Section \ref{sec:mediation} and the principal stratification also relied on the 2,918 students assigned to control.
+Together, the 5,009 students were nested within 116 teachers, in 43 schools across five states.
+%The observational study (Section \ref{sec:observational} and part of the mediation analysis \ref{sec:mediation} used only the treatment (CTA1) group.
 
@@
 
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Mon Feb 10 10:49:52 2020
+% latex table generated in R 3.6.2 by xtable 1.8-4 package
+% Tue Feb 11 09:12:59 2020
 \begin{table}[ht]
@@
 \end{tabular}
-\caption{Estimates and sensitivity analysis of the effect of hint usage.} 
+\caption{Estimates, standard errors, and sensitivity analysis of the weighted average effect of hint usage on posttest scores, under different weighting schemes.} 
 \label{tab:matchResults}
@@
 \section{Covariate Balance Before and After Matching}
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Mon Feb 10 10:56:04 2020
+% latex table generated in R 3.6.2 by xtable 1.8-4 package
+% Tue Feb 11 09:12:59 2020
 \begin{table}[ht]

### 2020-02-21 R/dataAnalysis.r
change heading on balance tables

@@ -101,12 +101,12 @@ hintAvgStud%>%
   geom_point()+
   facet_wrap(~xaxis,scales="free_x")
 
-runFull <- FALSE
+#runFull <- FALSE
 ### rasch mixture model?
 ##NOT RUN
 if(runFull) source('R/raschMixture.r') else load('output/raschMixOutput.RData')
 
-runFull <- TRUE
+#runFull <- TRUE
 
 mixProp <- round(mean(hintAvgStud$classProb),1)
 
@@ -133,7 +133,7 @@ if(runFull){
 dist3 <- match_on(M~predict(psmod3.2,type='link'),within=exactMatch(M~schoolid2,data=dat1),data=dat1)
 m3 <- fullmatch(dist3,data=dat1)
 
-balMod <- xBalance(M~grade+race+xirt+spec+esl+state,strata=list(unmatched=NULL,matched=~m3),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
+balMod <- xBalance(M~grade+race+xirt+spec+esl+state,strata=list(`Before Matching`=NULL,`After Matching`=~m3),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
 
 pdf('output/balancePlot.pdf')
 plot(balMod,colors=subwayPalette)

### 2020-02-24 R/dataAnalysis.r
change name of pretest variable for balance table

@@ -133,7 +133,9 @@ if(runFull){
 dist3 <- match_on(M~predict(psmod3.2,type='link'),within=exactMatch(M~schoolid2,data=dat1),data=dat1)
 m3 <- fullmatch(dist3,data=dat1)
 
-balMod <- xBalance(M~grade+race+xirt+spec+esl+state,strata=list(`Before Matching`=NULL,`After Matching`=~m3),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
+dat1$pretest <- dat1$xirt
+
+balMod <- xBalance(M~grade+race+pretest+spec+esl+state,strata=list(`Before Matching`=NULL,`After Matching`=~m3),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
 
 pdf('output/balancePlot.pdf')
 plot(balMod,colors=subwayPalette)

### 2020-04-29 paper.Rnw
Revised intro, issue #1

@@
 standard deviation.
-
-Over the course of the CTA1 study, the software administrators
+As educational technology (EdTech) booms, so do RCTs in the mold of the RAND
+CTA1 study. A recent systematic review \citep{escueta2017education} cites 29
+published reports of RCTs of ``computer assisted learning'' programs,
+all but one of which was published since 2001. We are aware of
+numerous other such RCTs planned or ongoing.
+
+RCTs are invaluable tools for estimating these programs'
+efficacy. However, the interventions they study are highly
+multifaceted and complex. Each classroom, and each student within
+those classrooms, may use an EdTech application in a different way;
+presumably, the application's efficacy depends on these usage
+decisions.
+Similarly, EdTech applications often include a number of optional
+features---to what extent do each of these drive the program's
+effectiveness?
+The data collected during typical education RCTs---data on treatment
+assignments, outcomes, a standard set of demographics, and possibly a
+pretest---offer limited options for studying treatment effect
+heterogeneity, and offer no information on how the intervention was
+implemented.
+
+However, with little extra effort, researchers studying EdTech can
+gather rich implementation data over the course of an RCT.
+For instance, the software administrators running the CTA1 RCT
 gathered computer log data for students in the treatment group.
@@
 number of errors committed for each problem.
-This log data is fine-grained and rich;
-education researchers could use it, and datasets like it,
-to gain insight into how, why, and when educational technology works.
-
-Along the same lines, log data presents a challenge, since it differs in
-structure and size from the type of data commonly encountered in
-studies of causal mechanisms.
-While statistical models of log data are nearly as old as the data itself
-\citep[e.g.][]{corbett1994knowledge}, data from an RCT differs in
-two principal ways from more traditional domains.
-RCTs feature randomization of treatment assignment, and a posttest
-(or some other outcome) administered to both users and non-users.
-These factors allow for causal modeling.
-
-As educational technology booms, so do RCTs in the mold of the RAND
-CTA1 study. A recent systematic review \citep{escueta2017education} cites 29
-published reports of RCTs of ``computer assisted learning'' programs,
-all but one of which was published since 2001. We are aware of
-numerous other such RCTs planned or ongoing.
-These RCTs produce log data automatically, as a bi-product.
-The current case-study is
-about the use of hints in CTA1, but the huge variation in products
-under study, each with its own unique design, opens a wide set of opportunities to understand how features of those
-products drive their effects. As one example, RAND is recently
+Analogous log data is (or may be) collected from other EdTech RCTs.
+As one example, RAND recently
 completed an efficacy study of a different algebra tutoring system,
-ALEKS. When students begin working on a particular topic in ALEKS, the
+ALEKS, and gathered student log data in order to study
+implementation.
+For instance, an ongoing research project is using this dataset to
+study instructional material within ALEKS: when students begin working on a particular topic in ALEKS, the
 system first presents instructional material, followed by problem
@@
 %estimating overall average treatment effects, but
-Our hope is that this paper can serve as a starting point
-for researchers interested in mining the rich resource of log data.
+% Our hope is that this paper can serve as a starting point
+% for researchers interested in mining the rich resource of log data to
+% answer these types of questions.
+
+Analysis of log data from EdTech RCTs could, in principle, lead to
+insights on the relationship between implementation and effectiveness.
+Using log data, researchers could investigate how the product was
+actually implemented, as well as what aspects of implementation
+correlate with higher (or lower) treatment effects, or which features
+of the program drive its effects.
+
+However, answering these questions requires new statistical tools.
+These tools must simultaneously surmount three challenges: first,
+implementation is not typically randomized, so the relationship
+between log data and outcomes is likely confounded. Second,
+log data is only gathered from students assigned to use the
+intervention, and not from the control group. Third, log datasets have
+a complex structure---often a large number of granular observations of several types
+gathered over the course of each student's engagement with the
+program.
 
@@
 of hints within the software.
-The first approach discards the control group, and analyzes
+We will attempt to answer two related questions: first, what role does
+the availability of hints play in driving the overall CTA1 treatment
+effect, and second, can requesting hints more often lead to a larger
+treatment effect.
+The first approach we consider discards the control group, and analyzes
 hint-requesting in the treatment group as an observational study,
@@
 \citet{vanderweele2008simple}, \citet{jo2008causal} and others have
-derived formal equivalences between parameters under certain
-conditions.
+given conditions under which certain principal stratification and mediational
+estimands coincide.
 Our contribution will build on this literature in two ways: first, by
@@
 context of studying implementation data from an randomized
-experiment studying educational technology (EdTech).
+experiment studying educational technology.
 
@@
 data.
-Section \ref{sec:synthesis} will compare the three methods directly,
+As it turns out, these analyses appear to give contradictory
+results---the observational study and mediation analysis estimate a
+negative effect of requesting hints, while the principal
+stratification analysis finds that students who request more hints
+experience larger overall treatment effects.
+Section \ref{sec:synthesis} will contrast these findings (including a
+possible reconciliation), compare the three methods generally,
 and discuss how they may be applied to data from other educational
@@
 
-The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{github.com/[redacted]}.
+The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{https://github.com/adamSales/logDataCaseStudy}.
 

### 2020-04-30 paper.Rnw
johns comment on intro

@@
 implementation.
-For instance, an ongoing research project is using this dataset to
+An ongoing research project is using this dataset to
 study instructional material within ALEKS: when students begin working on a particular topic in ALEKS, the

### 2020-05-05 paper.Rnw
took out overview (1.1)
given planned changes in response to #5 this subsection seemed redundant
also, luke didn't seem to like figure (#7)

added some TOC to the end of the intro (#1)

@@
 
-After a brief overview of the three causal approaches,
-the next section will provide background for the case-study,
+The following section will provide background for the case-study,
 describing the CTA1 curriculum, the RAND study, and our focus on
@@
 The following three sections will present the observational study,
-causal mediation, and principal stratification approaches to the hint
-data.
+which discards the control group altogether and estimates the effect
+of hint requests on posttest scores;
+causal mediation, which decomposes the overall treatment effect into a
+component driven by hint requests and another driven by other
+mechanisms; and principal stratification, which estimates separate
+treatment effects for groups of students who would request hints
+differently, if assigned to treatment.
 As it turns out, these analyses appear to give contradictory
@@
 The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{https://github.com/adamSales/logDataCaseStudy}.
-
-\subsection{Brief Overview of Three Causal Approaches}
-\begin{figure}
-\centering
-\begin{tikzpicture}[scale=1,transform shape]
-	%\draw[help lines] (0,0) grid (15,3);
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (1,0) {$H$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(0:3){$Y$};
-	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
-	%\draw[->,line width=2pt](x)--(z);
-	%\draw[->,line width=2pt](x)--(y);
-	\draw[->,line width=2pt](z)--(y);
-	\node at (2.5,-1) {(a)};
-
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (5,0) {$Z$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(8,0){$Y$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (m) at (6.5,2){$H$};
-	\draw[->,line width=2pt](z)--(m);
-	\draw[->,line width=2pt](m)--(y);
-	\draw[->,line width=2pt](z)--(y);
-	\node at (5.4,1.2) {(1)};
-	\node at (7.6,1.2) {(2)};
-	\node at (6.5,.3) {(3)};
-	\node at (6.5,-1) {(b)};
-
-        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (10,0) {$Z$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(12,0){$Y$};
-        \node (mid) at (11,0) {};
-        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (h)
-        at (11,2) {$H(1)$};
-	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
-	%\draw[->,line width=2pt](x)--(z);
-	%\draw[->,line width=2pt](x)--(y);
-	\draw[->,line width=2pt](z)--(y);
-        \draw[->,line width=2pt](h)--(mid);
-	\node at (11,-1) {(c)};
-
-\end{tikzpicture}
-\caption{A schematic representation of three causal models for
-  log data: (a) an observational study estimating the effects of hints
-  $H$ on posttest scores $Y$, (b) a mediation analysis estimating
-  natural direct effects of treatment assignment $Z$ on $Y$ and
-  indirect effects mediated by $H$, and (c) principal stratification.}
-\label{mediationFig}
-\end{figure}
-
-Figure \ref{mediationFig} is a schematic representation of the three
-approaches to causal modeling of log data that we will discuss in this
-paper.
-First, an observational study to estimate the effect of hint usage $H$
-on posttest scores $Y$, is represented by panel (a).
-This approach, which we will discuss in Section
-\ref{sec:observational}, discards the control group from the RCT and
-uses techniques from observational studies to determine if and how requesting
-a greater number of hints affects posttest scores.
-For the sake of illustration, we will use techniques similar to those
-discussed in \citet{rosenbaum2002observational},
- \citet{hansen2009b}, and \citet{ho:etal:2007}, combining a propensity
- score matching design with methods for analyzing experiments and
- parametric modeling.
-The conceptual and methodological issues we will discuss extend to
-other covariate control methods as well.
-
-Next, as represented in panel (b), we will conduct a mediation
-analysis, estimating natural
-indirect effects of CTA1 treatment assignment $Z$, as mediated by
-$H$. That is, we will estimate the component of $Z$'s effect on $Y$ that comes about
-due to $Z$'s effect on $H$---path (1)---and $H$'s effect on $Y$, path
-(2).
-Notice that path (2) is identical to panel (a)---the effect of $H$ on
-$Y$.
-We will use this equivalence to estimate and interpret indirect
-effects.
-As part of the mediation analysis, we will also estimate path (3), the
-natural direct effect of $Z$ on $Y$---that is, the component of $Z$'s
-effect on $Y$ that is not due to hints $H$.
-
-Finally, we will estimate a principal stratification model, as represented in panel
-(c).
-In the principal stratification analysis, we do not model $H$ as an
-intervention that could, in theory, be randomized.
-Instead, we posit a variable ``$H_i(1)$,''  subject $i$'s penchant for
-requesting hints \emph{if} $i$ is assigned to the treatment
-condition.
-Unlike $H$, $H(1)$ is well defined (if
-unobserved) for all participants in the RCT.
-Principal stratification treats $H(1)$ as an effect moderator or
-interaction variable: $H(1)$ does not affect $Y$, but the effect of
-$Z$ on $Y$ may vary as a function of $H(1)$.
-This analysis will largely follow the method of \citet{aoas}, which in
-turn built heavily on \citet{schwartz2011bayesian} and
-\citet{jin2008principal}.
-
-All three of these approaches are designed to model intermediate
-variables---those variables, like hint usage,
-that were possibly affected by $Z$ but were measured prior to $Y$.
-Though they were developed for more traditional data scenarios, the
-novel scenario of computer log data from an educational technology RCT
-presents new and interesting challenges and opportunities.
 

### 2020-05-05 paper.Rnw
Added intro summaries to Obs Study and Mediation sections
partially addressed #5 and #8
also added subsection titles
put back (most) of the citations removed in 6ab06c2b5877d2ca531533d85ad188ed37961ae3 (in different places)
updated paper.tex

@@
 hints.
-The following three sections will present the observational study,
+The following three sections will present an observational study,
 which discards the control group altogether and estimates the effect
@@
 
+This section will illustrate an observational study to answer these
+questions.
+Generally speaking, the observational study approach discards the
+control group entirely and uses statistical confounder control (in our
+case, propensity score matching) to estimate causal effects of a usage parameter on posttest scores.
+Implementing the method requires data from members of the treatment group: usage (in our case, $H$), posttest scores or another outcome of interest, and a set of baseline covariates to control for confounding.
+Since the control group is not used, identification does not depend on randomization---in fact, this same approach could be (and is) used with non-experimental data.
+However, outside of a planned experiment, posttest scores may not be available.
+Our usage variable $H$ is binary, which is necessary for classical propensity score matching; alternatve methods are available for usage variable that are continuous \citep[e.g.][]{hirano2004propensity}, ordered \citep[e.g.][]{leon2005mixed}, categorical \citep[e.g.][]{lopez2017estimation} or other types.
+Like all observational studies, it depends on the strong, untestable assumption of no unmeasured confounding.
+
+The approach we take here, following \citet{rosenbaum2002observational},
+ \citet{hansen2009b}, and \citet{ho:etal:2007}, has three broad steps.
+In the first, we construct a "match," identifying groups of students, some of whom have $H=0$ and others with $H=1$, but who are otherwise comparable.
+The hope is that the distribution of $H$ within this match resembles what might have been seen if (counterfactually) $H$ had been randomized within matched groups.
+This stage is iterative: we construct a match, evaluate it (most importantly by checking if matched students have similar baseline covariate distributions), and if necessary construct a new match.
+In the second step, we estimate the average effect of $H$ on posttest scores, by comparing posttest scores between matched students with $H=1$ and $H=0$.
+In the third and final step, we account for the possiblity of unmeasured confounding in a sensitivity analysis.
+
+The next subsection will give more formal background on propensity score matching, including notation and idenfication assumptions.
+Following subsections will illustrate each of the three steps of propensity score matching.
+
+\subsection{Observational Study: Background}
 Let $Y_i$ represent subject $i$'s posttest score, and let $Z_i$ represent
@@
 
+\subsection{Constructing and Evaluating a Match}
+
 We constructed an optimal full matching design \citep{hansen2004} to
@@
 
+\subsection{Sensitivity Analysis: Assessing the Role of Unmeasured Confounding}
+
 None of these estimates addressed the possibility of confounding from
@@
 
+The method we present here builds on the results from the observational study in the previous section.
+Estimating the indirect effect is a way to interpret the estimated effect of $H$ in the context of the overall treatment effect of the EdTech intervention.
+In fact, in our analysis estimating indirect effects requires no more data than the observational study.
+Estimating direct effects requires, additionally, data from the control group: posttest scores and (ideally) baseline covariates.
+The randomization of the intervention in the RCT now plays an important role in identification, since the overall effect of the intervention on posttest scores is at issue, as does the fact that the control group does not have access to the program.
+Finaly, the assumption of no unmeasured confounding remains important.
+
+The following subsection reviews the formal definitions of direct and indirect effects.
+The next two subsections illustrate the estimation of direct and indirect effects, respectively, in our study of hints.
+It turns out that indirect effects are are nearly unidentifiable, without some acrobatics; however, under the identification scheme we present here, estimating indirect effects is exceedingly simple.
+Therefore, Section \ref{sec:indirectEst} will mostly consist of a discussion of identification.
+Section \ref{sec:directEst} builds on the framework of the previous subsections to estimate direct effects with a regression model.
+
+\subsection{Review: Defining Direct and Indirect Effect}
 More precisely, express a subject's potential outcomes as a function
@@
 
-\subsection{Identification and Estimation of Direct and Indirect
-  Effects}
+\subsection{Identification and Estimation Log Data  Indirect
+  Effects}\label{sec:indirectEst}
 
@@
 because of hints, but despite them.
+
+\subsection{Estimating Log Data Direct Effects}\label{sec:directEst}
 
@@
 However, $\bar{h}$ has some disadvantages as a continuous measure of
-hint usage.
+hint usage \citep{aoas}.
 For one, since it is essentially a sample mean over a subset of each student's

### 2020-05-05 paper.tex
change heading on balance tables
Added intro summaries to Obs Study and Mediation sections
partially addressed #5 and #8
also added subsection titles
put back (most) of the citations removed in 6ab06c2b5877d2ca531533d85ad188ed37961ae3 (in different places)
updated paper.tex

@@
 number of errors committed for each problem.
-This log data is fine-grained and rich;
-education researchers could use it, and datasets like it,
-to gain insight into how, why, and when educational technology works.
-
-Along the same lines, log data presents a challenge, since it differs in
-structure and size from the type of data commonly encountered in
-studies of causal mechanisms.
-While statistical models of log data are nearly as old as the data itself
-\citep[e.g.][]{corbett1994knowledge}, data from an RCT differs in
-two principal ways from more traditional domains.
-RCTs feature randomization of treatment assignment, and a posttest
-(or some other outcome) administered to both users and non-users.
-These factors allow for causal modeling.
-
-As educational technology booms, so do RCTs in the mold of the RAND
-CTA1 study. A recent systematic review \citep{escueta2017education} cites 29
-published reports of RCTs of ``computer assisted learning'' programs,
-all but one of which was published since 2001. We are aware of
-numerous other such RCTs planned or ongoing.
-These RCTs produce log data automatically, as a bi-product.
-The current case-study is
-about the use of hints in CTA1, but the huge variation in products
-under study, each with its own unique design, opens a wide set of opportunities to understand how features of those
-products drive their effects. As one example, RAND is recently
+Analogous log data is (or may be) collected from other EdTech RCTs.
+As one example, RAND recently
 completed an efficacy study of a different algebra tutoring system,
-ALEKS. When students begin working on a particular topic in ALEKS, the
+ALEKS, and gathered student log data in order to study
+implementation.
+An ongoing research project is using this dataset to
+study instructional material within ALEKS: when students begin working on a particular topic in ALEKS, the
 system first presents instructional material, followed by problem
@@
 context of studying implementation data from an randomized
-experiment studying educational technology (EdTech).
-
-After a brief overview of the three causal approaches,
-the next section will provide background for the case-study,
+experiment studying educational technology.
+
+The following section will provide background for the case-study,
 describing the CTA1 curriculum, the RAND study, and our focus on
 hints.
-The following three sections will present the observational study,
-causal mediation, and principal stratification approaches to the hint
-data.
-Section \ref{sec:synthesis} will compare the three methods directly,
+The following three sections will present an observational study,
+which discards the control group altogether and estimates the effect
+of hint requests on posttest scores;
+causal mediation, which decomposes the overall treatment effect into a
+component driven by hint requests and another driven by other
+mechanisms; and principal stratification, which estimates separate
+treatment effects for groups of students who would request hints
+differently, if assigned to treatment.
+As it turns out, these analyses appear to give contradictory
+results---the observational study and mediation analysis estimate a
+negative effect of requesting hints, while the principal
+stratification analysis finds that students who request more hints
+experience larger overall treatment effects.
+Section \ref{sec:synthesis} will contrast these findings (including a
+possible reconciliation), compare the three methods generally,
 and discuss how they may be applied to data from other educational
@@
 
-The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{github.com/[redacted]}.
-
-\subsection{Brief Overview of Three Causal Approaches}
-\begin{figure}
-\centering
-\begin{tikzpicture}[scale=1,transform shape]
-	%\draw[help lines] (0,0) grid (15,3);
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (1,0) {$H$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(0:3){$Y$};
-	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
-	%\draw[->,line width=2pt](x)--(z);
-	%\draw[->,line width=2pt](x)--(y);
-	\draw[->,line width=2pt](z)--(y);
-	\node at (2.5,-1) {(a)};
-
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (5,0) {$Z$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(8,0){$Y$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (m) at (6.5,2){$H$};
-	\draw[->,line width=2pt](z)--(m);
-	\draw[->,line width=2pt](m)--(y);
-	\draw[->,line width=2pt](z)--(y);
-	\node at (5.4,1.2) {(1)};
-	\node at (7.6,1.2) {(2)};
-	\node at (6.5,.3) {(3)};
-	\node at (6.5,-1) {(b)};
-
-        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (z) at (10,0) {$Z$};
-	\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (y) at +(12,0){$Y$};
-        \node (mid) at (11,0) {};
-        \node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (h)
-        at (11,2) {$H(1)$};
-	%\node[draw,shape=circle, inner sep=0pt,minimum size=1cm] (x) at (2.5,3.5){$X$};
-	%\draw[->,line width=2pt](x)--(z);
-	%\draw[->,line width=2pt](x)--(y);
-	\draw[->,line width=2pt](z)--(y);
-        \draw[->,line width=2pt](h)--(mid);
-	\node at (11,-1) {(c)};
-
-\end{tikzpicture}
-\caption{A schematic representation of three causal models for
-  log data: (a) an observational study estimating the effects of hints
-  $H$ on posttest scores $Y$, (b) a mediation analysis estimating
-  natural direct effects of treatment assignment $Z$ on $Y$ and
-  indirect effects mediated by $H$, and (c) principal stratification.}
-\label{mediationFig}
-\end{figure}
-
-Figure \ref{mediationFig} is a schematic representation of the three
-approaches to causal modeling of log data that we will discuss in this
-paper.
-First, an observational study to estimate the effect of hint usage $H$
-on posttest scores $Y$, is represented by panel (a).
-This approach, which we will discuss in Section
-\ref{sec:observational}, discards the control group from the RCT and
-uses techniques from observational studies to determine if and how requesting
-a greater number of hints affects posttest scores.
-For the sake of illustration, we will use techniques similar to those
-discussed in \citet{rosenbaum2002observational},
- \citet{hansen2009b}, and \citet{ho:etal:2007}, combining a propensity
- score matching design with methods for analyzing experiments and
- parametric modeling.
-The conceptual and methodological issues we will discuss extend to
-other covariate control methods as well.
-
-Next, as represented in panel (b), we will conduct a mediation
-analysis, estimating natural
-indirect effects of CTA1 treatment assignment $Z$, as mediated by
-$H$. That is, we will estimate the component of $Z$'s effect on $Y$ that comes about
-due to $Z$'s effect on $H$---path (1)---and $H$'s effect on $Y$, path
-(2).
-Notice that path (2) is identical to panel (a)---the effect of $H$ on
-$Y$.
-We will use this equivalence to estimate and interpret indirect
-effects.
-As part of the mediation analysis, we will also estimate path (3), the
-natural direct effect of $Z$ on $Y$---that is, the component of $Z$'s
-effect on $Y$ that is not due to hints $H$.
-
-Finally, we will estimate a principal stratification model, as represented in panel
-(c).
-In the principal stratification analysis, we do not model $H$ as an
-intervention that could, in theory, be randomized.
-Instead, we posit a variable ``$H_i(1)$,''  subject $i$'s penchant for
-requesting hints \emph{if} $i$ is assigned to the treatment
-condition.
-Unlike $H$, $H(1)$ is well defined (if
-unobserved) for all participants in the RCT.
-Principal stratification treats $H(1)$ as an effect moderator or
-interaction variable: $H(1)$ does not affect $Y$, but the effect of
-$Z$ on $Y$ may vary as a function of $H(1)$.
-This analysis will largely follow the method of \citet{aoas}, which in
-turn built heavily on \citet{schwartz2011bayesian} and
-\citet{jin2008principal}.
-
-All three of these approaches are designed to model intermediate
-variables---those variables, like hint usage,
-that were possibly affected by $Z$ but were measured prior to $Y$.
-Though they were developed for more traditional data scenarios, the
-novel scenario of computer log data from an educational technology RCT
-presents new and interesting challenges and opportunities.
+The \texttt{R} and Stan code used to for all of the analyses in the paper can be found in a github repository, \url{https://github.com/adamSales/logDataCaseStudy}.
 
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Tue Feb 11 09:12:59 2020
+% Mon Apr 27 14:45:08 2020
 \begin{table}[ht]
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Tue Feb 11 09:12:59 2020
+% Mon Apr 27 14:45:09 2020
 \begin{table}[ht]

### 2020-05-06 paper.Rnw
added some introductory material to principal stratification section #5

@@
 treatment effect.
-Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
+Principal stratification \citep[PS;][]{frangakis} takes a different
+approach entirely, more akin to moderation than to mediation.
+
+The goal of principal stratification is to stratify all students in
+the RCT---in both the treatment and control arms---based on how they
+would use the software were they (perhaps counterfactually) assigned
+to the treatment condition, and then to estimate the average effect of
+assignment to the treatment condition separately in different stratum.
+Principal stratification analysis requires data on outcomes
+(e.g. posttest scores) in both tretment arms, as well as data on
+implementation.
+In classical principal stratification
+\citep[e.g.][]{page2012principal,feller2016compared}, the intermediate
+variable, implementation, is discrete or categorical, leading to
+discrete strata.
+Other approachs \citep[e.g.]{gilbertHudgens,jin2008principal} allow
+the intermediate variable to be continuous.
+The method we use here is based on \citet{aoas}, in which the variable
+defining principal strata is latent---specifically, the ``ability''
+parameter is \eqref{eq:rasch}.
+For this approach, implementation is measured with a series of binary
+measurements for each student, with each measurement taken from a
+specific problem or section.
+Other implementation data structures may call for different
+measurement models.
+Finally, covariates may not be strictly necessary for principal
+stratification, but they can be extremely helpful.
+
 In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
@@
 Our approach to PS estimation is model-based and Bayesian, following
-\citet{page2012principal}, \citet{feller2016compared},
+\citet{jin2008principal},
 \citet{schwartz2011bayesian} and others.

### 2020-05-07 paper.Rnw
finished (draft of) intro to principal stratification, addressing #8, #5

@@
 
-The goal of principal stratification is to stratify all students in
+The goal of PS is to stratify all students in
 the RCT---in both the treatment and control arms---based on how they
@@
 assignment to the treatment condition separately in different stratum.
-Principal stratification analysis requires data on outcomes
+In an RCT, identification of these average effects relies on
+randomization of treatment assignment,
+rather than confounding control or assumptions about
+unmeasured confounders.
+PS analysis requires data on outcomes
 (e.g. posttest scores) in both tretment arms, as well as data on
 implementation.
-In classical principal stratification
+In classical PS
 \citep[e.g.][]{page2012principal,feller2016compared}, the intermediate
@@
 measurement models.
-Finally, covariates may not be strictly necessary for principal
-stratification, but they can be extremely helpful.
-
+Finally, covariates may not be strictly necessary for PS, but they can be extremely helpful.
+
+Below, we describe a Bayesian PS model fit in one step.
+However, the full process of PS analysis is a much more involved
+procedure.
+After a formal introduction in Section \ref{sec:psIntro},
+\S~\ref{sec:psModel} describes a set of sub-models for student hint
+requests, posttest scores, and treatment effects.
+Ideally, each of these models should be carefully tailored to each
+data application, and each must be rigorously tested.
+Model checking includes examination of residual data plots, fake data
+simulation, and fitting a range of alternative specifications in a
+sensitivity analysis.
+More details for checking PS models can be found in \citet{aoas},
+\citet{psTutorial}, and for Bayesian models in general in
+\citet{gelman2014bayesian} and \citet{mcelreath2020statistical}.
+
+\subsection{Principal Effects for EdTech Log Data}\label{sec:psIntro}
 In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
@@
 
-The goal of principal stratification is to estimate ``principal
+The goal of PS is to estimate ``principal
 effects'':
@@
 \eqref{eq:mixture}.
-With this change, the principal stratification estimand became
+With this change, the PS estimand became
 \begin{equation*}
@@
 the treatment effect for students whose potential hint usage, if assigned to the treatment condition $\eta(1)=r$.
-A full treatment of this variant of principal stratification, including a
+A full treatment of this variant of PS, including a
 discussion of identification and estimation, may be found in
@@
 
-\subsection{Identification and Estimation}
+\subsection{Specifying PS Models}\label{sec:psModel}
 Estimating $\tau(r)$ required estimating $\EE[Y(0)|\eta(1)=r]$,
@@
 normal.
-The rest of the parameters received Stan's default uniform priors.
+The rest of the parameters received standard reference priors.
 In all cases, we expected true parameter values to be much smaller in
 magnitude than the prior standard deviation.
+
+\subsection{Estimating Principal Effects}\label{sec:PSest}
 

### 2020-05-08 paper.Rnw
edits to intros of mediation and ps sections to clarify research Qs

@@
 mediator, which itself, in turn, affected the outcome.
-In contrast, the direct effect is the component of the effect that
-operates via other mechanisms.
+Assignment to the treatment condition allowed students to request hints during their work---how did the availability of hints affect their posttest scores?
+The direct effect is the component of the overall effect that operates via other mechanisms. 
+For instance, how would assignment to CTA1 affect posttest scores were hints not available (or, at least, limited)?
 
 The method we present here builds on the results from the observational study in the previous section.
-Estimating the indirect effect is a way to interpret the estimated effect of $H$ in the context of the overall treatment effect of the EdTech intervention.
 In fact, in our analysis estimating indirect effects requires no more data than the observational study.
@@
 \section{Principal Stratification}\label{sec:principalStratification}
-The observational study approach and mediation analysis both rely on
-an ignorability assumption to estimate the effect of $H$ on $Y$, in
-the latter case framing this effect as one component of the overall
-treatment effect.
-Principal stratification \citep[PS;][]{frangakis} takes a different
-approach entirely, more akin to moderation than to mediation.
-
-The goal of PS is to stratify all students in
+
+In a principal stratification \citep[PS;][]{frangakis}  analysis, the researcher stratifies all students in
 the RCT---in both the treatment and control arms---based on how they
-would use the software were they (perhaps counterfactually) assigned
-to the treatment condition, and then to estimate the average effect of
+\emph{would} use the software were they (perhaps counterfactually) assigned
+to the treatment condition.
+The goal is to estimate the average effect of
 assignment to the treatment condition separately in different stratum.
-In an RCT, identification of these average effects relies on
+In our analysis, we imagine that each student has his or her own proclivity to request hints, when possible: if assigned to treatment, each student would tend to request hints at a different rate.
+How does the effect of assignment to the CTA1 condition vary with these proclivities?
+Would the benefit of being assigned to the CTA1 condition be higher or lower for students who would tend to request more hints than for those who would tend to request fewer?
+
+In an RCT, identification of these varying effects relies on
 randomization of treatment assignment,
@@
 unmeasured confounders.
+Unlike the observational study or mediation analysis, PS in an RCT requires no assumptions about unmeasured confounding. 
 PS analysis requires data on outcomes

### 2020-05-08 paper.tex
edits to intros of mediation and ps sections to clarify research Qs

@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Mon Apr 27 14:45:08 2020
+% Fri May  8 16:01:57 2020
 \begin{table}[ht]
@@
 \section{Principal Stratification}\label{sec:principalStratification}
-The observational study approach and mediation analysis both rely on
-an ignorability assumption to estimate the effect of $H$ on $Y$, in
-the latter case framing this effect as one component of the overall
-treatment effect.
-Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
+
+In a principal stratification \citep[PS;][]{frangakis}  analysis, the researcher stratifies all students in
+the RCT---in both the treatment and control arms---based on how they
+\emph{would} use the software were they (perhaps counterfactually) assigned
+to the treatment condition.
+The goal is to estimate the average effect of
+assignment to the treatment condition separately in different stratum.
+In our analysis, we imagine that each student has his or her own proclivity to request hints, when possible: if assigned to treatment, each student would tend to request hints at a different rate.
+How does the effect of assignment to the CTA1 condition vary with these proclivities?
+Would the benefit of being assigned to the CTA1 condition be higher or lower for students who would tend to request more hints than for those who would tend to request fewer?
+
+In an RCT, identification of these varying effects relies on
+randomization of treatment assignment,
+rather than confounding control or assumptions about
+unmeasured confounders.
+Unlike the observational study or mediation analysis, PS in an RCT requires no assumptions about unmeasured confounding. 
+PS analysis requires data on outcomes
+(e.g. posttest scores) in both tretment arms, as well as data on
+implementation.
+In classical PS
+\citep[e.g.][]{page2012principal,feller2016compared}, the intermediate
+variable, implementation, is discrete or categorical, leading to
+discrete strata.
+Other approachs \citep[e.g.]{gilbertHudgens,jin2008principal} allow
+the intermediate variable to be continuous.
+The method we use here is based on \citet{aoas}, in which the variable
+defining principal strata is latent---specifically, the ``ability''
+parameter is \eqref{eq:rasch}.
+For this approach, implementation is measured with a series of binary
+measurements for each student, with each measurement taken from a
+specific problem or section.
+Other implementation data structures may call for different
+measurement models.
+Finally, covariates may not be strictly necessary for PS, but they can be extremely helpful.
+
+Below, we describe a Bayesian PS model fit in one step.
+However, the full process of PS analysis is a much more involved
+procedure.
+After a formal introduction in Section \ref{sec:psIntro},
+\S~\ref{sec:psModel} describes a set of sub-models for student hint
+requests, posttest scores, and treatment effects.
+Ideally, each of these models should be carefully tailored to each
+data application, and each must be rigorously tested.
+Model checking includes examination of residual data plots, fake data
+simulation, and fitting a range of alternative specifications in a
+sensitivity analysis.
+More details for checking PS models can be found in \citet{aoas},
+\citet{psTutorial}, and for Bayesian models in general in
+\citet{gelman2014bayesian} and \citet{mcelreath2020statistical}.
+
+\subsection{Principal Effects for EdTech Log Data}\label{sec:psIntro}
 In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Mon Apr 27 14:45:09 2020
+% Fri May  8 16:01:58 2020
 \begin{table}[ht]

### 2020-05-08 paper.Rnw
added discussion of hint model #3

@@
 \citep{rasch} to hint request data.
-Specifically, we modeled the probability of a hint request in each problem
+Specifically, we modeled the probability of at least one hint request in each problem
 $p$, worked by student $i$, in which $i$ requested a hint and/or
@@
 \end{equation}
-where $h_{ip}$ is the event that student $i$ requests a hint on
+where $h_{ip}$ is the event that student $i$ requests at least one hint on
 problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
@@
 problems in section $s$.
-This essentially assumes that all problems within each section are equally
+This essentially models all problems within each section as equally
 likely to elicit a hint request.
+
+With even more granular data, \eqref{eq:rasch} could be further elaborated. 
+In the Cognitive Tutor, problems often have multiple parts, and each part offers its own hints.
+Researchers in posession of data on each part of each worked problem could specify a model at the part-level, instead of at the problem-level like \eqref{rasch}. 
+Perhaps an even better option (suggested by an anonymous reviewer) would incorporate data on the specific skills required to solve each problem.
+While sections of the tutor differ from each other in the skills they teach, not every problem in each section includes the same set of skills; different problems requiring the same skill set will be similar in difficulty \citep[see, e.g.]{goldin2012learner,pavlik2009performance}.
 

### 2020-05-13 paper.Rnw
revised statements of applied conclusions #2

@@
 Therefore, a sensitivity analysis considering an omitted covariate as important as pretest scores may be too pessimistic.
-Instead,
-the final column of Table \ref{tab:matchResults} gives sensitivity intervals for the omission of a covariate that, at most, predicts $Y$ and $H$ as well as ethnicity dummy variables, the second most important of our observed covariates.
+On the other hand, it is likely that pretest scores do not fully capture between-student variance in academic ability; that is, there may be important components of prior ability (such as the ability to learn algebra) that are correlated with hint requests and posttest scores but not reflected in pretest scores.
+A hypothetical measurement of one of those components may indeed have constituted a more important covariate than pretest scores.
+
+A more optimistic scenario is reflected in the final column of Table \ref{tab:matchResults}.
+That column gives sensitivity intervals for the omission of a covariate that, at most, predicts $Y$ and $H$ as well as ethnicity dummy variables, the second most important of our observed covariates.
 These sensitivity intervals are also wide, but the interval corresponding
@@
 cannot explain the estimated OLS-weighted ATE.
-Allowing for the possible omission of such a covariate, we infer that requesting a
+Allowing for the possible omission of such a covariate, the results suggest that requesting a
 large number of hints hurts students' posttest scores.
@@
 a standard error of \Sexpr{round(effects$tot_eff$std.error*(mixProp),2)}.
-This analysis suggests that the availability of hints
+Like the $TOT$ estimate on which it is based, this estimate of indirect effects assumes matched ignorabilty---no confounding between $H$ and $Y$ within matched sets.
+With this important caveat, this analysis suggests that the availability of hints
 actually \emph{lowers} the treatment effect, that CTA1 works not
@@
 
-Table \ref{tab:mediation} summarizes our mediation results.
+Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availablity of hints (an indirect effect of \Sexpr{round(mixProp*effects[['tot_eff']][['coefficients']],2)} standard deviations), and the effect due to CTA1's other mechanisms is \Sexpr{round(dir2.1t[1],2)}.
 
@@
 covariates.
-As far as we can tell, the optimal strategy is to request few, if any, hints.
-While our estimate is subject to bias from an omitted confounder, the omitted variable must be a relatively strong predictor of hint usage and of posttest scores in order to explain the observed relationship.
+Taken at face value, this suggests that the optimal strategy is to request few, if any, hints.
+%While our estimate is subject to bias from an omitted confounder, the omitted variable must be a relatively strong predictor of hint usage and of posttest scores in order to explain the observed relationship.
+However, this conclusion assumes that there were no omitted confounders.
 An omitted confounder that is as important as pretest could explain the relationship, but an omitted confounder as important as ethnicity---our second most important covariate---could not (so long as we estimate an OLS-weighted average effect).
@@
 overall CTA1 treatment effect.
-Specifically, our analysis showed that CTA1 affected students'
+Specifically, the analysis in \ref{sec:mediation} showed that CTA1 affected students'
 posttest scores in (at least) two opposite ways: by allowing hints, it

### 2020-05-13 paper.tex
revised statements of applied conclusions #2

@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Fri May  8 16:01:57 2020
+% Wed May 13 17:14:14 2020
 \begin{table}[ht]
@@
 a standard error of 0.02.
-This analysis suggests that the availability of hints
+Like the $TOT$ estimate on which it is based, this estimate of indirect effects assumes matched ignorabilty---no confounding between $H$ and $Y$ within matched sets.
+With this important caveat, this analysis suggests that the availability of hints
 actually \emph{lowers} the treatment effect, that CTA1 works not
@@
 
-Table \ref{tab:mediation} summarizes our mediation results.
+Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availablity of hints (an indirect effect of -0.04 standard deviations), and the effect due to CTA1's other mechanisms is 0.18.
 
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Fri May  8 16:01:58 2020
+% Wed May 13 17:14:15 2020
 \begin{table}[ht]

### 2020-05-20 paper.Rnw
simplified table 3 #2

@@
 \begin{table}
-\begin{tabular}{llll}
+\centering
+\begin{tabular}{lll}
 \hline
-Effect&Method&Estimate&CI\\
+Effect&Estimate&CI\\
 \hline
@@
 cat(
-  '$1-p_0=\\EE[H(1)-H(0)|Z=1]$ & By Definition&',round(mixProp,2),'&n/a\\\\
-$TOT=\\EE[Y(Z=1,H=1)-Y(Z=1,H=0)|Z=1]$&Matching&',round(effects[['tot_eff']][['coefficients']],2),'&',printci(with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
-$\\Delta=\\EE[Y(1,H(1))-Y(1,H(0))|Z=1]$&$(1-p_0)TOT$&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
-$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \\eqref{eq:directEffReg}&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
-$\\Xi=\\EE[Y(1,H(0))-Y(0,H(0))]$& \\eqref{eq:directEffReg} + covariates&',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\\\\\n')
+'Avg. Indirect Effect ($\\Delta$)&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
+Avg. Direct Effect ($\\Xi$)&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
+Avg. Direct Effect, covariate adjusted ($\\Xi$) &',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\\\\\n')
 @

### 2020-05-20 paper.tex
simplified table 3 #2

@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Wed May 13 17:14:14 2020
+% Wed May 20 10:38:55 2020
 \begin{table}[ht]
@@
 \begin{table}
-\begin{tabular}{llll}
+\centering
+\begin{tabular}{lll}
 \hline
-Effect&Method&Estimate&CI\\
+Effect&Estimate&CI\\
 \hline
-$1-p_0=\EE[H(1)-H(0)|Z=1]$ & By Definition& 0.3 &n/a\\
-$TOT=\EE[Y(Z=1,H=1)-Y(Z=1,H=0)|Z=1]$&Matching& -0.14 & [-0.25,-0.03] \\
-$\Delta=\EE[Y(1,H(1))-Y(1,H(0))|Z=1]$&$(1-p_0)TOT$& -0.04 & [-0.07,-0.01] \\
-$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$&Regression \eqref{eq:directEffReg}& 0.14 & [-0.08,0.37] \\
-$\Xi=\EE[Y(1,H(0))-Y(0,H(0))]$& \eqref{eq:directEffReg} + covariates& 0.18 & [-0.01,0.38] \\
+Avg. Indirect Effect ($\Delta$)& -0.04 & [-0.07,-0.01] \\
+Avg. Direct Effect ($\Xi$)& 0.14 & [-0.08,0.37] \\
+Avg. Direct Effect, covariate adjusted ($\Xi$) & 0.18 & [-0.01,0.38] \\
 
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Wed May 13 17:14:15 2020
+% Wed May 20 10:38:57 2020
 \begin{table}[ht]

### 2020-05-27 paper.Rnw
issue #6, more-concrete advice adjudicating btw methods

@@
 
+In summary: in situations where the average treatment effect of an
+implementation variable can be estimated with little confounding bias
+(due to the availability of high-quality covariates, good reason to
+believe in random, or haphazard, variation in implementation, or other
+factors), there is much to recommend the observational study approach.
+A wide variety of estimation techniques are available---such as the propensity score
+matching illustrated here---many of which are straightforward to carry
+out and highly transparent.
+The result---the ATE of implementation on the outcome---is an
+intuitive estimand.
+
+Mediation analysis can complement the observational study if,
+additionally, one well-represented level of the implementation
+variable is interpretable in the control condition.
+For instance, in our example, $H=0$ included both low hint usage in
+the treatment group and no access to hints in the control group.
+In that case, mediation analysis can contextualize the estimated
+effect in the observational study in terms of the overall treatment
+effect, so that implementation can be interpreted as a causal
+mechanism.
+
+However, if unobserved confounding is a serious threat, and complex
+statistical modeling (and model checking) is not a practical barrier,
+PS may be a better choice.
 % \begin{table}

### 2020-05-28 paper.Rnw
Suggestions, comments on #6
Adam, here are some suggestions, but please check carefully and feel free to reject.

--- 2020-05-20 paper.Rnw
@@
 
+In summary: there is much to recommend the observational study approach
+if risks of confounding bias appear limited
+(due to the availability of high-quality covariates, good reason to
+believe variation in implementation is random or haphazard, or other
+factors).
+A wide variety of straightforward and transparent estimation techniques 
+are available, such as the propensity score
+matching illustrated here, optionally also including covariates in the 
+outcomes model for additional protection against bias.
+The result---the ATE of implementation on the outcome---is an
+intuitive estimand.
+
+Mediation analysis may have stronger standing where
+at least one well-represented level of the implementation
+variable is interpretable in the control condition.
+In our example, $H=0$ included both low hint usage in
+the treatment group and no access to hints in the control group.
+When employed, mediation analysis can contextualize the estimated
+implementation effect from the observational study in terms of the 
+experiment's overall treatment
+effect, enabling implementation to be interpreted as a causal
+mechanism.
+
+However, where unobserved confounding is a serious threat, and complex
+statistical modeling (and model checking) is not a practical barrier,
+PS may be a better choice because no ignorability assumptions are necessary.
 % \begin{table}

### 2020-06-01 paper.Rnw
stuff about problem vs section level difficulty parameter #7, #3

@@
 function.
+
 Typically, the student parameter $\eta_i$ measures student ability;
 here, instead, it measures a student's proclivity to request a hint.
-The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
-dataset many problems were worked by very few students, so problem
-level parameters would be hard to estimate.
+The typical Rasch model includes a problem-level ``difficulty''
+parameter; here, it would measure the likelihood an individual problem
+elicits a hint request.
+That paramter is important in our context because the
+specific problems a student works on may influence his or her hint
+requests.
+For instance, two students with the same underlying proclivity to request
+hints $\eta$ may actually differ in their observed hint requests, because one
+worked harder problems than the other.
+However, in our dataset many problems were worked by very few students, so problem-level parameters would be hard to estimate.
 Instead, we included the section-level parameter $\delta_{s[p]}$,
@@
 problems in section $s$.
-This essentially models all problems within each section as equally
-likely to elicit a hint request.
-
-With even more granular data, \eqref{eq:rasch} could be further elaborated.
+Ignoring the differences between problems within a section could allow for wider
+variance in the average problem difficulty each student experiences
+than is captured by the section level parameters $\delta_{s[p]}$.
+That could, in turn, induce a difference in hint requests between two
+students with similar $\eta$ or vice-versa, and lead
+to bias in our estimate of $\eta$.
+Since the Cognitive Tutor selects problems for students based on estimates
+of their current skill mastery \citep{steveFaied2020}, this
+possibility is not out of the question.
+
+With even more granular data, \eqref{eq:rasch} could be further
+elaborated, avoiding this assumption.
 In the Cognitive Tutor, problems often have multiple parts, and each part offers its own hints.

### 2020-06-01 paper.Rnw
stuff about problem vs section level difficulty parameter #7, #3

@@
 
-Then, in order to dichotomize hint request behavior, we modeled
+In order to dichotomize hint request behavior, we modeled
 student effects $\eta$ as
@@
 Further, log data for sections that were not part of the
-standard CTA1 Algebra I curriculum and problems worked by fewer than 100
+standard CTA1 Algebra I curriculum and sections worked by fewer than 100
 students were omitted from the
-dataset.
+dataset.\footnote{The model was re-run without dropping sections, and
+  after dropping sections worked by fewer than 500 students, with
+  similar results.}
 

### 2020-06-01 paper.Rnw
#7 up to p12

@@
 Making an error on a problem without having requested a hint reflects
-a hesitance to request a hint when one might have been warranted.
+a hesitance to request a hint when one might have been warranted (or a
+careless mistake).
 $\bar{h}$ ignores the class of challenging problems on which students
@@
 Of course, the distribution of $H_i$ is unknown, and is presumably a
-function of $i$'s individual characteristics.
+function of $i$'s individual characteristics (which, as we shall see,
+differed at baseline between those students with $H=1$ and those with $H=0$).
 However, if these relevant characteristics were all measured and

### 2020-06-02 paper.Rnw
address Lukes comment about overlap #7

@@
 treatment group with $H=0$.
-If $Pr(H_i=1)$ were known for each $i$ (as would be the case if $H$
+If $Pr(H_i=1)$ were known for each $i$, and bounded away from 1 and 0, (as would be the case if $H$
 were randomized) then the average treatment effect (ATE) of $H$ on $Y$,
@@
 equivalent to conditioning on $\bm{x}$.
-We estimated $\pi$ using a Bayesian multilevel logistic regression,
+We estimated $\pi$ using a multilevel logistic regression,
 using the \texttt{lme4} package in
@@
 We constrained the match so that students could only be matched within
-schools.
-This was for two reasons: first, schools determine a number of factors
+schools, since schools determine a number of factors
 that in turn may impact posttest scores, including baseline
 student characteristics, pedagogical styles, and CTA1 usage patterns
-\citep[see][]{descriptivePaper}, and second, doing so facilitated the mediation
-analysis in the following section.
+\citep[see][]{descriptivePaper}.%, and second, doing so facilitated the mediation
+%analysis in the following section.
 Each matched set was allowed to contain any positive number of $H=1$
@@
 single $H=1$ subject.
-All students in the CTA1 group of the dataset were included in the match.
+
+This strategy allowed every student in the CTA1 arm our dataset to be included in the match.
+In contrast, many matching studies discard subjects in the data sample
+with estimated propensity scores close to 1 or 0, focusing on the
+``region of common support'' \citep[e.g.]{caliendo2008some,shadish2010primer}.
+Doing so requires modifying the causal estimand---studies can only
+estimate the ATE for included subjects and those who resemble them.
+We chose, instead, to include every subject in order to simplify the mediation
+analysis in the following section, and validated the match by
+inspecting covariate balance.
+This came at the cost of imprecision in estimating the overall ATE,
+which we discuss in more detail below.
 

### 2020-06-02 paper.Rnw
#7 up to p.19 (changed discussion of dichotomous H in mediaiton analysis)

@@
 decreases posttest scores by roughly a fifth of a standard deviation.
-That is, requesting hints hurts posttest scores---teachers should
-recommend against students' requesting hints.
+That is, they suggest that requesting hints hurts posttest scores.
 
@@
 group, whereas TOT estimates an effect for high hint users. Had these
-been very different, we might wonder why the effect of hint requests
+been very different, one might wonder why the effect of hint requests
 varies so much between high and low hint users, and if that suggests
@@
 both.
+(Recall that such imbalanced matched sets were the result of the
+decision to include every subject in the analysis.)
 It is larger in magnitude and much less noisy than the other two
@@
 
-Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
-the causal question.
-To make sense of the respective roles of $H$ and $\bar{h}$, the
-continuous measure on which it is based (see Section
-\ref{sec:dichotomizingHintUsage}),
-we may imagine a process in
-which students are assigned $H$, and under that constraint choose a
-value for $\bar{h}$.
-In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
-in the treatment group \Sexpr{round((1-mixProp)*100)}\% of students have $H=1$ and \Sexpr{round((mixProp)*100)}\% have
-$H=0$, and $\bar{h}$ varies within those groups.
-That structure implies a ``treatment mediator interaction''
-where the effect of the mediator on the outcome varies by treatment
-status.
-Of course, in our example treatment mediator interactions are not
-identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
-few treatment students asked for 0 hints.
-In any event, dichotomizing hint requests likely under-sells the role
-of hints, but allows a tractable mediation analysis.
+Dichotomizing hint requests into $H$ provides a
+way forward.
+The experimental design ensures that $H(0)=0$ for all experimental
+participants, since
+CTA1 software was unavailable to the control group.
+Therefore, $Y(1,H(0))\equiv Y(1,0)$.
+Specifically, for treated subjects $i$ for whom $H_i=0$, then, treatment assignment
+did not affect $H$, so $H_i(0)=H_i(1)=0$.
+%In other words, for these subjects,
+%\begin{equation*}
+%Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
+%\end{equation*}
+In other words, when $Z_i=1$ and $H_i=0$,
+$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
+indirect effect.
+
+Of course, $H=0$ in the treatment group typically means requesting
+\emph{some} hints, but not many, whereas subjects in the control group
+request exactly zero hints.
+One way to make sense of mediational estimands involving $H$ is to
+imagine an experiment in which students are assigned to different
+values of $H$.
+\Sexpr{round((1-mixProp)*100)}\% of students in the treatment group
+are randomly assigned to $H=1$ and \Sexpr{round((mixProp)*100)}\% to
+$H=0$.
+This assignment constrains $\bar{h}$, the proportion of problems on which they
+may request hints, so that if $H_i=0$,
+$\bar{h}_i$ is constrained to be less than
+\Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))},
+but if $H_i=1$ $\bar{h}_i$ must be greater than or equal to
+\Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
+In contrast, all students in the control group are assigned $H=0$ and $\bar{h}=0$.
 
 In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
-individual level, as $\delta(1)_i$ and $\xi(0)_i$.
-However, as in the usual causal inference case, individual effects are
-not identified without strong assumptions, and the target of
+individual level, as $\delta_i(1)$ and $\xi_i(0)$.
+However, individual effects are
+not identified without strong assumptions, so the target of
 estimation is instead taken to be super-population average effects,
-$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$. Our approach to estimating these
+$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$.
+Our approach to estimating these
 is to estimate the means of each of the component potential outcomes,
@@
 randomization of treatment assignment.
-
-On the other hand, randomization alone does not identify the strictly counterfactual
-potential outcome $\EE Y(1,H(0))$.
-In a typical mediation analysis of an RCT, identification of the
-distribution of $H(0)$ may be estimated based on the observed
-distribution of $H$ in the control group, due to randomization.
-In contrast, in the CTA1 study, $H$ is not observed in the control
-group.
-However, $H(0)$ was determined by the experimental design: since the
-CTA1 software was unavailable to the control group, $H(0)=0$ for all
-experimental participants, and $Y(1,H(0))\equiv Y(1,0)$.
-
-For treated subjects $i$ for whom $H_i=0$, then, treatment assignment
-did not affect $H$, so $H_i(0)=H_i(1)=0$.
-%In other words, for these subjects,
-%\begin{equation*}
-%Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
-%\end{equation*}
-Therefore, when $Z_i=1$ and $H_i=0$,
-$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
-indirect effect.
-
-Treated subjects with $H=1$ present more of a challenge.
+Additionally, as described above, $\delta_i(1)=0$ as long as $Z_i=1$ and $H_i=0$.
+
+Treated subjects with $H_i=1$ present more of a challenge.
 To estimate indirect effects for these subjects, we may rely on the
@@
 is independent of $H$.
-Therefore, if subject $i$ is matched to subject $j$, $H_i=1$, and
+Therefore, if subject $i$ is matched to subject $j$, with $H_i=1$, and
 $H_j=0$, we may estimate $Y_i(1,0)$ based on the observed

### 2020-06-02 paper.tex
#7 up to p.19 (changed discussion of dichotomous H in mediaiton analysis)

@@
 
-Dichotomizing hint requests into $H$, instead of $\bar{h}$, re-defines
-the causal question.
-To make sense of the respective roles of $H$ and $\bar{h}$, the
-continuous measure on which it is based (see Section
-\ref{sec:dichotomizingHintUsage}),
-we may imagine a process in
-which students are assigned $H$, and under that constraint choose a
-value for $\bar{h}$.
-In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
-in the treatment group 70\% of students have $H=1$ and 30\% have
-$H=0$, and $\bar{h}$ varies within those groups.
-That structure implies a ``treatment mediator interaction''
-where the effect of the mediator on the outcome varies by treatment
-status.
-Of course, in our example treatment mediator interactions are not
-identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
-few treatment students asked for 0 hints.
-In any event, dichotomizing hint requests likely under-sells the role
-of hints, but allows a tractable mediation analysis.
+Dichotomizing hint requests into $H$ provides a
+way forward.
+The experimental design ensures that $H(0)=0$ for all experimental
+participants, since
+CTA1 software was unavailable to the control group.
+Therefore, $Y(1,H(0))\equiv Y(1,0)$.
+Specifically, for treated subjects $i$ for whom $H_i=0$, then, treatment assignment
+did not affect $H$, so $H_i(0)=H_i(1)=0$.
+%In other words, for these subjects,
+%\begin{equation*}
+%Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
+%\end{equation*}
+In other words, when $Z_i=1$ and $H_i=0$,
+$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
+indirect effect.
+
+Of course, $H=0$ in the treatment group typically means requesting
+\emph{some} hints, but not many, whereas subjects in the control group
+request exactly zero hints.
+One way to make sense of mediational estimands involving $H$ is to
+imagine an experiment in which students are assigned to different
+values of $H$.
+70\% of students in the treatment group
+are randomly assigned to $H=1$ and 30\% to
+$H=0$.
+This assignment constrains $\bar{h}$, the proportion of problems on which they
+may request hints, so that if $H_i=0$,
+$\bar{h}_i$ is constrained to be less than
+0.6,
+but if $H_i=1$ $\bar{h}_i$ must be greater than or equal to
+0.6.
+In contrast, all students in the control group are assigned $H=0$ and $\bar{h}=0$.
 
 In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
-individual level, as $\delta(1)_i$ and $\xi(0)_i$.
-However, as in the usual causal inference case, individual effects are
-not identified without strong assumptions, and the target of
+individual level, as $\delta_i(1)$ and $\xi_i(0)$.
+However, individual effects are
+not identified without strong assumptions, so the target of
 estimation is instead taken to be super-population average effects,
-$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$. Our approach to estimating these
+$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$.
+Our approach to estimating these
 is to estimate the means of each of the component potential outcomes,

### 2020-06-03 paper.Rnw
took out conditioning on Z=1 for indirect effect fixed mistake in eq before (5)

@@
 In the Cognitive Tutor, problems often have multiple parts, and each part offers its own hints.
-Researchers in posession of data on each part of each worked problem could specify a model at the part-level, instead of at the problem-level like \eqref{rasch}.
+Researchers in posession of data on each part of each worked problem could specify a model at the part-level, instead of at the problem-level like \eqref{eq:rasch}.
 Perhaps an even better option (suggested by an anonymous reviewer) would incorporate data on the specific skills required to solve each problem.
@@
 \begin{equation*}
- \hat{\Delta}\equiv\widehat{\EE[\delta(0)|Z=1]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
+ \hat{\Delta}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
 \end{equation*}
-(Typically, the overall average indirect effect, $\EE[\delta(0)]$, is
-estimated instead of the average indirect effect of the treatment
-group, $\EE[\delta(0)|Z=1]$.
-We estimated the latter since data from the control group was not used
-in our estimation strategy, and estimating the overall average would
-simply complicate our standard error estimation.)
 
@@
 \begin{equation*}
-TOT=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]
+TOT=\EE[Y(1,1)-Y(1,0)|H=1]
 \end{equation*}
@@
 \begin{equation*}
-\Delta=\EE[Y(1,H(1))-Y(1,0)|Z=1]
+\Delta=\EE[Y(1,H(1))-Y(1,0)]
 \end{equation*}
@@
 \begin{split}
-\Delta&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
-&\qquad +\EE[Y(1,0)-Y(1,0)|Z=1,H=0]Pr(H=0)
+\Delta&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
+&\qquad +\EE[Y(1,0)-Y(1,0)|H=0]Pr(H=0)
 \end{split}
 \\[2ex]
-&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
+&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
 &=Pr(H=1)TOT
 \end{align*}
-Since $Pr(H=1)=$\Sexpr{round(mixProp,2)} by definition,
+Since, by definition, $Pr(H=1)=$\Sexpr{round(mixProp,2)}, we have that
 $\hat{\Delta}=$\Sexpr{round(mixProp,2)}$TOT$.
@@
 Like the $TOT$ estimate on which it is based, this estimate of indirect effects assumes matched ignorabilty---no confounding between $H$ and $Y$ within matched sets.
-With this important caveat, this analysis suggests that the availability of hints
+With this important caveat, this analysis suggests that the
+wide availability of a hints
 actually \emph{lowers} the treatment effect, that CTA1 works not
@@
 \begin{equation*}
-\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)\hat{Y}(1,0)
+\tilde{Y}\equiv Z\hat{Y}(1,0)+(1-Z)Y(0,0)
 \end{equation*}

### 2020-06-03 paper.tex
took out conditioning on Z=1 for indirect effect fixed mistake in eq before (5)

@@
 \begin{split}
-\Delta&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
-&\qquad +\EE[Y(1,0)-Y(1,0)|Z=1,H=0]Pr(H=0)
+\Delta&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
+&\qquad +\EE[Y(1,0)-Y(1,0)|H=0]Pr(H=0)
 \end{split}
 \\[2ex]
-&=\EE[Y(1,1)-Y(1,0)|Z=1,H=1]Pr(H=1)\\
+&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
 &=Pr(H=1)TOT
 \end{align*}
-Since $Pr(H=1)=$0.3 by definition,
+Since, by definition, $Pr(H=1)=$0.3, we have that
 $\hat{\Delta}=$0.3$TOT$.

### 2020-06-05 R/dataAnalysis.r
used clubsandwich ci estimate not just SE

@@ -238,13 +238,16 @@ dat2 <- rbind(dat1[,names(datC)],datC)
 
 dir1.1 <- lm(yhat~treatment+pair,data=dat2)
 dir1.1vcv <- vcovCR(dir1.1,dat2$schoolid2,type='CR2')
-dir1.1t <- coeftest(dir1.1,dir1.1vcv)[2,]
-dir1.1ci <- coefci(dir1.1,'treatment',vcov.=dir1.1vcv)
+dir1.1t <- coef_test(dir1.1,dir1.1vcv,coefs='treatment')
+dir1.1ci <- conf_int(dir1.1,dir1.1vcv,coefs='treatment')
+dir1.1ci <- c(dir1.1ci$CI_L[1],dir1.1ci$CI_U[1])
 
 dir2.1 <- update(dir1.1,.~.+poly(xirt,2)+race+grade+esl+frl+sex)
 dir2.1vcv <- vcovCR(dir2.1,dat2$schoolid2,type='CR2')
-dir2.1t <- coeftest(dir2.1,dir2.1vcv)[2,]
-dir2.1ci <- coefci(dir2.1,'treatment',vcov.=dir2.1vcv)
+dir2.1t <- coef_test(dir2.1,dir2.1vcv,coefs='treatment')
+dir2.1ci <- conf_int(dir2.1,dir2.1vcv,coefs='treatment')
+dir2.1ci <- c(dir2.1ci$CI_L[1],dir2.1ci$CI_U[1])
+
 
 
 

### 2020-06-05 paper.Rnw
lotta changes to mediation sec 4, #7

@@
 hints.
-The following three sections will present an observational study,
+The next three sections will present an observational study,
 which discards the control group altogether and estimates the effect
@@
 
-\subsection{Estimating the Effects of $H$ on Posttest Scores}
+\subsection{Estimating the Effects of $H$ on Posttest Scores}\label{sec:obsEst}
 
@@
 mediator, which itself, in turn, affected the outcome.
-Assignment to the treatment condition allowed students to request hints during their work---how did the availability of hints affect their posttest scores?
+Assignment to the treatment condition allowed students to request many
+hints during their work---how did the wide availability of hints affect their posttest scores?
 The direct effect is the component of the overall effect that operates via other mechanisms.
-For instance, how would assignment to CTA1 affect posttest scores were hints not available (or, at least, limited)?
+For instance, how would assignment to CTA1 affect posttest scores were
+hints more limited?
 
@@
 
+It is worth noting here that conducting mediation analysis requires
+measuring or positing a value of the mediator in the control arm of
+the study.
+We know that CTA1 hints were not available to the control group, so we
+set $H=0$.
+However, other facets of log data may not be defined, or coherent, in
+the control group; for instance, students' proclivity to master the
+skills of one CTA1 section before moving on to the next \citep{aoas}
+or teachers' practice of reassigning students to new CTA1 sections
+\citep{reassignmentEffect}.
+In cases such as these, the mediation framework may not apply.
+
 The following subsection reviews the formal definitions of direct and indirect effects.
-The next two subsections illustrate the estimation of direct and indirect effects, respectively, in our study of hints.
-It turns out that indirect effects are are nearly unidentifiable, without some acrobatics; however, under the identification scheme we present here, estimating indirect effects is exceedingly simple.
-Therefore, Section \ref{sec:indirectEst} will mostly consist of a discussion of identification.
-Section \ref{sec:directEst} builds on the framework of the previous subsections to estimate direct effects with a regression model.
+Subsection \ref{sec:medID} discusses identification, and
+\ref{sec:indirectEst} and \ref{sec:directEst} discuss estimation of
+indirect and direct effects, respectively.
 
 \subsection{Review: Defining Direct and Indirect Effect}
-More precisely, express a subject's potential outcomes as a function
+Express a subject's potential outcomes as a function
 of two variables, treatment assignment $Z$ and hint requests $H$:
@@
 building blocks of causal mediation analysis
-\citep[e.g.][]{vanderweele2015explanation,sales2017mediation}.
+\citep[e.g.][]{vanderweele2015explanation}.
 Let $Y(z,H(z'))$
@@
 assignment to CTA1 affects a student's posttest score by (possibly)
-causing him or her to request hints.
+causing him or her to request many hints.
 $\xi(0)$ gives the effect of assignment to CTA1 if hints were,
@@
 
-\subsection{Identification and Estimation Log Data  Indirect
-  Effects}\label{sec:indirectEst}
+Rather than attempt to estimate individual direct and indirect effects
+$\xi_i$ and $\delta_i$, our goal will be to estimate their means,
+$\EE \xi$ and $\EE \delta$.
+
+\subsection{Identification of Direct and Indirect
+  Effects}\label{sec:medID}
 
@@
 way forward.
-The experimental design ensures that $H(0)=0$ for all experimental
-participants, since
-CTA1 software was unavailable to the control group.
-Therefore, $Y(1,H(0))\equiv Y(1,0)$.
-Specifically, for treated subjects $i$ for whom $H_i=0$, then, treatment assignment
-did not affect $H$, so $H_i(0)=H_i(1)=0$.
-%In other words, for these subjects,
-%\begin{equation*}
-%Y_i(1,H(0))=Y_i(1,0)=Y_i(1,H(1))=Y_i
-%\end{equation*}
-In other words, when $Z_i=1$ and $H_i=0$,
-$\delta_i(1)=Y_i(1,H(1))-Y_i(1,H(0))=0$---there is no
-indirect effect.
-
-Of course, $H=0$ in the treatment group typically means requesting
+Since the experimental design ensures that $H(0)=0$, $Y(1,H(0))\equiv
+Y(1,0)$; this potential outcome is observed for subjects in
+the treatment arm with $H=0$.
+As above, $Y(0,H(0))=Y(0,0)$ and $Y(1,H(1))$ are the observed outcomes
+for subjects in the control and treatment arms, respectively.
+Therefore, we observe each of the potential outcomes in the definitions of
+$\delta(1)$ and $\xi(0)$ for at least some non-trivial
+portion of subjects in the study sample.
+
+Of course, $H=0$ means something different in the two treatment arms:
+in the treatment group $H=0$ typically means requesting
 \emph{some} hints, but not many, whereas subjects in the control group
@@
 
-In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
-individual level, as $\delta_i(1)$ and $\xi_i(0)$.
-However, individual effects are
-not identified without strong assumptions, so the target of
-estimation is instead taken to be super-population average effects,
-$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$.
-Our approach to estimating these
-is to estimate the means of each of the component potential outcomes,
-$\EE Y(1,H(1))$, $\EE Y(1,H(0))$, and $\EE Y(0,H(0))$.
-The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
-Y(0,H(0))=\EE Y(Z=0)$ are non-parametrically identified due to the
-randomization of treatment assignment.
-Additionally, as described above, $\delta_i(1)=0$ as long as $Z_i=1$ and $H_i=0$.
-
-Treated subjects with $H_i=1$ present more of a challenge.
-To estimate indirect effects for these subjects, we may rely on the
-same causal setup as in the observational approach:
-matched ignorability and the matching design of Section
-\ref{sec:observational}.
-Each member of the treatment group with $H=1$ is matched to other
-treatment-group subjects with $H=0$.
-Matched ignorability states that within these matched sets, $Y(1,0)$
-is independent of $H$.
-Therefore, if subject $i$ is matched to subject $j$, with $H_i=1$, and
-$H_j=0$, we may estimate $Y_i(1,0)$ based on the observed
-$Y_j=Y_j(1,0)$.
-When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
-$Y_i(1,0)$ is estimated with the average of their observed outcomes.
-
-Specifically, let
+Average potential outcomes $\EE Y(1,H(1))$ and $\EE Y(0,H(0))$ are
+identified due to the randomization of $Z$, but estimating $\EE
+Y(1,H(0))$ requires imputing the
+potential outcome $Y(1,0)$ for treated subjects with $H_i=1$.
+However, it turns out that the matching estimators from
+\ref{sec:observational} already solved this problem:
+under matched ignorability, within matched sets, $Y(1,0)$
+is independent of $H$, implying that $\EE[Y(1,0)|M,H=1]=\EE[Y(1,0)|M,H=0]$.
+That suggests imputations
 \begin{equation}\label{eq:estY10}
 \hat{Y}_i(1,0)=\begin{cases}
- Y_i & H_i=0\\[3ex]
-\frac{\displaystyle\sum_{j:M_j=M_i}
-  Y_j(1-H_j)}{\displaystyle\sum_{j:M_j=M_i} 1-H_j} & H_i=1
+ Y_i & H_i=0\\
+\bar{Y}_{M=m_i,H=0} & H_i=1
 \end{cases}
 \end{equation}
-For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
-average outcome of their matched pairs with $H=0$.
-
-Now, for each member $i$ of the treatment group we have observed
-$Y_i(1,H(1))=Y_i$, and estimated or inferred
-$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
-Then, we can estimate $\Delta$, the average indirect effect in the
-treatment group, by averaging over the treatment group:
+That is, when $Z_i=1$ and $H_i=0$, no imputation is necessary, since
+$Y_i(1,0)=Y_i$, the observed outcome. When $Z_i=1$ and $H_i=1$,
+impute $Y(1,0)$ as the average of the observed outcomes from the $H=0$ subjects
+matched to $i$.
+Matched ignorability implies that the imputations will be unbiased.
+
+\subsection{Estimating Log Data Indirect Effects}\label{sec:indirectEst}
+
+Since $\EE \delta(1)$ involves only $Z=1$ potential outcomes, and
+$H(0)=0$ is known based on the study design, $\EE\delta(1)$ may be
+estimated using data from the treatment group only.
+
+Using the imputations from \eqref{eq:estY10}, we estimate
+$\EE\delta(1)$ as
 \begin{equation*}
- \hat{\Delta}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
+\widehat{\EE\delta(1)}=\widehat{\EE Y(Z=1)}-\widehat{\EE
+  Y(1,0)}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,0)
 \end{equation*}
-
-This estimate is a close cousin of the ``TOT'' estimate in Table
-\ref{tab:matchResults} from the previous section.
-Recall that the TOT effect is the average effect of $H$ on treatment
-group subjects with $H=1$.
-Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
-data from only the treatment group.
-TOT is the average difference between the realized outcomes for students
-with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
-That is,
+where $n_T=\sum_i Z_i$, the number of subjects in the treatment
+group.
+Since $Y_i=\hat{Y}_i(1,0)$ whenever $H_i=0$, the summand is non-zero
+only when $H_i=1$.
+Further, note that when multiple $H=1$ subjects share a
+matched set, they also share imputations $\hat{Y}(1,0)$.
+Re-writing $\widehat{\EE\delta(1)}$ in terms of matched sets $M=m$,
+\begin{align}
+\widehat{\EE\delta(1)}&=\frac{1}{n_T}\displaystyle\sum_m\displaystyle\sum_{i:H_i=1,M_i=m}
+                        Y_i-\bar{Y}_{M=m_i,H=0} \nonumber \\
+&=\frac{1}{n_T}\displaystyle\sum_m
+  n_{1m}(\bar{Y}_{M=m_i,H=1}-\bar{Y}_{m=m_i,H=0}) \nonumber \\
+&=\frac{1}{n_T}\displaystyle\sum_mn_{1m}\hat{\tau}_m=\frac{\sum_iH_i}{n_T}\widehat{TOT} \label{eq:indTot1}
+\end{align}
+where $n_{1m}$ and $\hat{\tau}_m$ were defined in Section
+\ref{sec:obsEst} as the number of $H=1$ subjects in match $m$ and the
+difference in the sample mean of $Y$ between $H=1$ and $H=0$ subjects
+in match $m$, respectively, and $\widehat{TOT}$ is the TOT estimate,
+\eqref{eq:tauWeightedMean} with $w_m=n_{1m}$.
+The upshot of \eqref{eq:indTot1} is that under matched ignorability,
+$\EE\delta(0)$ can be estimated as the TOT times the proportion of the
+treatment group with $H=1$.
+
+Actually, this relationship between $\EE\delta(0)$ and the TOT holds
+in the population, regardless of how the TOT is estimated.
+This is because when $H_i=0$, then $H_i(1)=H_i(0)=0$, so the individual direct effect
+$\delta_i(1)=0$, and when $H_i=1$,
+ $\delta_i(1)=Y_i(H=1)-Y_i(H=0)$, the effect of $H$ on $Y$.
+Therefore,
 \begin{equation*}
-TOT=\EE[Y(1,1)-Y(1,0)|H=1]
+\EE \delta(1)=\EE\EE[\delta(1)|H]=\EE[\delta(1)|H=1]Pr(H=1)=TOTPr(H=1)
 \end{equation*}
 
-$\hat{\Delta}$ estimates the difference between the average outcome
-for students in the entire treatment group, and the average outcome
-for the treatment group were $H=0$ for all students.
-That is
-\begin{equation*}
-\Delta=\EE[Y(1,H(1))-Y(1,0)]
-\end{equation*}
-This can be re-written:
-\begin{align*}
-\begin{split}
-\Delta&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
-&\qquad +\EE[Y(1,0)-Y(1,0)|H=0]Pr(H=0)
-\end{split}
-\\[2ex]
-&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
-&=Pr(H=1)TOT
-\end{align*}
-Since, by definition, $Pr(H=1)=$\Sexpr{round(mixProp,2)}, we have that
-$\hat{\Delta}=$\Sexpr{round(mixProp,2)}$TOT$.
-Estimated thusly, the indirect effect is negative: \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
+In the hint study, the definition of $H$ implies that
+$Pr(H=1)=\sum_iH_i/n_T=$\Sexpr{round(mixProp,2)}.
+Therefore, our estimate of the average indirect effect is
+$\widehat{\EE \delta(1)}=$\Sexpr{round(mixProp,2)}$\widehat{TOT}$.
+In Table \ref{tab:matchResults}, we estimated the TOT as
+\Sexpr{round(effects[['tot_eff']][['coefficients']],2)} with a
+standard error of \Sexpr{round(effects$tot_eff$std.error,2)}.
+That implies an average indirect effect of \Sexpr{round(effects[['tot_eff']][['coefficients']]*(mixProp),2)}, with
 a standard error of \Sexpr{round(effects$tot_eff$std.error*(mixProp),2)}.
-Like the $TOT$ estimate on which it is based, this estimate of indirect effects assumes matched ignorabilty---no confounding between $H$ and $Y$ within matched sets.
+
+It is worth repeating that these estimates assume matched ignorabilty---no confounding between
+$H$ and $Y$ within matched sets.
 With this important caveat, this analysis suggests that the
@@
 What would the effect be were all students to request few hints?
-To answer that question, we estimate ``direct'' effects,
-$\Xi=\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
-Like indirect effects, this expression includes the average potential
-outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
-potential outcomes under the control condition.
-Since the main contrast of $\Xi$ is due to treatment assignment,
-construction of the estimator must account for the randomization
-scheme.
-This may be done with a regression estimator.
-One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
-subjects in the control group and $\hat{Y}(1,0)$ (of Eq. \ref{eq:estY10}) for subjects in the
-treatment group:
+To answer that question, we estimate average direct effects,
+$\EE[\xi(0)]=\EE[Y(1,H(0))-Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
+  is equivalent to the ``controlled direct effect,'' $CDE(0)$
+  e.g. \citet[][p. 57]{vanderweele2015explanation}.}
+In principal, this could be estimated as the difference in sample
+means of $\hat{Y}(1,0)$ in the treatment group and observed $Y=Y(0,0)$
+in the control group.
+In general, if we let
 \begin{equation*}
@@
 \end{equation*}
-Then, regress it on a dummy variable for treatment group and dummy
-variables for randomization block:
+then we can estimate $\EE[\xi(0)]$ as the ATE of treatment assignment
+$Z$ on $\tilde{Y}$.
+Just like estimates of the ATE of $Z$ on $Y$, this estimate should
+account for the experimental design: a paired, group-randomized trial
+with students clustered in schools and schools randomized within
+pairs.
+
+For the sake of simplicity, we will use an OLS regression estimator,
+with cluster-robust standard errors \citep{pustejovsky2018small},
+clustered at the school level.
+(Since in Section \ref{sec:observational} subjects were matched
+within schools, school-level cluster-robust standard errors also account for the
+dependence of $\hat{Y}(1,0)$ between matched treated students.)
+Specifically, we regress:
 \begin{equation}\label{eq:directEffReg}
-\tilde{Y}_i=\alpha_{b[i]}+\hat{\Xi}Z_i+\epsilon_i
+\tilde{Y}_i=\alpha_{p[i]}+\widehat{\EE[\xi(0)]}Z_i+\epsilon_i
 \end{equation}
-where $\alpha_{b[i]}$ is a fixed intercept for subject $i$'s randomization block, $\hat{\Xi}$ is an estimate of $\Xi$ and $\epsilon_i$ is a regression error.
-The clustering of subjects within schools can be accounted for in a
-number of ways, such as including random intercepts for school or
-computing cluster-robust standard error estimates.
-Following the latter route, and using the \texttt{clubSandwich}
-package in \texttt{R} \citep{clubsandwich}, we estimated a direct
+where $\alpha_{p[i]}$ is a fixed intercept for the randomization pair
+of subject $i$'s school and $\epsilon_i$ is a regression error.
+Using the \texttt{clubSandwich} package in \texttt{R}
+\citep{clubsandwich} to estimate standard errors, and 95\% confidence intervals, we estimated a direct
 effect of
@@
 cat(
-'Avg. Indirect Effect ($\\Delta$)&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
-Avg. Direct Effect ($\\Xi$)&',round(dir1.1t[1],2),'&',printci(dir1.1ci),'\\\\
-Avg. Direct Effect, covariate adjusted ($\\Xi$) &',round(dir2.1t[1],2),'&',printci(dir2.1ci),'\\\\\n')
+'Avg. Indirect Effect ($\\EE \\delta(1)$)&',round(mixProp*effects[['tot_eff']][['coefficients']],2),'&',printci(mixProp*with(effects[['tot_eff']],c(conf.low,conf.high))),'\\\\
+Avg. Direct Effect ($\\EE\\xi$)&',round(dir1.1t[1,1],2),'&',printci(dir1.1ci),'\\\\
+Avg. Direct Effect, covariate adjusted ($\\EE\\xi$) &',round(dir2.1t[1,1],2),'&',printci(dir2.1ci),'\\\\\n')
 @
@@
 \end{document}
+
+When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
+$Y_i(1,0)$ is estimated with the average of their observed outcomes.
+
+Now, for each member $i$ of the treatment group we have observed
+$Y_i(1,H(1))=Y_i$, and estimated or inferred
+$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
+Then, we can estimate $\EE \delta(1)$, the average indirect effect in the
+treatment group, by averaging over the treatment group:
+\begin{equation*}
+ \hat{\EE \delta(1)}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
+\end{equation*}
+
+This estimate is a close cousin of the

### 2020-06-05 paper.tex
lotta changes to mediation sec 4, #7

@@
 
-In principle, effects $\delta(1)$ and $\xi(0)$ are defined at the
-individual level, as $\delta_i(1)$ and $\xi_i(0)$.
-However, individual effects are
-not identified without strong assumptions, so the target of
-estimation is instead taken to be super-population average effects,
-$\Delta\equiv\EE[\delta(1)]$ and $\Xi\equiv\EE[\xi(0)]$.
-Our approach to estimating these
-is to estimate the means of each of the component potential outcomes,
-$\EE Y(1,H(1))$, $\EE Y(1,H(0))$, and $\EE Y(0,H(0))$.
-The first and last of these, $\EE Y(1,H(1))=\EE Y(Z=1)$ and $\EE
-Y(0,H(0))=\EE Y(Z=0)$ are non-parametrically identified due to the
-randomization of treatment assignment.
-Additionally, as described above, $\delta_i(1)=0$ as long as $Z_i=1$ and $H_i=0$.
-
-Treated subjects with $H_i=1$ present more of a challenge.
-To estimate indirect effects for these subjects, we may rely on the
-same causal setup as in the observational approach:
-matched ignorability and the matching design of Section
-\ref{sec:observational}.
-Each member of the treatment group with $H=1$ is matched to other
-treatment-group subjects with $H=0$.
-Matched ignorability states that within these matched sets, $Y(1,0)$
-is independent of $H$.
-Therefore, if subject $i$ is matched to subject $j$, with $H_i=1$, and
-$H_j=0$, we may estimate $Y_i(1,0)$ based on the observed
-$Y_j=Y_j(1,0)$.
-When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
-$Y_i(1,0)$ is estimated with the average of their observed outcomes.
-
-Specifically, let
+Average potential outcomes $\EE Y(1,H(1))$ and $\EE Y(0,H(0))$ are
+identified due to the randomization of $Z$, but estimating $\EE
+Y(1,H(0))$ requires imputing the
+potential outcome $Y(1,0)$ for treated subjects with $H_i=1$.
+However, it turns out that the matching estimators from
+\ref{sec:observational} already solved this problem:
+under matched ignorability, within matched sets, $Y(1,0)$
+is independent of $H$, implying that $\EE[Y(1,0)|M,H=1]=\EE[Y(1,0)|M,H=0]$.
+That suggests imputations
 \begin{equation}\label{eq:estY10}
 \hat{Y}_i(1,0)=\begin{cases}
- Y_i & H_i=0\\[3ex]
-\frac{\displaystyle\sum_{j:M_j=M_i}
-  Y_j(1-H_j)}{\displaystyle\sum_{j:M_j=M_i} 1-H_j} & H_i=1
+ Y_i & H_i=0\\
+\bar{Y}_{M=m_i,H=0} & H_i=1
 \end{cases}
 \end{equation}
-For treated subjects with $H=1$, the estimated $\hat{Y}(1,0)$ is the
-average outcome of their matched pairs with $H=0$.
-
-Now, for each member $i$ of the treatment group we have observed
-$Y_i(1,H(1))=Y_i$, and estimated or inferred
-$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
-Then, we can estimate $\Delta$, the average indirect effect in the
-treatment group, by averaging over the treatment group:
+That is, when $Z_i=1$ and $H_i=0$, no imputation is necessary, since
+$Y_i(1,0)=Y_i$, the observed outcome. When $Z_i=1$ and $H_i=1$,
+impute $Y(1,0)$ as the average of the observed outcomes from the $H=0$ subjects
+matched to $i$.
+Matched ignorability implies that the imputations will be unbiased.
+
+\subsection{Estimating Log Data Indirect Effects}\label{sec:indirectEst}
+
+Since $\EE \delta(1)$ involves only $Z=1$ potential outcomes, and
+$H(0)=0$ is known based on the study design, $\EE\delta(1)$ may be
+estimated using data from the treatment group only.
+
+Using the imputations from \eqref{eq:estY10}, we estimate
+$\EE\delta(1)$ as
 \begin{equation*}
- \hat{\Delta}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
+\widehat{\EE\delta(1)}=\widehat{\EE Y(Z=1)}-\widehat{\EE
+  Y(1,0)}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,0)
 \end{equation*}
-
-This estimate is a close cousin of the ``TOT'' estimate in Table
-\ref{tab:matchResults} from the previous section.
-Recall that the TOT effect is the average effect of $H$ on treatment
-group subjects with $H=1$.
-Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
-data from only the treatment group.
-TOT is the average difference between the realized outcomes for students
-with $Z=1$ and $H=1$, and their counterfactual outcomes were $Z=1$ and $H=0$.
-That is,
+where $n_T=\sum_i Z_i$, the number of subjects in the treatment
+group.
+Since $Y_i=\hat{Y}_i(1,0)$ whenever $H_i=0$, the summand is non-zero
+only when $H_i=1$.
+Further, note that when multiple $H=1$ subjects share a
+matched set, they also share imputations $\hat{Y}(1,0)$.
+Re-writing $\widehat{\EE\delta(1)}$ in terms of matched sets $M=m$,
+\begin{align}
+\widehat{\EE\delta(1)}&=\frac{1}{n_T}\displaystyle\sum_m\displaystyle\sum_{i:H_i=1,M_i=m}
+                        Y_i-\bar{Y}_{M=m_i,H=0} \nonumber \\
+&=\frac{1}{n_T}\displaystyle\sum_m
+  n_{1m}(\bar{Y}_{M=m_i,H=1}-\bar{Y}_{m=m_i,H=0}) \nonumber \\
+&=\frac{1}{n_T}\displaystyle\sum_mn_{1m}\hat{\tau}_m=\frac{\sum_iH_i}{n_T}\widehat{TOT} \label{eq:indTot1}
+\end{align}
+where $n_{1m}$ and $\hat{\tau}_m$ were defined in Section
+\ref{sec:obsEst} as the number of $H=1$ subjects in match $m$ and the
+difference in the sample mean of $Y$ between $H=1$ and $H=0$ subjects
+in match $m$, respectively, and $\hat{TOT}$ is the TOT estimate,
+\eqref{eq:tauWeightedMean} with $w_m=n_{1m}$.
+The upshot of \eqref{eq:indTot1} is that under matched ignorability,
+$\EE\delta(0)$ can be estimated as the TOT times the proportion of the
+treatment group with $H=1$.
+
+Actually, this relationship between $\EE\delta(0)$ and the TOT holds
+in the population, regardless of how the TOT is estimated.
+This is because when $H_i=0$, then $H_i(1)=H_i(0)=0$, so the individual direct effect
+$\delta_i(1)=0$, and when $H_i=1$,
+ $\delta_i(1)=Y_i(H=1)-Y_i(H=0)$, the effect of $H$ on $Y$.
+Therefore,
 \begin{equation*}
-TOT=\EE[Y(1,1)-Y(1,0)|H=1]
+\EE \delta(1)=\EE\EE[\delta(1)|H]=\EE[\delta(1)|H=1]Pr(H=1)=TOTPr(H=1)
 \end{equation*}
 
-$\hat{\Delta}$ estimates the difference between the average outcome
-for students in the entire treatment group, and the average outcome
-for the treatment group were $H=0$ for all students.
-That is
-\begin{equation*}
-\Delta=\EE[Y(1,H(1))-Y(1,0)]
-\end{equation*}
-This can be re-written:
-\begin{align*}
-\begin{split}
-\Delta&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
-&\qquad +\EE[Y(1,0)-Y(1,0)|H=0]Pr(H=0)
-\end{split}
-\\[2ex]
-&=\EE[Y(1,1)-Y(1,0)|H=1]Pr(H=1)\\
-&=Pr(H=1)TOT
-\end{align*}
-Since, by definition, $Pr(H=1)=$0.3, we have that
-$\hat{\Delta}=$0.3$TOT$.
-Estimated thusly, the indirect effect is negative: -0.04, with
+In the hint study, the definition of $H$ implies that
+$Pr(H=1)=\sum_iH_i/n_T=$0.3.
+Therefore, our estimate of the average indirect effect is
+$\widehat{\EE \delta(1)}=$0.3$\widehat{TOT}$.
+In Table \ref{tab:matchResults}, we estimated the TOT as
+-0.14 with a
+standard error of 0.06.
+Thefore, we estimate an average indirect effect of -0.04, with
 a standard error of 0.02.
-Like the $TOT$ estimate on which it is based, this estimate of indirect effects assumes matched ignorabilty---no confounding between $H$ and $Y$ within matched sets.
+
+It is worth repeating that these estimates assume matched ignorabilty---no confounding between
+$H$ and $Y$ within matched sets.
 With this important caveat, this analysis suggests that the
@@
 What would the effect be were all students to request few hints?
-To answer that question, we estimate ``direct'' effects,
-$\Xi=\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
-Like indirect effects, this expression includes the average potential
-outcome $\EE[Y(1,H(0))]$; unlike indirect effects, it includes a term for
-potential outcomes under the control condition.
-Since the main contrast of $\Xi$ is due to treatment assignment,
-construction of the estimator must account for the randomization
-scheme.
-This may be done with a regression estimator.
-One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
-subjects in the control group and $\hat{Y}(1,0)$ (of Eq. \ref{eq:estY10}) for subjects in the
-treatment group:
+To answer that question, we estimate average direct effects,
+$\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
+  is equivalent to the ``controlled direct effect,'' $CDE(0)$
+  e.g. \citet[][p. 57]{vanderweele2015explanation}.}
+In principal, this could be estimated as the difference in sample
+means of $\hat{Y}(1,0)$ in the treatment group and observed $Y=Y(0,0)$
+in the control group.
+In general, if we let
 \begin{equation*}
@@
 \end{equation*}
-Then, regress it on a dummy variable for treatment group and dummy
-variables for randomization block:
+then we can estimate $\EE[\xi(0)]$ as the ATE of treatment assignment
+$Z$ on $\tilde{Y}$.
+Just like estimates of the ATE of $Z$ on $Y$, this estimate should
+account for the experimental design: a paired, group-randomized trial
+with students clustered in schools and schools randomized within
+pairs.
+
+For the sake of simplicity, we will use an OLS regression estimator,
+with cluster-robust standard errors \citep{pustejovsky2018small},
+clustered at the school level.
+(Since in Section \ref{sec:observational} subjects were matched
+within schools, school-level cluster-robust standard errors also account for the
+dependence of $\hat{Y}(1,0)$ between matched treated students.)
+Specifically:
 \begin{equation}\label{eq:directEffReg}
-\tilde{Y}_i=\alpha_{b[i]}+\hat{\Xi}Z_i+\epsilon_i
+\tilde{Y}_i=\alpha_{p[i]}+\hat{\EE\xi}Z_i+\epsilon_i
 \end{equation}
-where $\alpha_{b[i]}$ is a fixed intercept for subject $i$'s randomization block, $\hat{\Xi}$ is an estimate of $\Xi$ and $\epsilon_i$ is a regression error.
-The clustering of subjects within schools can be accounted for in a
-number of ways, such as including random intercepts for school or
-computing cluster-robust standard error estimates.
-Following the latter route, and using the \texttt{clubSandwich}
-package in \texttt{R} \citep{clubsandwich}, we estimated a direct
+where $\alpha_{p[i]}$ is a fixed intercept for the randomization pair
+of subject $i$'s school, $\hat{\EE\xi}$ is an estimate of $\EE\xi$ and $\epsilon_i$ is a regression error.
+Using the \texttt{clubSandwich} package in \texttt{R}
+\citep{clubsandwich} to estimate standard errors, and 95\% confidence intervals, we estimated a direct
 effect of
-0.14$\pm$0.23.
+0.14$\pm$0.25.
 Adding covariates to the regression \eqref{eq:directEffReg} estimates a direct effect of
-0.18$\pm$0.19.
+0.18$\pm$0.21.
 
@@
 \hline
-Avg. Indirect Effect ($\Delta$)& -0.04 & [-0.07,-0.01] \\
-Avg. Direct Effect ($\Xi$)& 0.14 & [-0.08,0.37] \\
-Avg. Direct Effect, covariate adjusted ($\Xi$) & 0.18 & [-0.01,0.38] \\
+Avg. Indirect Effect ($\EE \delta(1)$)& -0.04 & [-0.07,-0.01] \\
+Avg. Direct Effect ($\EE\xi$)& 0.14 & [-0.10,0.39] \\
+Avg. Direct Effect, covariate adjusted ($\EE\xi$) & 0.18 & [-0.03,0.39] \\
 

### 2020-06-09 paper.Rnw
put eq. 9 into eq. 8 #7

@@
 procedure.
-After a formal introduction in Section \ref{sec:psIntro},
-\S~\ref{sec:psModel} describes a set of sub-models for student hint
+After a formal introduction in Subsection \ref{sec:psIntro},
+Subsection \ref{sec:psModel} describes a set of sub-models for student hint
 requests, posttest scores, and treatment effects.
@@
 $\eta(1)$ is defined for all the students in the study.
-Second, a modeling difference: since we did not dichotomize $\eta(1)$
-in this analysis, we modeled it as normal , not as a normal mixture, as in
-\eqref{eq:mixture}.
+Second, a modeling difference: instead of the normal mixture model
+\eqref{eq:mixture}, we modeled $\eta(1)$ as normal.
 With this change, the PS estimand became
@@
  Y|\left(Z_i,\bm{x}_i,\bm{\theta},\eta(1)\right) \sim  \mathcal{N}\left(
-\gamma_{0b[i]}+\bm{x}_i^T\bm{\gamma}+a\eta(1)+Z_i\tau(\eta(1))+\zeta^{tch}_{t[i]}+\zeta^{scl}_{s[i]},\omega_{Z[i]}\right)
+\gamma_{0b[i]}+\bm{x}_i^T\bm{\gamma}+a\eta(1)+Z_i(b_0+b_1\eta(1))+\zeta^{tch}_{t[i]}+\zeta^{scl}_{s[i]},\omega_{Z[i]}\right)
 \end{equation}
@@
 
-Finally, we modeled treatment effects
-$\tau(\eta(1))$ as linear:
-\begin{equation}\label{eq:tauModel}
-\EE[Y(1)-Y(0)|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)
-\end{equation}
+Model \eqref{eq:outcomeSubmodel} implies that treatment effects are
+linear in $\eta(1)$,
+$\EE[Y(1)-Y(0)|\eta(1)]=\tau({\eta(1)})=b_0+b_1\eta(1)$
+
 While more complex models for $\tau(\eta(1))$ are theoretically
@@
 model), in our experience non-linear effect models do not perform as
-well in model checks as the the linear model (\ref{eq:tauModel}).
+well in model checks as the the linear model.
 

### 2020-06-09 R/dataAnalysis.r
Luke's figure suggestions #7

@@ -23,11 +23,11 @@ cat('Warning: runFull=',runFull,'\n')
 cat('----------------------------------------------------\n\n')
 
 source('R/helperFunctions.r')
-source('~/Box Sync/rcode/matchingFunctions.r')
-source('~/Box Sync/rcode/hhh.r')
-source('~/Box Sync/rcode/printXbal.r')
+source('~/Box/rcode/matchingFunctions.r')
+source('~/Box/rcode/hhh.r')
+source('~/Box/rcode/printXbal.r')
 
-load('~/Box Sync/CT/data/RANDstudyData/HSdata.RData') ## for "error"
+load('~/Box/CT/data/RANDstudyData/HSdata.RData') ## for "error"
 if(runFull) source('R/prelimStan.r') else load('data/hintsData.RData')
 
 ## for printing nice CIs
@@ -281,32 +281,32 @@ cat('
 #########################
 ',as.character(Sys.time()),'\n')
 
-pdMain <- pdMod(psmod1)
-## pdMain <- within(pdMain,
-## {
-##     b0 <- b0/pooledSD
-##     b1 <- b1/pooledSD*iqr
-##     xmin <- xmin/mean(iqr)
-##     xmax <- xmax/mean(iqr)
-##     ymin <- ymin/pooledSD
-##     ymax <- ymax/pooledSD
-## }
-## )
+pooledSD <- with(sdat, sqrt(((sum(Z)-1)*var(Y[Z==1])+(sum(1-Z)-1)*var(Y[Z==0]))/(nstud-2)))
+samp <- seq(1,length(draws$b1),length=1000)
+Usamp <- draws$studEff[samp,]
+studEff95 <- quantile(Usamp,c(0.025,0.975))
 
-#tikzDevice::tikz('mainEffects.tex', standAlone=T,
-#     width=6,height=5)
-p1 <- ggplot(pdMain)+
-    geom_abline(aes(intercept=b0,slope=b1,group=id),color='red')+
-    coord_cartesian(xlim=c(min(pdMain$xmin),max(pdMain$xmax)),
-                    ylim=c(min(pdMain$ymin),max(pdMain$ymax)),expand=FALSE)+
-    geom_line(aes(x=x,y=y,group=truthOrAvg,linetype=truthOrAvg,color=truthOrAvg,alpha=truthOrAvg),size=1.5)+
-    xlab('$\\eta(1)$')+ylab('CTA1 Treatment Effect')+
-    labs(group=NULL,color=NULL,linetype=NULL)+
[…]

### 2020-06-09 paper.Rnw
last edits #7 summarizing discussion +diagnostic plots

@@
 elicits a hint request.
-That paramter is important in our context because the
+That parameter is important in our context because the
 specific problems a student works on may influence his or her hint
@@
 In the Cognitive Tutor, problems often have multiple parts, and each part offers its own hints.
-Researchers in posession of data on each part of each worked problem could specify a model at the part-level, instead of at the problem-level like \eqref{eq:rasch}.
+Researchers in possession of data on each part of each worked problem could specify a model at the part-level, instead of at the problem-level like \eqref{eq:rasch}.
 Perhaps an even better option (suggested by an anonymous reviewer) would incorporate data on the specific skills required to solve each problem.
@@
 However, outside of a planned experiment, posttest scores may not be available.
-Our usage variable $H$ is binary, which is necessary for classical propensity score matching; alternatve methods are available for usage variable that are continuous \citep[e.g.][]{hirano2004propensity}, ordered \citep[e.g.][]{leon2005mixed}, categorical \citep[e.g.][]{lopez2017estimation} or other types.
+Our usage variable $H$ is binary, which is necessary for classical propensity score matching; alternative methods are available for usage variable that are continuous \citep[e.g.][]{hirano2004propensity}, ordered \citep[e.g.][]{leon2005mixed}, categorical \citep[e.g.][]{lopez2017estimation} or other types.
 Like all observational studies, it depends on the strong, untestable assumption of no unmeasured confounding.
@@
 In the second step, we estimate the average effect of $H$ on posttest scores, by comparing posttest scores between matched students with $H=1$ and $H=0$.
-In the third and final step, we account for the possiblity of unmeasured confounding in a sensitivity analysis.
-
-The next subsection will give more formal background on propensity score matching, including notation and idenfication assumptions.
+In the third and final step, we account for the possibility of unmeasured confounding in a sensitivity analysis.
+
+The next subsection will give more formal background on propensity score matching, including notation and identification assumptions.
 Following subsections will illustrate each of the three steps of propensity score matching.
@@
 
-It is worth repeating that these estimates assume matched ignorabilty---no confounding between
+It is worth repeating that these estimates assume matched ignorability---no confounding between
 $H$ and $Y$ within matched sets.
@@
 
-Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availablity of hints (an indirect effect of \Sexpr{round(mixProp*effects[['tot_eff']][['coefficients']],2)} standard deviations), and the effect due to CTA1's other mechanisms is \Sexpr{round(dir2.1t[1],2)}.
+Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availability of hints (an indirect effect of \Sexpr{round(mixProp*effects[['tot_eff']][['coefficients']],2)} standard deviations), and the effect due to CTA1's other mechanisms is \Sexpr{round(dir2.1t[1],2)}.
 
@@
 PS analysis requires data on outcomes
-(e.g. posttest scores) in both tretment arms, as well as data on
+(e.g. posttest scores) in both treatment arms, as well as data on
 implementation.
@@
 discrete strata.
-Other approachs \citep[e.g.]{gilbertHudgens,jin2008principal} allow
+Other approaches \citep[e.g.]{gilbertHudgens,jin2008principal} allow
 the intermediate variable to be continuous.
@@
 $\hat{\tau}(\cdot)$ function were positive, implying a posterior
-probability of 0.88 that students who asked for hints
+probability of \Sexpr{sprintf('%1.2f',probPos)}
+that students who asked for hints
 with greater regularity benefited more from the CTA1 curriculum.
@@
 
-These results suggest that students who, if assigned to the treatment
-condition, would be more likely to request hints on challenging
-problems, tend to experience higher treatment effects.
-That is, it suggests that frequent hint requesters benefit more from
-assignment to the CTA1 condition than infrequent hint requesters.
+These results suggest that students who, under treatment, would tend
+to ask for hints on challenging problems, are most responsive to
+treatment.
 
 \section{Comparing Strategies}\label{sec:synthesis}
-The goals, assumptions, and processes of matching, mediation analysis,
-and principal stratification, when applied to implementation data from
-an EdTech RCT, each differ from each other, but also overlap in
-interesting ways.
+% The goals, assumptions, and processes of matching, mediation analysis,
+% and principal stratification, when applied to implementation data from
+% an EdTech RCT, each differ from each other, but also overlap in
+% interesting ways.
 In the case discussed here, the observational study and mediation
 approaches appear to give the opposite conclusion from principal
-stratification---matching and mediation suggest that exessive hint
+stratification---matching and mediation suggest that excessive hint
 requests hurt students' posttest scores, while principal
 stratification suggests that students who request more hints
-experiance higher treatment effects.
+experience higher treatment effects.
 This section will compare and contrast the three methods in terms of
@@
 are more likely to request hints are also more likely to struggle with
-the material, and hence score lower on the posttest.
+the material.
 We addressed this concern in three ways: first, we focused our attention
@@
 
-Mediation analysis builds on the matching design, and also requires an
+Mediation analysis builds on the matching design, and requires the same
 ignorability assumption; it must
@@
 and in the treatment group.
-Our approach of dichotomizing hint request behavior as $H$ provides a
-way forward, at the cost of ignoring information that could be gleaned
-from more continuous measures of hint requests.
+Under our approach of dichotomizing hint request behavior, mediation
+analysis is essentially a way to interpret the results of the
+observational study in the context of the full RCT.
 
@@
 match, and diagnose potential problems with resulting estimates.
-Though PS does not depend on ignorability, it does depend on a highly
+Though principal stratification does not depend on ignorability, it does depend on a highly
 intricate and parametric modeling structure.
@@
 models can give severely biased estimates.
-Extensive model interrogation and checking is necessary for PS
-analysis, but the complexity of the PS model makes this process
+Extensive model interrogation and checking is necessary for principal stratification
+analysis, but the complexity of the principal stratification model makes this process
 particularly difficult.
@@
 between hint requesting and posttest scores.
-Students who request more hints need more help, and thus score lower
-on the posttest.
-Fortunately, the PS analysis also suggests that these students
-receive more help----that is, they benefit more from CTA1---than do their peers who request fewer hints.
-However, it is unclear whether their larger treatment effects are due to their higher rate of hint requests, or to some other correlated factor.
+Those students who request more hints are those who need more help of
+some form, and tend to score lower on the posttest.
+Fortunately, the principal stratification analysis also suggests that these students
+may benefit more from CTA1 than their peers who request fewer hints.
+However, it is unclear whether their larger treatment effects are due
+to their higher rate of hint requests, or to some other related characteristic.
 
@@
 Does requesting hints help or hurt?
-Technically, PS has nothing to say about the effect of hints---$\eta(1)$
+Technically, principal stratification has nothing to say about the effect of hints---$\eta(1)$
 is modeled as a student characteristic, not an intervention.
@@
 That is, it may be that the observational and mediational results are
-correct---that requesting hints lowers test scores---but that the PS
+correct---that requesting hints lowers test scores---but that the principal stratification
 results are also correct---students who request more hints tend to
 experience higher treatment effects, due to other mechanisms.
+(Finally, we should note that the principal stratification model was
+uncertain about the sign of $b_1$---there is a decent (though not large) probability that the
+the slope between $\eta(1)$ and treatment effects is negative.)
 
@@
 In their most straightforward interpretations, principal stratification
-is a method for examining treatment effect heterogeneity, whereas
-mediation analysis is a method for assessing causal mechanisms.
-That said, PS may also shed light on potential causal mechanisms---if,
+is a method for examining treatment effect heterogeneity (are students
+benefiting differently?), whereas
+mediation analysis is a method for assessing causal mechanisms (is the
+wide availability of hints in CTA1 beneficial to students?).
+That said, principal stratification may also shed light on potential causal mechanisms---if,
 indeed, treatment effects are higher for students who request more
@@
 Conversely, researchers unwilling to consider untestable ignorability
-assumptions will find PS an attractive alternative.
+assumptions will find principal stratification an attractive alternative.
 
@@
 models for a dataset, and perhaps devise tests of model fit tailored
-to a specific problem, may be more confident in the fit of a PS model.
+to a specific problem, may be more confident in the fit of a principal stratification model.
 Researchers who prefer transparency and non-parametric or
@@
 Additional philosophical concerns come in to play as well.
-For instance, unlike observational studies and PS, mediation analysis
+For instance, unlike observational studies and principal stratification, mediation analysis
 depends on strictly counterfactual quantities, such as $\EE Y(0,H(1))$,
 that can never occur.
-Analogous quantities in PS, such as $\EE [Y(0)|H(1)]$ are
+Analogous quantities in principal stratification, such as $\EE [Y(0)|H(1)]$ are
 unobservable, but nevertheless refer to averages of potential outcomes
@@
 are available, such as the propensity score
-matching illustrated here, optionally also including covariates in the
-outcomes model for additional protection against bias.
+matching illustrated here.
 The result---the ATE of implementation on the outcome---is an
@@
 
-Mediation analysis may have stronger standing where
-at least one well-represented level of the implementation
-variable is interpretable in the control condition.
-In our example, $H=0$ included both low hint usage in
-the treatment group and no access to hints in the control group.
+% Mediation analysis may have stronger standing where
+% at least one well-represented level of the implementation
+% variable is interpretable in the control condition.
+% In our example, $H=0$ included both low hint usage in
+% the treatment group and no access to hints in the control group.
 When employed, mediation analysis can contextualize the estimated
@@
 statistical modeling (and model checking) is not a practical barrier,
-PS may be a better choice because no ignorability assumptions are necessary.
+principal stratification may be a better choice because no
+ignorability assumptions are necessary.
+
+Of course, there are approaches to modeling implementation beyond
+the three considered here. For instance, principal stratification
+based on non-parametric bounding \citep{bounding} or randomization
+inference \citep{nolen} would avoid the parametric assumptions of
+model-based principal stratification, as would a standard moderation
+analysis based on predicted implementation, $\hat{\eta}(1)$ instead of
+partially observed $\eta(1)$.
+Future work (by us, and, we hope, others) may develop these methods,
+and still others.
+
 % \begin{table}
@@
 Second, our decision to dichotomize hint request rates in the observational study and mediation analysis allowed us to use of a traditional matching estimator and to identify mediational estimands.
-Third, our decision to use a Rasch model to measure hint usage in the PS analysis accounted for differences between students in both the number and difficulty of worked CT problems.
+Third, our decision to use a Rasch model to measure hint usage in the principal stratification analysis accounted for differences between students in both the number and difficulty of worked CT problems.
 
@@
 %\clearpage
-\section{Stan Code for PS Model}
+\section{Stan Code for Principal Stratification Model}
 
@@
 \subsection{Posterior Predictive Density Plots}
+Figures \ref{fig:ppdPooled}, \ref{fig:ppdTrt}, and \ref{fig:ppdCtl}
+compare the density of observed outcomes $y$ to 1000 random draws of
+the model-implied density, $y_{rep}$ for the full (``pooled'') sample,
+the treatment group, and the control group, respectively.
+They show a close, but not perfect, correspondence between the
+observed data and the model.
+The most noticeable difference is the left-truncation of the observed
+outcomes---the lower tail is lighter than would be expected from a
+normal distribution---suggesting floor effects on the posttest not
+considered in the model.
+The misspecification appears similar in both treatment groups.
+A model accounting for floor effects may fit the data more closely.
+These plots were made with the \texttt{bayesplot} package
+\citep{bayesplot} in \texttt{R}.
+
 \begin{figure}[!h]
@@
 \includegraphics[width=0.5\textwidth]{output/ppcYoverall.jpg}
+\caption{A comparison of the estimated density of observed outcomes
+  $y$ with 1000 random draws of the model-implied outcome density,
+  for the full dataset.}
+\label{fig:ppdPooled}
 \end{figure}
@@
 \includegraphics[width=0.5\textwidth]{output/ppcYtrt.jpg}
+\caption{A comparison of the estimated density of observed outcomes
+  $y$ with 1000 random draws of the model-implied outcome density,
+  for the treatment group.}
+\label{fig:ppdTrt}
 \end{figure}
@@
 \includegraphics[width=0.5\textwidth]{output/ppcYctl.jpg}
+\caption{A comparison of the estimated density of observed outcomes
+  $y$ with 1000 random draws of the model-implied outcome density,
+  for the control group.}
+\label{fig:ppdCtl}
 \end{figure}
@@
 \subsection{Residual Plots}
+Figures \ref{fig:residFull}, \ref{fig:residTrt}, and
+\ref{fig:residCtl} plot residuals $Y_i-\hat{Y}_i$ against fitted
+values $\hat{Y}_i$ for the full sample, the treatment group, and the
+control group, respectively.
+Here, fitted values $\hat{Y}$ are posterior means---averages over all
+of the draws from the model.
+The residual plots reflect the floor effects apparent in the posterior
+predictive density plots---a subset of points appears to fall along a
+straight line in the bottom left of the plot.
+Otherwise, little structure is apparent.
+
+Figure \ref{fig:binnedResids} assesses the hint sub-model,
+\eqref{eq:rasch} and \eqref{eq:rasch2} by comparing the binned model-implied
+probabilities of hint requests on individual worked problems to residuals---the
+differences between these probabilities and the proportions of worked
+problems within each bin that resulted in an
+actual hint request \citep[c.f.][]{gelman2014bayesian}.
+Nine random draws from the posterior distribution are shown.
+Little structure is apparent in the binned residuals, suggesting that
+the model for hint requests fits well.
+
 \begin{figure}[!h]
 \centering
-\includegraphics[width=0.5\textwidth]{output/residPlotOverall.jpg}
+\includegraphics[width=\textwidth]{output/residPlotOverall.jpg}
+\caption{A residual plot for the full sample.}
+\label{fig:residFull}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{output/residPlotTrt.jpg}
+\includegraphics[width=\textwidth]{output/residPlotTrt.jpg}
+\caption{A residual plot for the treatment group.}
+\label{fig:residTrt}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{output/residPlotCtl.jpg}
+\includegraphics[width=\textwidth]{output/residPlotCtl.jpg}
+\caption{A residual plot for the control group}
+\label{fig:residCtl}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{output/binnedplot.pdf}
+\includegraphics[width=\textwidth]{output/binnedplot.pdf}
 \caption{Binned residual plot for model of hint requests.}
+\label{fig:binnedResids}
 \end{figure}
+
+\subsection{Fake Data Simulation}
+Figure \ref{fig:fakeSim} implements a model assessment for principal
+stratification suggested in \citet{aoas}.
+In this method, we replace data from the real control group with duplicated
+outcome and covariate data from the treatment group (now re-labeled as
+control).
+In other words, each member of the treatment group appears in the
+dataset twice, once in the ``treatment'' group and once in the
+``control'' group.
+Only the ``treatment'' observations are linked with hint data.
+We also re-labeled teachers and schools in the ``control'' group so
+that the were distinct from the corresponding labels in the
+``treatment'' group.
+
+The upper left plot of \ref{fig:fakeSim} fits the same principal
+stratification model described in Section \ref{sec:psModel} to the new
+dataset, in which there is exactly no effect.
+In the other three panels of \ref{fig:fakeSim}, the model is fit to
+the fake data with simulated treatment effects added: a random
+treatment effect uncorrelated with $\eta(1)$, a treatment effect
+linear in $\eta(1)$ (as we estimated) and a treatment effect quadratic
+in $\eta(1)$.
+These fake datasets contain the same quirks as the real treatment
+group data (and hence the same model misspecification) but with known
+treatment effects.
+In the first three plots, our model recovered the truth, albeit with
+considerable uncertainty.
+In the bottom left (quadratic) plot, our model could not recover the
+true relationship between $\eta(1)$ and $\tau$ because we modeled
+$\tau$ as linear in $\eta(1)$. However, the model did not estimate
+a spurious positive or negative associations between $\eta(1)$ and $\tau$.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{output/qqPlots.pdf}
-\caption{Q-Q plots for outcomes and fitted values}
+\includegraphics[width=\textwidth]{output/fakePlots.pdf}
+\caption{Comparisons between principal stratification model estimates
+  and true treatment effects when the treatment group is replicated
+  (without corresponding hint request data) and relabeled as control,
+  and (clockwise from upper left) no treatment effect, a
+  random treatment effect uncorrelated with $\eta(1)$, a treatment
+  effect linear in $\eta(1)$, or a treatment effect quadratic in
+  $\eta(1)$ is added to the original treatment group.}
+\label{fig:fakeSim}
 \end{figure}
@@
 \end{document}
-
-When $i$, with $H_i=1$, is matched to several subjects with $H=0$,
-$Y_i(1,0)$ is estimated with the average of their observed outcomes.
-
-Now, for each member $i$ of the treatment group we have observed
-$Y_i(1,H(1))=Y_i$, and estimated or inferred
-$\hat{Y}_i(1,H(0))=\hat{Y}_i(1,0)$ from \eqref{eq:estY10}.
-Then, we can estimate $\EE \delta(1)$, the average indirect effect in the
-treatment group, by averaging over the treatment group:
-\begin{equation*}
- \hat{\EE \delta(1)}\equiv\widehat{\EE[\delta(0)]}=\frac{1}{n_T}\displaystyle\sum_{i:Z_i=1} Y_i-\hat{Y}_i(1,H(0))
-\end{equation*}
-
-This estimate is a close cousin of the

### 2020-06-09 paper.tex
last edits #7 summarizing discussion +diagnostic plots

@@
 difference in the sample mean of $Y$ between $H=1$ and $H=0$ subjects
-in match $m$, respectively, and $\hat{TOT}$ is the TOT estimate,
+in match $m$, respectively, and $\widehat{TOT}$ is the TOT estimate,
 \eqref{eq:tauWeightedMean} with $w_m=n_{1m}$.
@@
 standard error of 0.06.
-Thefore, we estimate an average indirect effect of -0.04, with
+That implies an average indirect effect of -0.04, with
 a standard error of 0.02.
 
-It is worth repeating that these estimates assume matched ignorabilty---no confounding between
+It is worth repeating that these estimates assume matched ignorability---no confounding between
 $H$ and $Y$ within matched sets.
@@
 To answer that question, we estimate average direct effects,
-$\EE[\xi(0)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
+$\EE[\xi(0)]=\EE[Y(1,H(0))-Y(0,H(0))]=\EE[Y(1,0)-Y(0,0)]$.\footnote{This
   is equivalent to the ``controlled direct effect,'' $CDE(0)$
@@
 dependence of $\hat{Y}(1,0)$ between matched treated students.)
-Specifically:
+Specifically, we regress:
 \begin{equation}\label{eq:directEffReg}
-\tilde{Y}_i=\alpha_{p[i]}+\hat{\EE\xi}Z_i+\epsilon_i
+\tilde{Y}_i=\alpha_{p[i]}+\widehat{\EE[\xi(0)]}Z_i+\epsilon_i
 \end{equation}
 where $\alpha_{p[i]}$ is a fixed intercept for the randomization pair
-of subject $i$'s school, $\hat{\EE\xi}$ is an estimate of $\EE\xi$ and $\epsilon_i$ is a regression error.
+of subject $i$'s school and $\epsilon_i$ is a regression error.
 Using the \texttt{clubSandwich} package in \texttt{R}
@@
 
-Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availablity of hints (an indirect effect of -0.04 standard deviations), and the effect due to CTA1's other mechanisms is 0.18.
+Table \ref{tab:mediation} summarizes our mediation results: assuming matched ignorability, assignment to CTA1 increases student test scores despite the availability of hints (an indirect effect of -0.04 standard deviations), and the effect due to CTA1's other mechanisms is 0.18.
 
@@
 $\hat{\tau}(\cdot)$ function were positive, implying a posterior
-probability of 0.88 that students who asked for hints
+probability of 0.75
+that students who asked for hints
 with greater regularity benefited more from the CTA1 curriculum.

### 2020-06-09 R/diagnosticPlots.r

@@ -38,8 +38,8 @@ qplot(fitd[sdat$Z==0],sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],xlab='Fitted Values',yla
 ggsave('output/residPlotCtl.jpg')
 
 
-pdf('output/qqPlots.pdf', height=3,width=6)
-par(mfrow=c(1,2))
+pdf('output/qqPlots.pdf', height=8,width=5)
+par(mfrow=c(2,1))
 qqnorm(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1],main='Treatment Group')
 qqline(sdat$Y[sdat$Z==1]-fitd[sdat$Z==1])
 qqnorm(sdat$Y[sdat$Z==0]-fitd[sdat$Z==0],main='Control Group')

### 2020-06-15 paper.Rnw
correct posterior prob b1>0, ci

@@
 library(english)
-
+cap <- function(x){
+  x <- as.character(x)
+  paste0(toupper(substr(x,1,1)),substr(x,2,nchar(x)))
+}
 @
@@
 distribution, showing estimation uncertainty.
-Eighty-eight percent of posterior draws of the slope of the
+\Sexpr{cap(english(round(probPos*100)))} percent of posterior draws of the slope of the
 $\hat{\tau}(\cdot)$ function were positive, implying a posterior
@@
 with greater regularity benefited more from the CTA1 curriculum.
+A 95\% highest-density credible interval for the slope parameter is
+\Sexpr{printci(b1ci)}.
+Therefore, the data are consistent with either a slightly negative or
+a positive relationship between hint requests and treatment
+effects, but the latter is more likely.
+
 The right hand panel plots one posterior draw of $\eta(1)$
@@
 requests hurt students' posttest scores, while principal
-stratification suggests that students who request more hints
+stratification suggests that students who request more hints may
 experience higher treatment effects.
@@
 
-Like in the observational study and in mediation analysis, the
-principal stratification analysis found a negative correlation
+The
+principal stratification analysis also found a negative correlation
 between hint requesting and posttest scores.
@@
 some form, and tend to score lower on the posttest.
-Fortunately, the principal stratification analysis also suggests that these students
+Fortunately, the principal stratification analysis also suggests (with
+probability \Sexpr{sprintf('%1.2f',probPos)}) that these students
 may benefit more from CTA1 than their peers who request fewer hints.
@@
 experience higher treatment effects, due to other mechanisms.
-(Finally, we should note that the principal stratification model was
-uncertain about the sign of $b_1$---there is a decent (though not large) probability that the
-the slope between $\eta(1)$ and treatment effects is negative.)
+(Of course, the data are also consistent with the possibilities that
+the relationship between hint requests and treatment effects is
+negative, or close to zero, in which case there is less that needs to
+be explained.)
 

### 2020-06-15 paper.tex
correct posterior prob b1>0, ci

@@
 distribution, showing estimation uncertainty.
-Eighty-eight percent of posterior draws of the slope of the
+Seventy-five percent of posterior draws of the slope of the
 $\hat{\tau}(\cdot)$ function were positive, implying a posterior
@@
 with greater regularity benefited more from the CTA1 curriculum.
+A 95\% highest-density credible interval for the slope parameter is
+[-0.06,0.12].
+Therefore, the data are consistent with either a slightly negative or
+a positive relationship between hint requests and treatment
+effects, but the latter is more likely.
+
 The right hand panel plots one posterior draw of $\eta(1)$
@@
 some form, and tend to score lower on the posttest.
-Fortunately, the principal stratification analysis also suggests that these students
+Fortunately, the principal stratification analysis also suggests (with
+probability 0.75) that these students
 may benefit more from CTA1 than their peers who request fewer hints.

### 2020-06-15 R/dataAnalysis.r

@@ -10,6 +10,8 @@ library(tidyverse)
 library(rstan)
 library(lme4)
 library(estimatr)
+library(HDInterval)
+
 traceplot <- rstan::traceplot
 select <- dplyr::select
 options(stringsAsFactors=FALSE)
@@ -38,6 +40,8 @@ map_dfr <- function(.x, .f, ...){
   out <- map(.x,.f,...)
   as.data.frame(do.call('rbind',out))
 }
+
+capitalize <- function(x) paste0(toupper(substr(x,1,1)),substr(x,2,nchar(x)))
 
 citopm <- function(ci) paste(sprintf('%.2f',c(median(ci),(ci[2]-ci[1])/2)),collapse='$\\pm$')
 
@@ -252,8 +256,6 @@ dir2.1ci <- c(dir2.1ci$CI_L[1],dir2.1ci$CI_U[1])
 
 
 
-runFull <- FALSE
-
 cat('
 ############################
 ########## principal stratification
@@ -274,6 +276,9 @@ if(runFull){
 
 cat('extracting ps model',as.character(Sys.time()),'\n')
 draws <- rstan::extract(psmod1)
+
+probPos <- mean(draws$b1>0)
+b1ci <- hdi(draws$b1)
 
 cat('
 ############################

### 2020-06-19 paper.Rnw
final version (famous last words)

@@
 implementation.
-An ongoing research project is using this dataset to
-study instructional material within ALEKS: when students begin working on a particular topic in ALEKS, the
+ %Researchers should not stop at
+%estimating overall average treatment effects, but
+% Our hope is that this paper can serve as a starting point
+% for researchers interested in mining the rich resource of log data to
+% answer these types of questions.
+
+Analysis of log data from EdTech RCTs could, in principle, lead to
+insights on the relationship between implementation and effectiveness.
+Using log data, researchers could investigate how the product was
+actually implemented, as well as what aspects of implementation
+correlate with higher (or lower) treatment effects, or which features
+of the program drive its effects.
+For instance, \citet{sales2016student,aoas,reassignmentEffect}
+investigated aspects of mastery learning with CTA1.
+An ongoing research project uses the ALEKS log data to
+study its instructional material: when students begin working on a particular topic in ALEKS, the
 system first presents instructional material, followed by problem
@@
 strategy for students in terms of learning algebra, or should ALEKS be
-redesigned to prevent such skipping? %Researchers should not stop at
-%estimating overall average treatment effects, but
-% Our hope is that this paper can serve as a starting point
-% for researchers interested in mining the rich resource of log data to
-% answer these types of questions.
-
-Analysis of log data from EdTech RCTs could, in principle, lead to
-insights on the relationship between implementation and effectiveness.
-Using log data, researchers could investigate how the product was
-actually implemented, as well as what aspects of implementation
-correlate with higher (or lower) treatment effects, or which features
-of the program drive its effects.
-
-However, answering these questions requires new statistical tools.
-These tools must simultaneously surmount three challenges: first,
+redesigned to prevent such skipping?
+
+Answering questions like these calls for new statistical tools, which must simultaneously surmount three challenges: first,
 implementation is not typically randomized, so the relationship
@@
 
-This paper will illustrate three approaches to this type of research
-using log data from the CTA1 study---specifically, data on students' use
-of hints within the software.
-We will attempt to answer two related questions: first, what role does
-the availability of hints play in driving the overall CTA1 treatment
-effect, and second, can requesting hints more often lead to a larger
-treatment effect.
+This paper will use log data from the CTA1 study---specifically, data on students' use
+of hints within the software---to illustrate three approaches to these
+challenges.
+We will attempt to answer two related questions: first, what is the role of
+the availability of hints in driving the overall CTA1 treatment
+effect? Second, can requesting hints more often lead to a larger
+treatment effect?
 The first approach we consider discards the control group, and analyzes
@@
 Finally, we conduct a principal stratification analysis
-\citep{frangakis} to learn if students requesting different amounts of
-hints experienced different treatment effects.
+\citep{frangakis} to learn if students requesting hints at different
+rates experienced different treatment effects.
 
@@
 hints.
-The next three sections will present an observational study,
-which discards the control group altogether and estimates the effect
-of hint requests on posttest scores;
-causal mediation, which decomposes the overall treatment effect into a
-component driven by hint requests and another driven by other
-mechanisms; and principal stratification, which estimates separate
-treatment effects for groups of students who would request hints
-differently, if assigned to treatment.
+The next three sections will present the observational study,
+causal mediation, and principal stratification, respectively.
 As it turns out, these analyses appear to give contradictory
@@
 negative effect of requesting hints, while the principal
-stratification analysis finds that students who request more hints
-experience larger overall treatment effects.
+stratification analysis finds that students who requested more hints
+may have experienced larger overall treatment effects.
 Section \ref{sec:synthesis} will contrast these findings (including a
@@
 %\marginpar{make sure this makes sense before ``Data'' section}
-Unlike previous studies of hint usage, the CTA1 study was not designed
-to estimate the effect of hint usage; its purpose was to estimate the
-effect of CTA1 as a package, including the associated textbook and
-recommended classroom practices.
-Access to the full CTA1 curriculum was randomized, rather than any
-specific aspect of usage, and effectiveness was measured with a
+Since the purpose of the CTA1 study was to estimate the overall effect
+of the CTA1 curriculum---including the associated textbook and
+recommended classroom practices---access to the full CTA1 curriculum
+was randomized, rather than any specific aspect of usage.
+Effectiveness was measured with a
 standardized posttest covering a broad range of Algebra I skills.
@@
 contexts in which students may request hints.
-This variability, in turn, allows researchers to fit flexible models
-and ask nuanced questions about hint usage, for instance, exploring differing
-patterns of usage among different students.
-This sort of analysis is beyond the scope of the current study, which
-focuses on comparing and contrasting causal modeling approaches for
-these data.
-
-For the current purpose, we narrow down the type
-of hint data we model to align with our research focus.
+
+In order to align with our research focus, we narrow the type
+of hint data we model.
 Since our posttest measured Algebra I skills, we only
@@
 hints $\eta$ may actually differ in their observed hint requests, because one
-worked harder problems than the other.
+worked on harder problems than the other.
 However, in our dataset many problems were worked by very few students, so problem-level parameters would be hard to estimate.
@@
 This definition agreed with the model's classification (based on
-$\eta$ as opposed to $\bar{h}$ in
+$\eta$ as opposed to $\bar{h}$) in
 approximately \Sexpr{round(agree*100)}\% of cases.
@@
 students were omitted from the
-dataset.\footnote{The model was re-run without dropping sections, and
+dataset.\footnote{The principal stratification model was re-run without dropping sections, and
   after dropping sections worked by fewer than 500 students, with
@@
 based on baseline covariates, and randomized between treatment and
-control condition within these randomization blocks.
+control condition within those randomization blocks.
 When we dropped a treatment school from our analysis, we also dropped
@@
 were dropped from the study.\footnote{Including these students in a
-  principal stratification model is straightforward \citep{aoas}; see
-  the online supplement for results from a model including them, which
-  were broadly similar to the results presented here. Including
+  principal stratification model is straightforward \citep{aoas}. Including
   subjects with missing log data in a mediation or observational study
@@
  \citet{hansen2009b}, and \citet{ho:etal:2007}, has three broad steps.
-In the first, we construct a "match," identifying groups of students, some of whom have $H=0$ and others with $H=1$, but who are otherwise comparable.
-The hope is that the distribution of $H$ within this match resembles what might have been seen if (counterfactually) $H$ had been randomized within matched groups.
-This stage is iterative: we construct a match, evaluate it (most importantly by checking if matched students have similar baseline covariate distributions), and if necessary construct a new match.
+In the first, we construct a ``match," identifying groups of students, some of whom have $H=0$ and others with $H=1$, but who are otherwise comparable.
+The hope is that the distribution of $H$ conditional on this match
+resembles what might have been seen if (counterfactually) $H$ had been
+randomized within matched groups.
+Thus we evaluate the success of the match by checking if matched
+students have similar baseline covariate distributions; if necessary,
+we may revise match.
 In the second step, we estimate the average effect of $H$ on posttest scores, by comparing posttest scores between matched students with $H=1$ and $H=0$.
@@
 Then the CTA1 treatment effect for student $i$ is $Y_i(Z=1)-Y_i(Z=0)$.
-A similar structure applies to hint request for students in the
-treatment group.
-For these students, let the potential outcomes be $Y_i(Z=1,H=1)$ and
-$Y_i(Z=1,H=0)$ or $Y_i(H=1)$ and $Y_i(H=0)$ for short.
+For students in the treatment group, define potential outcomes $Y_i(Z=1,H=1)$ and
+$Y_i(Z=1,H=0)$, or $Y_i(H=1)$ and $Y_i(H=0)$ for short, corresponding
+to $i$'s posttest score were $H_i=1$ or 0, respectively.
+Then the effect of $H$ on $Y$ for student $i$ is $Y_i(H=1)-Y_i(H=0)$.
 % For a study measuring hint requests as a continuous variable, we may
@@
 differed at baseline between those students with $H=1$ and those with $H=0$).
-However, if these relevant characteristics were all measured and
-available, the ATE of $H$ on $Y$ would be identified.
 
@@
 a natural spline with five degrees of freedom for pretest.
-State, school, and class, were each included using normally
-distributed random intercepts.
+State, school, and class were each included using normally
+distributed random intercepts, $\epsilon^{state}$,
+$\epsilon^{school}$, and $\epsilon^{class}$.
 
@@
 student characteristics, pedagogical styles, and CTA1 usage patterns
-\citep[see][]{descriptivePaper}.%, and second, doing so facilitated the mediation
+\citep[see][]{descriptivePaper}. %, and second, doing so facilitated the mediation
 %analysis in the following section.
@@
 with estimated propensity scores close to 1 or 0, focusing on the
-``region of common support'' \citep[e.g.]{caliendo2008some,shadish2010primer}.
+``region of common support'' \citep[e.g.][]{caliendo2008some,shadish2010primer}.
 Doing so requires modifying the causal estimand---studies can only
@@
 The p-value from an omnibus covariate balance test
-\citep{covBal} is \Sexpr{prettyNum(round(balMod[['overall']]['matched','p.value'],2))}.
+\citep{covBal} is \Sexpr{prettyNum(round(balP,2))}.
 
@@
 The three estimates all roughly agree that being a high hint user
-decreases posttest scores by roughly a fifth of a standard deviation.
+decreases posttest scores by around 0.15 standard deviations.
 That is, they suggest that requesting hints hurts posttest scores.
@@
 unmeasured covariates; instead, they relied on matched ignorability.
-This would be a particular concern if pretest scores do not completely
-capture students' mathematics ability, and if weaker students ask for hints at
-higher rates, and also score lower on the posttest.
-Then, whatever aspects of student ability are not captures by the
-pretest, and correlate with both hint requests and posttest scores,
-would confound the estimates in the upper part of Table
-\ref{tab:matchResults}.
-
 To account for the possibility of unmeasured confounding, we conducted
@@
 accounting for the observed covariates in the model, and $T_Z$ is the
-t-statistic from an ordinary least squares regression of $H$ on $U$, along with the observed covariates.
+t-statistic for the coefficient on $U$ from an ordinary least squares
+regression of $H$ on $U$ and the observed covariates.
 These sensitivity parameters may be benchmarked by estimating their
@@
 On the other hand, it is likely that pretest scores do not fully capture between-student variance in academic ability; that is, there may be important components of prior ability (such as the ability to learn algebra) that are correlated with hint requests and posttest scores but not reflected in pretest scores.
-A hypothetical measurement of one of those components may indeed have constituted a more important covariate than pretest scores.
+Hypothetical measurements of these components could, perhaps,
+constitute unobserved covariates with similar or greater importance
+than observed pretest scores.
 
@@
 Estimating direct effects requires, additionally, data from the control group: posttest scores and (ideally) baseline covariates.
-The randomization of the intervention in the RCT now plays an important role in identification, since the overall effect of the intervention on posttest scores is at issue, as does the fact that the control group does not have access to the program.
+The randomization of the intervention in the RCT now plays an
+important role in identification, since the effects of the
+intervention on both hint usage and posttest scores are at issue; the
+fact that the control group does not have access to the program
+likewise plays a crucial role.
 Finaly, the assumption of no unmeasured confounding remains important.
@@
 We know that CTA1 hints were not available to the control group, so we
-set $H=0$.
-However, other facets of log data may not be defined, or coherent, in
+set $H=0$ for all members of the control group.
+However, other facets of log data may not be coherently defined in
 the control group; for instance, students' proclivity to master the
@@
 also be nearly impossible.
-This is because students assigned to the control group request exactly
+This is because students assigned to the control group requested exactly
 zero hints, whereas merely
@@
 way forward.
-Since the experimental design ensures that $H(0)=0$, $Y(1,H(0))\equiv
+Since the experimental design ensures that $H(0)=0$, we have that $Y(1,H(0))\equiv
 Y(1,0)$; this potential outcome is observed for subjects in
@@
 potential outcome $Y(1,0)$ for treated subjects with $H_i=1$.
-However, it turns out that the matching estimators from
+However, it turns out that the matching estimators from Section
 \ref{sec:observational} already solved this problem:
@@
 $Y_i(1,0)=Y_i$, the observed outcome. When $Z_i=1$ and $H_i=1$,
-impute $Y(1,0)$ as the average of the observed outcomes from the $H=0$ subjects
-matched to $i$.
+impute the average of the observed outcomes from the $H=0$ subjects
+matched to $i$ for $Y_i(1,0)$.
 Matched ignorability implies that the imputations will be unbiased.
@@
 The goal is to estimate the average effect of
-assignment to the treatment condition separately in different stratum.
+assignment to the treatment condition separately in each stratum.
 In our analysis, we imagine that each student has his or her own proclivity to request hints, when possible: if assigned to treatment, each student would tend to request hints at a different rate.
@@
 randomization of treatment assignment,
-rather than confounding control or assumptions about
+rather than on assumptions about
 unmeasured confounders.
@@
 defining principal strata is latent---specifically, the ``ability''
-parameter is \eqref{eq:rasch}.
+parameter in \eqref{eq:rasch}.
 For this approach, implementation is measured with a series of binary
@@
 sensitivity analysis.
-More details for checking PS models can be found in \citet{aoas},
+More details for checking PS models can be found in \citet{aoas} and
 \citet{psTutorial}, and for Bayesian models in general in
@@
 hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
-Whereas the goal of the observational study and mediation
-analysis was the relationship between $\bar{h}$ and posttest scores, the
-goal of PS is the relationship between $\bar{h}(1)$ and treatment effects.
+The goal of PS is to estimate the relationship between $\bar{h}(1)$
+and treatment effects; this contrasts with the observational study and mediation
+analysis whose goal was the relationship between $\bar{h}$ and posttest scores.
 Notably, although $Y(0)$ and $\bar{h}(1)$ and $Y(0)$ are never
@@
 Second, a modeling difference: instead of the normal mixture model
-\eqref{eq:mixture}, we modeled $\eta(1)$ as normal.
+\eqref{eq:mixture}, we modeled $\eta(1)$ with a normal regression model.
 With this change, the PS estimand became
@@
 \end{equation*}
-the treatment effect for students whose potential hint usage, if assigned to the treatment condition $\eta(1)=r$.
+the treatment effect for students whose potential hint usage, if assigned to the treatment condition, $\eta(1)=r$.
 A full treatment of this variant of PS, including a
@@
 Estimating $\tau(r)$ required estimating $\EE[Y(0)|\eta(1)=r]$,
-despite the fact that hint requests never observed at the same time as
+despite the fact that hint requests are never observed at the same time as
 $Y(0)$ (in fact $\eta(1)$ isn't directly observed in either the
 treatment or control group).
-However, hint requests are observed, and the conditional distribution
-$\eta(1)|\bm{x}$ is identified, among those students with $Z=1$;
+However, among those students with $Z=1$, hint requests are observed
+and the conditional distribution $\eta(1)|\bm{x}$ is identified;
 because of randomization of $Z$, these inferences extend to the
-control group as well.
+control group as well \citep{feller2016compared}.
 
@@
 $r=\tau(1)$ varies; the black line shows the estimate (i.e. posterior
-mean) and the red lines are random draws from the posterior
+mean) and the magenta lines are random draws from the posterior
 distribution, showing estimation uncertainty.
@@
 $H$ is not an exposure, that ``happens to'' subjects, but rather its
-potential value $H_T$ is a characteristic of the subjects.
+potential value $H(1)$ is a characteristic of the subjects.
 Observed $H$ reveals what was there all along.
@@
 While any model specification can, and should, be checked against
-observed data, this process can only narrow space of acceptable
+observed data, this process can only narrow the space of acceptable
 models, it cannot be used to determine the correct form of
@@
 qualitatively different conclusions.
-In practice, analysts make the untestable assumption that $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ is
+In practice, we make the untestable assumption that $p(Y|\bm{x},Z=0,\bm{\theta},\eta(1))$ is
 of the same form as $p(Y|\bm{x},Z=1,\bm{\theta},\eta(1))$.
@@
 
-Can these results be reconciled with the results from the previous two
-sections?
+Can these three sets of results be reconciled?
 Does requesting hints help or hurt?
@@
 For instance, unlike observational studies and principal stratification, mediation analysis
-depends on strictly counterfactual quantities, such as $\EE Y(0,H(1))$,
+depends on strictly counterfactual quantities, such as $Y(0,H(1))$,
 that can never occur.
-Analogous quantities in principal stratification, such as $\EE [Y(0)|H(1)]$ are
+Analogous quantities in principal stratification, such as $\EE [Y(0)|\eta(1)]$ are
 unobservable, but nevertheless refer to averages of potential outcomes
@@
 partially observed $\eta(1)$.
-Future work (by us, and, we hope, others) may develop these methods,
+We hope that future work will develop these methods,
 and still others.

### 2020-06-19 paper.tex
final version (famous last words)

@@
 This definition agreed with the model's classification (based on
-$\eta$ as opposed to $\bar{h}$ in
+$\eta$ as opposed to $\bar{h}$) in
 approximately 93\% of cases.
@@
 based on baseline covariates, and randomized between treatment and
-control condition within these randomization blocks.
+control condition within those randomization blocks.
 When we dropped a treatment school from our analysis, we also dropped
 the control schools in its randomization block.
-Of the remaining 2,390 students, 87\% (2,091) had log
+Of the remaining 2,390 students, 88\% (2,108) had log
 data.
@@
 
-All told, the analyses presented here were all based on 2,091 students assigned to the CTA1 condition; the estimation of direct effects in Section \ref{sec:mediation} and the principal stratification also relied on the 2,918 students assigned to control.
-Together, the 5,009 students were nested within 116 teachers, in 43 schools across five states.
+All told, the analyses presented here were all based on 2,108 students assigned to the CTA1 condition; the estimation of direct effects in Section \ref{sec:mediation} and the principal stratification also relied on the 2,918 students assigned to control.
+Together, the 5,026 students were nested within 116 teachers, in 43 schools across five states.
 %The observational study (Section \ref{sec:observational} and part of the mediation analysis \ref{sec:mediation} used only the treatment (CTA1) group.
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Wed May 20 10:38:55 2020
+% Mon Jun 15 15:02:47 2020
 \begin{table}[ht]
@@
 % latex table generated in R 3.6.2 by xtable 1.8-4 package
-% Wed May 20 10:38:57 2020
+% Mon Jun 15 15:02:49 2020
 \begin{table}[ht]
@@
   raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
-  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
+  pretest & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
   spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 

### 2020-06-19 R/covariateTable.r

@@ -21,8 +21,8 @@ miss['xirt','Imputation Error'] <- sqrt(miss['xirt','Imputation Error'])/sd(covs
 balTest <- balanceTest(treatment~grade+esl+frl+race+sex+spec+xirt+cluster(schoolid2)+strata(pair),data=dat,report=c('adj.means','std.diffs','chisquare.test'),include.NA.flags=T)
 
 ## use "unstrat" for group means cuz no one knows/cares about "adjusted means"
-trtMeans <- round(balTest$results[,'Treatment','Unstrat']*100)
-ctlMeans <- round(balTest$results[,'Control','Unstrat']*100)
+trtMeans <- round(balTest$results[,'Treatment','--']*100)
+ctlMeans <- round(balTest$results[,'Control','--']*100)
 stdDiff <- sprintf('%.2f',balTest$results[,'std.diff','pair'])
 names(stdDiff) <- dimnames(balTest$results)[[1]]
 
@@ -83,8 +83,8 @@ for(vv in varbs){
     }
   } else cat0(
     varNames[vv],'&',miss[vv,2],'\\%&',impErr[vv],'& &',
-    sprintf('%.2f',balTest$results[vv,'Control','Unstrat']),'& ',
-    sprintf('%.2f',balTest$results[vv,'Treatment','Unstrat']),'& ',stdDiff[vv],'\\\\\n')
+    sprintf('%.2f',balTest$results[vv,'Control','--']),'& ',
+    sprintf('%.2f',balTest$results[vv,'Treatment','--']),'& ',stdDiff[vv],'\\\\\n')
   cat0('\\hline\n')
 }
 cat0('&&&\\multicolumn{4}{c}{Overall Covariate Balance: p=',results$balP,'}\\\\

### 2020-06-19 R/dataAnalysis.r

@@ -1,7 +1,12 @@
+devtools::dev_mode()
+#devtools::install_github("markmfredrickson/RItools@version_0.3.0")
+library(RItools)
+devtools::dev_mode()
+
+
 library(splines)
 library(gridExtra)
 library(optmatch)
-library(RItools)
 library(sandwich)
 library(clubSandwich)
 library(lmtest)
@@ -11,6 +16,7 @@ library(rstan)
 library(lme4)
 library(estimatr)
 library(HDInterval)
+
 
 traceplot <- rstan::traceplot
 select <- dplyr::select
@@ -140,6 +146,8 @@ m3 <- fullmatch(dist3,data=dat1)
 dat1$pretest <- dat1$xirt
 
 balMod <- xBalance(M~grade+race+pretest+spec+esl+state,strata=list(`Before Matching`=NULL,`After Matching`=~m3),report=c('std.diff','z.score','chisquare.test'),data=droplevels(dat1))
+
+balP <- balMod$overall['After.Matching','p.value']
 
 pdf('output/balancePlot.pdf')
 plot(balMod,colors=subwayPalette)

### 2020-06-19 R/helperFunctions.r

@@ -9,6 +9,7 @@ makeStanDat <- function(dat,x,xInteract=FALSE,missingUsage=TRUE){
  dat <- droplevels(dat)
 
  x$section <- as.factor(x$section)
+ x$unit <- as.factor(x$unit)
 
  stanDat$nsecWorked <- nrow(x)
  stanDat$nstud <- nrow(dat)
@@ -16,12 +17,15 @@ makeStanDat <- function(dat,x,xInteract=FALSE,missingUsage=TRUE){
  stanDat$nschool <- nlevels(dat$schoolid2)
  stanDat$nsec <- nlevels(x$section)
  stanDat$npair <- nlevels(dat$pair)
+ stanDat$nunit <- nlevels(x$unit)
 
  stanDat$teacher <- as.numeric(as.factor(dat$teachid2))
  stanDat$pair <- as.numeric(dat$pair)
  stanDat$school <- as.numeric(dat$school)
  stanDat$studentM <- seq(stanDat$nstud)[match(x$field_id,dat$field_id)]
  stanDat$section <- as.numeric(x$section)
+ stanDat$unit <- as.numeric(x$unit)
+
 
  stanDat$grad <- as.numeric(x$hint)
 
@@ -30,10 +34,53 @@ makeStanDat <- function(dat,x,xInteract=FALSE,missingUsage=TRUE){
  stanDat$X <- scale(X)
  stanDat$ncov <- ncol(X)
 
+ stanDat$studID <- dat$field_id
 
  stanDat$Z <- as.numeric(dat$treatment)
 
  stanDat$Y <- dat$Y
+
+ stanDat
+}
+
+makeStanDatTrt <- function(dat,x,xInteract=FALSE,missingUsage=TRUE){
+  stanDat <- list()
+
+  dat <- filter(dat,treatment==1)
+
+  if(!missingUsage)
+    dat <- filter(dat,field_id%in%x$field_id|dat$treatment==0)
+
+  x <- droplevels(x)
+ dat <- droplevels(dat)
+
+ x$section <- as.factor(x$section)
+ x$unit <- as.factor(x$unit)
+
+ stanDat$nsecWorked <- nrow(x)
+ stanDat$nstud <- nrow(dat)
+ stanDat$nteacher <- nlevels(dat$teachid2)
+ stanDat$nschool <- nlevels(dat$schoolid2)
+ stanDat$nsec <- nlevels(x$section)
[…]

### 2020-08-23 paper.Rnw
last resubmission? a couple tweaks

@@
 \section{Introduction}
-From 2007 through 2010, the RAND corporation
-conducted one of the first effectiveness trials awarded through
-competitive grant programs sponsored by the US Department of
-Education's Institute of Education Sciences.  The randomized
-controlled trial (RCT) was
-designed to estimate the effect of school-wide adoption of the
+As educational technology (EdTech) booms, so do randomized controlled
+trials (RCTs) designed to evaluate edTech
+applications. A recent systematic review \citep{escueta2017education} cites 29
+published reports of RCTs evaluating ``computer assisted learning'' programs,
+all but one of which was published since 2001, and we are aware of
+numerous other such RCTs planned or ongoing.
+For instance, from 2007 through 2010, the RAND corporation
+conducted %one of the first effectiveness trials awarded through
+% competitive grant programs sponsored by the US Department of
+% Education's Institute of Education Sciences.  The
+an
+RCT %was
+designed to estimate the effectiveness of school-wide adoption of the
 Cognitive Tutor Algebra I (CTA1) curriculum, whose centerpiece is a
@@
 standard deviation.
-As educational technology (EdTech) booms, so do RCTs in the mold of the RAND
-CTA1 study. A recent systematic review \citep{escueta2017education} cites 29
-published reports of RCTs of ``computer assisted learning'' programs,
-all but one of which was published since 2001. We are aware of
-numerous other such RCTs planned or ongoing.
 
@@
 
-This section will illustrate an observational study to answer these
+This section will illustrate an observational study, in an attempt to answer these
 questions.
@@
 \section{Mediation Analysis}\label{sec:mediation}
-The goal of causal mediation analysis is to decompose a treatment
+What role did hint requests play in the CTA1 effect?
+Assignment to the treatment condition allowed students to request many
+hints during their work---how did the wide availability of hints affect their posttest scores?
+Conversely, how would assignment to CTA1 have affected posttest scores had
+hints been more limited?
+
+This section will illustrate mediation analysis, in an attempt to answer these
+questions.
+Mediation analysis decomposes the treatment
 effect into an ``indirect'' or ``mediated''
 effect and a ``direct'' effect.
-The indirect effect captures the component of the effect that is due
-to the mediator: the treatment affected the outcome by affecting the
-mediator, which itself, in turn, affected the outcome.
-Assignment to the treatment condition allowed students to request many
-hints during their work---how did the wide availability of hints affect their posttest scores?
-The direct effect is the component of the overall effect that operates via other mechanisms.
-For instance, how would assignment to CTA1 affect posttest scores were
-hints more limited?
+The indirect effect captures the component of the treatment effect that was due
+to a mediator, in this case hint requests.
+That is, the treatment affected the
+mediator---i.e. allowed students to request many hints---which, in
+turn, affected the outcome.
+The direct effect is the component of the overall effect that operated via other mechanisms.
 
@@
 \section{Principal Stratification}\label{sec:principalStratification}
-
-In a principal stratification \citep[PS;][]{frangakis}  analysis, the researcher stratifies all students in
+Was CTA1 more beneficial for students who tended to request more
+hints, or for those who tended to request fewer?
+
+This section will illustrate principal stratification
+\citep[PS;][]{frangakis}, in an attempt to answer this question.
+In PS, the researcher stratifies all students in
 the RCT---in both the treatment and control arms---based on how they
@@
 assignment to the treatment condition separately in each stratum.
-In our analysis, we imagine that each student has his or her own proclivity to request hints, when possible: if assigned to treatment, each student would tend to request hints at a different rate.
-How does the effect of assignment to the CTA1 condition vary with these proclivities?
-Would the benefit of being assigned to the CTA1 condition be higher or lower for students who would tend to request more hints than for those who would tend to request fewer?
+In our analysis, we imagine that each student has his or her own
+proclivity to request hints, when possible: if assigned to treatment,
+each student would tend to request hints at a different rate.
+Then, we estimate varying treatment effects as a function of those
+proclivities.
 
